---
title: 'Heteroskedasticity'
subtitle: 'EC421, Set 04'
author:
  - name: Ed Rubin
    email: edwardr@uoregon.edu
editor:
  render-on-save: true
format:
  clean-revealjs:
    slide-level: 2
    theme: [../my.scss, default]
    highlight-style: gruvbox
    embed-resources: true
    slide-number: c/t
    incremental: false
    controls-layout: edges
    transition: none
    preview-links: auto
    html-math-method: mathjax
    code-line-numbers: false
execute:
  warning: false
  message: false
---

## {.center}

[Prologue]{.bigger .it .slab}

```{r}
#| label: "setup"
#| include: false
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(
  broom, latex2exp, ggplot2, ggthemes, viridis, extrafont,
  dplyr,
  magrittr, knitr, parallel
)
# Define pink color
red_pink = "#e64173"
grey_light = "grey70"
grey_mid = "grey50"
grey_dark = "grey20"
# Dark slate grey: #314f4f
# Notes directory
dir_slides = "~/Dropbox/UO/Teaching/EC421W19/LectureNotes/02Review/"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = FALSE,
  message = FALSE
)
# A blank theme for ggplot
theme_empty =
  theme_bw() +
  theme(
    line = element_blank(),
    rect = element_blank(),
    strip.text = element_blank(),
    axis.text = element_blank(),
    plot.title = element_blank(),
    axis.title = element_blank(),
    plot.margin = structure(c(0, 0, -0.5, -1), unit = "lines", valid.unit = 3L, class = "unit"),
    legend.position = "none"
  )
theme_axes =
  theme_void(base_family = "Fira Sans Condensed") +
  theme(
    text =
      element_text(family = "Fira Sans Condensed"),
    axis.title =
      element_text(size = 20, family = "Fira Sans Condensed"),
    axis.title.x =
      element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines"), family = "Fira Sans Condensed"),
    axis.title.y =
      element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines"), family = "Fira Sans Condensed"),
    axis.line =
      element_line(
        color = "grey70",
        size = 0.25,
        arrow = arrow(angle = 30, length = unit(0.15, "inches")
      )),
    plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
    legend.position = "none"
  )
```

```{css}
#| echo: false

@media print {
  .has-continuation {
    display: block !important;
  }
}
```

---

## [R]{.mono } showcase


**[[Quarto]{.mono }](https://quarto.org/)**
- Simple mark-up language for for combining/creating documents, equations, figures, [R]{.mono }, and more
- [Basics of [Markdown]{.mono }](https://rmarkdown.rstudio.com/authoring_basics.html)
- _E.g._, `**I'm bold**`, `*I'm italic*`, `I = "code"`

Interactive data visualization beyond [`shiny`](https://shiny.posit.co/): [`plotly`](https://plotly.com/examples/), [`crosstalk`](https://rstudio.github.io/crosstalk/), [`echarts4r`](https://echarts4r.john-coene.com/), [`highcharter`](https://jkunst.com/highcharter/).

**[Econometrics with [R]{.mono }](https://www.econometrics-with-r.org)**
- (Currently) free, online textbook
- Written and published using [R]{.mono } (and probably [R Markdown]{.mono })
- *Warning:* I haven't read the full book.

Related: Tyler Ransom has a [great cheatsheet for econometrics](https://github.com/tyleransom/EconometricsLabs/blob/master/econometricsCheatSheet.pdf).

---

## Schedule


### Last Time

We wrapped up our review.

### Today

Heteroskedasticity

---

## Schedule


### This week

First assignment!

Submit **one file** (HTML, PDF, or Word) that includes 

1. Your written answers
1. Figures and regression output
1. The [R]{.mono } code that generated your answers.

This file should be a rendered RMarkdown or Quarto file (`.rmd` or `.qmd`).

[Important]{.b .pink }

- We should be able to easily find your answers for each question.
- **Do not copy.** (You [will]{.attn } receive a zero.)

---

## {.center}

[Prerequisite material]{.bigger .it .slab}

---

## Prerequisite material
### The $\chi^2$ distribution

Some test statistics are distributed as $\chi^2$ random variables.

The $\chi^2$ distribution is just another example of a common (named) distribution (like the Normal distribution, the $t$ distribution, and the $F$).

The shape of the $\chi^2$ distribution depends on a single parameter:
- We will call this parameter $k$
- Our test statistics will refer to $k$ as *degrees of freedom*.

---

## Prerequisite material
### The $\chi^2$ distribution

Three examples of $\chi_k^2$: $\textcolor{#314f4f}{k = 1}$, $\textcolor{#e64173}{k = 2}$, and $\textcolor{orange}{k = 9}$

```{r}
#| label: "chisq1"
#| echo: false
#| dev: "svg"
#| fig-height: 5.5
ggplot(data = tibble(x = c(0, 20)), aes(x)) +
  geom_area(
    stat = "function", fun = dchisq, args = list(df = 2),
    fill = "darkslategrey", alpha = 0.3
  ) +
  stat_function(
    fun = dchisq, args = list(df = 2), n = 1e3,
    color = "darkslategrey"
  ) +
  geom_area(
    stat = "function", fun = dchisq, args = list(df = 3),
    fill = red_pink, alpha = 0.3
  ) +
  stat_function(
    fun = dchisq, args = list(df = 3), n = 1e3,
    color = red_pink
  ) +
  geom_area(
    stat = "function", fun = dchisq, args = list(df = 9),
    fill = "orange", alpha = 0.3
  ) +
  stat_function(
    fun = dchisq, args = list(df = 9), n = 1e3,
    color = "orange"
  ) +
  labs(x = "x", y = "f") +
  theme_axes
```

---

## Prerequisite material
### The $\chi^2$ distribution

Probability of observing a more extreme test statistic $\widehat{\text{LM}}$ under H[0]{.sub }

```{r}
#| label: "chisq2"
#| echo: false
#| dev: "svg"
#| fig-height: 5.5
ggplot(data = tibble(x = c(0, 8)), aes(x)) +
  geom_area(
    stat = "function", fun = dchisq, args = list(df = 2),
    fill = "darkslategrey", alpha = 0.05
  ) +
  geom_area(
    stat = "function", fun = dchisq, args = list(df = 2),
    fill = red_pink, alpha = 0.85,
    xlim = c(5, 8)
  ) +
  stat_function(
    fun = dchisq, args = list(df = 2), n = 1e3,
    color = "darkslategrey"
  ) +
  geom_vline(xintercept = 5, color = grey_dark, size = 0.5, linetype = "dotted") +
  annotate("text", x = 5, y = 1.55 * dchisq(5, df = 2), label = TeX("$\\widehat{LM}$"), family = "MathJax_Main", size = 7) +
  labs(x = "x", y = "f") +
  theme_axes
```

---

## {.center}

[Heteroskedasticity]{.bigger .it .slab}

---

## Heteroskedasticity

Let's write down our **current assumptions**

1. Our sample (the $x_k$'s and $y_i$) was [randomly drawn]{.b .pink } from the population.

2. $y$ is a [linear function]{.b .pink } of the $\beta_k$'s and $u_i$.

3. There is no perfect [multicollinearity]{.b .pink } in our sample.

4. The explanatory variables are [exogenous]{.b .pink }: $\mathop{\boldsymbol{E}}\left[ u \middle| X \right] = 0 \left(\implies \mathop{\boldsymbol{E}}\left[ u \right] = 0\right)$.

5. The disurbances have [constant variance]{.b .pink } $\sigma^2$ and [zero covariance]{.b .pink }, _i.e._,
    - $\mathop{\boldsymbol{E}}\left[ u_i^2 \middle| X \right] = \mathop{\text{Var}} \left( u_i \middle| X \right) = \sigma^2 \implies \mathop{\text{Var}} \left( u_i \right) = \sigma^2$
    - $\mathop{\text{Cov}} \left( u_i, \, u_j \middle| X \right) = \mathop{\boldsymbol{E}}\left[ u_i u_j \middle| X \right] = 0$ for $i\neq j$

6. The disturbances come from a [Normal]{.b .pink } distribution, _i.e._, $u_i \overset{\text{iid}}{\sim} \mathop{\text{N}}\left( 0, \sigma^2 \right)$.

---

## Heteroskedasticity

Today we're focusing on assumption \#5:

5\. The disurbances have [constant variance]{.b .pink } $\sigma^2$ and [zero covariance]{.b .pink }, _i.e._,

  - $\mathop{\boldsymbol{E}}\left[ u_i^2 \middle| X \right] = \mathop{\text{Var}} \left( u_i \middle| X \right) = \sigma^2 \implies \mathop{\text{Var}} \left( u_i \right) = \sigma^2$
  - $\mathop{\text{Cov}} \left( u_i, \, u_j \middle| X \right) = \mathop{\boldsymbol{E}}\left[ u_i u_j \middle| X \right] = 0$ for $i\neq j$

::: {.fragment}
We will focus on the assumption of [constant variance]{.b .pink }<br>(also known as *homoskedasticity*).
:::

::: {.fragment}
**Violation of this assumption:**

[Heteroskedasticity:]{.b .pink } $\mathop{\text{Var}} \left( u_i \right) = \sigma^2_i$ and $\sigma^2_i \neq \sigma^2_j$ for some $i\neq j$.
:::

::: {.fragment}
In other words: Our disturbances have different variances.
:::

---

## Heteroskedasticity

Classic example of heteroskedasticity: The funnel

Variance of $u$ increases with $x$

```{r}
#| label: "het-ex1"
#| dev: "svg"
#| echo: false
#| fig-height: 5.5
set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 4 + 1.5 * x)
), aes(x = x, y = e)) +
geom_point(color = "darkslategrey", size = 2.75, alpha = 0.5) +
labs(x = "x", y = "u") +
theme_axes
```

---

## Heteroskedasticity

Another example of heteroskedasticity: (double funnel?)

Variance of $u$ increasing at the extremes of $x$

```{r}
#| label: "het-ex2"
#| dev: "svg"
#| echo: false
#| fig-height: 5.5
set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 2 + x^2)
), aes(x = x, y = e)) +
geom_point(color = "darkslategrey", size = 2.75, alpha = 0.5) +
labs(x = "x", y = "u") +
theme_axes
```

---

## Heteroskedasticity

Another example of heteroskedasticity:

Differing variances of $u$ by group

```{r}
#| label: "het-ex3"
#| dev: "svg"
#| echo: false
#| fig-height: 5.5
set.seed(12345)
ggplot(data = tibble(
  g = sample(c(F,T), 1e3, replace = T),
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 0.5 + 2 * g)
), aes(x = x, y = e, color = g, shape = g, alpha = g)) +
geom_point(size = 2.75) +
scale_color_manual(values = c("darkslategrey", red_pink)) +
scale_shape_manual(values = c(16, 1)) +
scale_alpha_manual(values = c(0.5, 0.8)) +
labs(x = "x", y = "u") +
theme_axes
```

---

## Heteroskedasticity

### What and why?

[Heteroskedasticity]{.b .pink } is present when the variance of $u$ changes with any combination of our explanatory variables $x_1$, through $x_k$ (henceforth: $X$).

::: {.fragment}
It's very common in practice—and should probably be our default.
:::

::: {.fragment}
[Why we care:]{.b .pink } Heteroskedasticity shows us how small violations of our assumptions can affect OLS's performance.
:::

---

## Heteroskedasticity

### Consequences

So what are the consquences of heteroskedasticity? Bias? Inefficiency?

First, let's check if it has consquences for the the unbiasedness of OLS.

::: {.fragment}
**Recall[1]{.sub }:** OLS being unbiased means $\mathop{\boldsymbol{E}}\left[ \hat{\beta}_k \middle| X \right] = \beta_k$ for all $k$.
:::

::: {.fragment}
**Recall[2]{.sub }:** We previously showed $\hat{\beta}_1 = \dfrac{\sum_i\left(y_i-\overline{y}\right)\left(x_i-\overline{x}\right)}{\sum_i\left(x_i -\overline{x}\right)^2}$
:::

::: {.fragment}
It will actually help us to rewrite this estimator as

$$ \hat{\beta}_1 = \beta_1 + \dfrac{\sum_i \left( x_i - \overline{x} \right) u_i}{\sum_i \left( x_i - \overline{x} \right)^2} $$
:::

---

## Heteroskedasticity

**Proof:** Assuming $y_i = \beta_0 + \beta_1 x_i + u_i$

$$
\begin{aligned}
  \hat{\beta}_1
  &= \dfrac{\sum_i\left(y_i-\overline{y}\right)\left(x_i-\overline{x}\right)}{\sum_i\left(x_i -\overline{x}\right)^2} \\
  &= \dfrac{\sum_i\left(\left[ \beta_0 + \beta_1 x_i + u_i \right]- \left[ \beta_0 + \beta_1 \overline{x} + \overline{u} \right] \right)\left(x_i-\overline{x}\right)}{\sum_i\left(x_i -\overline{x}\right)^2} \\
  &= \dfrac{\sum_i\left(\beta_1 \left[ x_i - \overline{x} \right] + \left[u_i - \overline{u}\right]  \right)\left(x_i-\overline{x}\right)}{\sum_i\left(x_i -\overline{x}\right)^2} \\
  &= \dfrac{\sum_i\left(\beta_1 \left[ x_i - \overline{x} \right]^2 + \left[ x_i - \overline{x} \right] \left[u_i - \overline{u}\right]\right)}{\sum_i\left(x_i -\overline{x}\right)^2} \\
  &= \beta_1 + \dfrac{\sum_i\left(x_i - \overline{x}\right) \left(u_i - \overline{u}\right)}{\sum_i\left(x_i -\overline{x}\right)^2}
\end{aligned}
$$

---

## Heteroskedasticity

$$
\begin{aligned}
  \hat{\beta}_1
  &= \cdots = \beta_1 + \dfrac{\sum_i\left(x_i - \overline{x}\right) \left(u_i - \overline{u}\right)}{\sum_i\left(x_i -\overline{x}\right)^2} \\
  &= \beta_1 + \dfrac{\sum_i\left(x_i - \overline{x}\right) u_i - \overline{u} \sum_i\left(x_i - \overline{x}\right)}{\sum_i\left(x_i -\overline{x}\right)^2} \\
  &= \beta_1 + \dfrac{\sum_i\left(x_i - \overline{x}\right) u_i - \overline{u} \left(\sum_i x_i - \sum_i \overline{x}\right)}{\sum_i\left(x_i -\overline{x}\right)^2} \\
  &= \beta_1 + \dfrac{\sum_i\left(x_i - \overline{x}\right) u_i - \overline{u} \left(\sum_i x_i - n \overline{x}\right)}{\sum_i\left(x_i -\overline{x}\right)^2} \\
  &= \beta_1 + \dfrac{\sum_i\left(x_i - \overline{x}\right) u_i - \overline{u} \textcolor{#e64173}{\left(\sum_i x_i - \sum_i x_i\right)}}{\sum_i\left(x_i -\overline{x}\right)^2} \\
  &= \beta_1 + \dfrac{\sum_i\left(x_i - \overline{x}\right) u_i}{\sum_i\left(x_i -\overline{x}\right)^2} \quad \text{😅}
\end{aligned}
$$

---

## Heteroskedasticity

### Consequences: Bias

We now want to see if heteroskedasticity biases the OLS estimator for $\beta_1$.

::: {.fragment}
$$
\begin{aligned}
  \mathop{\boldsymbol{E}}\left[ \hat{\beta}_1 \middle| X \right]
  &= \mathop{\boldsymbol{E}}\left[ \beta_1 + \dfrac{\sum_i\left(x_i - \overline{x}\right) u_i}{\sum_i\left(x_i -\overline{x}\right)^2} \middle| X \right] \\[0.5em]
  &= \beta_1 + \mathop{\boldsymbol{E}}\left[ \dfrac{\sum_i\left(x_i - \overline{x}\right) u_i}{\sum_i\left(x_i -\overline{x}\right)^2} \middle| X \right] \\[0.5em]
  &= \beta_1 + \dfrac{\sum_i\left(x_i - \overline{x}\right)}{\sum_i\left(x_i -\overline{x}\right)^2} \textcolor{#e64173}{\underbrace{\mathop{\boldsymbol{E}}\left[ u_i \middle| X \right]}_{=0}} \\[0.5em]
  &= \beta_1
\end{aligned}
$$
:::

::: {.fragment}
Phew. [OLS is still unbiased]{.b .pink } for the $\beta_k$.
:::

---

## {.center}

OLS's **efficiency** and **inference** do not survive heteroskedasticity.

---

## Heteroskedasticity

### Consequences: Efficiency


In the presence of heteroskedasticity, OLS is [no longer the most efficient]{.b .pink } (best) linear unbiased estimator.

::: {.fragment}
It would be more informative (efficient) to [weight observations]{.b .pink } inversely to their $u_i$'s variance.

- Downweight high-variance $u_i$'s (too noisy to learn much).
- Upweight observations with low-variance $u_i$'s (more 'trustworthy').
- Now you have the idea of weighted least squares (WLS)
:::

---

## Heteroskedasticity

### Consequences: Inference

OLS [standard errors are biased]{.b .pink } in the presence of heteroskedasticity.

- Wrong confidence intervals

- Problems for hypothesis testing (both $t$ and $F$ tests)

- It's hard to learn much without sound inference.

---

## Heteroskedasticity

### Solutions


1. **Tests** to determine whether heteroskedasticity is present.

2. **Remedies** for (1) efficiency and (2) inference

---

## {.center}

[Testing for heteroskedasticity]{.bigger .it .slab}

---

## Testing for heteroskedasticity

While we *might* have solutions for heteroskedasticity, the efficiency of our estimators depends upon whether or not heteroskedasticity is present.

1. The **Goldfeld-Quandt test**

1. The **White test**

::: {.fragment}
Each of these tests^[There are many other options for testing, *e.g.*, the **Breusch-Pagan test**.] centers on the fact that we can [use the OLS residual]{.b .pink } $\textcolor{#e64173}{e_i}$ [to estimate the population disturbance]{.b .pink } $\textcolor{#e64173}{u_i}$.
:::

---

## Testing for heteroskedasticity
### The Goldfeld-Quandt test

Focuses on a specific type of heteroskedasticity: whether the variance of $u_i$ differs [between two groups]{.b .pink }.^[The G-Q test was one of the early tests of heteroskedasticity (1965).]

Remember how we used our residuals to estimate the $\sigma^2$?

$$ s^2 = \dfrac{\text{SSE}}{n-1} = \dfrac{\sum_i e_i^2}{n-1} $$

We will use this same idea to determine whether there is evidence that our two groups differ in the variances of their disturbances, effectively comparing $s^2_1$ and $s^2_2$ from our two groups.

---

## Testing for heteroskedasticity
### The Goldfeld-Quandt test

Operationally,

::: {.pseudocode-small}

1. Order your the observations by $x$

2. Split the data into two groups of size n[⭑]{.super }
  - G[1]{.sub }: The first third
  - G[2]{.sub }: The last third

3. Run separate regressions of $y$ on $x$ for G[1]{.sub } and G[2]{.sub }

4. Record SSE[1]{.sub } and SSE[2]{.sub }

5. Calculate the G-Q test statistic

:::

---

## Testing for heteroskedasticity
### The Goldfeld-Quandt test

The G-Q test statistic

$$ F_{\left(n^{\star}-k,\, n^{\star}-k\right)} = \dfrac{\text{SSE}_2/(n^\star-k)}{\text{SSE}_1/(n^\star-k)} = \dfrac{\text{SSE}_2}{\text{SSE}_1} $$

follows an $F$ distribution (under the null hypothesis) with $n^{\star}-k$ and $n^{\star}-k$ degrees of freedom.^[Goldfeld and Quandt suggested $n^{\star}$ of $(3/8)n$. $k$ gives number of estimated parameters (_i.e._, $\hat{\beta}_j$'s).]

::: {.fragment}
**Notes**

- The G-Q test requires the disturbances follow normal distributions.
- The G-Q assumes a very specific type/form of heteroskedasticity.
- Performs very well if we know the form of potentially heteroskedasticity.
:::

---

## Testing for heteroskedasticity
### The Goldfeld-Quandt test

```{r}
#| label: "gq1a"
#| echo: false
#| dev: "svg"
#| fig-height: 3.75

set.seed(12345)
# Data
gq_df = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 4 + 1.5 * x),
  y = 1 + 3 * x + e
)
# Quantiles
gq_x = quantile(gq_df$x, probs = c(3/8, 5/8))
# Regressions
sse1 = lm(y ~ x, data = gq_df %>% filter(x < gq_x[1])) %>%
  residuals() %>% raise_to_power(2) %>% sum()
sse2 = lm(y ~ x, data = gq_df %>% filter(x > gq_x[2])) %>%
  residuals() %>% raise_to_power(2) %>% sum()
ggplot(data = gq_df, aes(x = x, y = e)) +
geom_point(color = "darkslategrey", size = 2.75, alpha = 0.5) +
labs(x = "x", y = "u") +
theme_axes
```

---

## Testing for heteroskedasticity
### The Goldfeld-Quandt test

```{r}
#| label: "gq1b"
#| echo: false
#| dev: "svg"
#| fig-height: 3.75

ggplot(data = gq_df, aes(
  x = x, y = e,
  color = cut(x, c(-Inf, gq_x, Inf)),
  alpha = cut(x, c(-Inf, gq_x, Inf)),
  shape = cut(x, c(-Inf, gq_x, Inf))
)) +
geom_vline(
  xintercept = gq_x,
  color = grey_mid,
  size = 0.25
) +
geom_point(size = 2.75) +
labs(x = "x", y = "u") +
scale_color_manual(values = c("darkslategrey", grey_mid, red_pink)) +
scale_shape_manual(values = c(19, 1, 19)) +
scale_alpha_manual(values = c(0.5, 0.8, 0.6)) +
theme_axes
```

$F_{375,\,375} = \dfrac{\textcolor{#e64173}{\text{SSE}_2 = `r format(round(sse2, 1), nsmall = 0L, big.mark = ",")`}}{\textcolor{#314f4f}{\text{SSE}_1 = `r format(round(sse1, 1), nsmall = 0L, big.mark = ",")`}} \approx `r format(round(sse2/sse1, 1), nsmall = 0L, big.mark = ",")` \implies$ *p*-value $< 0.001$

$\therefore$ We reject H[0]{.sub }: $\sigma^2_1 = \sigma^2_2$ and conclude there is statistically significant evidence of heteroskedasticity.

---

## Testing for heteroskedasticity
### The Goldfeld-Quandt test

The problem...

---

## Testing for heteroskedasticity
### The Goldfeld-Quandt test

```{r}
#| label: "gq2a"
#| echo: false
#| dev: "svg"
#| fig-height: 4
set.seed(12345)
# Data
gq2_df = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 2 + x^2),
  y = 1 + 3 * x + e
)
# Quantiles
gq_x = quantile(gq2_df$x, probs = c(3/8, 5/8))
# Regressions
sse1b = lm(y ~ x, data = gq2_df %>% filter(x < gq_x[1])) %>%
  residuals() %>% raise_to_power(2) %>% sum()
sse2b = lm(y ~ x, data = gq2_df %>% filter(x > gq_x[2])) %>%
  residuals() %>% raise_to_power(2) %>% sum()
ggplot(data = gq2_df, aes(
  x = x, y = e
)) +
geom_point(color = "darkslategrey", size = 2.75, alpha = 0.5, shape = 19) +
labs(x = "x", y = "u") +
theme_axes
```

---

## Testing for heteroskedasticity
### The Goldfeld-Quandt test

```{r}
#| label: "gq2b"
#| echo: false
#| dev: "svg"
#| fig-height: 4
ggplot(data = gq2_df, aes(
  x = x, y = e,
  color = cut(x, c(-Inf, gq_x, Inf)),
  alpha = cut(x, c(-Inf, gq_x, Inf)),
  shape = cut(x, c(-Inf, gq_x, Inf))
)) +
geom_vline(
  xintercept = gq_x,
  color = grey_mid,
  size = 0.25
) +
geom_point(size = 2.75) +
labs(x = "x", y = "u") +
scale_color_manual(values = c("darkslategrey", grey_mid, red_pink)) +
scale_shape_manual(values = c(19, 1, 19)) +
scale_alpha_manual(values = c(0.5, 0.8, 0.6)) +
theme_axes
```

$F_{375,\,375} = \dfrac{\textcolor{#e64173}{\text{SSE}_2 = `r format(round(sse2b, 1), nsmall = 0L, big.mark = ",")`}}{\textcolor{#314f4f}{\text{SSE}_1 = `r format(round(sse1b, 1), nsmall = 0L, big.mark = ",")`}} \approx `r format(round(sse2b/sse1b, 1), nsmall = 0L, big.mark = ",")` \implies$ *p*-value $\approx `r round(pf(sse2b/sse1b, 375, 375, lower.tail = F), 3)`$

$\therefore$ We fail to reject H[0]{.sub }: $\sigma^2_1 = \sigma^2_2$ while heteroskedasticity is present.

---

## Testing for heteroskedasticity
### The White test

**Breusch and Pagan (1981)** attempted to solve this issue of being too specific with the functional form of the heteroskedasticity.

- Regress $e_i^2$ on $X = \left[ 1,\, x_1,\, x_2,\, \ldots,\, x_k \right]$ and test for joint significance.
- Allows the data to show if/how the variance of $u_i$ correlates with $X$.
- If $\sigma_i^2$ correlates with $X$, then we have heteroskedasticity.

::: {.fragment}
However, we actually want to know if

$$ \sigma_1^2 = \sigma_2^2 = \cdots = \sigma_n^2 $$

**Q:** Can't we just test this hypothesis?
:::

::: {.fragment}
 **A:** Sort of.
:::

---

## Testing for heteroskedasticity
### The White test

Toward this goal, Hal White took advantage of the fact that we can [replace the homoskedasticity requirement with a weaker assumption]{.b .pink }:

- **Old:** $\mathop{\text{Var}} \left( u_i \middle| X \right) = \sigma^2$

- **New:** $u^2$ is *uncorrelated* with the explanatory variables (_i.e._,  $x_j$ for all $j$), their squares (_i.e._, $x_j^2$), and the first-degree interactions (_i.e._, $x_j x_h$).

::: {.fragment}
This new assumption is easier to explicitly test (*hint:* regression).
:::

---

## Testing for heteroskedasticity
### The White test

An outline of White's test for heteroskedasticity:

::: {.pseudocode-small}

1\. Regress y on x[1]{.sub }, x[2]{.sub }, …, x[k]{.sub }. Save residuals e.

2\. Regress squared residuals on all explanatory variables, their squares, and interactions.

$$ e^2 = \alpha_0 + \sum_{h=1}^k \alpha_h x_h + \sum_{j=1}^k \alpha_{k+j} x_j^2 + \sum_{\ell = 1}^{k-1} \sum_{m = \ell + 1}^k \alpha_{\ell,m} x_\ell x_m + v_i $$

3\. Record R[e]{.sub }[2]{.super }.

4\. Calculate test statistic to test H[0]{.sub }: $\alpha_p = 0$ for all $p\neq0$.

:::

---

## Testing for heteroskedasticity
### The White test

White's test statistic is

$$ \text{LM} = n \times R_e^2 \qquad \text{Under H}_0,\, \text{LM} \overset{\text{d}}{\sim} \chi_k^2 $$

where $R^2_e$ comes from the regression of $e^2$ on the explanatory variables, their squares, and their interactions.

$$ e^2 = \alpha_0 + \underbrace{\sum_{h=1}^k \alpha_h x_h}_{\text{Expl. variables}} + \underbrace{\sum_{j=1}^k \alpha_{k+j} x_j^2}_{\text{Squared terms}} + \underbrace{\sum_{\ell = 1}^{k-1} \sum_{m = \ell + 1}^k \alpha_{\ell,m} x_\ell x_m}_{\text{Interactions}} + v_i $$

**Note:** The $k$ (for our $\chi_k^2$) equals the number of estimated parameters in the regression above (the $\alpha_j$), excluding the intercept $\left( \alpha_0 \right)$.

---

## Testing for heteroskedasticity
### The White test

**Practical note:** If a variable is equal to its square (_e.g._, binary variables), then you don't (can't) include it. The same rule applies for interactions.

---

## Testing for heteroskedasticity
### The White test

*Example:* Consider the model^[To simplify notation here, I'm dropping the $i$ subscripts.] $y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + u$

**Step 1:** Estimate the model; obtain residuals $(e)$.

**Step 2:** Regress $e^2$ on explanatory variables, squares, and interactions.
$$
\begin{aligned}
  e^2 =
  &\alpha_0 + \alpha_1 x_1 + \alpha_2 x_2 + \alpha_3 x_3 + \alpha_4 x_1^2 + \alpha_5 x_2^2 + \alpha_6 x_3^2 \\
  &+ \alpha_7 x_1 x_2 + \alpha_8 x_1 x_3 + \alpha_9 x_2 x_3 + v
\end{aligned}
$$

Record the R[2]{.super } from this equation (call it $R_e^2$).

**Step 3:** Test H[0]{.sub }: $\alpha_1 = \alpha_2 = \cdots = \alpha_9 = 0$ using $\text{LM} = n R^2_e \overset{\text{d}}{\sim} \chi_9^2$.

---

## Testing for heteroskedasticity
### The White test

```{r}
#| label: "white1"
#| echo: false
#| dev: "svg"
#| fig-height: 4
set.seed(12345)
# Data
bp_df = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 2 + x^2),
  y = 1 + 3 * x + e
)
# Regressions
lm_bp1 = lm(residuals(lm(y ~ x, bp_df))^2 ~ 1 + bp_df$x) %>%
  summary() %$% r.squared %>% multiply_by(1e3)
lm_bp2 = lm(residuals(lm(y ~ x, bp_df))^2 ~ 1 + bp_df$x + I(bp_df$x^2)) %>%
  summary() %$% r.squared %>% multiply_by(1e3)
# The figure
ggplot(data = bp_df, aes(x = x, y = e)) +
geom_point(size = 2.75, color = "darkslategrey", alpha = 0.5) +
labs(x = "x", y = "u") +
theme_axes
```

The White test for this simple linear regression.

$$
\begin{aligned}
 e_i^2 &= \hat{\alpha}_0 + \hat{\alpha}_1 x_{1i} \textcolor{#e64173}{+ \hat{\alpha}_2 x^2_{1i}} & \widehat{\text{LM}} &= `r round(lm_bp2, 2)` &\mathit{p}\text{-value} < 0.001
\end{aligned}
$$

---

## {.center}

[Testing for Heteroskedasticity]{.bigger .it .slab}

[Examples]{.it .slab .grey-vlight}

---

## Testing for heteroskedasticity
### Examples

**Goal:** Estimate the relationship between standardized test scores (outcome variable) and (1) student-teacher ratio and (2) income, _i.e._,

$$ \left(\text{Test score}\right)_i = \beta_0 + \beta_1 \text{Ratio}_{i} + \beta_2 \text{Income}_{i} + u_i \tag{1} $$

**Potential issue:** Heteroskedasticity... and we do not observe $u_i$.

**Solution:**

1. Estimate the relationship in $(1)$ using OLS.
2. Test for heteroskedasticity.
  - Goldfeld-Quandt
  - White

---

## Testing for heteroskedasticity
### Examples

We will use testing data from the dataset `Caschool` in the `Ecdat` [R]{.mono } package.

```{r}
#| label: "ex-data"
#| echo: true

# Load packages
library(pacman)
p_load(tidyverse, Ecdat)
# Select and rename desired variables; assign to new dataset
test_df = select(Caschool, test_score = testscr, ratio = str, income = avginc)
# Format as tibble
test_df = as_tibble(test_df)
# View first 2 rows of the dataset
head(test_df, 2)
```

---

## Testing for heteroskedasticity
### Examples

Let's begin by estimating our model

$$ \left(\text{Test score}\right)_i = \beta_0 + \beta_1 \text{Ratio}_{i} + \beta_2 \text{Income}_{i} + u_i $$

```{r}
#| label: "ex-gq1"
#| echo: true

# Estimate the model
est_model = lm(test_score ~ ratio + income, data = test_df)
# Summary of the estimate
tidy(est_model)
```

---

## Testing for heteroskedasticity
### Examples

Now, let's see what the residuals suggest about heteroskedasticity

```{r}
#| label: "ex-gq2"
#| echo: true

# Add the residuals to our dataset
test_df$e = residuals(est_model)
```

```{r}
#| label: "gq3"
#| echo: false
#| dev: "svg"
#| fig-height: 4.5

# Plot residuals against income
plot1 = ggplot(data = test_df, aes(x = income, y = e)) +
  geom_point(size = 2.5, alpha = 0.5, color = red_pink) +
  labs(x = "Income", y = TeX("\\textit{e}")) +
  theme_axes
# Plot residuals against student-teacher ratio
plot2 = ggplot(data = test_df, aes(x = ratio, y = e)) +
  geom_point(size = 2.5, alpha = 0.5, color = "darkslategrey") +
  labs(x = "Student-to-teacher ratio", y = TeX("\\textit{e}")) +
  theme_axes
# The grid
gridExtra::grid.arrange(plot1, plot2, nrow = 1)
```

---

## Testing for heteroskedasticity
### Example: Goldfeld-Quandt

Income looks potentially heteroskedastic; let's test via Goldfeld-Quandt.

```{r}
#| label: "ex-gq4"
#| echo: true
#| eval: false

# Arrange the data by income
test_df = arrange(test_df, income)
```

---

## Testing for heteroskedasticity
### Example: Goldfeld-Quandt

Income looks potentially heteroskedastic; let's test via Goldfeld-Quandt.

```{r}
#| label: "ex-gq5"
#| echo: true
#| eval: false

# Arrange the data by income
test_df = arrange(test_df, income)
# Re-estimate the model for the last and first 158 observations
est_model1 = lm(test_score ~ ratio + income, data = tail(test_df, 158))
est_model2 = lm(test_score ~ ratio + income, data = head(test_df, 158))
```

---

## Testing for heteroskedasticity
### Example: Goldfeld-Quandt

Income looks potentially heteroskedastic; let's test via Goldfeld-Quandt.

```{r}
#| label: "ex-gq6"
#| echo: true
#| eval: false

# Arrange the data by income
test_df = arrange(test_df, income)
# Re-estimate the model for the last and first 158 observations
est_model1 = lm(test_score ~ ratio + income, data = tail(test_df, 158))
est_model2 = lm(test_score ~ ratio + income, data = head(test_df, 158))
# Grab the residuals from each regression
e_model1 = residuals(est_model1)
e_model2 = residuals(est_model2)
```

---

## Testing for heteroskedasticity
### Example: Goldfeld-Quandt

Income looks potentially heteroskedastic; let's test via Goldfeld-Quandt.

```{r}
#| label: "ex-gq7"
#| echo: true

# Arrange the data by income
test_df = arrange(test_df, income)
# Re-estimate the model for the last and first 158 observations
est_model1 = lm(test_score ~ ratio + income, data = tail(test_df, 158))
est_model2 = lm(test_score ~ ratio + income, data = head(test_df, 158))
# Grab the residuals from each regression
e_model1 = residuals(est_model1)
e_model2 = residuals(est_model2)
# Calculate SSE for each regression
(sse_model1 = sum(e_model1^2))
(sse_model2 = sum(e_model2^2))
```

---

## Testing for heteroskedasticity
### Example: Goldfeld-Quandt

Remember the Goldfeld-Quandt test statistic?

$F_{n^\star-k,\,n^\star-k} = \dfrac{\text{SSE}_2}{\text{SSE}_1}$
[$\approx\dfrac{`r round(sse_model2, 2) %>% format(big.mark = ",")`}{`r round(sse_model1, 2) %>% format(big.mark = ",")`}$]{.fragment}
[$\approx`r round(sse_model2/sse_model1, 2)`$]{.fragment}

::: {.fragment}
Test via $F_{158-3,\,158-3}$
:::

::: {.fragment}
```{r}
#| label: "ex-gq8"
#| echo: true

# G-Q test statistic
(f_gq = sse_model2/sse_model1)
```
:::

::: {.fragment}
```{r}
#| label: "ex-gq9"
#| echo: true

# p-value
pf(q = f_gq, df1 = 158-3, df2 = 158-3, lower.tail = F)
```
:::

---

## Testing for heteroskedasticity
### Example: Goldfeld-Quandt

The Goldfeld-Quandt test statistic and its null distribution

```{r}
#| label: "ex-gq10"
#| echo: false
#| dev: "svg"
#| fig-height: 5
ggplot(data = tibble(x = c(0, f_gq * 1.3)), aes(x)) +
  geom_area(
    stat = "function", fun = df, args = list(df1 = 158, df2 = 158),
    fill = "darkslategrey", alpha = 0.05
  ) +
  geom_area(
    stat = "function", fun = df, args = list(df1 = 158, df2 = 158),
    fill = red_pink, alpha = 0.85,
    xlim = c(f_gq, f_gq * 1.3)
  ) +
  stat_function(
    fun = df, args = list(df1 = 158, df2 = 158), n = 1e3,
    color = "darkslategrey"
  ) +
  geom_vline(xintercept = f_gq, color = grey_dark, size = 0.5, linetype = "dotted") +
  annotate(
    "text", x = f_gq, y = 5 * df(f_gq, df1 = 158, df2 = 158),
    label = TeX("$\\widehat{F}=1.53$"),
    family = "MathJax_Main", size = 7
  ) +
  labs(x = "x", y = "f") +
  theme_axes
```

---

## Testing for heteroskedasticity
### Example: Goldfeld-Quandt

Putting it all together:

H[0]{.sub }: $\sigma_1^2 = \sigma_2^2$ *vs.* H[A]{.sub }: $\sigma_1^2 \neq \sigma_2^2$

Goldfeld-Quandt test statistic: $F \approx `r round(f_gq, 2)`$

*p*-value $\approx `r round(pf(q = f_gq, df1 = 158-3, df2 = 158-3, lower.tail = F), 5)`$

∴ Reject H[0]{.sub } (*p*-value is less than 0.05).

**Conclusion:** We find statistically significant evidence that $\sigma_1^2 \neq \sigma_2^2$.<br>There is statistically significant evidence of heteroskedasticity (at the 5\% level).

---

## Testing for heteroskedasticity
### Example: Goldfeld-Quandt

What if we had chosen to focus on student-to-teacher ratio?

::: {.fragment}
```{r}
#| label: "ex-gq11"
#| echo: true

# Arrange the data by ratio
test_df = arrange(test_df, ratio)
# Re-estimate the model for the last and first 158 observations
est_model3 = lm(test_score ~ ratio + income, data = tail(test_df, 158))
est_model4 = lm(test_score ~ ratio + income, data = head(test_df, 158))
# Grab the residuals from each regression
e_model3 = residuals(est_model3)
e_model4 = residuals(est_model4)
# Calculate SSE for each regression
(sse_model3 = sum(e_model3^2))
(sse_model4 = sum(e_model4^2))
```
:::

---

## Testing for heteroskedasticity
### Example: Goldfeld-Quandt

$F_{n^\star-k,\,n^\star-k} = \dfrac{\text{SSE}_4}{\text{SSE}_3} \approx\dfrac{`r round(sse_model4, 2) %>% format(big.mark = ",")`}{`r round(sse_model3, 2) %>% format(big.mark = ",")`} \approx`r round(sse_model4 / sse_model3, 2)`$

which has a *p*-value of approximately `r round(pf(sse_model4 / sse_model3, 158-3, 158-3, lower.tail = F), 4)`.

::: {.fragment}
∴ We would have failed to reject H[0]{.sub }, concluding that we failed to find statistically significant evidence of heteroskedasticity.
:::

::: {.fragment}
**Lesson:** Understand the limitations of estimators, tests, *etc.*
:::

---

## Heteroskedasticity
### Example: White

Let's test the same model and data with the White test.

*Recall:* We saved our residuals as `e` in our dataset, _i.e._,

```{r}
#| label: "white-recall"
#| eval: false
#| echo: true

# Estimate the model
est_model = lm(test_score ~ ratio + income, data = test_df)
# Add the residuals to our dataset
test_df$e = residuals(est_model)
```

---

## Heteroskedasticity
### Example: White

The [**White test** adds squared terms and interactions]{.pink } to initial regression specification (the right-hand side)
$$
\begin{align}
 \textcolor{#314f4f}{u_{i}^2} =& \textcolor{#314f4f}{\alpha_{0} + \alpha_{1} \text{Ratio}_{i} + \alpha_{2} \text{Income}_{i} } \\
 &+ \textcolor{#e64173}{\alpha_{3} \text{Ratio}_{i}^2 + \alpha_{4} \text{Income}_{i}^2 + \alpha_{5} \text{Ratio}_{i}\times\text{Income}_{i}} \\
 &+ \textcolor{#314f4f}{w_{i}}
\end{align}
$$

The [**White test**]{.pink } tests the null hypothesis
<br>[H[0]{.sub }:]{.pink } $\textcolor{#e64173}{\alpha_1 = \alpha_2 = \alpha_3 = \alpha_4 = \alpha_5 = 0}$

We just need to write some [R]{.mono } code to test [H[0]{.sub }]{.pink }.

---

## Heteroskedasticity
### Example: White

*Aside:* [R]{.mono } has funky notation for squared terms and interactions in `lm()`:

- **Squared terms** use `I()`, _e.g._, `lm(y ~ I(x^2))`

- **Interactions** use `:` between the variables, _e.g._, `lm(y ~ x1:x2)`

*Example:* Regress `y` on quadratic of `x1` and `x2`:

```{r}
#| label: "ex-quad"
#| eval: false
#| echo: true

# Pretend quadratic regression w/ interactions
lm(y ~ x1 + x2 + I(x1^2) + I(x2^2) + x1:x2, data = pretend_df)
```

---

## Heteroskedasticity
### Example: White

**Step 1:** Regress $e_i^2$ on 1[st]{.super } degree, 2[nd]{.super } degree, and interactions

```{r}
#| label: "ex-white1"
#| eval: false
#| echo: true

# Regress squared residuals on quadratic of explanatory variables
white_model = lm(
  I(e^2) ~ ratio + income + I(ratio^2) + I(income^2) + ratio:income,
  data = test_df
)
# Grab the R-squared
(white_r2 = summary(white_model)$r.squared)
```

---

## Heteroskedasticity
### Example: White

**Step 2:** Collect $R_e^2$ from the regression.

```{r}
#| label: "ex-white2"
#| eval: true
#| echo: true

# Regress squared residuals on quadratic of explanatory variables
white_model = lm(
  I(e^2) ~ ratio + income + I(ratio^2) + I(income^2) + ratio:income,
  data = test_df
)
# Grab the R-squared
(white_r2 = summary(white_model)$r.squared)
```

---

## Heteroskedasticity
### Example: White

**Step 3:** Calculate White test statistic $\text{LM} = n \times R_e^2 \approx 420 \times `r round(white_r2, 3)`$

```{r}
#| label: "ex-white3"
#| eval: true
#| echo: true

# Regress squared residuals on quadratic of explanatory variables
white_model = lm(
  I(e^2) ~ ratio + income + I(ratio^2) + I(income^2) + ratio:income,
  data = test_df
)
# Grab the R-squared
white_r2 = summary(white_model)$r.squared
# Calculate the White test statistic
(white_stat = 420 * white_r2)
```

---

## Heteroskedasticity
### Example: White

**Step 4:** Calculate the associated *p*-value (where $\text{LM} \overset{d}{\sim} \chi_k^2$); here, $k=5$

```{r}
#| label: "ex-white4"
#| eval: true
#| echo: true

# Regress squared residuals on quadratic of explanatory variables
white_model = lm(
  I(e^2) ~ ratio + income + I(ratio^2) + I(income^2) + ratio:income,
  data = test_df
)
# Grab the R-squared
white_r2 = summary(white_model)$r.squared
# Calculate the White test statistic
white_stat = 420 * white_r2
# Calculate the p-value
pchisq(q = white_stat, df = 5, lower.tail = F)
```

---

## Heteroskedasticity
### Example: White

Putting everything together...

[H[0]{.sub }: $\alpha_1 = \alpha_2 = \alpha_3 = \alpha_4 = \alpha_5 = 0$]{.fragment fragment-index=0}
[ *vs.* H[A]{.sub }: $\alpha_i \neq 0$ for some $i \in \{1,\, 2,\,\ldots,\, 5\}$]{.fragment fragment-index=1}

::: {.fragment fragment-index=1}
from the model
$$
\begin{align}
u_i^2 =& \alpha_0 + \alpha_1 \text{Ratio}_i + \alpha_2 \text{Income}_i \\
&+ \alpha_3 \text{Ratio}_i^2 + \alpha_4 \text{Income}_i^2 \\
&+ \alpha_5 \text{Ratio}_i \times \text{Income}_i + w_i
\end{align}
$$
:::

::: {.fragment fragment-index=2}
Our White test statistic: $\text{LM} = n \times R_e^2 \approx 420 \times `r round(white_r2, 3)` \approx `r round(white_stat, 2)`$
:::

::: {.fragment fragment-index=3}
Under the $\chi_5^2$ distribution, this $\widehat{\text{LM}}$ has a *p*-value less than 0.001.
:::

::: {.fragment fragment-index=4}
∴ We [reject H[0]{.sub }]{.b .pink } and conclude there is [statistically significant evidence of heteroskedasticity]{.b .pink } (at the 5-percent level).
:::

---

## Heteroskedasticity
### Example: White

The White test statistic and its null distribution

```{r}
#| label: "ex-white-plot"
#| echo: false
#| dev: "svg"
#| fig-height: 5
ggplot(data = tibble(x = c(0, white_stat * 1.1)), aes(x)) +
  geom_area(
    stat = "function", fun = dchisq, args = list(df = 5),
    fill = "darkslategrey", alpha = 0.05
  ) +
  geom_area(
    stat = "function", fun = dchisq, args = list(df = 5),
    fill = red_pink, alpha = 0.85,
    xlim = c(white_stat, white_stat * 1.1)
  ) +
  stat_function(
    fun = dchisq, args = list(df = 5), n = 1e3,
    color = "darkslategrey"
  ) +
  geom_vline(xintercept = white_stat, color = grey_dark, size = 0.5, linetype = "dotted") +
  annotate(
    "text", x = white_stat, y = 0.01,
    label = TeX("$\\widehat{LM}=30.8$"),
    family = "MathJax_Main", size = 7,
    hjust = 0.5
  ) +
  labs(x = "x", y = "f") +
  theme_axes
```

---

## Heteroskedasticity
### Review questions

::: {.fragment}
- **Q:** What is the definition of heteroskedasticity?

- **Q:** Why are we concerned about heteroskedasticity?

- **Q:** Does plotting $y$ against $x$, tell us anything about heteroskedasticity?

- **Q:** Does plotting $e$ against $x$, tell us anything about heteroskedasticity?

- **Q:** Since we cannot observe the $u_i$'s, what do we use to *learn about* heteroskedasticity?

- **Q:** Which test do you recommend to test for heteroskedasticity? Why?
:::

---

## Heteroskedasticity
### Review questions

- **Q:** What is the definition of heteroskedasticity?

::: {.fragment}
- **A:**
<br>[Math:]{.b .pink } $\mathop{\text{Var}} \left( u_i | X \right) \neq \mathop{\text{Var}} \left( u_j | X \right)$ for some $i\neq j$.
<br>[Words:]{.b .pink } There is a systematic relationship between the variance of $u_i$ and our explanatory variables.
:::

---

## Heteroskedasticity
### Review questions

::: {.grey-vlight}

- **Q:** What is the definition of heteroskedasticity?

:::

- **Q:** Why are we concerned about heteroskedasticity?

::: {.fragment}
- **A:** It biases our standard errors—wrecking our statistical tests and confidence intervals. Also: OLS is no longer the most efficient (best) linear unbiased estimator.
:::

---

## Heteroskedasticity
### Review questions

::: {.grey-vlight}

- **Q:** What is the definition of heteroskedasticity?

- **Q:** Why are we concerned about heteroskedasticity?

:::

- **Q:** Does plotting $y$ against $x$, tell us anything about heteroskedasticity?

::: {.fragment}
- **A:** It's not exactly what we want, but since $y$ is a function of $x$ and $u$, it can still be informative. If $y$ becomes more/less disperse as $x$ changes, we likely have heteroskedasticity.
:::

---

## Heteroskedasticity
### Review questions

::: {.grey-vlight}

- **Q:** What is the definition of heteroskedasticity?

- **Q:** Why are we concerned about heteroskedasticity?

- **Q:** Does plotting $y$ against $x$, tell us anything about heteroskedasticity?

:::

- **Q:** Does plotting $e$ against $x$, tell us anything about heteroskedasticity?

::: {.fragment}
- **A:** Yes. The spread of $e$ depicts its variance—and tells us something about the variance of $u$. Trends in this variance, along $x$, suggest heteroskedasticity.
:::

---

## Heteroskedasticity
### Review questions

::: {.grey-vlight}

- **Q:** What is the definition of heteroskedasticity?

- **Q:** Why are we concerned about heteroskedasticity?

- **Q:** Does plotting $y$ against $x$, tell us anything about heteroskedasticity?

- **Q:** Does plotting $e$ against $x$, tell us anything about heteroskedasticity?

:::

- **Q:** Since we cannot observe the $u_i$'s, what do we use to *learn about* heteroskedasticity?

::: {.fragment}
- **A:** We use the $e_i$'s to predict/learn about the $u_i$'s. This trick is key for almost everything we do with heteroskedasticity testing/correction.
:::

---

## Heteroskedasticity
### Review questions

::: {.grey-vlight}

- **Q:** What is the definition of heteroskedasticity?

- **Q:** Why are we concerned about heteroskedasticity?

- **Q:** Does plotting $y$ against $x$, tell us anything about heteroskedasticity?

- **Q:** Does plotting $e$ against $x$, tell us anything about heteroskedasticity?

- **Q:** Since we cannot observe the $u_i$'s, what do we use to *learn about* heteroskedasticity?

:::

- **Q:** Which test do you recommend to test for heteroskedasticity? Why?

::: {.fragment}
- **A:** I like White. Fewer assumptions. Fewer issues.
:::

---

## {.center}

*Next time:* Living/working with heteroskedasticity.

---

## {.center}

[Appendix]{.bigger .it .slab}

---

## {.clear}

One more test...

---

## Testing for heteroskedasticity
### The Breusch-Pagan test

Breusch and Pagan (1981) attempted to solve this issue of being too specific with the functional form of the heteroskedasticity.

- Allows the data to show if/how the variance of $u_i$ correlates with $X$.

- If $\sigma_i^2$ correlates with $X$, then we have heteroskedasticity.

- Regresses $e_i^2$ on $X = \left[ 1,\, x_1,\, x_2,\, \ldots,\, x_k \right]$ and tests for joint significance.

---

## Testing for heteroskedasticity
### The Breusch-Pagan test

How to implement:

::: {.pseudocode-small}

1\. Regress y on an intercept, x[1]{.sub }, x[2]{.sub }, …, x[k]{.sub }.

2\. Record residuals e.

3\. Regress e[2]{.super } on an intercept, x[1]{.sub }, x[2]{.sub }, …, x[k]{.sub }.

$$ e_i^2 = \alpha_0 + \alpha_1 x_{1i} + \alpha_2 x_{2i} + \cdots + \alpha_k x_{ki} + v_i $$

4\. Record R[2]{.super }.

5\. Test hypothesis H[0]{.sub }: $\alpha_1 = \alpha_2 = \cdots = \alpha_k = 0$

:::

---

## Testing for heteroskedasticity
### The Breusch-Pagan test

The B-P test statistic^[This specific form of the test statistic actually comes form Koenker (1981).] is

$$ \text{LM} = n \times R^2_{e} $$

where $R^2_e$ is the $R^2$ from the regression

$$ e_i^2 = \alpha_0 + \alpha_1 x_{1i} + \alpha_2 x_{2i} + \cdots + \alpha_k x_{ki} + v_i $$

Under the null, $\text{LM}$ is asymptotically distributed as $\chi^2_k$.

::: {.fragment}
This test statistic tests H[0]{.sub }: $\alpha_1 = \alpha_2 = \cdots = \alpha_k = 0$.

Rejecting the null hypothesis implies evidence of heteroskedasticity.
:::

---

## Testing for heteroskedasticity
### The Breusch-Pagan test

**Problem:** We're still assuming a fairly restrictive [functional form]{.b .pink } between our explanatory variables $X$ and the variances of our disturbances $\sigma^2_i$.

::: {.fragment}
**Result:** B-P *may* still miss fairly simple forms of heteroskedasticity.
:::

---

## Testing for heteroskedasticity
### The Breusch-Pagan test

Breusch-Pagan tests are still [sensitive to functional form]{.b .pink }.

```{r}
#| label: "bp1"
#| echo: false
#| dev: "svg"
#| fig-height: 3.75

set.seed(12345)
# Data
bp_df = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 2 + x^2),
  y = 1 + 3 * x + e
)
# Regressions
lm_bp1 = lm(residuals(lm(y ~ x, bp_df))^2 ~ 1 + bp_df$x) %>%
  summary() %$% r.squared %>% multiply_by(1e3)
lm_bp2 = lm(residuals(lm(y ~ x, bp_df))^2 ~ 1 + bp_df$x + I(bp_df$x^2)) %>%
  summary() %$% r.squared %>% multiply_by(1e3)
# The figure
ggplot(data = bp_df, aes(x = x, y = e)) +
geom_point(size = 2.75, color = "darkslategrey", alpha = 0.5) +
labs(x = "x", y = "u") +
theme_axes
```

$$
\begin{aligned}
  e_i^2 &= \hat{\alpha}_0 + \hat{\alpha}_1 x_{1i} & \widehat{\text{LM}} &= `r round(lm_bp1, 2)` &\mathit{p}\text{-value} \approx `r round(pchisq(lm_bp1, 1, lower.tail = F), 3)` \\
  e_i^2 &= \hat{\alpha}_0 + \hat{\alpha}_1 x_{1i} \textcolor{#e64173}{+ \hat{\alpha}_2 x^2_{1i}} & \widehat{\text{LM}} &= `r round(lm_bp2, 2)` &\mathit{p}\text{-value} < 0.001
\end{aligned}
$$

---

## Testing for heteroskedasticity
### Example: Breusch-Pagan

Let's test the same model with the Breusch Pagan.

*Recall:* We saved our residuals as `e` in our dataset, _i.e._,

```{r}
#| label: "bp-recall"
#| eval: false
#| echo: true

test_df$e = residuals(est_model)
```

---

## Testing for heteroskedasticity
### Example: Breusch-Pagan

In B-P, we first regress $e_i^2$ on the explanatory variables,

```{r}
#| label: "ex-bp1"
#| echo: true

# Regress squared residuals on explanatory variables
bp_model = lm(I(e^2) ~ ratio + income, data = test_df)
```

---

## Testing for heteroskedasticity
### Example: Breusch-Pagan

and use the resulting $R^2$ to calculate a test statistic.

```{r}
#| label: "ex-bp2"
#| echo: true

# Regress squared residuals on explanatory variables
bp_model = lm(I(e^2) ~ ratio + income, data = test_df)
# Grab the R-squared
(bp_r2 = summary(bp_model)$r.squared)
```

---

## Testing for heteroskedasticity
### Example: Breusch-Pagan

The Breusch-Pagan test statistic is

$\text{LM} = n \times R^2_e$

::: {.fragment}
 $\approx 420 \times `r bp_r2 %>% round(7) %>% format(nsmall = 7, scientific = F)`$
:::

::: {.fragment}
 $\approx `r round(nrow(test_df) * bp_r2, 4)`$

which we test against a $\chi_k^2$ distribution (here: $k=2$).^[$k$ is the number of explanatory variables (excluding the intercept).]
:::

::: {.fragment}
```{r}
#| label: "ex-bp3"
#| echo: true

# B-P test statistic
bp_stat = 420 * bp_r2
# Calculate the p-value
pchisq(q = bp_stat, df = 2, lower.tail = F)
```
:::

---

## Testing for heteroskedasticity
### Example: Breusch-Pagan

H[0]{.sub }: $\alpha_1 = \alpha_2 = 0$ *vs.* H[A]{.sub }: $\alpha_1 \neq 0$ and/or $\alpha_2 \neq 0$
for the model $u_i^2 = \alpha_0 + \alpha_1 \text{Ratio}_i + \alpha_2 \text{Income}_i + w_i$

Breusch-Pagan test statistic: $\widehat{\text{LM}} \approx `r round(bp_stat, 3)`$

*p*-value $\approx `r pchisq(q = bp_stat, df = 2, lower.tail = F) %>% round(3)`$

∴ Fail to reject H[0]{.sub } (the *p*-value is greater than 0.05)

**Conclusion:** We do not find statistically significant evidence of heteroskedasticity at the 5-percent level.

(We find no evidence of a *linear* relationship between $u_i^2$ and the explanatory variables.)

---

## Testing for heteroskedasticity
### Example: Breusch-Pagan

The Breusch-Pagan test statistic and its null distribution

```{r}
#| label: "ex-bp-plot"
#| echo: false
#| dev: "svg"
#| fig-height: 5
ggplot(data = tibble(x = c(0, 7)), aes(x)) +
  geom_area(
    stat = "function", fun = dchisq, args = list(df = 2),
    fill = "darkslategrey", alpha = 0.05
  ) +
  geom_area(
    stat = "function", fun = dchisq, args = list(df = 2),
    fill = red_pink, alpha = 0.85,
    xlim = c(bp_stat, 7)
  ) +
  stat_function(
    fun = dchisq, args = list(df = 2), n = 1e3,
    color = "darkslategrey"
  ) +
  geom_vline(xintercept = bp_stat, color = grey_dark, size = 0.5, linetype = "dotted") +
  annotate(
    "text", x = bp_stat, y = 5 * dchisq(bp_stat, df = 2),
    label = TeX("$\\widehat{LM}=0.014$"),
    family = "MathJax_Main", size = 7,
    hjust = 0
  ) +
  labs(x = "x", y = "f") +
  theme_axes
```
