---
title: 'Heteroskedasticity, Part II'
subtitle: 'EC 421, Set 05'
author:
  - name: Ed Rubin
    email: edwardr@uoregon.edu
editor:
  render-on-save: true
format:
  clean-revealjs:
    slide-level: 2
    theme: [../my.scss, default]
    highlight-style: gruvbox
    embed-resources: true
    slide-number: c/t
    incremental: false
    controls-layout: edges
    transition: none
    preview-links: auto
    html-math-method: mathjax
    code-line-numbers: false
knitr:
  opts_chunk:
    comment: "#>"
    fig-align: "center"
    fig-height: 7
    fig-width: 10.5
    warning: false
    message: false
    fig-format: svg
execute:
  warning: false
  message: false
---

## [R]{.mono } showcase

```{r}
#| label: "setup"
#| include: false

# General options
options(htmltools.dir.version = FALSE)
# Load package
library(pacman)
p_load(
  broom,
  latex2exp, ggplot2, ggthemes, viridis, extrafont,
  kableExtra,
  dplyr, magrittr, knitr, parallel
)
# Define colors
red_pink   = "#e64173"
turquoise  = "#20B2AA"
orange     = "#FFA500"
red        = "#fb6107"
blue       = "#3b3b9a"
green      = "#8bb174"
grey_light = "grey70"
grey_mid   = "grey50"
grey_dark  = "grey20"
purple     = "#6A5ACD"
slate      = "#23373b"
# Knitr options
# opts_chunk$set(
#   comment = "#>",
#   fig.align = "center",
#   fig.height = 7,
#   fig.width = 10.5,
#   warning = FALSE,
#   message = FALSE
# )
# opts_chunk$set(dev = "svg")
# options(device = function(file, width, height) {
#   svg(tempfile(), width = width, height = height)
# })
# A simple theme for math plots
theme_axes =
  theme_void(base_family = "Fira Sans Condensed") +
  theme(
    text =
      element_text(family = "Fira Sans Condensed"),
    axis.title =
      element_text(size = 20, family = "Fira Sans Condensed"),
    axis.title.x =
      element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines"), family = "Fira Sans Condensed"),
    axis.title.y =
      element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines"), family = "Fira Sans Condensed"),
    axis.line =
      element_line(
        color = "grey70",
        size = 0.25,
        arrow = arrow(angle = 30, length = unit(0.15, "inches")
      )),
    plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
    legend.position = "none"
  )
```

```{css}
#| echo: false

@media print {
  .has-continuation {
    display: block !important;
  }
}
```


**[Posit Cloud](https://posit.cloud)**

- Cloud-based RStudio IDE.
- Free tier available. Also student plans.
- No installation of R/RStudio required.
- Easily share projects.
- Runs in your browser. Code on your phone!

---

## {.center}

[Prologue]{.bigger .it .slab}

---

## Schedule


### Last Time

Heteroskedasticity: Issues and tests

### Today

- [First assignment]{.b .pink } due Sunday.
- Living with heteroskedasticity.

### Upcoming

- [Second assignment]{.b .pink } coming soon.

---

## EC 421

### Goals

- Develop [intuition]{.b .pink } for econometrics.
- Learn how to [*apply*]{.b .pink } econometrics—strengths, weaknessed, *etc.*
- Learn [[R]{.mono }]{.b .pink }.

::: {.fragment}
[R]{.mono } does the calculations and has already memorized the formulas.

I want you to know what the formulas mean, when/why we use them, and when they fail/work.
:::

::: {.fragment}
This course has the potential to be one of the most useful/valuable/applicable/marketable classes that you take at UO.
:::

---

## {.center}

[Heteroskedasticity]{.bigger .it .slab}

[Review]{.it .slab .grey-vlight}

---

## Heteroskedasticity
### Review

Three review questions

**Question 1:** What is the difference between $u_i$ and $e_i$?

**Question 2:** We spend *a lot* of time discussing $u_i^2$. Why?

**Question 3:** We also spend *a lot* of time discussing $e_i^2$. Why?

---

## Heteroskedasticity
### Review

**Question 1:** What is the difference between $u_i$ and $e_i$?

**Answer 1:** [$\textcolor{#e64173}{u_i}$ gives the [population disturbance]{.b .pink } for the $i$[th]{.super } observation.]{.fragment}

::: {.fragment}
 $u_i$ measures how far the $i$[th]{.super } observation is from the [population]{.b .pink } line, _i.e._,

$$ u_i = y_i - \textcolor{#e64173}{\underbrace{\left(\beta_0 + \beta_1 x_i\right)}_{\text{Population line}}} $$
:::

::: {.fragment}
$\textcolor{#6A5ACD}{e_i}$ gives the [regression residual (error)]{.b .purple } for the $i$[th]{.super } observation.
:::

::: {.fragment}
 $e_i$ measures how far the $i$[th]{.super } observation is from the [sample regression]{.b .purple } line, _i.e._,

 $$ e_i = y_i - \textcolor{#6A5ACD}{\underbrace{\left(\hat{\beta}_0 + \hat{\beta}_1 x_i\right)}_{\text{Sample reg. line} = \hat{y}}} = y_i - \textcolor{#6A5ACD}{\hat{y}_i} $$
:::

---

## Heteroskedasticity
### Review

**Question 2:** We spend *a lot* of time discussing $u_i^2$. Why?

**Answer 2:**

One of major assumptions is that our disturbances (the $u_i$'s) are homoskedastic (they have constant variance), _i.e._, $\mathop{\text{Var}} \left( u_i \middle| x_i \right) = \sigma^2$.

We also assume that the mean of these disturbances is zero, $\textcolor{#e64173}{\mathop{\boldsymbol{E}}\left[ u_i \middle| x_i \right] = 0}$.

By definition, $\mathop{\text{Var}} \left( u_i \middle| x_i \right) = \mathop{\boldsymbol{E}}\Big[ u_i^2 - \underbrace{\textcolor{#e64173}{\mathop{\boldsymbol{E}}\left[ u_i \middle| x_i \right]}^2}_{\textcolor{#e64173}{=0}} \Big| x_i \Big] = \mathop{\boldsymbol{E}}\left[ u_i^2 \middle| x_i \right]$

Thus, if we want to learn about the variance of $u_i$, we can focus on $u_i^2$.

---

## Heteroskedasticity
### Review

**Question 3:** We also spend *a lot* of time discussing $e_i^2$. Why?

**Answer 3:**

We cannot observe $u_i$ (or $u_i^2$).

But $u_i^2$ tells us about the variance of $u_i$.

We use $e_i^2$ to learn about $u_i^2$ and, consequently, $\sigma_i^2$.

---

## Heteroskedasticity
### Review: Current assumptions

::: {.fragment}
1. Our sample (the $x_k$'s and $y_i$) was [randomly drawn]{.b .pink } from the population.
:::

::: {.fragment}
2. $y$ is a [linear function]{.b .pink } of the $\beta_k$'s and $u_i$.
:::

::: {.fragment}
3. There is no perfect [multicollinearity]{.b .pink } in our sample.
:::

::: {.fragment}
4. The explanatory variables are [exogenous]{.b .pink }: $\mathop{\boldsymbol{E}}\left[ u \middle| X \right] = 0 \left(\implies \mathop{\boldsymbol{E}}\left[ u \right] = 0\right)$.
:::

::: {.fragment}
5. The disurbances have [constant variance]{.b .pink } $\sigma^2$ and [zero covariance]{.b .pink }, _i.e._,
    - $\mathop{\boldsymbol{E}}\left[ u_i^2 \middle| X_i \right] = \mathop{\text{Var}} \left( u_i \middle| X_i \right) = \sigma^2 \implies \mathop{\text{Var}} \left( u_i \right) = \sigma^2$
    - $\mathop{\text{Cov}} \left( u_i, \, u_j \middle| X_i,\,X_j \right) = \mathop{\boldsymbol{E}}\left[ u_i u_j \middle| X_i,\,X_j \right] = 0$ for $i\neq j$
:::

::: {.fragment}
6. The disturbances come from a [Normal]{.b .pink } distribution, _i.e._, $u_i \overset{\text{iid}}{\sim} \mathop{\text{N}}\left( 0, \sigma^2 \right)$.
:::

---

## Heteroskedasticity
### Review

We're currently focused on assumption \#5:

5\. The disurbances have [constant variance]{.b .pink } $\sigma^2$ and [zero covariance]{.b .pink }, _i.e._,

  - $\mathop{\boldsymbol{E}}\left[ u_i^2 \middle| X_i \right] = \mathop{\text{Var}} \left( u_i \middle| X \right) = \sigma^2 \implies \mathop{\text{Var}} \left( u_i \right) = \sigma^2$
  - $\mathop{\text{Cov}} \left( u_i, \, u_j \middle| X_i,\,X_j \right) = \mathop{\boldsymbol{E}}\left[ u_i u_j \middle| X_i,\,X_j \right] = 0$ for $i\neq j$

Specifically, we're discussing the assumption of [constant variance]{.b .pink } (also known as *homoskedasticity*).

**Violation of this assumption:** Our disturbances have different variances.
<br>
[Heteroskedasticity:]{.b .pink } $\mathop{\text{Var}} \left( u_i \right) = \sigma^2_i$ and $\sigma^2_i \neq \sigma^2_j$ for some $i\neq j$.

---

## Heteroskedasticity
### Review

Classic example of heteroskedasticity: The funnel

Variance of $u$ increases with $x$

```{r}
#| label: "het-ex1"
#| dev: "svg"
#| echo: false
#| fig-height: 4.5
set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 4 + 1.5 * x)
), aes(x = x, y = e)) +
geom_point(color = "darkslategrey", size = 2.75, alpha = 0.5) +
labs(x = "x", y = "u") +
theme_axes
```

---

## Heteroskedasticity
### Review

Another example of heteroskedasticity: (double funnel?)

Variance of $u$ increasing at the extremes of $x$

```{r}
#| label: "het-ex2"
#| dev: "svg"
#| echo: false
#| fig-height: 4.5
set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 2 + x^2)
), aes(x = x, y = e)) +
geom_point(color = "darkslategrey", size = 2.75, alpha = 0.5) +
labs(x = "x", y = "u") +
theme_axes
```

---

## Heteroskedasticity
### Review

Another example of heteroskedasticity:

Differing variances of $u$ by group

```{r}
#| label: "het-ex3"
#| dev: "svg"
#| echo: false
#| fig-height: 4.5
set.seed(12345)
ggplot(data = tibble(
  g = sample(c(F,T), 1e3, replace = T),
  x = runif(1e3, -3, 3),
  e = rnorm(1e3, 0, sd = 0.5 + 2 * g)
), aes(x = x, y = e, color = g, shape = g, alpha = g)) +
geom_point(size = 2.75) +
scale_color_manual(values = c("darkslategrey", red_pink)) +
scale_shape_manual(values = c(16, 1)) +
scale_alpha_manual(values = c(0.5, 0.8)) +
labs(x = "x", y = "u") +
theme_axes
```

---

## Heteroskedasticity
### Review

[Heteroskedasticity]{.b .pink } is present when the variance of $u$ changes with any combination of our explanatory variables $x_1$ through $x_k$.

---

## Testing for heteroskedasticity

We have some tests that may help us detect heteroskedasticity.

- Goldfeld-Quandt
- White
- (There are others, *e.g.*, Breusch-Pagan)

::: {.fragment}
What do we do if we detect it?
:::

---

## {.center}

[Living with heteroskedasticity]{.bigger .it .slab}

---

## Living with heteroskedasticity

In the presence of heteroskedasticity, OLS is

- still [unbiased]{.b .pink }
- [no longer the most efficient]{.b .pink } unbiased linear estimator

On average, we get the right answer but with more noise (less precision).
<br> *Also:* Our standard errors are biased.

::: {.fragment}
**Options:**

1. Check regression [specification]{.b .pink }.
2. Find a new, more efficient [unbiased estimator]{.b .pink } for $\beta_j$'s.
3. Live with OLS's inefficiency; find a [new variance estimator]{.b .pink } to fix<br> standard errors, conf. intervals, hyp. tests.
:::

---

## Living with heteroskedasticity 
### Misspecification

As we've discussed, the specification^[*Specification:* Functional form and included variables.] of your regression model matters a lot for the unbiasedness and efficiency of your estimator.

**Response \#1:** Ensure your specification doesn't cause heteroskedasticity.

---

## Living with heteroskedasticity
### Misspecification

*Example:* Let the population relationship be $$ y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + u_i $$

with $\mathop{\boldsymbol{E}}\left[ u_i \middle| x_i \right] = 0$ and $\mathop{\text{Var}} \left( u_i \middle| x_i \right) = \sigma^2$.

However, we omit $x^2$ and estimate $$ y_i = \gamma_0 + \gamma_1 x_i + w_i $$

Then $$ w_i = u_i + \beta_2 x_i^2 \implies \mathop{\text{Var}} \left( w_i \right) = f(x_i) $$

_I.e._, the variance of $w_i$ changes systematically with $x_i$ (heteroskedasticity).

## Living with heteroskedasticity {auto-animate=true}
### Misspecification

```{r}
#| label: "spec-data"
#| include: false
# Set the seed
set.seed(1234)
# Generate data
spec_df = tibble(x = runif(1e3, 0, 3), y = exp(0.5 + 0.6 * x + rnorm(1e3, sd = 0.3)))
# Add residuals ('w': wrong specification; 'c': correct specification)
spec_df %<>% mutate(
  e_w = lm(y ~ x, data = spec_df) %>% residuals(),
  e_c = lm(log(y) ~ x, data = spec_df) %>% residuals()
)
```

[Truth:]{.pink } $\textcolor{#e64173}{\log\left(y_i\right) = \beta_0 + \beta_1 x_i + u_i}$  [**Misspecification:**]{.slate } $\textcolor{#314f4f}{y_i = \beta_0 + \beta_1 x_i + v_i}$

::: {data-id="spec-plot"}
```{r}
#| label: "spec-plot1"
#| echo: false
#| dev: "svg"
#| fig-height: 5
ggplot(data = spec_df, aes(x = x)) +
  geom_point(aes(y = e_w), color = "darkslategrey", size = 2.75, alpha = 0.5, shape = 16) +
  geom_point(aes(y = e_c), color = red_pink, size = 2.5, alpha = 0, shape = 19) +
  labs(x = "x", y = "e") +
  theme_axes
```
:::

## Living with heteroskedasticity {auto-animate=true}
### Misspecification

[**Truth:**]{.pink } $\textcolor{#e64173}{\log\left(y_i\right) = \beta_0 + \beta_1 x_i + u_i}$  [Misspecification:]{.slate } $\textcolor{#314f4f}{y_i = \beta_0 + \beta_1 x_i + v_i}$

::: {data-id="spec-plot"}
```{r}
#| label: "spec-plot2"
#| echo: false
#| dev: "svg"
#| fig-height: 5
ggplot(data = spec_df, aes(x = x)) +
  geom_point(aes(y = e_w), color = "darkslategrey", size = 2.75, alpha = 0.25, shape = 1) +
  geom_point(aes(y = e_c), color = red_pink, size = 2.5, alpha = 0.5, shape = 19) +
  labs(x = "x", y = "e") +
  theme_axes
```
:::

---

## Living with heteroskedasticity
### Misspecification

More generally:

**Misspecification problem:** Incorrect specification of the regression model can cause heteroskedasticity (among other problems).

::: {.fragment}
**Solution:** 💡 Get it right (_e.g._, don't omit $x^2$).
:::

::: {.fragment}
**New problems:**

- We often don't know the *right* specification.
- We'd like a more formal process for addressing heteroskedasticity.
:::

::: {.fragment}
**Conclusion:** Specification often will not "solve" heteroskedasticity.
<br> However, correctly specifying your model is still really important.
:::

---

## Living with heteroskedasticity
### Weighted least squares

Weighted least squares (WLS) presents another approach.

**Response \#2:** Increase efficiency by weighting our observations.

::: {.fragment}
Let the true population relationship be

$$
\begin{align}
  y_i = \beta_0 + \beta_1 x_{i} + u_i \tag{1}
\end{align}
$$

with $u_i \sim \mathop{N} \left( 0,\, \sigma_i^2 \right)$.
:::

::: {.fragment}
Now transform $(1)$ by dividing each observation's data by $\sigma_i$, _i.e._,

$$
\begin{align}
  \dfrac{y_i}{\sigma_i} &= \beta_0 \dfrac{1}{\sigma_i} + \beta_1 \dfrac{x_{i}}{\sigma_i} + \dfrac{u_i}{\sigma_i} \tag{2}
\end{align}
$$
:::

---

## Living with heteroskedasticity
### Weighted least squares

$$
\begin{align}
  y_i &= \beta_0 + \beta_1 x_{i} + u_i \tag{1} \\[1em]
  \dfrac{y_i}{\sigma_i} &= \beta_0 \dfrac{1}{\sigma_i} + \beta_1 \dfrac{x_{i}}{\sigma_i} + \dfrac{u_i}{\sigma_i} \tag{2}
\end{align}
$$

Whereas $(1)$ is heteroskedastic,

::: {.fragment}
 $\textcolor{#e64173}{(2)}$ [is homoskedastic]{.b .pink }.

∴ OLS is efficient and unbiased for estimating the $\beta_k$ in $(2)$!
:::

::: {.fragment}
Why is $(2)$ homoskedastic?
:::

[$\mathop{\text{Var}} \left( \dfrac{u_i}{\sigma_i} \middle| x_i \right) =$]{.fragment}
[$\dfrac{1}{\sigma_i^2} \mathop{\text{Var}} \left( u_i \middle| x_i \right) =$]{.fragment}
[$\dfrac{1}{\sigma_i^2} \sigma_i^2 =$]{.fragment}
[$1$]{.fragment}

---

## Living with heteroskedasticity
### Weighted least squares

WLS is great, but we need to know $\sigma_i^2$, which is generally unlikely.

We can *slightly* relax this requirement—instead requiring

1. $\mathop{\text{Var}} \left( u_i | x_i \right) = \sigma_i^2 = \sigma^2 h(x_i)$

2. We know $h(x)$.

::: {.fragment}
As before, we transform our heteroskedastic model into a homoskedastic model. This time we divide each observation's data^[Divide *all* of the data by $\sqrt{h(x_i)}$, including the intercept.] by $\sqrt{h(x_i)}$.
:::

---

## Living with heteroskedasticity
### Weighted least squares

$$
\begin{align}
  y_i &= \beta_0 + \beta_1 x_{i} + u_i \tag{1} \\[1em]
  \dfrac{y_i}{\sqrt{h(x_i)}} &= \beta_0 \dfrac{1}{\sqrt{h(x_i)}} + \beta_1 \dfrac{x_{i}}{\sqrt{h(x_i)}} + \dfrac{u_i}{\sqrt{h(x_i)}} \tag{2}
\end{align}
$$
with $\mathop{\text{Var}} \left( u_i | x_i \right) = \sigma^2 h(x_i)$.

Now let's check that $(2)$ is indeed homoskedastic.

[$\mathop{\text{Var}} \left( \dfrac{u_i}{\sqrt{h(x_i)}} \middle| x_i \right) =$]{.fragment}
[$\dfrac{1}{h(x_i)} \mathop{\text{Var}} \left( u_i \middle| x_i \right) =$]{.fragment}
[$\dfrac{1}{h(x_i)} \sigma^2 h(x_i) =$]{.fragment}
[$\textcolor{#e64173}{\sigma^2}$]{.fragment}

[Homoskedasticity!]{.b .pink .fragment}

---

## Living with heteroskedasticity
### Weighted least squares

[Weighted least squares]{.b .pink } (WLS) estimators are a special class of<br>[generalized least squares]{.b .pink } (GLS) estimators focused on heteroskedasticity.

::: {.fragment}
$$
  y_i = \beta_0 + \beta_1 x_{1i} + u_i \quad \textcolor{#e64173}{\text{ vs. }} \quad
  \dfrac{y_i}{\sigma_i} = \beta_0 \dfrac{1}{\sigma_i} + \beta_1 \dfrac{x_{1i}}{\sigma_i} + \dfrac{u_i}{\sigma_i}
$$

*Notes:*

1. WLS **transforms** a heteroskedastic model into a homoskedastic model.
2. **Weighting:** WLS downweights observations with higher variance $u_i$'s.
3. **Big requirement:** WLS requires that we *know* $\sigma_i^2$ for each observation.
4. WLS is generally **infeasible**. *Feasible* GLS (FGLS) offers a solution.
5. Under its assumptions: WLS is the **best linear unbiased estimator**.
:::

---

## Living with heteroskedasticity
### Heteroskedasticity-robust standard errors

**Response \#3:**

- Ignore OLS's inefficiency (in the presence of heteroskedasticity).
- Focus on [unbiased estimates for our standard errors]{.b .pink }.
- In the process: Correct inference.

::: {.fragment}
**Q:** What is a standard error?
:::

::: {.fragment}
<br>**A:** The [standard deviation of an estimator's distribution]{.b .pink }.
:::

::: {.fragment}
Estimators (like $\hat{\beta}_1$) are random variables, so they have distributions.

Standard errors give us a sense of how much variability is in our estimator.
:::

---

## Living with heteroskedasticity
### Heteroskedasticity-robust standard errors

*Recall*: We can write the OLS estimator for $\beta_1$ as

$$ \hat{\beta}_1 = \beta_1 + \dfrac{\sum_i \left( x_i - \overline{x} \right) u_i}{\sum_i \left( x_i - \overline{x} \right)^2} = \beta_1 + \dfrac{\sum_i \left( x_i - \overline{x} \right) u_i}{\text{SST}_x} \tag{3} $$

::: {.fragment}
Let $\mathop{\text{Var}} \left( u_i \middle| x_i \right) = \sigma_i^2$.
:::

::: {.fragment}
We can use $(3)$ to write the variance of $\hat{\beta}_1$, _i.e._,

$$ \mathop{\text{Var}} \left( \hat{\beta}_1 \middle| x_i \right) = \dfrac{\sum_i \left( x_i - \overline{x} \right)^2 \sigma_i^2}{\text{SST}_x^2} \tag{4} $$
:::

---

## Living with heteroskedasticity
### Heteroskedasticity-robust standard errors

If we want unbiased estimates for our standard errors, we need an unbiased estimate for

$$ \dfrac{\sum_i \left( x_i - \overline{x} \right)^2 \sigma_i^2}{\text{SST}_x^2} $$

Our old friend Hal White provided such an estimator:^[This specific equation is for simple linear regression.]

$$ \widehat{\mathop{\text{Var}}} \left( \hat{\beta}_1 \right) = \dfrac{\sum_i \left( x_i - \overline{x} \right)^2 e_i^2}{\text{SST}_x^2} $$

where the $e_i$ comes from the OLS regression of interest.

---

## Living with heteroskedasticity
### Heteroskedasticity-robust standard errors

Our heteroskedasticity-robust estimators for the standard error of $\beta_j$.

[Case 1]{.b .pink } Simple linear regression, $y_i = \beta_0 + \beta_1 x_i + u_i$

$$ \widehat{\mathop{\text{Var}}} \left( \hat{\beta}_1 \right) = \dfrac{\sum_i \left( x_i - \overline{x} \right)^2 e_i^2}{\text{SST}_x^2} $$

[Case 2]{.b .pink } Multiple (linear) regression, $y_i = \beta_0 + \beta_1 x_{1i} + \cdots + \beta_k x_{ki} + u_i$

$$ \widehat{\mathop{\text{Var}}} \left( \hat{\beta}_j \right) = \dfrac{\sum_i \hat{r}_{ij}^2 e_i^2}{\text{SST}_{x_j^2}} $$

where $\hat{r}_{ij}$ denotes the i[th]{.super } residual from regressing $x_j$ on all other explanatory variables.

---

## Living with heteroskedasticity
### Heteroskedasticity-robust standard errors

With these standard errors, we can return to correct statistical inferencel

_E.g._, we can update our previous $t$ statistic formula with our new heteroskedasticity-robust standard erros.

$$ t = \dfrac{\text{Estimate} - \text{Hypothesized value}}{\text{Standard error}} $$

---

## Living with heteroskedasticity
### Heteroskedasticity-robust standard errors

**Notes**

- We are still using [OLS estimates for]{.b .pink } $\textcolor{#e64173}{\beta_j}$
- Our het.-robust standard errors use a [different estimator]{.b .pink }.
- Homoskedasticity
  - Plain OLS variance estimator is more efficient.
  - Het.-robust is still unbiased.
- Heteroskedasticity
  - Plain OLS variance estimator is biased.
  - Het.-robust variance estimator is unbiased.

---

## Living with heteroskedasticity
### Heteroskedasticity-robust standard errors

These standard errors go by many names

- Heteroskedasticity-robust standard errors
- Het.-robust standard errors
- White standard errors
- Eicker-White standard errors
- Huber standard errors
- Eicker-Huber-White standards errors
- (some other combination of Eicker, Huber, and White)

**Do not say:** "Robust standard errors". The problem: "robust" to what?

---

## {.center}

[Living with heteroskedasticity]{.bigger .it .slab}

[Examples]{.it .slab .grey-vlight}

---

## Living with heteroskedasticity

### Examples


Back to our test-scores dataset…

```{r}
#| label: "ex-test-data"
#| echo: true
# Load packages
library(pacman)
p_load(tidyverse, Ecdat)
# Select and rename desired variables; assign to new dataset; format as tibble
test_df = Caschool %>% select(
  test_score = testscr, ratio = str, income = avginc, enrollment = enrltot
) %>% as_tibble()
# View first 2 rows of the dataset
head(test_df, 2)
```

---

## Living with heteroskedasticity
### Example: Model specification

We found significant evidence of heteroskedasticity.

Let's check if it was due to misspecifying our model.

---

## Living with heteroskedasticity
### Example: Model specification

Model[1]{.sub }: $\text{Score}_i = \beta_0 + \beta_1 \text{Ratio}_i + \beta_2 \text{Income}_i + u_i$
<br>`lm(test_score ~ ratio + income, data = test_df)`


```{r}
#| label: "ex-spec1"
#| echo: false
#| dev: "svg"
#| fig-height: 4.75
# Model 1: test ~ ratio + income
test_df %<>% mutate(e1 = lm(test_score ~ ratio + income, data = test_df) %>% residuals())
# Plot
ggplot(data = test_df, aes(x = income, y = e1)) +
geom_point(size = 3, alpha = 0.5, color = red_pink) +
labs(x = "Income", y = TeX("\\textit{e}")) +
theme_axes
```

---

## Living with heteroskedasticity
### Example: Model specification

Model[2]{.sub }: $\log\left(\text{Score}_i\right) = \beta_0 + \beta_1 \text{Ratio}_i + \beta_2 \text{Income}_i + u_i$
<br>`lm(log(test_score) ~ ratio + income, data = test_df)`


```{r}
#| label: "ex-spec2"
#| echo: false
#| dev: "svg"
#| fig-height: 4.75
# Model 1: test ~ ratio + income
test_df %<>% mutate(e2 = lm(log(test_score) ~ ratio + income, data = test_df) %>% residuals())
# Plot
ggplot(data = test_df, aes(x = income)) +
geom_point(aes(y = e2), size = 3, alpha = 0.5, color = red_pink) +
labs(x = "Income", y = TeX("\\textit{e}")) +
theme_axes
```

---

## Living with heteroskedasticity
### Example: Model specification

Model[3]{.sub }: $\log\left(\text{Score}_i\right) = \beta_0 + \beta_1 \text{Ratio}_i + \beta_2 \log\left(\text{Income}_i\right) + u_i$
<br>`lm(log(test_score) ~ ratio + log(income), data = test_df)`


```{r}
#| label: "ex-spec3"
#| echo: false
#| dev: "svg"
#| fig-height: 4.75
# Model 1: test ~ ratio + income
test_df %<>% mutate(e3 = lm(log(test_score) ~ ratio + log(income), data = test_df) %>% residuals())
# Plot
ggplot(data = test_df, aes(x = log(income))) +
geom_point(aes(y = e3), size = 3, alpha = 0.5, color = red_pink) +
labs(x = "Log(Income)", y = TeX("\\textit{e}")) +
theme_axes
```

---

## Living with heteroskedasticity
### Example: Model specification

Let's test this new specification with the White test for heteroskedasticity.

::: {.center}
Model[3]{.sub }: $\log\left(\text{Score}_i\right) = \beta_0 + \beta_1 \text{Ratio}_i + \beta_2 \log\left(\text{Income}_i\right) + u_i$
:::

```{r}
#| label: "ex-spec-test"
#| include: false
white_r2_spec = lm(e3^2 ~
  ratio * log(income) + I(ratio^2) + I(log(income)^2),
  data = test_df
) %>% summary() %$% r.squared
white_stat_spec = white_r2_spec %>% multiply_by(420)
```

::: {.fragment}
The regression for the White test
:::

::: {.fragment}
$$
\begin{align}
  e_i^2 = &\alpha_0 + \alpha_1 \text{Ratio}_i + \alpha_2 \log\left(\text{Income}_i\right) + \alpha_3 \text{Ratio}_i^2 + \alpha_4 \left(\log\left(\text{Income}_i\right)\right)^2 \\
  &+ \alpha_5 \left(\text{Ratio}_i\times\log\left(\text{Income}_i\right)\right) + v_i
\end{align}
$$
:::

[yields $R_e^2\approx`r round(white_r2_spec, 3)`$]{.fragment}
[and test statistic of]{.fragment}
[$\widehat{\text{LM}} = n\times R_e^2 \approx `r round(white_stat_spec, 1)`$.]{.fragment}

[Under H[0]{.sub }, $\text{LM}$ is distributed as]{.fragment}
[$\chi_5^2$ $\implies$ *p*-value $\approx$ `r pchisq(white_stat_spec, 5, lower.tail = F) %>% round(3)`.]{.fragment}

[∴ [Reject H[0]{.sub }.]{.b .pink } [Conclusion:]{.b .pink }]{.fragment}
[There is (still) statistically significant evidence of heteroskedasticity at the 5\%-percent level.]{.fragment}

---

## Living with heteroskedasticity
### Example: Model specification

Okay, we tried adjusting our specification, but there is still evidence of heteroskedasticity.

**Next:** In general, you will turn to [heteroskedasticity-robust standard errors]{.b .pink}.

- OLS is still unbiased for the [coefficients]{.b .pink } (the $\beta_j$'s)

- Heteroskedasticity-robust standard errors are unbiased for the<br>[standard errors]{.b .pink } of the $\hat{\beta}_j$'s, _i.e._, $\sqrt{\mathop{\text{Var}} \left( \hat{\beta}_j \right)}$.

---

## Living with heteroskedasticity
### Example: Het.-robust standard errors

Let's return to our model

$$ \text{Score}_i = \beta_0 + \beta_1 \text{Ratio}_i + \beta_2 \text{Income}_i + u_i $$

We can use the `fixest` package in [R]{.mono } to calculate standard errors.

---

## Living with heteroskedasticity
### Example: Het.-robust standard errors

$$ \text{Score}_i = \beta_0 + \beta_1 \text{Ratio}_i + \beta_2 \text{Income}_i + u_i $$

1\. Run the regression with `feols()` (instead of `lm()`)
```{r}
#| label: "fixest1"
#| echo: true
# Load 'fixest' package
p_load(fixest)
# Regress log score on ratio and log income
test_reg = feols(test_score ~ ratio + income, data = test_df)
```

::: {.fragment}
*Notice* that `feols()` uses the same syntax as `lm()` for this regression.
:::

---

## Living with heteroskedasticity
### Example: Het.-robust standard errors

$$ \text{Score}_i = \beta_0 + \beta_1 \text{Ratio}_i + \beta_2 \text{Income}_i + u_i $$

2\. Estimate het.-robust standard errors with `vcov = 'hetero'` option in `summary()`

```{r}
#| label: "fixest2a"
#| eval: false
#| echo: true
# Het-robust standard errors with 'vcov = 'hetero''
summary(test_reg, vcov = 'hetero')
```

```{r}
#| label: "fixest2b"
#| echo: false
test_het_out = summary(test_reg, vcov = 'hetero') %>% capture.output()
test_het_out[4:7] %>% paste0("\n") %>% cat()
```

---

## Living with heteroskedasticity
### Example: Het.-robust standard errors

Ceofficients and **heteroskedasticity-robust standard errors**:

```{r}
#| label: "fixest3"
#| eval: false
#| echo: true
summary(test_reg, vcov = 'hetero')
```

```{r}
#| label: "fixest4"
#| echo: false
test_het_out = summary(test_reg, vcov = 'hetero') %>% capture.output()
test_het_out[4:7] %>% paste0("\n") %>% cat()
```

Ceofficients and **plain OLS standard errors** (assumes homoskedasticity):

```{r}
#| label: "fixest5"
#| eval: false
#| echo: true
summary(test_reg, vcov = 'iid')
```

```{r}
#| label: "fixest6"
#| echo: false
test_hom_out = summary(test_reg, vcov = 'iid') %>% capture.output()
test_hom_out[4:7] %>% paste0("\n") %>% cat()
```

---

## Living with heteroskedasticity
### Example: WLS

We mentioned that WLS is often not possible—we need to know the functional form of the heteroskedasticity—either

**A**\. $\sigma_i^2$

or

**B**\. $h(x_i)$, where $\sigma_i^2 = \sigma^2 h(x_i)$

::: {.fragment}
There *are* occasions in which we can know $h(x_i)$.
:::

---

## Living with heteroskedasticity
### Example: WLS

Imagine individuals in a population have homoskedastic disturbances.

However, instead of observing individuals' data, we observe (in data) groups' averages (_e.g._, cities, counties, school districts).

If these groups have different sizes, then our dataset will be heteroskedastic—in a predictable fashion.

**Recall:** The variance of the sample mean depends upon the sample size,
$$ \mathop{\text{Var}} \left( \overline{x} \right) = \dfrac{\sigma_x^2}{n} $$

::: {.fragment}
**Example:**  Our school testing data is averaged at the school level.
:::

---

## Living with heteroskedasticity
### Example: WLS

*Example:*  Our school testing data is averaged at the school level.

Even if individual students have homoskedastic disturbances, the schools would have heteroskedastic disturbances, _i.e._,

**Individual-level model:** $\quad \text{Score}_i = \beta_0 + \beta_1 \text{Ratio}_i + \beta_2 \text{Income}_i + u_i$

**School-level model:** $\quad \overline{\text{Score}}_s = \beta_0 + \beta_1 \overline{\text{Ratio}}_s + \beta_2 \overline{\text{Income}}_s + \overline{u}_s$

where the $s$ subscript denotes an individual school (just as $i$ indexes an individual person).

$$ \mathop{\text{Var}} \left( \overline{u}_s \right) = \dfrac{\sigma^2}{n_s} $$

---

## Living with heteroskedasticity
### Example: WLS

For WLS, we're looking for a function $h(x_s)$ such that $\mathop{\text{Var}} \left( \overline{u}_s | x_s \right) = \sigma^2 h(x_s)$.

::: {.fragment}
We just showed^[Assuming the individuals' disturbances are homoskedastic.] that $\mathop{\text{Var}} \left( \overline{u}_s |x_s \right) = \dfrac{\sigma^2}{n_s}$.
:::

::: {.fragment}
Thus, $h(x_s) = 1/n_s$, where $n_s$ is the number of students in school $s$.
:::

::: {.fragment}
To implement WLS, we divide each observation's data by $1/\sqrt{h(x_s)}$, meaning we need to multiply each school's data by $\sqrt{n_s}$.
:::

::: {.fragment}
The variable [enrollment]{.mono } in the [test_df]{.mono } dataset is our $n_s$.
:::

---

## Living with heteroskedasticity
### Example: WLS

Using WLS to estimate $\text{Score}_i = \beta_0 + \beta_1 \text{Ratio}_i + \beta_2 \text{Income}_i + u_i$

**Step 1:** Multiply each variable by $1/\sqrt{h(x_i)} = \sqrt{\text{Enrollment}_i}$

```{r}
#| label: "wls1"
#| echo: true
# Create WLS transformed variables, multiplying by sqrt of 'pop'
test_df = mutate(test_df,
  test_score_wls = test_score * sqrt(enrollment),
  ratio_wls      = ratio * sqrt(enrollment),
  income_wls     = income * sqrt(enrollment),
  intercept_wls  = 1 * sqrt(enrollment)
)
```

Notice that we are creating a transformed intercept.

---

## Living with heteroskedasticity
### Example: WLS

Using WLS to estimate $\text{Score}_i = \beta_0 + \beta_1 \text{Ratio}_i + \beta_2 \text{Income}_i + u_i$

**Step 2:** Run our WLS (transformed) regression

```{r}
#| label: "wls2"
#| echo: true
# WLS regression
wls_reg = lm(
  test_score_wls ~ -1 + intercept_wls + ratio_wls + income_wls,
  data = test_df
)
```

::: {.fragment}
*Note:* The `-1` in our regression tells [R]{.mono } not to add an intercept, since we are adding a transformed intercept (`intercept_wls`).
:::

---

## Living with heteroskedasticity
### Example: WLS

The [WLS estimates and standard errors:]{.b .pink }
```{r}
#| label: "wls3"
#| echo: false
# Grab the summary
test_wls_out = summary(wls_reg) %>% capture.output()
# Print the coefficients
test_wls_out[11:14] %>% paste0("\n") %>% cat()
```

::: {.fragment}
<br>
The [OLS estimates]{.b .pink } and [het.-robust standard errors:]{.b .pink }
```{r}
#| label: "wls4"
#| echo: false
# Print the coefficients
test_het_out[4:7] %>% paste0("\n") %>% cat()
```
:::

---

## Living with heteroskedasticity
### Example: WLS

Alternative to doing your own weighting: feed `lm()` some `weights`.

```{r}
#| label: "wls5"
#| eval: false
#| echo: true

lm(test_score ~ ratio + income, data = test_df, weights = enrollment)
```

```{r}
#| label: "wls6"
#| echo: false

test_wls_out2 = lm(test_score ~ ratio + income, data = test_df, weights = enrollment)
# Grab the summary
test_wls_out = summary(wls_reg) %>% capture.output()
# Print the coefficients
test_wls_out[11:14] %>% paste0("\n") %>% cat()
```

---

## Living with heteroskedasticity

In this example

- [Heteroskedasticity-robust standard errors]{.b .pink } did not change our standard errors very much (relative to plain OLS standard errors).

- [WLS]{.b .pink } changed our answers a bit—coefficients and standard errors.

::: {.fragment}
These examples highlighted a few things:

1. Using the correct estimator for your standard errors really matters.^[Sit in on an economics seminar, and you will see what I mean.]

2. Econometrics doesn't always offer an obviously *correct* route.
:::

::: {.fragment}
To see \#1, let's run a simulation.
:::

---

## Living with heteroskedasticity
### Simulation

```{r}
#| label: "sim-params"
#| include: false
b0 = 1L
b1 = 10L
```

Let's examine a simple linear regression model with heteroskedasticity.

$$ y_i = \underbrace{\beta_0}_{=`r b0`} + \underbrace{\beta_1}_{=`r b1`} x_i + u_i $$

where $\mathop{\text{Var}} \left( u_i | x_i \right) = \sigma_i^2 = \sigma^2 x_i^2$.

::: {.r-stack}

::: {.fragment .fade-out fragment-index=1}
```{r}
#| label: "sim-plot-y"
#| echo: false
#| fig-height: 4
#| fig-width: 10.5

set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, 0.5, 1.5),
  y = b0 + b1 * x + rnorm(1e3, 0, sd = x^2)
), aes(x = x, y = y)) +
geom_point(color = "darkslategrey", size = 2.75, alpha = 0.5) +
geom_abline(intercept = b0, slope = b1, color = "orange", size = 1.5, alpha = 0.85) +
labs(x = "x", y = "y") +
theme_axes
```
:::

::: {.fragment .fade-in fragment-index=1}
```{r}
#| label: "sim-plot-u"
#| echo: false
#| fig-height: 4
#| fig-width: 10.5

set.seed(12345)
ggplot(data = tibble(
  x = runif(1e3, 0.5, 1.5),
  u = rnorm(1e3, 0, sd = x^2)
), aes(x = x, y = u)) +
geom_point(color = "darkslategrey", size = 2.75, alpha = 0.5) +
labs(x = "x", y = "u") +
theme_axes
```
:::

:::

---

## Living with heteroskedasticity
### Simulation

*Note regarding WLS:*

Since $\mathop{\text{Var}} \left( u_i | x_i \right) = \sigma^2 x_i^2$,

$$ \mathop{\text{Var}} \left( u_i | x_i \right) = \sigma^2 h(x_i) \implies h(x_i) = x_i^2 $$

WLS multiplies each variable by $1/\sqrt{h(x_i)} = 1/x_i$.

---

## Living with heteroskedasticity
### Simulation

In this simulation, we want to compare

The **efficiency** of

+ OLS
+ WLS with correct weights: $h(x_i) = x_i$
+ WLS with incorrect weights: $h(x_i) = \sqrt{x_i}$

How well our **standard errors** perform (via confidence intervals) with

+ Plain OLS standard errors
+ Heteroskedasticity-robust standard errors
+ WLS standard errors

---

## Living with heteroskedasticity
### Simulation

The simulation plan:

::: {.pseudocode-small}

Do 10,000 times:

1. Generate a sample of size 30 from the population

2. Calculate/save OLS and WLS (×2) estimates for β[1]{.sub }

3. Calculate/save standard errors for β[1]{.sub } using
  - Plain OLS standard errors
  - Heteroskedasticity-robust standard errors
  - WLS (correct)
  - WLS (incorrect)

:::

---

## Living with heteroskedasticity
### Simulation

**For one iteration of the simulation:**

Code to generate the data...

```{r}
#| label: "sim-one-iteration-1"
#| eval: false
#| echo: true
# Parameters
b0 = 1
b1 = 10
s2 = 1
# Sample size
n = 30
# Generate data
sample_df = tibble(
  x = runif(n, 0.5, 1.5),
  y = b0 + b1 * x + rnorm(n, 0, sd = s2 * x^2)
)
```

---

## Living with heteroskedasticity
### Simulation

**For one iteration of the simulation:**

Code to estimate our coefficients and standard errors...

```{r}
#| label: "sim-one-iteration-2"
#| eval: false
#| echo: true
# OLS
ols = feols(y ~ x, data = sample_df)
# WLS: Correct weights
wls_t = lm(y ~ x, data = sample_df, weights = 1/x^2)
# WLS: Correct weights
wls_f = lm(y ~ x, data = sample_df, weights = 1/x)
# Coefficients and standard errors
summary(ols, vcov = 'iid')
summary(ols, vcov = 'hetero')
summary(wls_t)
summary(wls_f)
```

Then save the results.

---

## Living with heteroskedasticity
### Simulation: Coefficients

```{r}
#| label: "sim-df"
#| include: false
#| cache: true
# Parameters
b0 = 1
b1 = 10
s2 = 1
# Sample size
n = 30
# Number of iterations
n_iter = 1e3
# Set seed
set.seed(1234)
# The simulation
sim_df = mclapply(X = 1:n_iter, FUN = function(i, size) {
  # Generate data
  sample_df = tibble(
    x = runif(size, 0.5, 1.5),
    y = b0 + b1 * x + rnorm(size, 0, sd = s2 * x^2)
  )
  # OLS
  ols = feols(y ~ x, data = sample_df)
  # WLS: Correct weights
  wls_t = lm(y ~ x, data = sample_df, weights = 1/x^2)
  # WLS: Correct weights
  wls_f = lm(y ~ x, data = sample_df, weights = 1/x)
  # Save results
  iter_df = rbind(
    summary(ols, vcov = 'iid') %>% coeftable() %>% magrittr::extract(2, 1:2),
    summary(ols, vcov = 'hetero') %>% coeftable() %>% magrittr::extract(2, 1:2),
    summary(wls_t) %>% coef() %>% magrittr::extract(2,1:2),
    summary(wls_f) %>% coef() %>% magrittr::extract(2,1:2)
  ) %>%
  as_tibble() %>%
  mutate(
    model = c("OLS Hom.", "OLS Het.", "WLS T", "WLS F"),
    iter = i
  )
  # Return the data
  return(iter_df)
}, mc.cores = detectCores()-1, size = n) %>% bind_rows()
# Change names
names(sim_df) = c("coef", "se", "model", "iter")
```

```{r}
#| label: "sim-plot-efficiency"
#| echo: false
#| fig-height: 6.25
ggplot(data = sim_df %>% filter(model != "OLS Hom."), aes(x = coef, color = model, fill = model)) +
geom_vline(xintercept = 10, linetype = "dashed") +
geom_density(alpha = 0.1) +
geom_hline(yintercept = 0) +
labs(x = "Estimated coefficient", y = "Density") +
scale_color_viridis_d("",
  labels = c("OLS", "WLS Incorrect", "WLS Correct"),
  end = 0.9, option = "C"
) +
scale_fill_viridis_d("",
  labels = c("OLS", "WLS Incorrect", "WLS Correct"),
  end = 0.9, option = "C"
) +
theme_pander(base_size = 20, base_family = "Fira Sans Condensed") +
theme(
  legend.position = c(.85, .85),
  # legend.background = element_blank(),
  legend.key.size = unit(1, "cm")
)
```

---

## Living with heteroskedasticity
### Simulation: Inference


```{r}
#| label: "sim-plot-t-stat"
#| echo: false
#| fig-height: 6.25
ggplot(data = sim_df, aes(x = (coef-10)/se, color = model, fill = model)) +
geom_vline(xintercept = qt(c(0.025, 0.975), df = 28), linetype = "dashed") +
geom_density(alpha = 0.1) +
geom_hline(yintercept = 0) +
labs(x = "t statistic testing the true value", y = "Density") +
scale_color_viridis_d("",
  labels = c("OLS + Het.-robust", "Plain OLS", "WLS Incorrect", "WLS Correct"),
  end = 0.9, option = "C"
) +
scale_fill_viridis_d("",
labels = c("OLS + Het.-robust", "Plain OLS", "WLS Incorrect", "WLS Correct"),
  end = 0.9, option = "C"
) +
theme_pander(base_size = 20, base_family = "Fira Sans Condensed") +
theme(
  legend.position = c(.85, .85),
  # legend.background = element_blank(),
  legend.key.size = unit(1, "cm")
)
```

---

## Living with heteroskedasticity
### Simulation: Results

Summarizing our simulation results (10,000 iterations)

:::: {.columns}

::: {.column width="48%"}
::: {.center}
**Estimation**: Summary of $\hat{\beta}_1$'s

```{r}
#| label: "sim-table-coef"
#| eval: true
#| echo: false
sim_df %>%
  filter(model != "OLS Hom.") %>%
  mutate(Estimator = recode(model,
    "OLS Het." = "OLS",
    "WLS F" = "WLS Incorrect",
    "WLS T" = "WLS Correct"
  )) %>%
  group_by(Estimator) %>%
  summarize(Mean = mean(coef) %>% round(3), "S.D." = sd(coef) %>% round(3)) %>%
  kable() %>%
  kable_styling(full_width = T)
```
:::
:::

::: {.column width="48%"}
::: {.fragment}
::: {.center}
**Inference:** % of times we reject $\beta_1$

```{r}
#| label: "sim-table-se"
#| eval: true
#| echo: false
sim_df %>%
  mutate(Estimators = recode(model,
    "OLS Hom." = "OLS + Homosk.",
    "OLS Het." = "OLS + Het.-robust",
    "WLS F" = "WLS Incorrect",
    "WLS T" = "WLS Correct"
  )) %>%
  group_by(Estimators) %>%
  summarize(`% Reject` = mean(abs(coef-10)/se > qt(0.975, 28)) %>% multiply_by(100) %>% round(1)) %>%
  kable() %>%
  kable_styling(full_width = T)
```
:::
:::
:::

::::

---

## {.center}

Going further...

---

## Similar violations
### Assumptions

Recall our old assumption that led to this heteroskedasticity discussion:

5\. The disurbances have [constant variance]{.b .slate } $\sigma^2$ and [zero covariance]{.b .pink }, _i.e._,

- $\textcolor{#314f4f}{\mathop{\boldsymbol{E}}\left[ u_i^2 \middle| X_i \right] = \mathop{\text{Var}} \left( u_i \middle| X_i \right) = \sigma^2 \implies \mathop{\text{Var}} \left( u_i \right) = \sigma^2}$
- $\textcolor{#e64173}{\mathop{\text{Cov}} \left( u_i, \, u_j \middle| X_i,\,X_j \right) = \mathop{\boldsymbol{E}}\left[ u_i u_j \middle| X_i,\,X_j \right] = 0}$ for $i\neq j$

::: {.fragment}
It's also possible (likely) that the [disturbances are correlated]{.b .pink }.

Ignoring this correlation is _even more_ problematic for inference.
:::

---

## Correlated disturbances
### The problem

In many cases, [observations' disturbances]{.b .slate } $\left(u_i,\,u_j\right)$ [may correlate]{.b .pink }.

[Remember]{.note }

- The [disturbance]{.b .slate } represents the un-included variables that affect $y$. 
- Some [observations]{.b .pink } in the sample may *relate* to other observations.

If these observation-level relationships extend to the variables in the disturbance, then disturbances can correlate.

$\implies \mathop{\text{Cov}} \left( u_i, \, u_j \middle| X_i,\,X_j \right) \textcolor{#e64173}{\neq} 0$.

::: {.fragment}
Ignoring this correlation can cause [big problems]{.b .pink} in your inference.
:::

---

## Correlated disturbances
### The intution


Why is ignoring this correlation problematic?

**False precision:** We can get "overconfident" in our knowledge.

When we teating correlated observations as independent, we OLS thinks we're learning more than we actually are. 

::: {.fragment}
**Extreme example:** If duplicate your dataset (stack it on top of itself), plain OLS standard errors would decrease every time you duplicated the dataset.
:::

---

## Hidden {visibility="hidden"}

```{r}
#| label: "dup-data"
#| echo: false
#| cache: true

# And an id
test_df %<>% mutate(id = 1:nrow(test_df))
# Stack the dataset a bunch of times
n_dup = 1e2
dup_df = lapply(X = 1:n_dup, FUN = function(x) mutate(test_df, dup = x)) %>% bind_rows()
# Now run a regression
dup_est = lapply(
  X = 1:n_dup,
  FUN = function(x) feols(test_score ~ ratio + income, data = dup_df %>% filter(dup <= x))
)
# Grab standard errors
dup_se = lapply(
  X = seq_along(dup_est),
  FUN = function(x) {
    data.frame(
      dups = x - 1,
      se_iid = dup_est[[x]] %>% summary(vcov = 'iid') %>% coeftable() %>% magrittr::extract(2,2),
      se_cl = dup_est[[x]] %>% summary(cluster = 'id') %>% coeftable() %>% magrittr::extract(2,2)
    )
  }
) %>% bind_rows()
```

---

## {auto-animate=true}

The effect of duplicating our data on the [OLS standard error]{.pink } of `ratio`.

::: {data-id="dup"}
```{r}
#| label: "dup-plot0"
#| echo: false
#| fig-height: 6.5

ggplot(
  data = dup_se[1:5, ],
  aes(x = dups, y = se_iid)
) +
geom_hline(yintercept = 0) +
geom_line(color = red_pink, size = 0.5) +
geom_point(color = red_pink, size = 3.5) +
labs(x = "Number of times dataset is duplicated", y = "Standard error") +
theme_pander(base_size = 22, base_family = "Fira Sans Condensed")
```
:::

## {auto-animate=true}

The effect of duplicating our data on the [OLS standard error]{.pink } of `ratio`.

::: {data-id="dup"}
```{r}
#| label: "dup-plot1"
#| echo: false
#| fig-height: 6.5

ggplot(
  data = dup_se[1:25, ],
  aes(x = dups, y = se_iid)
) +
geom_hline(yintercept = 0) +
geom_line(color = red_pink, size = 0.5) +
geom_point(color = red_pink, size = 3.5) +
labs(x = "Number of times dataset is duplicated", y = "Standard error") +
theme_pander(base_size = 22, base_family = "Fira Sans Condensed")
```
:::

## {auto-animate=true}

The effect of duplicating our data on the [OLS standard error]{.pink } of `ratio`.

::: {data-id="dup"}
```{r}
#| label: "dup-plot2"
#| echo: false
#| fig-height: 6.5

ggplot(
  data = dup_se,
  aes(x = dups, y = se_iid)
) +
geom_hline(yintercept = 0) +
geom_line(color = red_pink, size = 0.5) +
geom_point(color = red_pink, size = 3.5) +
labs(x = "Number of times dataset is duplicated", y = "Standard error") +
theme_pander(base_size = 22, base_family = "Fira Sans Condensed")
```
:::

## {auto-animate=true}

Correcting our [standard errors for clustering]{.orange } (observations' correlation).

::: {data-id="dup"}
```{r}
#| label: "dup-plot3"
#| echo: false
#| fig-align: "center"
#| fig-height: 6.5

ggplot(
  data = dup_se,
  aes(x = dups)
) +
geom_hline(yintercept = 0) +
geom_line(aes(y = se_iid), color = red_pink, size = 0.5, alpha = 0.25) +
geom_point(aes(y = se_iid), color = red_pink, size = 3.5, alpha = 0.25) +
geom_line(aes(y = se_cl), color = orange, size = 0.5) +
geom_point(aes(y = se_cl), color = orange, size = 3.5) +
labs(x = "Number of times dataset is duplicated", y = "Standard error") +
theme_pander(base_size = 22, base_family = "Fira Sans Condensed")
```
:::

## Correlated disturbances
### Examples


"Real" examples where disturbances might [correlate]{.b .purple }:

- students in a [classroom]{.purple} (share teacher, curriculum, etc.);
- counties in a [state]{.purple} (share state-level policies/laws);
- businesses in a [city]{.purple} (share local economic shocks);
- [consecutive]{.purple} days in a sample (share events, weather, etc.).

## Correlated disturbances
### The solution


Just like we calculate *heteroskedasticity*-robust standard errors,
<br>we can also calculate standard errors robust to correlated disturbances.

People call these [*cluster-robust standard errors*]{.pink } (or just *clustered*).

From `fixest` package:

`feols(y ~ x, data = fake_data, cluster = 'cluster_var')`

or even

`feols(y ~ x, data = fake_data, cluster = c('cluster1', 'cluster2'))`

## Final word
### Better inference

You should

1. default to assuming your data are [heteroskedastic]{.b };
1. consider how explanatory vars/disturbances [corr. across observations]{.b }.

[Failure]{.pink } to address these issues can lead to

- [overconfident inference]{.b .pink },
- [false positives]{.b .pink },
- [incorrect conclusions/decisions]{.b .pink }.
