---
title: 'Autocorrelation'
subtitle: 'EC 421, Set 8'
author:
  - name: Ed Rubin
    email: edwardr@uoregon.edu
editor:
  render-on-save: true
format:
  clean-revealjs:
    slide-level: 2
    theme: [../my.scss, default]
    highlight-style: gruvbox
    embed-resources: true
    slide-number: c/t
    incremental: false
    controls-layout: edges
    transition: none
    preview-links: auto
    html-math-method: mathjax
    code-line-numbers: false
execute:
  warning: false
  message: false

---

## {.center}


[Prologue]{.bigger .it .slab}

---

## Schedule


```{r}
#| label: "setup"
#| include: false
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(
  broom, here, tidyverse,
  latex2exp, ggplot2, ggthemes, viridis, extrafont, gridExtra,
  kableExtra,
  dplyr,
  lubridate,
  magrittr, knitr, parallel
)
# Define pink color
red_pink = "#e64173"
turquoise = "#20B2AA"
grey_light = "grey70"
grey_mid = "grey50"
grey_dark = "grey20"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -0.5, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_simple <- theme_bw() + theme(
  line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_text(size = 18, family = "STIXGeneral"),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  # plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_math <- theme_void() + theme(
  text = element_text(family = "MathJax_Math"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_serif <- theme_void() + theme(
  text = element_text(family = "MathJax_Main"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes <- theme_void() + theme(
  text = element_text(family = "Fira Sans Book"),
  axis.title = element_text(size = 18),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = grey_light,
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_set(theme_gray(base_size = 20))
# A few extras
```

```{css}
#| echo: false

@media print {
  .has-continuation {
    display: block !important;
  }
}
```

### Last Time

- Introduction to time series

### Today

Autocorrelation

---

## [R]{.mono } showcase


### Functions

Writing your own functions.

---

## Writing functions


### Functions are everywhere

Everything you do in [R]{.mono } involves some sort of function, _e.g._,

- `mean()`
- `lm()`
- `summary()`
- `read_csv()`
- `ggplot()`
- `+`

The basic idea in [R]{.mono } is doing things to objects with functions.

---

## Writing functions


### Functions can help

We write functions to make life easier. Instead of copying and pasting the same line of code a million times, you can write one function.

In [R]{.mono }, you use the `function()` function to write functions.<sup>[†]{.pink }</sup>

^[[†]{.pink } Meta.]

```{r}
#| label: "ex-function-1"
#| echo: true
#| eval: false
# Our first function
the_name <- function(arg1, arg2) {
  # Insert code that involves arg1 and arg2 (this is where the magic happens)
}
```
- `the_name`: The name we are giving to our new function.
- `arg1`: The first argument of our function.
- `arg2`: The second argument of our function.

---

## Writing functions


### Our first real function

Let's write a function that multiplies two numbers. (It needs two arguments.)

```{r}
#| label: "ex-function-2"
#| echo: true
# Create our function
the_product <- function(x, y) {
  x * y
}
```

::: {.fragment}


Did it work?

:::

::: {.fragment}


```{r}
#| label: "ex-function-3"
#| echo: true
the_product(7, 15)
```

[💪]{.bigger }

:::

---

## Writing functions


### Functions can do anything

... that you tell them.

If you are going to repeat a task (_e.g._, a simulation), then you have a good situation for writing your own function.



[R]{.mono } offers many functions (via its many packages), but you will sometimes find a scenario for which no one has written a function.

Now you know how to write your own.

```{r}
#| label: "ex-function-4"
#| echo: true
# An ad lib function
ad_lib <- function(noun1, verb1, noun2) {
  paste("The next", noun1, "of our lecture", verb1, noun2)
}
```

---

## Writing functions


```{r}
#| label: "ex-function-5"
#| echo: true
ad_lib(noun1 = "part", verb1 = "reviews", noun2 = "time series.")
```

---

## {.center}

[Time series]{.bigger .it .slab}

[Review]{.it .slab .grey-vlight}

---

## Time series
### Review


Changes to our model/framework.

- Our model now has $\textcolor{#e64173}{t}$ subscripts for [time periods]{.b .pink }.

- [Dynamic models]{.b .pink } allow [lags]{.b .pink } of explanatory and/or outcome variables.

- We changed our **exogeneity** assumption to [contemporaneous exogeneity]{.b .pink }, _i.e._, $\mathop{\boldsymbol{E}}\left[ u_t \middle| X_t \right] = 0$

- Including [lags of outcome variables]{.b .orange } can lead to [biased coefficient estimates]{.b .pink } from OLS.

- [Lagged explanatory variables]{.b .orange } make [OLS inefficient]{.b .pink }.

---

## {.center}

[Autocorrelation]{.bigger .it .slab}

---

## Autocorrelation
### What is it?

[Autocorrelation]{.b .pink } occurs when our disturbances are correlated over time, _i.e._, $\mathop{\text{Cov}} \left( u_t,\, u_s \right) \neq 0$ for $t\neq s$.

::: {.fragment}


Another way to think about: If the *shock* from disturbance $t$ correlates with "nearby" shocks in $t-1$ and $t+1$.

:::

::: {.fragment}


*Note:* **Serial correlation** and **autocorrelation** are the same thing.

:::

::: {.fragment}


Why is autocorrelation prevalent in time-series analyses?

:::

---

[Positive autocorrelation]{.b .slate }: Disturbances $\left( u_t \right)$ over time
```{r}
#| label: "positive-auto-u"
#| echo: false
#| fig-height: 7.25
# Number of observations
T <- 1e2
# Rho
rho <- 0.95
# Set seed and starting point
set.seed(1234)
start <- rnorm(1)
# Generate the data
ar_df <- tibble(
  t = 1:T,
  x = runif(T, min = 0, max = 1),
  e = rnorm(T, mean = 0, sd = 2),
  u = NA
)
for (x in 1:T) {
  if (x == 1) {
    ar_df$u[x] <- rho * start + ar_df$e[x]
  } else {
    ar_df$u[x] <- rho * ar_df$u[x-1] + ar_df$e[x]
  }
}
ar_df %<>% mutate(y = 1 + 3 * x + u)
# Plot disturbances over time
ggplot(data = ar_df,
  aes(t, u)
) +
geom_line(color = "grey85", size = 0.35) +
geom_point(color = red_pink, size = 2.25) +
ylab("u") +
xlab("t") +
# theme_pander(base_family = "Fira Sans Book", base_size = 20)
theme_axes_math
```

---

[Positive autocorrelation]{.b .slate }: Outcomes $\left( y_t \right)$ over time
```{r}
#| label: "positive-auto-y"
#| echo: false
#| fig-height: 7.25
# Plot outcomes over time
ggplot(data = ar_df,
  aes(t, y)
) +
geom_line(color = "grey85", size = 0.35) +
geom_point(color = red_pink, size = 2.25) +
ylab("y") +
xlab("t") +
# theme_pander(base_family = "Fira Sans Book", base_size = 20)
theme_axes_math
```

---

[Negative autocorrelation]{.b .slate }: Disturbances $\left( u_t \right)$ over time
```{r}
#| label: "negative-auto-u"
#| echo: false
#| fig-height: 7.25
# Number of observations
T <- 1e2
# Rho
rho <- -0.95
# Set seed and starting point
set.seed(1234)
start <- rnorm(1)
# Generate the data
ar_df <- tibble(
  t = 1:T,
  x = runif(T, min = 0, max = 1),
  e = rnorm(T, mean = 0, sd = 2),
  u = NA
)
for (x in 1:T) {
  if (x == 1) {
    ar_df$u[x] <- rho * start + ar_df$e[x]
  } else {
    ar_df$u[x] <- rho * ar_df$u[x-1] + ar_df$e[x]
  }
}
ar_df %<>% mutate(y = 1 + 3 * x + u)
# Plot disturbances over time
ggplot(data = ar_df,
  aes(t, u)
) +
geom_line(color = "grey85", size = 0.35) +
geom_point(color = red_pink, size = 2.25) +
ylab("u") +
xlab("t") +
# theme_pander(base_family = "Fira Sans Book", base_size = 20)
theme_axes_math
```

---

[Negative autocorrelation]{.b .slate }: Outcomes $\left( y_t \right)$ over time
```{r}
#| label: "negative-auto-y"
#| echo: false
#| fig-height: 7.25
# Plot outcomes over time
ggplot(data = ar_df,
  aes(t, y)
) +
geom_line(color = "grey85", size = 0.35) +
geom_point(color = red_pink, size = 2.25) +
ylab("y") +
xlab("t") +
# theme_pander(base_family = "Fira Sans Book", base_size = 20)
theme_axes_math
```

---

## Autocorrelation
### In static time-series models


Let's start with a very common model: a static time-series model whose disturbances exhibit [first-order autocorrelation]{.b .pink }, *a.k.a.* [AR(1)]{.pink }:

$$
\begin{align}
  \text{Births}_t &= \beta_0 + \beta_1 \text{Income}_t + u_t
\end{align}
$$
where
$$
\begin{align}
  \textcolor{#e64173}{u_t} = \textcolor{#e64173}{\rho \, u_{t-1}} + \varepsilon_t
\end{align}
$$
and the $\varepsilon_t$ are independently and identically distributed (*i.i.d.*).

::: {.fragment}


[Second-order autocorrelation]{.b .purple }, or [AR(2)]{.purple }, would be

$$
\begin{align}
  \textcolor{#6A5ACD}{u_t} = \textcolor{#6A5ACD}{\rho_1 u_{t-1}} + \textcolor{#6A5ACD}{\rho_2 u_{t-2}} + \varepsilon_t
\end{align}
$$

:::

---

## Autocorrelation
### In static time-series models


An [AR(p)]{.turquoise } model/process has a disturbance structure of

$$
\begin{align}
  u_t = \sum_{j=1}^\textcolor{#20B2AA}{p} \rho_j u_{t-j} + \varepsilon_t
\end{align}
$$

allowing the current disturbance to correlated with up to $\textcolor{#20B2AA}{p}$ of its lags.

---

## Autocorrelation
### OLS

For **static models** or **dynamic models with lagged explanatory variables**, in the presence of autocorrelation

1. OLS provides [**unbiased** estimates for the coefficients.]{.pink }

1. OLS creates [**biased** estimates for the standard errors.]{.pink }

1. OLS is [**inefficient**.]{.pink }

*Recall:* Same implications as heteroskedasticity.

Autocorrelation get trickier with lagged outcome variables.

---

## Autocorrelation
### OLS and lagged outcome variables


Consider a model with one lag of the outcome variable—ADL(1, 0)—model with AR(1) disturbances

$$
\begin{align}
  \text{Births}_t = \beta_0 + \beta_1 \text{Income}_t + \beta_2 \text{Births}_{t-1} + u_t
\end{align}
$$
where
$$
\begin{align}
  u_t = \rho u_{t-1} + \varepsilon_t
\end{align}
$$

::: {.fragment}


**Problem:**

:::

::: {.fragment}

 Both $\text{Births}_{t-1}$ (a regressor in the model for time $t$) and $u_{t}$ (the disturbance for time $t$) depend upon $u_{t-1}$. *I.e.*, a regressor is correlated with its contemporaneous disturbance.

:::

::: {.fragment}


**Q:** Why is this a problem?

:::

<br>[**A:** It violates [contemporaneous exogeneity]{.pink }]{.fragment}[, *i.e.*, $\mathop{\text{Cov}} \left( x_t,\,u_t \right) \neq 0$.]{.fragment}

---

## Autocorrelation
### OLS and lagged outcome variables


To see this problem, first write out the model for $t$ and $t-1$:
$$
\begin{align}
  \text{Births}_t &= \beta_0 + \beta_1 \text{Income}_t + \beta_2 \text{Births}_{t-1} + u_t \\
  \text{Births}_{t-1} &= \beta_0 + \beta_1 \text{Income}_{t-1} + \beta_2 \text{Births}_{t-2} + u_{t-1}
\end{align}
$$
and now note that $u_t = \rho u_{t-1} + \varepsilon_t$. Substituting...
$$
\begin{align}
  \text{Births}_t &= \beta_0 + \beta_1 \text{Income}_t + \beta_2 \textcolor{#6A5ACD}{\text{Births}_{t-1}} + \overbrace{\left( \rho \textcolor{#e64173}{u_{t-1}} + \varepsilon_t \right)}^{u_t} \tag{1} \\
  \textcolor{#6A5ACD}{\text{Births}_{t-1}} &= \beta_0 + \beta_1 \text{Income}_{t-1} + \beta_2 \text{Births}_{t-2} + \textcolor{#e64173}{u_{t-1}} \tag{2}
\end{align}
$$

In $(1)$, we can see that $u_t$ depends upon (covaries with) $\textcolor{#e64173}{u_{t-1}}$.
<br> In $(2)$, we can see that $\textcolor{#6A5ACD}{\text{Births}_{t-1}}$, a regressor in $(1)$, also covaries with $u_{t-1}$.

::: {.fragment}


∴ This model violates our contemporaneous exogeneity requirement.

:::

---

## Autocorrelation
### OLS and lagged outcome variables


*Implications:* For models with **lagged outcome variables** and **autocorrelated disturbances**

1. The models [**violate contemporaneous exogeneity**.]{.pink }

1. OLS is [**biased and inconsistent** for the coefficients.]{.pink }

---

## Autocorrelation
### OLS and lagged outcome variables


*Intuition?* Why is OLS inconsistent and biased when we violate exogeneity?

Think back to omitted-variable bias...

$$
\begin{align}
  y_t &= \beta_0 + \beta_1 x_t + u_t
\end{align}
$$

When $\mathop{\text{Cov}} \left( x_t,\, u_t \right)\neq0$, we cannot separate the effect of $u_t$ on $y_t$ from the effect of $x_t$ on $y_t$. Thus, we get inconsistent estimates for $\beta_1$.

::: {.fragment}

 Similarly,

$$
\begin{align}
  \text{Births}_t &= \beta_0 + \beta_1 \text{Income}_t + \beta_2 \text{Births}_{t-1} + \overbrace{\left( \rho u_{t-1} + \varepsilon_t \right)}^{u_t} \tag{1}
\end{align}
$$

we cannot separate the effects of $u_t$ on $\text{Births}_t$ from $\text{Births}_{t-1}$ on $\text{Births}_{t}$, because both $u_t$ and $\text{Births}_{t-1}$ depend upon $u_{t-1}$.

:::

::: {.fragment}

 $\textcolor{#e64173}{\hat{\beta}_2}$ [is **biased** (w/ OLS).]{.pink }

:::

---

## Autocorrelation and bias
### Simulation


To see how this bias can look, let's run a simulation.

$$
\begin{align}
  y_t = 1 + 2 x_t + 0.5 y_{t-1} + u_t \\
  u_t &= 0.9 u_{t-1} + \varepsilon_t
\end{align}
$$


One (easy) way generate 100 disturbances from AR(1), with $\rho = 0.9$:

```{r}
#| label: "arima.sim"
#| echo: true
#| eval: false
arima.sim(model = list(ar = c(0.9)), n = 100)
```

::: {.fragment}


We are going to run 10,000 iterations with $T=100$.

:::

::: {.fragment}


**Q:**  Will this simulation tell us about *bias* or *consistency*?

:::

::: {.fragment}

<br>**A:** Bias. We would need to let $T\rightarrow\infty$ to consider consistency.

:::

---

## Autocorrelation and bias
### Simulation


Outline of our simulation:

::: {.pseudocode-small}
1. Generate T=100 values of x
2. Generate T=100 values of u

   - Generate T=100 values of ε
   - Use ε and ρ=0.9 to calculate u[t]{.sub } = ρ u[t-1]{.sub } + ε[t]{.sub }

3. Calculate y[t]{.sub } = β[0]{.sub } + β[1]{.sub } x[t]{.sub } + β[2]{.sub } y[t-1]{.sub } + u[t]{.sub }
4. Regress y on x; record estimates

Repeat 1-4 10,000 times
:::

---

```{r}
#| label: "sim-setup"
#| echo: false
# Define parameters
set.seed(1234)
rho <- 0.9
b0 <- 1
b1 <- 2
b2 <- 0.5
t <- 100
```

```{r}
#| label: "sim-gen"
#| echo: false
#| eval: false
# Function for one iteration of the simulation
sim_fun <- function(x, rho, b0, b1, b2, t) {
  # Start generating data (initialize y as 1)
  # NOTE: u and u2 are both AR(1): u is manual; u2 uses canned function 'arima.sim'
  data_x <- tibble(
    e = rnorm(t),
    u = rnorm(1),
    u2 = arima.sim(model = list(ar = c(0.9)), n = t),
    x = rnorm(t),
    y = 1,
    y2 = 1
  )
  # Calculate u and y, iteratively
  for (j in 2:t) {
    data_x$u[j] <- rho * data_x$u[j-1] + data_x$e[j]
    data_x$y[j] <- b0 + b1 * data_x$x[j] + b2 * data_x$y[j-1] + data_x$u[j]
    data_x$y2[j] <- b0 + b1 * data_x$x[j] + b2 * data_x$y2[j-1] + data_x$u2[j]
  }
  # Regression
  lm(y2 ~ x + lag(y2), data = data_x) %>% tidy()
}
# Run the simulation 1,000 times
sim_df <- mclapply(
  X = 1:1e4,
  FUN = sim_fun,
  mc.cores = 8,
  rho = rho, b0 = b0, b1 = b1, b2 = b2, t = t
) %>% bind_rows()
# Save
saveRDS(
  object = sim_df,
  file = "sim.rds"
)
```

::: columns
::: {.column width="55%"}
[Distribution of OLS estimates,]{.b .slate } $\hat{\beta}_2$
:::
::: {.column width="45%"}
$y_t = 1 + 2 x_t + \textcolor{#e64173}{0.5} y_{t-1} + u_t$
:::
:::

```{r}
#| label: "sim-density-b2"
#| echo: false
#| fig-height: 6.75
ggplot(data = readRDS("sim.rds") %>% filter(term == "lag(y2)"), aes(x = estimate)) +
  geom_density(color = NA, fill = red_pink, alpha = 0.95) +
  geom_vline(xintercept = b2, linetype = "solid") +
  geom_vline(
    xintercept =
      readRDS("sim.rds") %>% filter(term == "lag(y2)") %>%
      select(estimate) %>% unlist() %>% mean(),
    linetype = "dashed") +
  geom_hline(yintercept = 0) +
  theme_pander(base_family = "Fira Sans Book", base_size = 20) +
  labs(x = "Estimate", y = "Density")
```

---

::: columns
::: {.column width="55%"}
[Distribution of OLS estimates,]{.b .slate } $\hat{\beta}_1$
:::
::: {.column width="45%"}
$y_t = 1 + \textcolor{#FFA500}{2} x_t + 0.5 y_{t-1} + u_t$
:::
:::

```{r}
#| label: "sim-density-b1"
#| echo: false
#| fig-height: 6.75
ggplot(data = readRDS("sim.rds") %>% filter(term == "x"), aes(x = estimate)) +
  geom_density(color = NA, fill = "orange", alpha = 0.95) +
  geom_vline(
    xintercept =
      readRDS("sim.rds") %>% filter(term == "x") %>%
      select(estimate) %>% unlist() %>% mean(),
    linetype = "dashed") +
  geom_vline(xintercept = b1, linetype = "solid") +
  geom_hline(yintercept = 0) +
  theme_pander(base_family = "Fira Sans Book", base_size = 20) +
  labs(x = "Estimate", y = "Density")
```

---

## {.center}

[Testing for autocorrelation]{.bigger .it .slab}

---

## Testing for autocorrelation
### Static models


Suppose we have the **static model**,
$$
\begin{align}
  \text{Births}_t = \beta_0 + \beta_1 \text{Income}_t + u_t \tag{A}
\end{align}
$$
and we want to test for an AR(1) process in our disturbances $u_t$.

::: {.fragment}


[Test for autocorrelation:]{.b .pink } Test for correlation in the lags of our residuals:

:::

::: {.fragment}


$$
\begin{align}
  e_t = \textcolor{#e64173}{\rho} e_{t-1} + v_t
\end{align}
$$

:::

::: {.fragment}


Does $\textcolor{#e64173}{\hat{\rho}}$ differ significantly from zero?

:::

::: {.fragment}


**Familiar idea:** Use residuals to learn about disturbances.

:::

---

## Testing for autocorrelation
### Static models


Specifically, to test for AR(1) disturbances in the static model
$$
\begin{align}
  \text{Births}_t = \beta_0 + \beta_1 \text{Income}_t + u_t \tag{A}
\end{align}
$$

::: {.pseudocode-small}
1. Estimate $(\text{A})$ via OLS.
2. Calculate residuals from the OLS regression in step 1.
3. Regress the residuals on their lags (without an intercept).

$$
e_t = \rho e_{t-1} + v_t
$$

4. Use a _t_ test to determine whether there is statistically significant evidence that $\rho$ differs from zero.
5. Rejecting H[0]{.sub } implies significant evidence of autocorrelation.
:::

---

For an example, let's return to our plot of negative autocorrelation.

---

[Negative autocorrelation]{.b .slate }: Disturbances $\left( u_t \right)$ over time
```{r}
#| label: "negative-auto-u-2"
#| echo: false
#| fig-height: 7.25
# Number of observations
T <- 1e2
# Rho
rho <- -0.95
# Set seed and starting point
set.seed(1234)
start <- rnorm(1)
# Generate the data
ar_df <- tibble(
  t = 1:T,
  x = runif(T, min = 0, max = 1),
  e = rnorm(T, mean = 0, sd = 2),
  u = NA
)
for (x in 1:T) {
  if (x == 1) {
    ar_df$u[x] <- rho * start + ar_df$e[x]
  } else {
    ar_df$u[x] <- rho * ar_df$u[x-1] + ar_df$e[x]
  }
}
ar_df %<>% mutate(y = 1 + 3 * x + u)
# Plot disturbances over time
ggplot(data = ar_df,
  aes(t, u)
) +
geom_line(color = "grey85", size = 0.35) +
geom_point(color = red_pink, size = 2.25) +
ylab("u") +
xlab("t") +
# theme_pander(base_family = "Fira Sans Book", base_size = 20)
theme_axes_math
```

---

## Testing for autocorrelation
### Example: Static model and AR(1)


**Step 1:** Estimate the static model $\left( y_t = \beta_0 + \beta_1 x_t + u_t \right)$ with OLS

```{r}
#| label: "ex-ar1-1"
#| echo: true
reg_est <- lm(y ~ x, data = ar_df)
```

::: {.fragment}


**Step 2:** Add the residuals to our dataset

```{r}
#| label: "ex-ar1-2"
#| echo: true
ar_df$e <- residuals(reg_est)
```

:::

::: {.fragment}


**Step 3:** Regress the residual on its lag (**no intercept**)

```{r}
#| label: "ex-ar1-3"
#| echo: true
reg_resid <- lm(e ~ -1 + lag(e), data = ar_df)
```

:::

---

## Testing for autocorrelation
### Example: Static model and AR(1)


**Step 4:** _t_ test for the estimated $(\hat{\rho})$ coefficient in step 3.
```{r}
#| label: "ex-ar1-4"
#| echo: true
tidy(reg_resid)
```

::: {.fragment}


That's a very small *p*-value—much smaller than 0.05.

:::

::: {.fragment}


[Reject H[0]{.sub }]{.b .pink } (H[0]{.sub } was $\rho=0$, _i.e._, no autocorrelation).

:::

::: {.fragment}


**Step 5:** Conclude. Statistically significant evidence of autocorrelation.

:::

---

## Testing for autocorrelation
### Example: Static model and AR(3)


What if we wanted to test for AR(3)?

- We add more lags of residuals to the regression in *Step 3*.

- We **jointly** test the significance of the coefficients (_i.e._, $\text{LM}$ or $F$).

Let's do it.

---

## Testing for autocorrelation
### Example: Static model and AR(3)


**Step 1:** Estimate the static model $\left( y_t = \beta_0 + \beta_1 x_t + u_t \right)$ with OLS

```{r}
#| label: "ex-ar3-1"
#| echo: true
reg_est <- lm(y ~ x, data = ar_df)
```

::: {.fragment}


**Step 2:** Add the residuals to our dataset

```{r}
#| label: "ex-ar3-2"
#| echo: true
ar_df$e <- residuals(reg_est)
```

:::

::: {.fragment}


**Step 3:** Regress the residual on its lag (**no intercept**)

```{r}
#| label: "ex-ar3-3"
#| echo: true
reg_ar3 <- lm(e ~ -1 + lag(e) + lag(e, 2) + lag(e, 3), data = ar_df)
```

:::

::: {.fragment}


*Note:* `lag(v, n)` from `dplyr` takes the [n]{.mono }[th]{.super } lag of the variable [v]{.mono }.

:::

---

## Testing for autocorrelation
### Example: Static model and AR(3)


**Step 4:** Calculate the $\text{LM} = n\times R_e^2$ test statistic—distributed $\chi_k^2$.
<br>
$k$ is the number of regressors in the regression in *Step 3* (here, $k=3$).

```{r}
#| label: "ex-ar3-4"
#| echo: true
# Grab R squared
r2_e <- summary(reg_ar3)$r.squared
# Calculate the LM test statistic: n times r2_e
(lm_stat <- 100 * r2_e)
# Calculate the p-value
(pchisq(q = lm_stat, df = 3, lower.tail = F))
```

---

## Testing for autocorrelation
### Example: Static model and AR(3)


**Step 5:** Conclude.

::: {.fragment}


*Recall:* Our hypotheses consider the model
$$
\begin{align}
  e_t = \rho_1 e_{t-1} + \rho_2 e_{t-2} + \rho_3 e_{t-3}
\end{align}
$$

:::

::: {.fragment}


which we are actually using to learn about the model
$$
\begin{align}
  u_t = \rho_1 u_{t-1} + \rho_2 u_{t-2} + \rho_3 u_{t-3}
\end{align}
$$

:::

::: {.fragment}


H[0]{.sub }: $\rho_1 = \rho_2 = \rho_3 = 0$
 *vs.* H[A]{.sub }: $\rho_j \neq 0$ for at least one $j$ in $\{1,\, 2,\, 3\}$

Our p-value is less than 0.05.

:::

::: {.fragment}

[Reject H[0]{.sub }.]{.b .pink }

:::

::: {.fragment}


Conclude there is statistically significant evidence of autocorrelation.

:::

---

## Testing for autocorrelation
### Dynamic models with lagged outcome variables


**Recall:** OLS is biased and inconsistent when our model has *both*

1. a lagged dependent variable

1. autocorrelated disturbances

::: {.fragment}


**Problem:** If OLS is biased for $\beta$, then it is also biased for $u_t$.

:::

::: {.fragment}


∴ We can't apply our nice trick of *just* using $e_t$ to learn about $u_t$.

:::

::: {.fragment}


**Solution:** [Breusch-Godfrey]{.b .purple } test includes the other explanatory variables,

$$
\begin{align}
  e_t = \textcolor{#6A5ACD}{\underbrace{\gamma_0 + \gamma_1 x_{1t} + \gamma_2 x_{2t} + \cdots}_{\text{Explanatory variables (RHS)}}} + \textcolor{#e64173}{\underbrace{\rho_1 e_{t-1} + \rho_2 e_{t-2} + \cdots}_{\text{Lagged residuals}}} + \varepsilon_t
\end{align}
$$

:::

---

## Testing for autocorrelation
### Dynamic models with lagged outcome variables


Specifically, to test for AR(2) disturbances in the ADL(1, 0) model
$$
\begin{align}
  \text{Births}_t = \beta_0 + \beta_1 \text{Income}_t + \beta_2 \text{Births}_{t-1} + u_t \tag{B}
\end{align}
$$

::: {.pseudocode-small}
1. Estimate $(\text{B})$ via OLS.
2. Calculate residuals (e[t]{.sub }) from the OLS regression in step 1.
3. Regress residuals on an intercept, explanatory variables, and lagged residuals.

$$
e_t = \gamma_0 + \gamma_1 \text{Income}_t + \gamma_3 \text{Births}_{t-1} + \rho_1 e_{t-1} + \rho_2 e_{t-2} + v_t
$$

4. Conduct LM or F test for ρ[1]{.sub } = ρ[2]{.sub } = 0.
5. Rejecting H[0]{.sub } implies significant evidence of AR(2).
:::

---

## Testing for autocorrelation
### Dynamic models with lagged outcome variables


For an example, let's consider the relationship between monthly presidential approval ratings and oil prices during President George W. Bush's<sup>[†]{.pink }</sup> presidency.

We will specify the process as ADL(1, 0) and test for an AR(2) process in our disturbances.

$$
\begin{align}
  \text{Approval}_t = \beta_0 + \beta_1 \text{Approval}_{t-1} + \beta_2 \text{Price}_t + u_t
\end{align}
$$

^[[†]{.pink } [Fun with approval ratings](https://projects.fivethirtyeight.com/trump-approval-ratings/).]

::: {.fragment}


*Note:* We're ignoring any other violations of exogeneity for the moment.


:::

---

## Hidden {visibility="hidden"}

```{r}
#| label: "approval-0"
#| include: false
# Load dataset
approval_df <- read_csv("approval.csv") %>% mutate(
  time = year + (month - 1) / 12
) %>%
  janitor::clean_names() %>%
  rename(price_oil = avg_price) %>%
  mutate(ap = approve)
```

---

[Monthly presidential approval ratings, 2001–2006]{.b .slate }
```{r}
#| label: "approval-1"
#| echo: false
ggplot(data = approval_df, aes(x = time, y = ap)) +
  geom_line(size = 0.5, alpha = 0.3) +
  geom_point(size = 2.5, color = red_pink) +
  xlab("Date") +
  ylab("Rating") +
  theme_axes_serif
```

---

[Approval rating vs. its one-month lag, 2001–2006]{.b .slate }
```{r}
#| label: "approval-2"
#| echo: false
ggplot(data = approval_df, aes(x = lag(approve), y = ap)) +
  geom_smooth(method = lm, se = F, linetype = "dashed", color = "darkslategrey") +
  geom_point(size = 2.5, color = red_pink) +
  xlab("Previous") +
  ylab("Current") +
  theme_axes_serif
```

---

[Approval rating vs. its two-month lag, 2001–2006]{.b .slate }
```{r}
#| label: "approval-3"
#| echo: false
ggplot(data = approval_df, aes(x = lag(ap, 2), y = ap)) +
  geom_smooth(method = lm, se = F, linetype = "dashed", color = "darkslategrey") +
  geom_point(size = 2.5, color = red_pink) +
  xlab("Previous") +
  ylab("Current") +
  theme_axes_serif
```

---

[Oil prices, 2001–2006]{.b .slate }
```{r}
#| label: "approval-4"
#| echo: false
ggplot(data = approval_df, aes(x = time, y = price_oil)) +
  geom_line(size = 0.5, alpha = 0.3) +
  geom_point(size = 2.5, color = red_pink) +
  xlab("Date") +
  ylab("$") +
  theme_axes_serif
```

---

[Approval rating vs. oil prices, 2001–2006]{.b .slate }
```{r}
#| label: "approval-5"
#| echo: false
ggplot(data = approval_df, aes(x = price_oil, y = ap)) +
  geom_smooth(method = lm, se = F, linetype = "dashed", color = "darkslategrey") +
  geom_point(size = 2.5, color = red_pink) +
  xlab("Oil price ($)") +
  ylab("Rating") +
  theme_axes_serif
```

---

## Testing for autocorrelation
### Example: Approval ratings and oil prices


**Step 1:** Estimate our ADL(1, 0) model with OLS.

```{r}
#| label: "bg1"
#| echo: true
# Estimate the model
ols_est <- lm(
  approve ~ lag(approve) + price_oil,
  data = approval_df
)
# Summary
tidy(ols_est)
```

---

## Testing for autocorrelation
### Example: Approval ratings and oil prices


**Step 2:** Record residuals from the OLS regression.
```{r}
#| label: "bg2"
#| echo: true
# Grab residuals
approval_df$e <- c(NA, residuals(ols_est))
```

::: {.fragment}


**Note:** We add an `NA` because we use a lag—the first element is missing.

:::

::: {.fragment}


*E.g.*,
<br>[{1, 2, 3, 4, 5, 6, 7, 8, 9} = x]{.mono }

:::

::: {.fragment}

<br>[{[?]{.pink }, 1, 2, 3, 4, 5, 6, 7, 8} = lag(x)]{.mono }

:::

::: {.fragment}

<br>[{[?]{.pink }, [?]{.pink }, 1, 2, 3, 4, 5, 6, 7} = lag(x, 2)]{.mono }

:::

::: {.fragment}

<br>[{[?]{.pink }, [?]{.pink }, [?]{.pink }, 1, 2, 3, 4, 5, 6} = lag(x, 3)]{.mono }

:::

---

## Testing for autocorrelation
### Example: Approval ratings and oil prices


```{r}
#| label: "bg2b"
#| echo: false
#| fig-height: 5.75
ggplot(data = approval_df, aes(x = time, y = e)) +
  geom_line(size = 0.5, alpha = 0.3) +
  geom_point(size = 2.5, color = red_pink) +
  xlab("Time") +
  ylab("e") +
  theme_axes_serif
```

---

## Testing for autocorrelation
### Example: Approval ratings and oil prices


**Step 3:** Regress residuals on an intercept, the explanatory variables, and lagged residuals.

```{r}
#| label: "bg3"
#| echo: true
# BG regression
bg_reg <- lm(
  e ~ lag(approve) + price_oil + lag(e) + lag(e, 2),
  data = approval_df
)
```

```{r}
#| label: "bg3b"
#| echo: false
summary(bg_reg) %>% capture.output() %>% extract(11:16) %>% paste0("\n") %>% cat()
```

---

## Testing for autocorrelation
### Example: Approval ratings and oil prices


**Step 4:** $F$ (or $\text{LM}$) test for $\rho_1=\rho_2=0$.

*Recall:* We can test joint significance using an $F$ test that compares the restricted (here: $\rho_1=\rho_2=0$) and unrestricted models.
$$
\begin{align}
  F_{q,\,n-p} = \dfrac{\left(\text{SSE}_r - \text{SSE}_u\right)\big/q}{\text{SSE}_u\big/\left( n-p \right)}
\end{align}
$$
where $q$ is the number of restrictions and $p$ is the number of parameters in our unrestricted model (include the intercept).

We can use the `waldtest()` function from the `lmtest` package for this test.

---

## Testing for autocorrelation
### Example: Approval ratings and oil prices


**Step 4:** $F$ (or $\text{LM}$) test for $\rho_1=\rho_2=0$.

```{r}
#| label: "bg4"
#| echo: true
#| eval: false
# BG regression
bg_reg <- lm(
  e ~ lag(approve) + price_oil + lag(e) + lag(e, 2),
  data = approval_df
)
# Test significance of the lags using 'waldtest' from 'lmtest' package
p_load(lmtest)
waldtest(bg_reg, c("lag(e)", "lag(e, 2)"))
```

Here, we're telling `waldtest` to test

- the model we specified in `bg_reg` (our [unrestricted model]{.b .pink })
- against a model without `lag(e)` and `lag(e, 2)` (our [restricted model]{.b .pink })

---

## Testing for autocorrelation
### Example: Approval ratings and oil prices


count: false

**Step 4:** $F$ (or $\text{LM}$) test for $\rho_1=\rho_2=0$.

```{r}
#| label: "bg4b"
#| echo: true
# BG regression
bg_reg <- lm(
  e ~ lag(approve) + price_oil + lag(e) + lag(e, 2),
  data = approval_df
)
# Test significance of the lags using 'waldtest' from 'lmtest' package
p_load(lmtest)
waldtest(bg_reg, c("lag(e)", "lag(e, 2)"))
```

---

## Testing for autocorrelation
### Example: Approval ratings and oil prices


**Step 5:** Conclusion of hypothesis test

With a *p*-value of ∼`r waldtest(bg_reg, c("lag(e)", "lag(e, 2)"))$"Pr(>F)"[2] %>% round(3)`, [we fail to reject the null hypothesis.]{.b .pink }

- We cannot reject $\rho_1=\rho_2=0$.

- We cannot reject "no autocorrelation".

::: {.fragment}


```{r}
#| label: "bg5"
#| include: false
# BG regression
bg_reg_ar1 <- lm(
  e ~ lag(approve) + price_oil + lag(e),
  data = approval_df
)
```

However, [we tested for a specific type of autocorrelation]{.b .pink }: AR(2).

We might get different answers with different tests.

The *p*-value for AR(1) is `r waldtest(bg_reg_ar1, c("lag(e)"))$"Pr(>F)"[2] %>% round(4)`—suggestive of first-order autocorrelation.

:::

---

## {.center}

[Living with autocorrelation]{.bigger .it .slab}

---

## Autocorrelation
### Working with it


Suppose we believe autocorrelation is present. What do we do?

::: {.fragment}


I'll give you three options.<sup>[†]{.pink }</sup>

1. [Misspecification]{.b .pink }

1. [Serial-correlation robust standard errors]{.b .pink } (a.k.a. *Newey-West*)

1. [FGLS]{.b .pink }

^[[†]{.pink } You should take [EC 422]{.mono } to go much deeper into time-series analysis/forecasting.]

:::

---

## Autocorrelation
### Option 1: Misspecification


[Misspecification]{.b .pink } with autocorrelation is very similar to our discussion with heteroskedasticity.

By incorrectly specifying your model, you can create autocorrelation.

Omitting variables that are correlated through time will cause your disturbances to be correlated through time.

---

## Autocorrelation
### Option 1: Misspecification


**Example:** Suppose births depend upon income and previous births
$$
\begin{align}
  \text{Births}_t = \beta_0 + \beta_1 \text{Births}_{t-1} + \beta_2 \text{Income}_t + u_t
\end{align}
$$

::: {.fragment}

but we write down the model as only depending upon previous births, _i.e._,
$$
\begin{align}
  \text{Births}_t = \beta_0 + \beta_1 \text{Births}_{t-1} + v_t
\end{align}
$$

:::

::: {.fragment}

Then our disturbance $v_t$ is
$$
\begin{align}
  v_t = \beta_2 \text{Income}_t + u_t
\end{align}
$$

:::

::: {.fragment}

which is likely autocorrelated, since income is correlated in time.

:::

::: {.fragment}


*Note:* This autocorrelation has nothing to do with $u_t$.

:::

---

## Autocorrelation
### Option 1: Misspecification


**"Proof"**

$$
\begin{align}
  v_{t} &= \beta_2 \text{Income}_{t} + u_{t} \\
  v_{t-1} &= \beta_2 \text{Income}_{t-1} + u_{t-1}
\end{align}
$$

$\mathop{\text{Cov}} \left( v_{t},\,v_{t-1} \right)$

::: {.fragment}


::: {.pad-left}
$= \mathop{\text{Cov}} \left( \beta_2 \text{Income}_{t} + u_{t},\, \beta_2 \text{Income}_{t-1} + u_{t-1} \right)$
:::

:::

::: {.fragment}


::: {.pad-left}
$= \textcolor{#e64173}{\mathop{\text{Cov}} \left( \beta_2 \text{Income}_{t},\, \beta_2 \text{Income}_{t-1} \right)} + \mathop{\text{Cov}} \left( \beta_2 \text{Income}_{t},\, u_t \right)$
<br> $\textcolor{#ffffff}{=} + \mathop{\text{Cov}} \left(u_{t},\, \beta_2 \text{Income}_{t-1}\right) + \mathop{\text{Cov}} \left( u_{t},\, u_{t-1} \right)$
:::

:::

::: {.fragment}


::: {.pad-left}
$\neq 0$ (in general) even if $u_t$ is exogenous and without autocorrelation.
:::


:::

---

## Autocorrelation
### Option 2: Newey-West standard errors


As was also the case with heteroskedasticity, you can still estimate consistent standard errors (and inference) in the presence of autocorrelation.

These standard errors are called *serial-correlation robust standard errors* (or *Newey-West standard errors*).

We are not going to derive the estimator for these standard errors.

---

## Autocorrelation
### Option 3: FGLS


If we do not have a lagged outcome variable, then [feasible generalized least squares]{.pink } ([FGLS]{.b .pink }) can give us efficient and consistent standard errors.

::: {.fragment}


Let's start with a simple static model that includes an AR(1) disturbance $u_t$.

$$
\begin{align}
  \text{Births}_t &= \beta_0 + \beta_1 \text{Income}_t + u_t \tag{1} \\
  u_t &= \rho u_{t-1} + \varepsilon_t \tag{2}
\end{align}
$$

:::

::: {.fragment}


Now our old trick: Write out $(1)$ for period $t-1$ (and then multiple by $\rho$)
$$
\begin{align}
  \text{Births}_{t-1} &= \beta_0 + \beta_1 \text{Income}_{t-1} + u_{t-1} \tag{3} \\
  \rho \text{Births}_{t-1} &= \rho \beta_0 + \rho \beta_1 \text{Income}_{t-1} + \rho u_{t-1} \tag{4}
\end{align}
$$

:::

::: {.fragment}


And now subtract $(4)$ from $(1)$…

:::

---

## Autocorrelation
### Option 3: FGLS


$$
\begin{align}
  \text{Births}_t - \rho \text{Births}_{t-1} =& \beta_0 \left( 1 - \rho \right) + \\
  & \beta_1 \text{Income}_t - \rho \beta_1 \text{Income}_{t-1} + \\
  & u_t - \rho u_{t-1}
\end{align}
$$

::: {.fragment}

which gives us a very specific dynamic model
$$
\begin{align}
  \text{Births}_t =& \beta_0 \left( 1 - \rho \right) + \rho \text{Births}_{t-1} +\\
  & \beta_1 \text{Income}_t - \rho \beta_1 \text{Income}_{t-1} + \\
  & \textcolor{#e64173}{\underbrace{u_t - \rho u_{t-1}}_{=\varepsilon_t}} \\
  \textcolor{#ffffff}{=}& \textcolor{#ffffff}{\beta_0 \left( 1 - \rho \right) + \rho \text{Births}_{t-1} +}\\
  &\textcolor{#ffffff}{ \beta_1 \text{Income}_t - \rho \beta_1 \text{Income}_{t-1} + \varepsilon_t}
\end{align}
$$

:::

---

## Autocorrelation
### Option 3: FGLS


count: false

$$
\begin{align}
  \text{Births}_t - \rho \text{Births}_{t-1} =& \beta_0 \left( 1 - \rho \right) + \\
  & \beta_1 \text{Income}_t - \rho \beta_1 \text{Income}_{t-1} + \\
  & u_t - \rho u_{t-1}
\end{align}
$$
which gives us a very specific dynamic model
$$
\begin{align}
  \text{Births}_t =& \beta_0 \left( 1 - \rho \right) + \rho \text{Births}_{t-1} +\\
  & \beta_1 \text{Income}_t - \rho \beta_1 \text{Income}_{t-1} + \\
  & \textcolor{#e64173}{\underbrace{u_t - \rho u_{t-1}}_{=\varepsilon_t}} \\
  =& \beta_0 \left( 1 - \rho \right) + \rho \text{Births}_{t-1} +\\
  & \beta_1 \text{Income}_t - \rho \beta_1 \text{Income}_{t-1} + \textcolor{#e64173}{\varepsilon_t}
\end{align}
$$
that happens to be [free of autocorrelation]{.b .pink }.

---

## Autocorrelation
### Option 3: FGLS


This [transformed model]{.b .pink } is free of autocorrelation.
$$
\begin{align}
  \text{Births}_t =& \beta_0 \left( 1 - \rho \right) + \rho \text{Births}_{t-1} +\\
  & \beta_1 \text{Income}_t - \rho \beta_1 \text{Income}_{t-1} + \textcolor{#e64173}{\varepsilon_t}
\end{align}
$$

**Q:** How do we actually estimate this model? (We don't know $\rho$.)

::: {.fragment}

<br>**A:** FGLS (of course)…

:::

::: {.fragment}


::: {.pseudocode-small}
1. Estimate the original (untransformed) model; save residuals.
2. Estimate $\rho$: Regress residuals on their lags (no intercept).
3. Estimate the **transformed model**, plugging in $\hat{\rho}$ for $\rho$.
:::

:::

---

## Table of contents

::: columns
::: {.column width="42%"}
#### Admin

::: {.smallest}
1. [Schedule](#schedule)
2. [[R]{.mono } showcase](#r_showcase)

   - [[ggplot2]{.mono }](#ggplot2)
   - [Writing functions](#functions)

3. [Review: Time series](#review)
:::
:::
::: {.column width="58%"}
#### Autocorrelation

::: {.smallest}
1. [Introduction](#intro)
2. [In static models](#static_models)
3. [OLS and bias/consistency](#ols_static)

   - [Static models](#ols_static)
   - [Dynamic models with lagged $y$](#ols_adl)

4. [Simulation: Bias](#ar2_simulation)
5. [Testing for autocorrelation](#testing)

   - [Static models](#testing_static)
   - [Dynamic models with lagged $y$](#testing_adl)

6. [Working with autocorrelation](#fixes)

   - [Misspecification](#misspecification)
   - [Newey-West standard errors](#newey)
   - [FGLS](#gls)
:::
:::
:::
