---
title: 'Non-Stationary Time Series'
subtitle: 'EC 421, Set 9'
author:
  - name: Ed Rubin
    email: edwardr@uoregon.edu
editor:
  render-on-save: true
format:
  clean-revealjs:
    slide-level: 2
    theme: [../my.scss, default]
    highlight-style: gruvbox
    embed-resources: true
    slide-number: c/t
    incremental: false
    controls-layout: edges
    transition: none
    preview-links: auto
    html-math-method: mathjax
    code-line-numbers: false
execute:
  warning: false
  message: false

---

## {.center}


[Prologue]{.bigger .it .slab}

---

## Schedule


```{r}
#| label: "setup"
#| include: false
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(
  broom, here, tidyverse,
  latex2exp, ggplot2, ggthemes, viridis, extrafont, gridExtra,
  kableExtra,
  data.table,
  dplyr,
  lubridate,
  magrittr, knitr, parallel
)
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -0.5, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_simple <- theme_bw() + theme(
  line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_text(size = 18, family = "STIXGeneral"),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  # plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_math <- theme_void() + theme(
  text = element_text(family = "MathJax_Math"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_serif <- theme_void() + theme(
  text = element_text(family = "MathJax_Main"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes <- theme_void() + theme(
  text = element_text(family = "Fira Sans Book"),
  axis.title = element_text(size = 18),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = grey_light,
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_set(theme_gray(base_size = 20))
# A few extras
```

```{css}
#| echo: false

@media print {
  .has-continuation {
    display: block !important;
  }
}
```

### Last Time

Autocorrelation

### Today

A brief introduction to nonstationarity

---

## {.center}

[Nonstationarity]{.bigger .it .slab}

---

## Nonstationarity


### Intro

Let's go back to our assumption of [weak dependence/persistence]{.b .pink }

> 1. **Weakly persistent outcomes**—essentially, $x_{t+k}$ in the distant period $t+k$ weakly correlates with $x_t$ (when $k$ is "big").

::: {.fragment}


We're essentially saying we need the time series $x$ to behave.

:::

::: {.fragment}


We'll define this *good behavior* as [stationarity]{.b .pink }.

:::

---

## Nonstationarity


### Stationarity

Requirements for [stationarity]{.b .pink } (a *stationary* time-series process):

::: {.fragment}


1. The [mean]{.b .pink } of the distribution is independent of time, _i.e._,

   $$
   \mathop{\boldsymbol{E}}\left[ x_t \right]  = \mathop{\boldsymbol{E}}\left[ x_{t-k} \right]
   $$
   for all $k$

:::

::: {.fragment}


2. The [variance]{.b .pink } of the distribution is independent of time, _i.e._,

   $$
   \mathop{\text{Var}} \left( x_t \right) = \mathop{\text{Var}} \left( x_{t-k} \right)
   $$
   for all $k$

:::

::: {.fragment}

3. The [covariance]{.b .pink } between $x_t$ and $x_{t-k}$ depends only on $k$—[not on]{.pink } $\textcolor{#e64173}{t}$, _i.e._,

   $$
   \mathop{\text{Cov}} \left( x_t,\,x_{t-k} \right) = \mathop{\text{Cov}} \left( x_s,\, x_{s-k} \right)
   $$
   for all $t$ and $s$

:::

---

## Nonstationarity


### Random walks

[Random walks]{.b .pink } are a famous example of a nonstationary process:

::: {.fragment}


$$
\begin{align}
  x_t = x_{t-1} + \varepsilon_t
\end{align}
$$

:::

::: {.fragment}


Why?

:::

::: {.fragment}



 $\mathop{\text{Var}} \left( x_t \right) = t \sigma_\varepsilon^2$, which [violates stationary variance]{.pink }.

:::

::: {.fragment}


$$
\begin{align}
   \mathop{\text{Var}} \left( x_t \right)
   &= \mathop{\text{Var}} \left( x_{t-1} + \varepsilon_t \right) \\
   &= \mathop{\text{Var}} \left( x_{t-2} + \varepsilon_{t-1} + \varepsilon_t \right) \\
   &= \mathop{\text{Var}} \left( x_{t-3} + \varepsilon_{t-2} + \varepsilon_{t-1} + \varepsilon_t \right) \\
   &\cdots \\
   &= \mathop{\text{Var}} \left( x_0 + \varepsilon_1 + \cdots + \varepsilon_{t_2} + \varepsilon_{t-1} + \varepsilon_t \right) \\
   &= \sigma^2_\varepsilon + \cdots + \sigma^2_\varepsilon + \sigma^2_\varepsilon + \sigma^2_\varepsilon \\
   &= t \sigma^2_\varepsilon
\end{align}
$$


:::

---

**Q:** What's the big deal with this violation?

---

[One 100-period random walk]{.b .slate }
```{r}
#| label: "walk1"
#| echo: false
set.seed(1246)
walk1 <- tibble(x = cumsum(rnorm(1e2)), t = 1:1e2, walk = "1")
walk2 <- tibble(x = cumsum(rnorm(1e2)), t = 1:1e2, walk = "2")
walk3 <- tibble(x = cumsum(rnorm(1e2)), t = 1:1e2, walk = "3")
walk4 <- tibble(x = cumsum(rnorm(1e2)), t = 1:1e2, walk = "4")
walk5 <- tibble(x = cumsum(rnorm(1e2)), t = 1:1e2, walk = "5")
ggplot(data = walk1, aes(x = t, y = x)) +
  geom_hline(yintercept = 0, color = "grey85", size = 1.25) +
  geom_path() +
  theme_axes_math
```

---

[Two 100-period random walks]{.b .slate }
```{r}
#| label: "walk2"
#| echo: false
ggplot(data = bind_rows(walk1, walk2), aes(x = t, y = x, color = walk)) +
  geom_hline(yintercept = 0, color = "grey85", size = 1.25) +
  geom_path() +
  scale_color_viridis(option = "magma", discrete = T, begin = 0.15, end = 0.85) +
  theme_axes_math
```

---

[Three 100-period random walks]{.b .slate }
```{r}
#| label: "walk3"
#| echo: false
ggplot(data = bind_rows(walk1, walk2, walk3), aes(x = t, y = x, color = walk)) +
  geom_hline(yintercept = 0, color = "grey85", size = 1.25) +
  geom_path() +
  scale_color_viridis(option = "magma", discrete = T, begin = 0.15, end = 0.85) +
  theme_axes_math
```

---

[Four 100-period random walks]{.b .slate }
```{r}
#| label: "walk4"
#| echo: false
ggplot(data = bind_rows(walk1, walk2, walk3, walk4), aes(x = t, y = x, color = walk)) +
  geom_hline(yintercept = 0, color = "grey85", size = 1.25) +
  geom_path() +
  scale_color_viridis(option = "magma", discrete = T, begin = 0.15, end = 0.85) +
  theme_axes_math
```

---

[Five 100-period random walks]{.b .slate }
```{r}
#| label: "walk5"
#| echo: false
ggplot(data = bind_rows(walk1, walk2, walk3, walk4, walk5), aes(x = t, y = x, color = walk)) +
  geom_hline(yintercept = 0, color = "grey85", size = 1.25) +
  geom_path() +
  scale_color_viridis(option = "magma", discrete = T, begin = 0.15, end = 0.85) +
  theme_axes_math
```

---

[Fifty 100-period random walks]{.b .slate }
```{r}
#| label: "walk50"
#| echo: false
#| cache: true
# Set seed
set.seed(1246)
# Generate data
walk_df <- lapply(X = 1:50, FUN = function(i) {
  tibble(x = cumsum(rnorm(1e2)), t = 1:1e2, walk = as.character(i))
}) %>% bind_rows()
# Plot
ggplot(data = walk_df, aes(x = t, y = x, color = walk)) +
  geom_hline(yintercept = 0, color = "grey85", size = 1.25) +
  geom_path(size = 0.25) +
  scale_color_viridis(option = "magma", discrete = T, end = 0.95) +
  theme_axes_math
```

---

[1,000 100-period random walks]{.b .slate }
```{r}
#| label: "walk1000"
#| echo: false
#| cache: true
# Set seed
set.seed(1246)
# Generate data
walk_df <- lapply(X = 1:1e3, FUN = function(i) {
  tibble(x = cumsum(rnorm(1e2)), t = 1:1e2, walk = as.character(i))
}) %>% bind_rows()
# Plot
ggplot(data = walk_df, aes(x = t, y = x, color = walk)) +
  geom_hline(yintercept = 0, color = "grey85", size = 1.25) +
  geom_path(size = 0.25) +
  scale_color_viridis(option = "magma", discrete = T, end = 0.95) +
  theme_axes_math
```

---

## Nonstationarity
### Problem

*One* problem is that nonstationary processes can lead to [spurious]{.b .pink } results.

::: {.fragment}


>**Defintion:** [Spurious]{.b .pink }
>- not being what it purports to be; false or fake
>- apparently but not actually valid

:::

::: {.fragment}


Back in 1974, Granger and Newbold showed that when they **generated random walks** and **regressed the random walks on each other**, [77/100 regressions were statistically significant]{.b .pink } at the 5% level (should have been approximately 5/100).

:::

---

[Granger and Newbold simulation example:]{.b .slate } _t_ statistic ≈ `r lm(filter(walk_df, walk == 1)$x ~ filter(walk_df, walk == 2)$x) %>% summary %>% coef %>% extract(2,3) %>% round(2)`

```{r}
#| label: "gb12"
#| echo: false
ggplot(data = walk_df %>% filter(walk %in% c(1:2)), aes(x = t, y = x, color = walk)) +
  geom_hline(yintercept = 0, color = "grey85", size = 1.25) +
  geom_path() +
  scale_color_viridis(option = "magma", discrete = T, begin = 0.15, end = 0.85) +
  theme_axes_math
```

---

[Granger and Newbold simulation example:]{.b .slate } _t_ statistic ≈ `r lm(filter(walk_df, walk == 3)$x ~ filter(walk_df, walk == 4)$x) %>% summary %>% coef %>% extract(2,3) %>% round(2)`

```{r}
#| label: "gb34"
#| echo: false
ggplot(data = walk_df %>% filter(walk %in% c(3:4)), aes(x = t, y = x, color = walk)) +
  geom_hline(yintercept = 0, color = "grey85", size = 1.25) +
  geom_path() +
  scale_color_viridis(option = "magma", discrete = T, begin = 0.15, end = 0.85) +
  theme_axes_math
```

---

[Granger and Newbold simulation example:]{.b .slate } _t_ statistic ≈ `r lm(filter(walk_df, walk == 5)$x ~ filter(walk_df, walk == 6)$x) %>% summary %>% coef %>% extract(2,3) %>% round(2)`

```{r}
#| label: "gb56"
#| echo: false
ggplot(data = walk_df %>% filter(walk %in% c(5:6)), aes(x = t, y = x, color = walk)) +
  geom_hline(yintercept = 0, color = "grey85", size = 1.25) +
  geom_path() +
  scale_color_viridis(option = "magma", discrete = T, begin = 0.15, end = 0.85) +
  theme_axes_math
```

---

## Nonstationarity


### Problem

```{r}
#| label: "gb_all"
#| echo: false
#| cache: true
walk_pv <- lapply(X = seq(1, 999, by = 2), FUN = function(i) {
  lm(filter(walk_df, walk == i)$x ~ filter(walk_df, walk == (i+1))$x) %>% summary() %>% coef() %>% extract(2,4)
}) %>% unlist()
```

In our data, `r 100*mean(walk_pv < 0.05)` percent of (independently generated) pairs reject the null hypothesis at the 5% level.

::: {.fragment}


**The point?**

:::

::: {.fragment}

 If our disturbance is nonstationary, we cannot trust plain OLS.

:::

::: {.fragment}


Random walks are only one example of [nonstationary processes]{.pink }...

[Random walk:]{.b .pink } $u_t = u_{t-1} + \varepsilon_t$

:::

::: {.fragment}


[Random walk with drift:]{.b .pink } $u_t = \alpha_0 + u_{t-1} + \varepsilon_t$

:::

::: {.fragment}


[Deterministic trend:]{.b .pink } $u_t = \alpha_0 + \beta_1 t + \varepsilon_t$

:::

---

## Nonstationarity
### A potential solution


Some processes are [difference stationary]{.b .pink }, which means we can get back to our stationarity (good behavior) requirement by taking the difference between $u_t$ and $u_{t-1}$.

::: {.fragment}


[Nonstationary:]{.b .slate } $u_t = u_{t-1} + \varepsilon_t$ [(a random walk)]{.slate }

:::

::: {.fragment}

<br>[Stationary:]{.b .pink } $u_t - u_{t-1} = u_{t-1} + \varepsilon_t - u_{t-1} = \textcolor{#e64173}{\varepsilon_t}$

:::

::: {.fragment}


So if we have good reason to believe that our disturbances follow a random walk, we can use OLS on the differences, _i.e._,

:::

::: {.fragment}

$$
\begin{align}
  y_t &= \beta_0 + \beta_1 x_t + u_t \\
  y_{t-1} &= \beta_0 + \beta_1 x_{t-1} + u_{t-1} \\
  y_t - y_{t-1} &= \beta_1 \left( x_t - x_{t-1} \right) + \left( u_t - u_{t-1} \right) \\
  \Delta y_t &= \beta_1 \Delta x_t + \Delta u_t
\end{align}
$$

:::

---

## Nonstationarity
### Testing

[Dickey-Fuller]{.pink } and [augmented Dickey-Fuller]{.pink } tests are popular ways to test for random walks and other forms of nonstationarity.

::: {.fragment}


[Dickey-Fuller tests]{.b .pink } compare

H[o]{.sub }: $y_t = \beta_0 + \beta_1 y_{t-1} + u_t$ with $|\beta_1|<1$ ([stationarity]{.b .pink })
<br>
H[a]{.sub }: $y_t = y_{t-1} + \varepsilon_t$ ([random walk]{.b .pink })

:::

::: {.fragment}


using a *t* test that $|\beta_1|<1$.<sup>[†]{.pink }</sup>

^[[†]{.pink } People often just test $\beta_1<1$.]

:::

---

## Table of contents

::: columns
::: {.column width="35%"}
#### Admin

::: {.smallest}
1. [Schedule](#schedule)
:::
:::
::: {.column width="65%"}
#### Nonstationarity

::: {.smallest}
1. [Introduction](#intro)
2. [Random walks](#walks)
3. [The actual problem](#problem)
4. [A potential solution](#solution)
5. [Dickey-Fuller tests](#test)
:::
:::
:::

---

## Hidden {visibility="hidden"}

```{r}
#| label: "print-pdfs"
#| echo: false
#| eval: false
pagedown::chrome_print(
  input = "09-nonstationarity.html",
  timeout = 60
)
```
