---
title: 'Asymptotics and consistency'
subtitle: 'EC 421, Set 6'
author:
  - name: Ed Rubin
    email: edwardr@uoregon.edu
editor:
  render-on-save: true
format:
  clean-revealjs:
    slide-level: 2
    theme: [../my.scss, default]
    highlight-style: gruvbox
    embed-resources: true
    slide-number: c/t
    incremental: false
    controls-layout: edges
    transition: none
    preview-links: auto
    html-math-method: mathjax
    code-line-numbers: false
execute:
  warning: false
  message: false

---

## {.center}

[Prologue]{.bigger .it .slab}

---

## Schedule

```{r}
#| label: "setup"
#| include: false
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(
  broom,
  latex2exp, ggplot2, ggthemes, viridis, extrafont, gridExtra,
  kableExtra,
  dplyr, magrittr, knitr, parallel
)
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -0.5, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
```

```{css}
#| echo: false

@media print {
  .has-continuation {
    display: block !important;
  }
}
```

### Last Time

Living with heteroskedasticity

### Today

Asymptotics and consistency

### Next

- New problem set
- _Next topic:_ Time series
- Midterm next week (May 7th)

---

## [R]{.mono } showcase

Need speed? [R]{.mono } allows essentially infinite parallelization.

Three popular packages:

- [`future`](https://github.com/HenrikBengtsson/future) and [`furrr`](https://github.com/DavisVaughan/furrr)

- `parallel`

- `foreach`

And here's a nice [tutorial](https://nceas.github.io/oss-lessons/parallel-computing-in-r/parallel-computing-in-r.html).

---

## {.center}

[Consistency]{.bigger .it .slab}

---

## Consistency
### Welcome to asymptopia


*Previously:* We examined estimators (_e.g._, $\hat{\beta}_j$) and their properties using

::: {.fragment}


1. The **mean**  of the estimator's distribution: $\mathop{\boldsymbol{E}}\left[ \hat{\beta}_j \right] = ?$

:::

::: {.fragment}


2. The **variance**  of the estimator's distribution: $\mathop{\text{Var}} \left( \hat{\beta}_j \right) = ?$

:::

::: {.fragment}


which tell us about the [tendency of the estimator]{.b .pink } if we took [∞ samples]{.b .pink }, each with [sample size]{.b .pink } $\textcolor{#e64173}{n}$.

:::

::: {.fragment}


This approach misses something.

:::

---

## Consistency
### Welcome to asymptopia


[New question:]{.b .pink }<br>How does our estimator behave as our sample gets larger (as $n\rightarrow\infty$)?

::: {.fragment}


This *new question* forms a new way to think about the properties of estimators: [asymptotic properties]{.b .pink } (or large-sample properties).

:::

::: {.fragment}


A "good" estimator will become indistinguishable from the parameter it estimates when $n$ is very large (close to $\infty$).

:::

---

## Consistency
### Probability limits


Just as the *expected value* helped us characterize **the finite-sample distribution of an estimator** with sample size $n$,

::: {.fragment}

the [*probability limit*]{.pink } helps us analyze [the asymptotic distribution of an estimator]{.b .pink } (the distribution of the estimator as $n$ gets "big"^[Here, "big" $n$ means $n\rightarrow\infty$. That's *really* big data.]).

:::

---

## Consistency
### Probability limits


Let $B_n$ be our estimator with sample size $n$.

Then the [probability limit]{.b .pink } of $B$ is $\alpha$ if

$$ \lim_{n\rightarrow\infty} \mathop{P}\left( \middle| B_n - \alpha \middle| > \epsilon \right) = 0 \tag{1} $$

for any $\varepsilon > 0$.

::: {.fragment}


The definition in $(1)$ *essentially* says that as the [sample size]{.pink } approaches infinity, the probability that $B_n$ differs from $\alpha$ by more than a very small number $(\epsilon)$ is zero.

:::

::: {.fragment}


*Practically:* $B$'s distribution collapses to a spike at $\alpha$ as $n$ approaches $\infty$.

:::

---

## Consistency
### Probability limits


Equivalent statements:

- The probability limit of $B_n$ is $\alpha$.

- $\text{plim}\: B = \alpha$

- $B$ converges in probability to $\alpha$.

---

## Consistency
### Probability limits


Probability limits have some nice/important properties:

- $\mathop{\text{plim}}\left( X \times Y \right) = \mathop{\text{plim}}\left( X \right) \times \mathop{\text{plim}}\left( Y \right)$

- $\mathop{\text{plim}}\left( X + Y \right) = \mathop{\text{plim}}\left( X \right) + \mathop{\text{plim}}\left( Y \right)$

- $\mathop{\text{plim}}\left( c \right) = c$, where $c$ is a constant

- $\mathop{\text{plim}}\left( \dfrac{X}{Y} \right) = \dfrac{\mathop{\text{plim}}\left( X \right)}{ \mathop{\text{plim}}\left( Y \right)}$

- $\mathop{\text{plim}}\!\big( f(X) \big) = \mathop{f}\!\big(\mathop{\text{plim}}\left( X \right)\big)$

---

## Consistency
### Consistent estimators


We say that [an estimator is consistent]{.b .pink } if

1. The estimator [has a prob. limit]{.b .pink } (its distribution collapses to a spike).

2. This spike is [located at the parameter]{.b .pink } the estimator predicts.

::: {.fragment}


*In other words...*

An estimator is consistent if its asymptotic distribution collapses to a spike located at the estimated parameter.

:::

::: {.fragment}


*In math:* The estimator $B$ is consistent for $\alpha$ if $\mathop{\text{plim}} B = \alpha$.

:::

::: {.fragment}


The estimator is *inconsistent* if $\mathop{\text{plim}} B \neq \alpha$.

:::

---

## Consistency
### Consistent estimators


*Example:* We want to estimate the population mean $\mu_x$ (where $X$∼Normal).

Let's compare the asymptotic distributions of two competing estimators:

1. The first observation: $X_{1}$
2. The sample mean: $\overline{X} = \dfrac{1}{n} \sum_{i=1}^n x_i$
3. Some other estimator: $\widetilde{X} = \dfrac{1}{n+1} \sum_{i=1}^n x_i$

Note that (1) and (2) are unbiased, but (3) is biased.

## Consistency
### Consistent estimators

To see which are unbiased/biased:

[$\mathop{\boldsymbol{E}}\left[ X_1 \right] =$]{.fragment} [$\mu_x$]{.fragment}

[$\mathop{\boldsymbol{E}}\left[ \overline{X} \right] =$]{.fragment}
[$\mathop{\boldsymbol{E}}\left[ \dfrac{1}{n} \sum_{i=1}^n x_i \right] =$]{.fragment}
[$\dfrac{1}{n} \sum_{i=1}^n \mathop{\boldsymbol{E}}\left[ x_i \right] =$]{.fragment}
[$\dfrac{1}{n} \sum_{i=1}^n \mu_x =$]{.fragment}
[$\dfrac{n}{n} \mu_x =$]{.fragment}
[$\mu_x$]{.fragment}

[$\mathop{\boldsymbol{E}}\left[ \widetilde{X} \right] =$]{.fragment}
[$\mathop{\boldsymbol{E}}\left[ \dfrac{1}{n+1} \sum_{i=1}^n x_i \right] =$]{.fragment}
[$\dfrac{1}{n+1} \sum_{i=1}^n \mathop{\boldsymbol{E}}\left[ x_i \right] =$]{.fragment}
[$\dfrac{n}{n+1}\mu_x$]{.fragment}
[$\neq \mu_x$]{.fragment .grey-light}

## Consistency

Distributions of $\textcolor{#FFA500}{X_1}$, $\textcolor{#e64173}{\overline{X}}$, and $\textcolor{#314f4f}{\widetilde{X}}$

$n = `r (n <- 2)`$

```{r}
#| label: "ex2"
#| echo: false
#| fig-height: 5.75
mu <- 10; v <- 1
ex1 <- mu
ex2 <- mu
ex3 <- n/(n+1) * mu
se1 <- sqrt(v)
se2 <- sqrt(v/n)
se3 <- sqrt(v/(n+1))
lb <- min(c(ex1, ex2, ex3)) -  max(c(se1, se2, se3)) * 3.5
ub <- max(c(ex1, ex2, ex3)) +  max(c(se1, se2, se3)) * 3.5
ggplot(data = tibble(x = c(lb, ub)), aes(x)) +
  stat_function(
    fun = dnorm, args = list(mean = ex1, sd = se1), n = 5e3,
    geom = "area", color = NA, fill = "orange", alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex2, sd = se2), n = 5e3,
    geom = "area", color = NA, fill = red_pink, alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex3, sd = se3), n = 5e3,
    geom = "area", color = NA, fill = "darkslategrey", alpha = 0.7
  ) +
  geom_vline(xintercept = mu, linetype = 5, size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0) +
  theme_empty
```

## Consistency

Distributions of $\textcolor{#FFA500}{X_1}$, $\textcolor{#e64173}{\overline{X}}$, and $\textcolor{#314f4f}{\widetilde{X}}$

$n = `r (n <- 5)`$

```{r}
#| label: "ex5"
#| echo: false
#| fig-height: 5.75
mu <- 10; v <- 1
ex1 <- mu
ex2 <- mu
ex3 <- n/(n+1) * mu
se1 <- sqrt(v)
se2 <- sqrt(v/n)
se3 <- sqrt(v/(n+1))
ggplot(data = tibble(x = c(lb, ub)), aes(x)) +
  stat_function(
    fun = dnorm, args = list(mean = ex1, sd = se1), n = 5e3,
    geom = "area", color = NA, fill = "orange", alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex2, sd = se2), n = 5e3,
    geom = "area", color = NA, fill = red_pink, alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex3, sd = se3), n = 5e3,
    geom = "area", color = NA, fill = "darkslategrey", alpha = 0.7
  ) +
  geom_vline(xintercept = mu, linetype = 5, size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0) +
  theme_empty
```

## Consistency

Distributions of $\textcolor{#FFA500}{X_1}$, $\textcolor{#e64173}{\overline{X}}$, and $\textcolor{#314f4f}{\widetilde{X}}$

$n = `r (n <- 10)`$

```{r}
#| label: "ex10"
#| echo: false
#| fig-height: 5.75
mu <- 10; v <- 1
ex1 <- mu
ex2 <- mu
ex3 <- n/(n+1) * mu
se1 <- sqrt(v)
se2 <- sqrt(v/n)
se3 <- sqrt(v/(n+1))
ggplot(data = tibble(x = seq(lb, ub, 0.001)), aes(x)) +
  stat_function(
    fun = dnorm, args = list(mean = ex1, sd = se1), n = 5e3,
    geom = "area", color = NA, fill = "orange", alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex2, sd = se2), n = 5e3,
    geom = "area", color = NA, fill = red_pink, alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex3, sd = se3), n = 5e3,
    geom = "area", color = NA, fill = "darkslategrey", alpha = 0.7
  ) +
  geom_vline(xintercept = mu, linetype = 5, size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0) +
  theme_empty
```

## Consistency

Distributions of $\textcolor{#FFA500}{X_1}$, $\textcolor{#e64173}{\overline{X}}$, and $\textcolor{#314f4f}{\widetilde{X}}$

$n = `r (n <- 30)`$

```{r}
#| label: "ex30"
#| echo: false
#| fig-height: 5.75
mu <- 10; v <- 1
ex1 <- mu
ex2 <- mu
ex3 <- n/(n+1) * mu
se1 <- sqrt(v)
se2 <- sqrt(v/n)
se3 <- sqrt(v/(n+1))
ggplot(data = tibble(x = c(lb, ub)), aes(x)) +
  stat_function(
    fun = dnorm, args = list(mean = ex1, sd = se1), n = 5e3,
    geom = "area", color = NA, fill = "orange", alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex2, sd = se2), n = 5e3,
    geom = "area", color = NA, fill = red_pink, alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex3, sd = se3), n = 5e3,
    geom = "area", color = NA, fill = "darkslategrey", alpha = 0.7
  ) +
  geom_vline(xintercept = mu, linetype = 5, size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0) +
  theme_empty
```

## Consistency

Distributions of $\textcolor{#FFA500}{X_1}$, $\textcolor{#e64173}{\overline{X}}$, and $\textcolor{#314f4f}{\widetilde{X}}$

$n = `r (n <- 50)`$

```{r}
#| label: "ex50"
#| echo: false
#| fig-height: 5.75
mu <- 10; v <- 1
ex1 <- mu
ex2 <- mu
ex3 <- n/(n+1) * mu
se1 <- sqrt(v)
se2 <- sqrt(v/n)
se3 <- sqrt(v/(n+1))
ggplot(data = tibble(x = c(lb, ub)), aes(x)) +
  stat_function(
    fun = dnorm, args = list(mean = ex1, sd = se1), n = 5e3,
    geom = "area", color = NA, fill = "orange", alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex2, sd = se2), n = 5e3,
    geom = "area", color = NA, fill = red_pink, alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex3, sd = se3), n = 5e3,
    geom = "area", color = NA, fill = "darkslategrey", alpha = 0.7
  ) +
  geom_vline(xintercept = mu, linetype = 5, size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0) +
  theme_empty
```

## Consistency

Distributions of $\textcolor{#FFA500}{X_1}$, $\textcolor{#e64173}{\overline{X}}$, and $\textcolor{#314f4f}{\widetilde{X}}$

$n = `r (n <- 100)`$

```{r}
#| label: "ex100"
#| echo: false
#| fig-height: 5.75
mu <- 10; v <- 1
ex1 <- mu
ex2 <- mu
ex3 <- n/(n+1) * mu
se1 <- sqrt(v)
se2 <- sqrt(v/n)
se3 <- sqrt(v/(n+1))
ggplot(data = tibble(x = c(lb, ub)), aes(x)) +
  stat_function(
    fun = dnorm, args = list(mean = ex1, sd = se1), n = 5e3,
    geom = "area", color = NA, fill = "orange", alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex2, sd = se2), n = 5e3,
    geom = "area", color = NA, fill = red_pink, alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex3, sd = se3), n = 5e3,
    geom = "area", color = NA, fill = "darkslategrey", alpha = 0.7
  ) +
  geom_vline(xintercept = mu, linetype = 5, size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0) +
  theme_empty
```

## Consistency

Distributions of $\textcolor{#FFA500}{X_1}$, $\textcolor{#e64173}{\overline{X}}$, and $\textcolor{#314f4f}{\widetilde{X}}$

$n = `r (n <- 500)`$

```{r}
#| label: "ex500"
#| echo: false
#| fig-height: 5.75
mu <- 10; v <- 1
ex1 <- mu
ex2 <- mu
ex3 <- n/(n+1) * mu
se1 <- sqrt(v)
se2 <- sqrt(v/n)
se3 <- sqrt(v/(n+1))
ggplot(data = tibble(x = c(lb, ub)), aes(x)) +
  stat_function(
    fun = dnorm, args = list(mean = ex1, sd = se1), n = 5e3,
    geom = "area", color = NA, fill = "orange", alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex2, sd = se2), n = 5e3,
    geom = "area", color = NA, fill = red_pink, alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex3, sd = se3), n = 5e3,
    geom = "area", color = NA, fill = "darkslategrey", alpha = 0.7
  ) +
  geom_vline(xintercept = mu, linetype = 5, size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0) +
  theme_empty
```

## Consistency

Distributions of $\textcolor{#FFA500}{X_1}$, $\textcolor{#e64173}{\overline{X}}$, and $\textcolor{#314f4f}{\widetilde{X}}$

$n = `r (n <- 1000)`$

```{r}
#| label: "ex1000"
#| echo: false
#| fig-height: 5.75
mu <- 10; v <- 1
ex1 <- mu
ex2 <- mu
ex3 <- n/(n+1) * mu
se1 <- sqrt(v)
se2 <- sqrt(v/n)
se3 <- sqrt(v/(n+1))
ggplot(data = tibble(x = c(lb, ub)), aes(x)) +
  stat_function(
    fun = dnorm, args = list(mean = ex1, sd = se1), n = 5e3,
    geom = "area", color = NA, fill = "orange", alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex2, sd = se2), n = 5e3,
    geom = "area", color = NA, fill = red_pink, alpha = 0.7
  ) +
  stat_function(
    fun = dnorm, args = list(mean = ex3, sd = se3), n = 5e3,
    geom = "area", color = NA, fill = "darkslategrey", alpha = 0.7
  ) +
  geom_vline(xintercept = mu, linetype = 5, size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0) +
  theme_empty
```

## Consistency


The distributions of $\textcolor{#314f4f}{\widetilde{X}}$

For $n$ in $\{\textcolor{#FCCE25}{2},\, \textcolor{#F89441}{5},\, \textcolor{#E16462}{10},\, \textcolor{#BF3984}{50},\, \textcolor{#900DA4}{100},\, \textcolor{#5601A4}{500}, \textcolor{#0D0887}{1000}\}$
```{r}
#| label: "ex-biased-consistent"
#| echo: false
#| fig-height: 5.75
mu <- 10; v <- 1
ggplot(data = tibble(x = c(6.67-2, 10)), aes(x)) +
  stat_function(
    fun = dnorm, args = list(mean = mu * 2/(2+1), sd = sqrt(v/(2+1))), n = 5e3,
    geom = "area", fill = plasma(7, end = 0.9)[7], color = NA, alpha = 0.7, size = 0.3
  ) +
  stat_function(
    fun = dnorm, args = list(mean = mu * 5/(5+1), sd = sqrt(v/(5+1))), n = 5e3,
    geom = "area", fill = plasma(7, end = 0.9)[6], color = NA, alpha = 0.7, size = 0.3
  ) +
  stat_function(
    fun = dnorm, args = list(mean = mu * 10/(10+1), sd = sqrt(v/(10+1))), n = 5e3,
    geom = "area", fill = plasma(7, end = 0.9)[5], color = NA, alpha = 0.7, size = 0.3
  ) +
  stat_function(
    fun = dnorm, args = list(mean = mu * 50/(50+1), sd = sqrt(v/(50+1))), n = 5e3,
    geom = "area", fill = plasma(7, end = 0.9)[4], color = NA, alpha = 0.7, size = 0.3
  ) +
  stat_function(
    fun = dnorm, args = list(mean = mu * 100/(100+1), sd = sqrt(v/(100+1))), n = 5e3,
    geom = "area", fill = plasma(7, end = 0.9)[3], color = NA, alpha = 0.7, size = 0.3
  ) +
  stat_function(
    fun = dnorm, args = list(mean = mu * 500/(500+1), sd = sqrt(v/(500+1))), n = 5e3,
    geom = "area", fill = plasma(7, end = 0.9)[2], color = NA, alpha = 0.7, size = 0.3
  ) +
  stat_function(
    fun = dnorm, args = list(mean = mu * 1000/(1000+1), sd = sqrt(v/(1000+1))), n = 5e3,
    geom = "area", fill = plasma(7, end = 0.9)[1], color = NA, alpha = 0.7, size = 0.3
  ) +
  geom_vline(xintercept = mu, linetype = 5, size = 0.6, alpha = 0.5) +
  geom_hline(yintercept = 0) +
  theme_empty
```

---

## Consistency


### The takeaway?

::: {.fragment}

- An estimator can be unbiased without being consistent (e.g., $\textcolor{#FFA500}{X_1}$).

:::

::: {.fragment}

- An estimator can be unbiased and consistent (e.g., $\textcolor{#e64173}{\overline{X}}$).

:::

::: {.fragment}

- An estimator can be biased but consistent (e.g., $\textcolor{#314f4f}{\widetilde{X}}$).

:::

::: {.fragment}

- An estimator can be biased and inconsistent (e.g., $\overline{X} - 50$).

:::

::: {.fragment}

**Best-case scenario:** The estimator is unbiased and consistent.

:::

---

## Consistency


### Why consistency (asymptotics)?

1. We cannot always find an unbiased estimator. In these situations, we generally (at least) want consistency.

2. Expected values can be hard/undefined. Probability limits are less constrained, _e.g._,
$$ \mathop{\boldsymbol{E}}\left[ g(X)h(Y) \right] \text{ vs. } \mathop{\text{plim}}\left( g(X)h(Y) \right) $$

3. Asymptotics help us move away from assuming the distribution of $u_i$.

::: {.fragment}


<br>

[Caution:]{.b .pink } As we saw, consistent estimators can be biased in small samples.

:::

---

## {.center}

[OLS in asymptopia]{.bigger .it .slab}

---

## OLS in asymptopia


OLS has two very nice asymptotic properties:

1. Consistency

2. Asymptotic Normality

::: {.fragment}


Let's prove \#1 for OLS with simple, linear regression, _i.e._,

$$ y_i = \beta_0 + \beta_1 x_i + u_i $$

:::

---

## OLS in asymptopia
### Proof of consistency


First, recall our previous derivation of of $\hat{\beta}_1$,

$$ \hat{\beta}_1 = \beta_1 + \dfrac{\sum_i \left( x_i - \overline{x} \right) u_i}{\sum_i \left( x_i - \overline{x} \right)^2} $$

::: {.fragment}


Now divide the numerator and denominator by $1/n$

:::

::: {.fragment}


$$ \hat{\beta}_1 = \beta_1 + \dfrac{\frac{1}{n} \sum_i \left( x_i - \overline{x} \right) u_i}{\frac{1}{n}\sum_i \left( x_i - \overline{x} \right)^2} $$

:::

---

## OLS in asymptopia
### Proof of consistency


We actually want to know the probability limit of $\hat{\beta}_1$, so

::: {.fragment}


$$ \mathop{\text{plim}} \hat{\beta}_1 = \mathop{\text{plim}}\left(\beta_1 + \dfrac{\frac{1}{n} \sum_i \left( x_i - \overline{x} \right) u_i}{\frac{1}{n}\sum_i \left( x_i - \overline{x} \right)^2} \right) $$

:::

::: {.fragment}


which, by the properties of probability limits, gives us

:::

::: {.fragment}


$$ = \beta_1 + \dfrac{\mathop{\text{plim}}\left(\frac{1}{n} \sum_i \left( x_i - \overline{x} \right) u_i \right)}{\mathop{\text{plim}}\left(\frac{1}{n}\sum_i \left( x_i - \overline{x} \right)^2 \right)} $$

:::

::: {.fragment}


The numerator and denominator are, in fact, population quantities

:::

::: {.fragment}


$$ = \beta_1 + \dfrac{\mathop{\text{Cov}} \left( x,\, u \right)}{\mathop{\text{Var}} \left( x \right)} $$

:::

---

## OLS in asymptopia
### Proof of consistency


So we have

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \dfrac{\mathop{\text{Cov}} \left( x,\, u \right)}{\mathop{\text{Var}} \left( x \right)} $$

::: {.fragment}


By our assumption of exogeneity (plus the law of total expectation)

:::

::: {.fragment}


$$ \mathop{\text{Cov}} \left( x,\,u \right) = 0 $$

:::

::: {.fragment}


Combining these two equations yields

:::

::: {.fragment}


$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \dfrac{0}{\mathop{\text{Var}} \left( x \right)} = \beta_1 \quad\text{🤓} $$

so long as $\mathop{\text{Var}} \left( x \right) \neq 0$ (which we've assumed).

:::

---

## OLS in asymptopia
### Asymptotic normality


Up to this point, we made a very specific assumption about the distribution of $u_i$—the $u_i$ came from a normal distribution.

We can relax this assumption—allowing the $u_i$ to come from any distribution (still assume exogeneity, independence, and homoskedasticity).

We will focus on the [asymptotic distribution]{.b .pink } of our estimators (how they are distributed as $n$ gets large), rather than their finite-sample distribution.

::: {.fragment}


As $n$ approaches $\infty$, the distribution of the OLS estimator converges to a normal distribution.

:::

---

## OLS in asymptopia
### Recap

With a more limited set of assumptions, OLS is [consistent]{.b .pink } and is [asymptotically normally distributed]{.b .pink }.

#### Current assumptions

1. Our data were **randomly sampled** from the population.
1. $y_i$ is a **linear function** of its parameters and disturbance.
1. There is **no perfect collinearity** in our data.
1. The $u_i$ have conditional mean of zero (**exogeneity**), $\mathop{\boldsymbol{E}}\left[ u_i \middle| X_i \right] = 0$.
1. The $u_i$ are **homoskedastic** with **zero correlation** between $u_i$ and $u_j$.

---

## {.center}

[Omitted-variable bias, redux]{.bigger .it .slab}

---

## Omitted-variable bias, redux
### Inconsistency?
Imagine we have a population whose true model is

$$
\begin{align}
 y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + u_i \tag{2}
\end{align}
$$

::: {.fragment}


*Recall[1]{.sub }:* [Omitted-variable bias]{.b .pink } occurs when we omit a variable in our linear regression model (_e.g._, leavining out $x_2$) such that

:::

::: {.fragment}


1. $x_{2}$ affects $y$, _i.e._, $\beta_2 \neq 0$.

:::

::: {.fragment}


2. Correlates with an included explanatory variable, _i.e._, $\mathop{\text{Cov}} \left( x_1,\, x_2 \right) \neq 0$.

:::

---

## Omitted-variable bias, redux
### Inconsistency?
Imagine we have a population whose true model is

$$
\begin{align}
 y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + u_i \tag{2}
\end{align}
$$

*Recall[2]{.sub }:* We defined the [bias]{.b .pink } of an estimator $W$ for parameter $\theta$

::: {.fragment}


$$ \mathop{\text{Bias}}_\theta \left( W \right) = \mathop{\boldsymbol{E}}\left[ W \right] - \theta $$

:::

---

## Omitted-variable bias, redux
### Inconsistency?
Imagine we have a population whose true model is

$$
\begin{align}
 y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + u_i \tag{2}
\end{align}
$$

We know that omitted-variable bias causes [biased estimates]{.pink }.

*Question:* Do *omitted variables* also cause [inconsistent estimates]{.pink }?

::: {.fragment}


*Answer:* Find $\mathop{\text{plim}} \hat{\beta}_1$ in a regression that omits $x_2$.

:::

---

## Omitted-variable bias, redux
### Inconsistency?
Imagine we have a population whose true model is

$$
\begin{align}
 y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + u_i \tag{2}
\end{align}
$$

but we instead specify the model as

$$
\begin{align}
 y_i = \beta_0 + \beta_1 x_{1i} + w_i \tag{3}
\end{align}
$$

where $w_i = \beta_2 x_{2i} + u_i$.

::: {.fragment}

 We estimate $(3)$ via OLS

$$
\begin{align}
 y_i = \hat{\beta}_0 + \hat{\beta}_1 x_{1i} + \hat{w}_i \tag{4}
\end{align}
$$

:::


[*Our question:* Is $\hat{\beta}_1$ consistent for $\beta_1$ when we omit $x_2$?]{.fragment}
[$\mathop{\text{plim}}\left( \hat{\beta}_1 \right) \overset{?}{=} \beta_1$]{.fragment}

---

## Omitted-variable bias, redux
### Inconsistency?

:::: {.columns}

::: {.column width="48%"}
[Truth:]{.b .pink } $y_i = \beta_0 + \beta_1 x_{1i} + \beta_2 x_{2i} + u_i$

:::

::: {.column width="48%"}
[Specified:]{.b .purple } $y_i = \beta_0 + \beta_1 x_{1i} + w_i$

:::

::::

We already showed $\mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \dfrac{\mathop{\text{Cov}} \left( x_1,\, w \right)}{\mathop{\text{Var}} \left( x_1 \right)}$

where $w$ is the disturbance.

::: {.fragment}

 Here, we know $w = \beta_2 x_2 + u$. Thus,

:::

::: {.fragment}

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \dfrac{\mathop{\text{Cov}} \left( x_1,\, \beta_2 x_2 + u \right)}{\mathop{\text{Var}} \left( x_1 \right)} $$

:::

::: {.fragment}

Now, we make use of $\mathop{\text{Cov}} \left( X,\, Y + Z \right) = \mathop{\text{Cov}} \left( X,\, Y \right) + \mathop{\text{Cov}} \left( X,\, Z \right)$

:::

::: {.fragment}

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \dfrac{\mathop{\text{Cov}} \left( x_1,\, \beta_2 x_2 \right) + \mathop{\text{Cov}} \left( x_1,\, u \right)}{\mathop{\text{Var}} \left( x_1 \right)} $$

:::

---

## Omitted-variable bias, redux
### Inconsistency?

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \dfrac{\mathop{\text{Cov}} \left( x_1,\, \beta_2 x_2 \right) + \mathop{\text{Cov}} \left( x_1,\, u \right)}{\mathop{\text{Var}} \left( x_1 \right)} $$

Now we use the fact that $\mathop{\text{Cov}} \left( X,\, cY \right) = c\mathop{\text{Cov}} \left( X,\,Y \right)$ for a constant $c$.

::: {.fragment}

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \dfrac{\beta_2 \mathop{\text{Cov}} \left( x_1,\, x_2 \right) + \mathop{\text{Cov}} \left( x_1,\, u \right)}{\mathop{\text{Var}} \left( x_1 \right)} $$

:::

::: {.fragment}

As before, our exogeneity (conditional mean zero) assumption implies $\mathop{\text{Cov}} \left( x_1,\, u \right) = 0$, which gives us

:::

::: {.fragment}

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \dfrac{\beta_2 \mathop{\text{Cov}} \left( x_1,\, x_2 \right)}{\mathop{\text{Var}} \left( x_1 \right)} $$

:::

---

## Omitted-variable bias, redux
### Inconsistency?

Thus, we find that

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \beta_2 \dfrac{ \mathop{\text{Cov}} \left( x_1,\, x_2 \right)}{\mathop{\text{Var}} \left( x_1 \right)} $$

In other words, an [omitted variable will cause OLS to be inconsistent if **both** of the following statements are true]{.pink }:

1. The omitted variable [affects our outcome]{.b .pink }, _i.e._, $\beta_2 \neq 0$.

2. The omitted variable [correlates with included explanatory variables]{.b .purple}, _i.e._, $\mathop{\text{Cov}} \left( x_1,\,x_2 \right) \neq 0$.

If both of these statements are true, then the OLS estimate $\hat{\beta}_1$ will not converge to $\beta_1$, even as $n$ approaches $\infty$.

---

## Omitted-variable bias, redux
### Signing the bias

Sometimes we're stuck with omitted variable bias.^[You will often hear the term "omitted-variable bias" when we're actually talking about inconsistency (rather than bias).]

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \beta_2 \dfrac{ \mathop{\text{Cov}} \left( x_1,\, x_2 \right)}{\mathop{\text{Var}} \left( x_1 \right)} $$

When this happens, we can often at least know the direction of the inconsistency.

---

## Omitted-variable bias, redux
### Signing the bias

Begin with

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \beta_2 \dfrac{ \mathop{\text{Cov}} \left( x_1,\, x_2 \right)}{\mathop{\text{Var}} \left( x_1 \right)} $$

We know $\textcolor{#20B2AA}{\mathop{\text{Var}} \left( x_1 \right) > 0}$. Suppose $\textcolor{#e64173}{\beta_2 > 0}$ and $\textcolor{#FFA500}{\mathop{\text{Cov}} \left( x_1,\,x_2 \right) > 0}$. Then

::: {.fragment}

$$
\begin{align}
 \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \textcolor{#e64173}{(+)} \dfrac{\textcolor{#FFA500}{(+)}}{\textcolor{#20B2AA}{(+)}} \implies \mathop{\text{plim}} \hat{\beta}_1 > \beta_1
\end{align}
$$
∴ In this case, OLS is **biased upward** (estimates are too large).

:::

::: {.fragment}

$$
\begin{matrix}
 \enspace & \textcolor{#FFA500}{\text{Cov}(x_1,\,x_2)> 0} & \textcolor{#FFA500}{\text{Cov}(x_1,\,x_2)< 0} \\
 \textcolor{#e64173}{\beta_2 > 0} & \text{Upward} &  \\
 \textcolor{#e64173}{\beta_2 < 0} &  &
\end{matrix}
$$

:::

---

## Omitted-variable bias, redux
### Signing the bias


Begin with

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \beta_2 \dfrac{ \mathop{\text{Cov}} \left( x_1,\, x_2 \right)}{\mathop{\text{Var}} \left( x_1 \right)} $$

We know $\textcolor{#20B2AA}{\mathop{\text{Var}} \left( x_1 \right) > 0}$. Suppose $\textcolor{#e64173}{\beta_2 < 0}$ and $\textcolor{#FFA500}{\mathop{\text{Cov}} \left( x_1,\,x_2 \right) > 0}$. Then

::: {.fragment}

$$
\begin{align}
 \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \textcolor{#e64173}{(-)} \dfrac{\textcolor{#FFA500}{(+)}}{\textcolor{#20B2AA}{(+)}} \implies \mathop{\text{plim}} \hat{\beta}_1 < \beta_1
\end{align}
$$
∴ In this case, OLS is **biased downward** (estimates are too small).

$$
\begin{matrix}
 \enspace & \textcolor{#FFA500}{\text{Cov}(x_1,\,x_2)> 0} & \textcolor{#FFA500}{\text{Cov}(x_1,\,x_2)< 0} \\
 \textcolor{#e64173}{\beta_2 > 0} & \text{Upward} &  \\
 \textcolor{#e64173}{\beta_2 < 0} & \text{Downward} &
\end{matrix}
$$

:::

---

## Omitted-variable bias, redux
### Signing the bias


Begin with

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \beta_2 \dfrac{ \mathop{\text{Cov}} \left( x_1,\, x_2 \right)}{\mathop{\text{Var}} \left( x_1 \right)} $$

We know $\textcolor{#20B2AA}{\mathop{\text{Var}} \left( x_1 \right) > 0}$. Suppose $\textcolor{#e64173}{\beta_2 > 0}$ and $\textcolor{#FFA500}{\mathop{\text{Cov}} \left( x_1,\,x_2 \right) < 0}$. Then

$$
\begin{align}
 \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \textcolor{#e64173}{(+)} \dfrac{\textcolor{#FFA500}{(-)}}{\textcolor{#20B2AA}{(+)}} \implies \mathop{\text{plim}} \hat{\beta}_1 < \beta_1
\end{align}
$$
∴ In this case, OLS is **biased downward** (estimates are too small).

$$
\begin{matrix}
 \enspace & \textcolor{#FFA500}{\text{Cov}(x_1,\,x_2)> 0} & \textcolor{#FFA500}{\text{Cov}(x_1,\,x_2)< 0} \\
 \textcolor{#e64173}{\beta_2 > 0} & \text{Upward} & \text{Downward} \\
 \textcolor{#e64173}{\beta_2 < 0} & \text{Downward} &
\end{matrix}
$$

---

## Omitted-variable bias, redux
### Signing the bias


Begin with

$$ \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \beta_2 \dfrac{ \mathop{\text{Cov}} \left( x_1,\, x_2 \right)}{\mathop{\text{Var}} \left( x_1 \right)} $$

We know $\textcolor{#20B2AA}{\mathop{\text{Var}} \left( x_1 \right) > 0}$. Suppose $\textcolor{#e64173}{\beta_2 < 0}$ and $\textcolor{#FFA500}{\mathop{\text{Cov}} \left( x_1,\,x_2 \right) < 0}$. Then

$$
\begin{align}
 \mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \textcolor{#e64173}{(-)} \dfrac{\textcolor{#FFA500}{(-)}}{\textcolor{#20B2AA}{(+)}} \implies \mathop{\text{plim}} \hat{\beta}_1 > \beta_1
\end{align}
$$
∴ In this case, OLS is **biased upward** (estimates are too large).

$$
\begin{matrix}
 \enspace & \textcolor{#FFA500}{\text{Cov}(x_1,\,x_2)> 0} & \textcolor{#FFA500}{\text{Cov}(x_1,\,x_2)< 0} \\
 \textcolor{#e64173}{\beta_2 > 0} & \text{Upward} & \text{Downward} \\
 \textcolor{#e64173}{\beta_2 < 0} & \text{Downward} & \text{Upward}
\end{matrix}
$$

---

## Omitted-variable bias, redux
### Signing the bias


Thus, in cases where we have a sense of

1. the sign of $\mathop{\text{Cov}} \left( x_1,\,x_2 \right)$

2. the sign of $\beta_2$

we know in which direction inconsistency pushes our estimates.

::: {.center}
**Direction of bias**

:::

$$
\begin{matrix}
 \enspace & \textcolor{#FFA500}{\text{Cov}(x_1,\,x_2)> 0} & \textcolor{#FFA500}{\text{Cov}(x_1,\,x_2)< 0} \\
 \textcolor{#e64173}{\beta_2 > 0} & \text{Upward} & \text{Downward} \\
 \textcolor{#e64173}{\beta_2 < 0} & \text{Downward} & \text{Upward}
\end{matrix}
$$

---

## Measurement error

[Measurement error]{.b .pink } in our explanatory variables presents another case in which OLS is inconsistent.

Consider the population model: $y_i = \beta_0 + \beta_1 z_i + u_i$

- We want to observe $z_i$ but cannot.

- Instead, we *measure* the variable $x_i$, which is $z_i$ plus some error (noise):
$$ x_i = z_i + \omega_i $$

- Assume $\mathop{\boldsymbol{E}}\left[ \omega_i \right] = 0$, $\mathop{\text{Var}} \left( \omega_i \right) = \sigma^2_\omega$, and $\omega$ is independent of $z$ and $u$.

::: {.fragment}

<br>
OLS regression of $y$ and $x$ will produce inconsistent estimates for $\beta_1$.

:::

---

## Measurement error
### Proof

$y_i = \beta_0 + \beta_1 z_i + u_i$
[<br> $\quad= \beta_0 + \beta_1 \left( x_i - \omega_i \right) + u_i$]{.fragment}
[<br> $\quad= \beta_0 + \beta_1 x_i + \left( u_i - \beta_1 \omega_i \right)$]{.fragment}
[<br> $\quad= \beta_0 + \beta_1 x_i + \varepsilon_i$]{.fragment}

[where $\varepsilon_i = u_i - \beta_1 \omega_i$]{.fragment}

::: {.fragment}

What happens when we estimate $y_i = \hat{\beta}_0 + \hat{\beta}_1 x_i + e_i$?

$\mathop{\text{plim}} \hat{\beta}_1 = \beta_1 + \dfrac{\mathop{\text{Cov}} \left( x,\,\varepsilon \right)}{\mathop{\text{Var}} \left( x \right)}$

We will derive the numerator and denominator separately...

:::

---

## Measurement error
### Proof


The covariance of our noisy variable $x$ and the disturbance $\varepsilon$.

$\mathop{\text{Cov}} \left( x,\, \varepsilon \right)$

[$\quad= \mathop{\text{Cov}} \left( \left[ z + \omega \right],\, \left[ u - \beta_1 \omega \right] \right)$]{.fragment}

[$\quad= \mathop{\text{Cov}} \left( z,\,u \right) -\beta_1 \mathop{\text{Cov}} \left( z,\,\omega \right) + \mathop{\text{Cov}} \left( \omega,\, u \right) - \beta_1 \mathop{\text{Var}} \left( \omega \right)$]{.fragment}

[$\quad= 0 + 0 + 0 - \beta_1 \sigma_\omega^2$]{.fragment}

[$\quad= - \beta_1 \sigma_\omega^2$]{.fragment}

---

## Measurement error
### Proof

Now for the denominator, $\mathop{\text{Var}} \left( x \right)$.

$\mathop{\text{Var}} \left( x \right)$

[$\quad = \mathop{\text{Var}} \left( z + \omega \right)$]{.fragment}

[$\quad= \mathop{\text{Var}} \left( z \right) + \mathop{\text{Var}} \left( \omega \right) + 2\mathop{\text{Cov}} \left( z,\,\omega \right)$]{.fragment}

[$\quad= \sigma_z^2 + \sigma_\omega^2$]{.fragment}

---

## Measurement error
### Proof

Putting the numerator and denominator back together,

$$
\begin{align}
 \mathop{\text{plim}} \hat{\beta}_1
 &= \beta_1 + \dfrac{\mathop{\text{Cov}} \left( x,\,\varepsilon \right)}{\mathop{\text{Var}} \left( x \right)} \\
 &= \beta_1 + \dfrac{-\beta_1 \sigma_\omega^2}{\sigma_z^2 + \sigma_\omega^2} \\
 &= \beta_1 - \beta_1 \dfrac{\sigma_\omega^2}{\sigma_z^2 + \sigma_\omega^2} \\
 &= \beta_1 \dfrac{\sigma_z^2 + \sigma_\omega^2}{\sigma_z^2 + \sigma_\omega^2} - \beta_1 \dfrac{\sigma_\omega^2}{\sigma_z^2 + \sigma_\omega^2} \\
 &= \beta_1 \dfrac{\sigma_z^2}{\sigma_z^2 + \sigma_\omega^2}
\end{align}
$$

---

## Measurement error
### Summary


∴ $\mathop{\text{plim}} \hat{\beta}_1 = \beta_1 \dfrac{\sigma_z^2}{\sigma_z^2 + \sigma_\omega^2}$.

What does this equation tell us?

::: {.fragment}


[Measurement error in our explanatory variables]{.b .pink } biases the coefficient estimates toward zero.

- This type of bias/inconsistency is often called [attenuation bias]{.b .pink }.

- If [the measurement error correlates with the explanatory variables]{.b .pink }, we have bigger problems with inconsistency/bias.


:::

---

## Measurement error
### Summary


What about **measurement in the outcome variable**?

It doesn't really matter—it just increases our standard errors.

## Measurement error
### It's everywhere

**General cases**

1. We cannot perfectly observe a variable.
1. We use one variable as a *proxy* for another.

**Specific examples**

- GDP
- Population
- Crime/police statistics
- Air quality
- Health data
- Ability

## Conclusions

### Wrapping up

_Asymptotic_ properties highlight some interesting features:

- [Consistency]{.b .purple} describes estimators' behavior as $n$ approaches $\infty$.
- [Omitted-variable bias]{.b .pink} causes OLS to be inconsistent.

$\implies$ Even if with [_infinite data_]{.purple}, you can get the [wrong answer]{.pink}.

::: {.fragment}
[Measurement error]{.b .slate} in explanatory variables also causes [inconsistency]{.b .pink}.
:::

::: {.fragment}
Again, even if with [_infinite data_]{.purple}, you can get the [wrong answer]{.pink}.
:::

::: {.fragment}
[Takeaway?]{.note} Assumptions and data quality matter more than size.
:::
