---
title: 'Causality'
subtitle: 'EC 421, Set 10'
author:
  - name: Ed Rubin
    email: edwardr@uoregon.edu
editor:
  render-on-save: true
format:
  clean-revealjs:
    slide-level: 2
    theme: [../my.scss, default]
    highlight-style: gruvbox
    embed-resources: true
    slide-number: c/t
    incremental: false
    controls-layout: edges
    transition: none
    preview-links: auto
    html-math-method: mathjax
    code-line-numbers: false
execute:
  warning: false
  message: false

---

## {.center}

[Prologue]{.bigger .it .slab}

---

## Schedule {#schedule}

```{r}
#| label: "setup"
#| include: false
# devtools::install_github("dill/emoGG")
library(pacman)
p_load(
  broom, here, tidyverse,
  emoGG, latex2exp, ggplot2, ggthemes, viridis, extrafont, gridExtra,
  kableExtra,
  data.table,
  dplyr,
  lubridate,
  magrittr, knitr, parallel
)
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -0.5, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_simple <- theme_bw() + theme(
  line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_text(size = 18, family = "STIXGeneral"),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  # plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_math <- theme_void() + theme(
  text = element_text(family = "MathJax_Math"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_serif <- theme_void() + theme(
  text = element_text(family = "MathJax_Main"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes <- theme_void() + theme(
  text = element_text(family = "Fira Sans Book"),
  axis.title = element_text(size = 18),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = grey_light,
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_set(theme_gray(base_size = 20))
# A few extras

if (!exists("geom_emoji", mode = "function")) {
  geom_emoji <- function(data = NULL, mapping = NULL, emoji = NULL, ..., inherit.aes = TRUE) {
    label <- intToUtf8(strtoi(emoji, 16L))
    ggplot2::geom_text(
      data = data,
      mapping = mapping,
      label = label,
      ...,
      inherit.aes = inherit.aes
    )
  }
}
```

### Last Time

Autocorrelation and nonstationarity

### Today

Causality

---

## Causality {#intro}

### Intro

Most tasks in econometrics boil down to one of two goals:

$$
\begin{align}
  y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_k x_k + u
\end{align}
$$

::: {.fragment}

1. [Prediction:]{.b .purple} Accurately and dependably [predict/forecast]{.purple} $\textcolor{#6A5ACD}{y}$ using on some set of explanatory variables—doesn't need to be $x_1$ through $x_k$. Focuses on $\textcolor{#6A5ACD}{\hat{y}}$. $\beta_j$ doesn't really matter.

:::

::: {.fragment}

1. [Causal estimation:]{.b .pink}[[†]{.pink}]{.super} Estimate the actual data-generating process—learning about the true, population model that explains [how]{.pink} $\textcolor{#e64173}{y}$ [changes when we change]{.pink} $\textcolor{#e64173}{x_j}$—focuses on $\textcolor{#e64173}{\beta_j}$. Accuracy of $\hat{y}$ is not important.

:::

^[[†]{.pink} Often called *causal identification*.]

::: {.fragment}

For the rest of the term, we will focus on [causally estimating]{.b .pink} $\textcolor{#e64173}{\beta_j}$.

:::

---

## Causality {#challenges}

### The challenges

As you saw in the data-analysis exercise, determining and estimating the true model can be pretty difficult—both [practically]{.purple} and [econometrically]{.pink}.

::: columns
::: {.column width="50%"}
::: {.fragment}

::: {.purple}
**Practical challenges**

- Which variables?
- Which functional form(s)?
- Do data exist? How much?
- Is the sample representative?
:::

:::
:::
::: {.column width="50%"}
::: {.fragment}

::: {.pink}
**Econometric challenges**

- Omitted-variable bias
- Reverse causality
- Measurement error
- How precise can/must we be?
:::

:::
:::
:::

::: {.fragment}

Many of these challenges relate to [exogeneity]{.b .slate}, _i.e._, $\textcolor{#314f4f}{\mathop{\boldsymbol{E}}\left[ u_i | X \right] = 0}$.

:::

::: {.fragment}

<br>Causality requires us to [hold all else constant]{.b .slate} (*ceterus paribus*).

:::

---

## Causality

### It's complicated

Occasionally, [*causal*]{.b .pink} relationships are simply/easily understood, _e.g._,

::: {.fragment}

- What [caused]{.pink} the forest fire?
- [How]{.pink} did this baby get here?

:::

::: {.fragment}

Generally, [*causal*]{.b .pink} relationships are complex and challenging to answer, _e.g._,

:::

::: {.fragment}

- What [causes]{.pink} some countries to grow and others to decline?
- What [caused]{.pink} the capital riot?
- Did lax regulation[cause]{.pink} Texas's recent energy problems?
- [How]{.pink} does the number of police officers affect crime?
- What is the [effect]{.pink} of better air quality on test scores?
- Do longer prison sentences [decrease]{.pink} crime?
- How did cannabis legalization [affect]{.pink} mental health/opioid addiction?

:::

---

## Causality

### Correlation ≠ Causation

You've likely heard the saying

> Correlation is not causation.

The saying is just pointing out that there are violations of exogeneity.

::: {.fragment}

Although correlation is not causation, [causation *requires* correlation]{.b .pink}.

:::

::: {.fragment}

[New saying:]{.b .slate}

:::

::: {.fragment}

> Correlation plus exogeneity is causation.

:::

---

## {.clear}

Let's work through a few examples.

---

## Causation {#fertilizer}

### Example: The causal effect of fertilizer[[†]{.pink}]{.super}

^[[†]{.pink} Many of the early statistical and econometric studies involved agricultural field trials.]

Suppose we want to know the causal effect of fertilizer on corn yield.

::: {.fragment}

**Q:** Could we simply regress yield on fertilizer?

:::

::: {.fragment}

<br>**A:** Probably not (if we want the causal effect).

:::

::: {.fragment}

<br><br>**Q:** Why not?

:::

::: {.fragment}

<br>**A:** Omitted-variable bias: Farmers may apply less fertilizer in areas that are already worse on other dimensions that affect yield (soil, slope, water).<br>[Violates *all else equal* (exogeneity). Biased and/or spurious results.]{.pink}

:::

::: {.fragment}

<br><br>**Q:** So what *should* we do?

:::

::: {.fragment}

<br>**A:** [Run an experiment!]{.b .pink}

:::

::: {.fragment}

💩

:::

---

## Causation

### Example: The causal effect of fertilizer

Randomized experiments help us maintain *all else equal* (exogeneity).

::: {.fragment}

We often call these experiments [*randomized control trials*]{.b .pink} (RCTs).[[†]{.pink}]{.super}

:::

^[[†]{.pink} Econometrics (and statistics) borrows this language from biostatistics and pharmaceutical trials.]

::: {.fragment}

Imagine an RCT where we have two groups:

:::

::: {.fragment}

- [Treatment:]{.b .slate} We apply fertilizer.
- [Control:]{.b .slate} We do not apply fertilizer.

:::

::: {.fragment}

By randomizing plots of land into [treatment]{.b .slate} or [control]{.b .slate}, we will, on average, include all kinds of land (soild, slope, water, *etc.*) in both groups.

:::

::: {.fragment}

*All else equal*!

:::

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot1"
#| echo: false
xd <- 9
yd <- 6
set.seed(123)
fert_df <- expand.grid(x = 1:xd, y = 1:yd) %>%
  mutate(
    q = x + y + 1.25 * rnorm(xd*yd),
    trt = sample(x = c(T, F), size = xd*yd, replace = T, prob = c(0.51, 0.49)),
  ) %>%
  arrange(-y, x) %>%
  mutate(id = 1:(xd*yd) %>% str_pad(2, "left", "0")) %>%
  arrange(y, x)
ggplot(data = fert_df, aes(x, y)) +
geom_tile(color = "grey40", fill = "grey93", size = 0.2) +
geom_text(aes(label = id), color = "grey45", size = 5, family = "Roboto Mono") +
theme_void() +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot2"
#| echo: false
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_1"
#| echo: false
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 1), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_2"
#| echo: false
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 2), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_3"
#| echo: false
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 3), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_4"
#| echo: false
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 4), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_5"
#| echo: false
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 5), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_6"
#| echo: false
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 6), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_7"
#| echo: false
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 7), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_8"
#| echo: false
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 8), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_9"
#| echo: false
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, 9), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_10"
#| echo: false
set.seed(246)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, round(xd * yd / 2, 0)), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_11"
#| echo: false
set.seed(248)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, round(xd * yd / 2, 0)), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## {.clear}

[54 equal-sized plots]{.b .slate}

```{r}
#| label: "fertilizer_plot3_12"
#| echo: false
set.seed(250)
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = sample_n(fert_df, round(xd * yd / 2, 0)), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## Causation

### Example: The causal effect of fertilizer

We can estimate the [causal effect]{.b .pink} of fertilizer on crop yield by comparing the average yield in the treatment group (💩) with the control group (no 💩).

$$
\begin{align}
  \overline{\text{Yield}}_\text{Treatment} - \overline{\text{Yield}}_\text{Control}
\end{align}
$$

::: {.fragment}

Alternatively, we can use the regression

:::

::: {.fragment}

$$
\begin{align}
  \text{Yield}_i = \beta_0 + \beta_1 \text{Trt}_i + u_i \tag{1}
\end{align}
$$

:::

::: {.fragment}

where $\text{Trt}_i$ is a binary variable (=1 if plot $i$ received the fertilizer treatment).

:::

::: {.fragment}

**Q:** Should we expect $(1)$ to satisfy exogeneity? Why?

:::

::: {.fragment}

<br>**A:** On average, [randomly assigning treatment should balance]{.b .pink} trt. and control across the other dimensions that affect yield (soil, slope, water).

:::

---

## Causality {#returns}
### Example: Returns to education

Labor economists, policy makers, parents, and students are all interested in the (monetary) *return to education.*

::: {.fragment}

[Thought experiment:]{.b .slate}

:::

::: {.fragment}

- Randomly select an individual.
- Give her an additional year of education.
- How much do her earnings increase?

:::

::: {.fragment}

This change in earnings gives the [causal effect]{.b .slate} of education on earnings.

:::

---

## Causality
### Example: Returns to education

**Q:** Could we simply regress earnings on education?

::: {.fragment}

<br>**A:** Again, probably not if we want the true, causal effect.

:::

::: {.fragment}

1. People *choose* education based upon many factors, *e.g.*, ability.
1. Education likely reduces experience (time out of the workforce).
1. Education is [*endogenous*]{.b .pink} (violates *exogeneity*).

:::

::: {.fragment}

The point (2) above also illustrates the difficulty in learning about educations while *holding all else constant*.

:::

::: {.fragment}

Many important variables have the same challenge—gender, race, income.

:::

---

## Causality
### Example: Returns to education

**Q:** So how can we estimate the returns to education?

::: {.fragment}

[Option 1:]{.b .slate}

:::

::: {.fragment}

- Randomly [assign education]{.pink} (might be difficult).
- Randomly [encourage education]{.pink} (might work).
- Randomly [assign programs]{.pink} that affect education (*e.g.*, mentoring).

:::

::: {.fragment}

[Option 2:]{.b .slate}

:::

::: {.fragment}

- Admissions [cutoffs]{.purple}
- [Lottery]{.purple} enrollment and/or capacity [constraints]{.purple}

:::

---

## Causality {#real}

### Real-world experiments

Both examples consider [real experiments]{.b .slate} that isolate causal effects.

[Characteristics]{.b .slate}

- [Feasible]{.purple}—we can actually (potentially) run the experiment.
- [Compare individuals]{.purple} randomized into treatment against individuals randomized into control.
- [Require "good" randomization]{.purple} to get *all else equal* (exogeneity).

::: {.fragment}

*Note:* Your experiment's results are only as good as your randomization.

:::

---

## {.clear}

[Unfortunate randomization]{.b .slate}

```{r}
#| label: "fertilizer_plot3_bad"
#| echo: false
ggplot(data = fert_df,
  aes(x, y)
) +
geom_tile(aes(fill = q), color = "white", size = 0.2) +
# geom_text(aes(label = trt)) +
geom_emoji(data = filter(fert_df, x < xd/3 | y <= yd/3 | (x == xd & y == yd)), emoji = "1f4a9") +
scale_fill_viridis(option = "magma", direction = -1) +
theme_void() +
theme(legend.position = "none") +
coord_equal()
```

---

## Causality {#ideal}
### The ideal experiment

The [ideal experiment]{.b .pink} would be subtly different.

Rather than comparing units randomized as [treatment]{.pink} vs. [control]{.pink}, the ideal experiment would compare treatment and control [for the same, exact unit]{.b .pink}.

::: {.fragment}

$$
\begin{align}
  y_{\text{Treatment},i} - y_{\text{Control},i}
\end{align}
$$

:::

::: {.fragment}

which we will write (for simplicity) as

:::

::: {.fragment}

$$
\begin{align}
  y_{1,i} - y_{0,i}
\end{align}
$$

:::

::: {.fragment}

This [*ideal experiment*]{.pink} is clearly infeasible[[†]{.pink}]{.super}, but it creates nice notation for causality (the Rubin causal model/Neyman potential outcomes framework).

:::

^[[†]{.pink} Without (1) God-like abilities and multiple universes or (2) a time machine.]

---

## Causality
### The ideal experiment

::: columns
::: {.column width="50%"}
The *ideal* data for 10 people

```{r}
#| label: "ideal_data"
#| echo: false
set.seed(3)
ideal_df <- data.frame(
  i = 1:10,
  trt = rep(c(1, 0), each = 5),
  y1i = c(runif(10, 4, 10) %>% round(2)),
  y0i = c(runif(10, 0, 5) %>% round(2))
)
ideal_df
```
:::
::: {.column width="50%"}
::: {.fragment}

Calculate the causal effect of trt.
$$
\begin{align}
  \tau_i = y_{1,i} -  y_{0,i}
\end{align}
$$
for each individual $i$.

:::
:::
:::

---

## Causality
### The ideal experiment

::: columns
::: {.column width="50%"}
The *ideal* data for 10 people

```{r}
#| label: "ideal_data_trt"
#| echo: false
ideal_df %>% mutate(effect_i = y1i - y0i)
```
:::
::: {.column width="50%"}
Calculate the causal effect of trt.
$$
\begin{align}
  \tau_i = y_{1,i} -  y_{0,i}
\end{align}
$$
for each individual $i$.
:::
:::

---

## Causality
### The ideal experiment

::: columns
::: {.column width="50%"}
The *ideal* data for 10 people

```{r}
#| label: "ideal_data_trt2"
#| echo: false
ideal_df %>% mutate(effect_i = y1i - y0i)
```
:::
::: {.column width="50%"}
Calculate the causal effect of trt.
$$
\begin{align}
  \tau_i = y_{1,i} -  y_{0,i}
\end{align}
$$
for each individual $i$.

The mean of $\tau_i$ is the<br>[average treatment effect]{.b .pink} ([ATE]{.pink}).

Thus, $\textcolor{#e64173}{\overline{\tau} = `r transmute(ideal_df, effect_i = y1i - y0i) %>% unlist %>% mean %>% round(2)`}$
:::
:::

---

## Causality
### The ideal experiment

This model highlights the fundamental problem of causal inference.
$$
\begin{align}
  \tau_i = \textcolor{#e64173}{y_{1,i}} &- \textcolor{#6A5ACD}{y_{0,i}}
\end{align}
$$

::: {.fragment}

[The challenge:]{.b .slate}

:::

::: {.fragment}

If we observe $\textcolor{#e64173}{y_{1,i}}$, then we cannot observe $\textcolor{#6A5ACD}{y_{0,i}}$.
<br>If we observe $\textcolor{#6A5ACD}{y_{0,i}}$, then we cannot observe $\textcolor{#e64173}{y_{1,i}}$.

:::

---

## Causality
### The ideal experiment

So a dataset that we actually observe for 6 people will look something like

::: columns
::: {.column width="50%"}
```{r}
#| label: "ideal_data_obs"
#| echo: false
obs_df <- ideal_df
obs_df$y0i[1:5] <- NA
obs_df$y1i[6:10] <- NA
obs_df
```
:::
::: {.column width="50%"}
::: {.fragment}

We can't observe $\textcolor{#e64173}{y_{1,i}}$ and $\textcolor{#6A5ACD}{y_{0,i}}$.

But, we do observe

<ul>
<li>$\textcolor{#e64173}{y_{1,i}}$ for $i$ in 1, 2, 3, 4, 5</li>
<li>$\textcolor{#6A5ACD}{y_{0,j}}$ for $j$ in 6, 7, 8, 9, 10</li>
</ul>

:::
:::
:::

::: {.fragment}

**Q:** How do we "fill in" the `NA`s and estimate $\overline{\tau}$?

:::

---

## Causality {#estimation}
### Causally estimating the treatment effect

[Notation:]{.b .slate}

- $\textcolor{#e64173}{D_i=1}$ [if individual]{.pink} $\textcolor{#e64173}{i}$ [is treated]{.pink}.
- $\textcolor{#6A5ACD}{D_i=0}$ [if individual]{.purple} $\textcolor{#6A5ACD}{i}$ [is not treated (*control* group).]{.purple}

::: {.fragment}

Then, rephrasing the previous slide,

:::

::: {.fragment}

- We only observe $\textcolor{#e64173}{y_{1,i}}$ when $\textcolor{#e64173}{D_{i}=1}$.
- We only observe $\textcolor{#6A5ACD}{y_{0,i}}$ when $\textcolor{#6A5ACD}{D_{i}=0}$.

:::

::: {.fragment}

**Q:** How can we estimate $\overline{\tau}$ using only $\left(\textcolor{#e64173}{y_{1,i}|D_i=1}\right)$ and $\left(\textcolor{#6A5ACD}{y_{0,i}|D_i=0}\right)$?

:::

---

## Causality
### Causally estimating the treatment effect

**Q:** How can we estimate $\overline{\tau}$ using only $\left(\textcolor{#e64173}{y_{1,i}|D_i=1}\right)$ and $\left(\textcolor{#6A5ACD}{y_{0,i}|D_i=0}\right)$?

::: {.fragment}

**Idea:** What if we compare the groups' means? _I.e._,
$$
\begin{align}
  \textcolor{#e64173}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}
\end{align}
$$

:::

::: {.fragment}

**Q:** When does this simple difference in groups' means provide information on the [causal effect]{.b .slate} of the treatment?

:::

::: {.fragment}

**Q[2.0]{.sub}:** Is $\textcolor{#e64173}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}$ a *good* estimator for $\overline{\tau}$?

:::

::: {.fragment}

Time for math! [🎉]{.bigger}

:::

---

## Causality
### Causally estimating the treatment effect

[Assumption:]{.b .slate}

This assumption says that the treatment effect is equal (constant) across all individuals $i$.

::: {.fragment}

[Note:]{.b .slate}

:::

::: {.fragment}

$$
\begin{align}
  \tau_i = \tau = \textcolor{#e64173}{y_{1,i}} - \textcolor{#6A5ACD}{y_{0,i}}
\end{align}
$$

:::

::: {.fragment}

which implies

:::

::: {.fragment}

$$
\begin{align}
   \textcolor{#e64173}{y_{1,i}} = \textcolor{#6A5ACD}{y_{0,i}} + \tau
\end{align}
$$

:::

---

## {.clear}

**Q[3.0]{.sub}:** Is $\textcolor{#e64173}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}$ a *good* estimator for $\tau$?

::: {.fragment}

Difference in groups' means

:::

::: {.fragment}

<br> $\quad \textcolor{#ffffff}{\Bigg|}=\textcolor{#e64173}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}$

:::

::: {.fragment}

<br> $\quad \textcolor{#ffffff}{\Bigg|}=\textcolor{#e64173}{\mathop{Avg}\left( y_{1,i}\mid D_i = 1 \right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}$

:::

::: {.fragment}

<br> $\quad \textcolor{#ffffff}{\Bigg|}=\textcolor{#e64173}{\mathop{Avg}\left( \textcolor{#000000}{\tau \: +} \: \textcolor{#6A5ACD}{y_{0,i}} \mid D_i = 1 \right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}$

:::

::: {.fragment}

<br> $\quad \textcolor{#ffffff}{\Bigg|}=\tau + \textcolor{#e64173}{\mathop{Avg}\left(\textcolor{#6A5ACD}{y_{0,i}} \mid D_i = 1 \right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}$

:::

::: {.fragment}

<br> $\quad \textcolor{#ffffff}{\Bigg|}= \text{Average causal effect} + \textcolor{#FFA500}{\text{Selection bias}}$

:::

::: {.fragment}

So our proposed group-difference estimator give us the sum of

:::

::: {.fragment}

1. $\tau$, the [causal, average treatment effect]{.b .slate} that we want
2. [Selection bias:]{.b .orange} How much trt. and control groups differ (on average).

:::

---

## {.clear}

[Next time:]{.b .slate}

---

## Table of contents

::: columns
::: {.column width="50%"}
#### Admin

::: {.smallest}
1. [Schedule](#schedule)
<!-- 1. [[R]{.mono} showcase](#r_showcase)
  - [Strategizing](#r_showcase)
  - [`gather`-ing](#gather)
  - [Results](#results) -->
:::
:::
::: {.column width="50%"}
#### Causality

::: {.smallest}
1. [Introduction](#intro)
1. [The challenges](#challenges)
1. Examples
  - [Fertilizer](#fertilizer)
  - [Returns to education](#returns)
1. [*Real* experiments](#real)
1. [The ideal experiment](#ideal)
1. [Estimation](#estimation)
1. [Derivation](#derivation)
:::
:::
:::

---

## Hidden {visibility="hidden"}

```{r}
#| label: "print-pdfs"
#| echo: false
#| eval: false
pagedown::chrome_print(
  input = "10-causality.html",
  timeout = 60
)
```
