---
title: 'Introduction and overview'
subtitle: 'EC421, Set 01'
author:
  - name: Ed Rubin
    email: edwardr@uoregon.edu
editor:
  render-on-save: true
format:
  clean-revealjs:
    slide-level: 2
    theme: [../my.scss, default]
    highlight-style: gruvbox 
    embed-resources: true
    slide-number: c/t
    incremental: false
    controls-layout: edges
    transition: none
    preview-links: auto
    html-math-method: mathjax
    code-line-numbers: false
    auto-animate-unmatched: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

options(htmltools.dir.version = FALSE)
library(ggplot2)
library(ggthemes)
red_pink = "#e64173"
```

## {.center}

[Prologue]{.bigger .it .slab}

---

## Why?

### Motivation

Let's start with a few __basic, general questions:__

::: {.fragment}
1. What is the goal of econometrics?

2. Why do economists (or other people) study or use econometrics?
:::

::: {.fragment}
__One simple answer:__ Learn about the world using data.
:::

::: {.fragment}
- _Learn about the world_ = Raise, answer, and challenge questions, theories, assumptions.

- _data_ = Plural of datum.
:::

---

## Why?

### Example

One might (reasonably) guess a company's [sales]{.purple} are a function of its [advertising spending, price, and intesity of competitors]{.pink}.

::: {.fragment}
So, one might hypothesize a model $\textcolor{#6A5ACD}{\text{Sales}} = f(\textcolor{#e64173}{\text{Ad}, \text{Price}, \text{Comp}})$

where

- $\textcolor{#e64173}{\text{Ad}}$ represents dollars spent on advertising,
- $\textcolor{#e64173}{\text{Price}}$ is the product's price,
- $\textcolor{#e64173}{\text{Comp}}$ gives the product's competition.
:::

::: {.fragment}
We _expect_ that [sales]{.purple} $\uparrow$ with [advertising]{.pink} and $\downarrow$ with [price]{.pink} and [competition]{.pink}.
:::

---

##

But who needs to [expect]{.grey .it}?

We can [test]{.orange .it} these hypotheses __using regression__.

::: {.fragment}

[More importantly:]{.grey-vlight .it} Regression estimates the [size]{.orange .it} of these effects

- *How much* does an additional dollar of [advertising]{.pink} increase [sales]{.purple}?
- *How much* does a one-dollar increase in [price]{.pink} decrease [sales]{.purple}?
- *How much* does an additional [competitor]{.pink} reduce [sales]{.purple}?

These (causal) questions are central to efficient decision-making<br>and are the bread and butter of econometrics.
:::

---

__Regression model:__

$$ \textcolor{#6A5ACD}{\text{Sales}}_i = \beta_0 + \beta_1 \textcolor{#e64173}{\text{Ad}}_i + \beta_2 \textcolor{#e64173}{\text{Price}}_i + \beta_3 \textcolor{#e64173}{\text{Comp}}_i + \varepsilon_i $$

---

__Regression model:__

$$ \textcolor{#6A5ACD}{\text{Sales}}_i = \beta_0 + \beta_1 \textcolor{#e64173}{\text{Ad}}_i + \beta_2 \textcolor{#e64173}{\text{Price}}_i + \beta_3 \textcolor{#e64173}{\text{Comp}}_i + \varepsilon_i $$


With this basic regression model, we can test/estimate/quantify the (linear) relationship between sales and advertising, price, and competition.

---

__Regression model:__

$$ \textcolor{#6A5ACD}{\text{Sales}}_i = \beta_0 + \beta_1 \textcolor{#e64173}{\text{Ad}}_i + \beta_2 \textcolor{#e64173}{\text{Price}}_i + \beta_3 \textcolor{#e64173}{\text{Comp}}_i + \varepsilon_i $$

#### (Review) Questions

::: {.fragment}
- __Q:__ How do we interpret $\beta_1$?
:::

::: {.fragment}
- __A:__ An additional dollar of [advertising]{.pink} corresponds with a $\beta_1$-unit change in [sales]{.purple} (holding [price]{.pink} and [competition]{.pink} fixed).
:::

---

__Regression model:__

$$ \textcolor{#6A5ACD}{\text{Sales}}_i = \beta_0 + \beta_1 \textcolor{#e64173}{\text{Ad}}_i + \beta_2 \textcolor{#e64173}{\text{Price}}_i + \beta_3 \textcolor{#e64173}{\text{Comp}}_i + \varepsilon_i $$

#### (Review) Questions

- __Q:__ Are the $\beta_k$ terms population parameters or sample statistics?

::: {.fragment}
- __A:__ Greek letters denote __population parameters__. Their estimates get hats, _e.g._, $\hat{\beta}_k$. Population parameters represent the __average__ behavior across the population.
:::

---

__Regression model:__

$$ \textcolor{#6A5ACD}{\text{Sales}}_i = \beta_0 + \beta_1 \textcolor{#e64173}{\text{Ad}}_i + \beta_2 \textcolor{#e64173}{\text{Price}}_i + \beta_3 \textcolor{#e64173}{\text{Comp}}_i + \varepsilon_i $$


#### (Review) Questions

- __Q:__ Can we interpret the estimates for $\beta_2$ as causal?

::: {.fragment}
- __A:__ Not without making more assumptions and/or knowing more about the data-generating process.
:::

---

__Regression model:__

$$ \textcolor{#6A5ACD}{\text{Sales}}_i = \beta_0 + \beta_1 \textcolor{#e64173}{\text{Ad}}_i + \beta_2 \textcolor{#e64173}{\text{Price}}_i + \beta_3 \textcolor{#e64173}{\text{Comp}}_i + \varepsilon_i $$

#### (Review) Questions

- __Q:__ What is $\varepsilon_i$?

::: {.fragment}
- __A:__ An individual's random deviation/disturbance from the population parameters.

Population parameters are averages; individuals are rarely average.
:::

---

__Regression model:__

$$ \textcolor{#6A5ACD}{\text{Sales}}_i = \beta_0 + \beta_1 \textcolor{#e64173}{\text{Ad}}_i + \beta_2 \textcolor{#e64173}{\text{Price}}_i + \beta_3 \textcolor{#e64173}{\text{Comp}}_i + \varepsilon_i $$

#### (Review) Questions

- __Q:__ Which assumptions do we impose when estimating with OLS?

::: {.fragment}
- __A:__
  - The relationship between the sales and the [explanatory variables]{.pink} is linear in parameters, and $\varepsilon$ enters additively.
  - The [explanatory variables]{.pink} are __exogenous__, _i.e._, $E[\varepsilon|X] = 0$.
  - You've also typically assumed something along the lines of:<br> $E[\varepsilon_i] = 0$, $E[\varepsilon_i^2] = \sigma^2$, $E[\varepsilon_i \varepsilon_j] = 0$ for $i \neq j$.
  - And (maybe) $\varepsilon_i$ is distributed normally.
:::

---

## Assumptions

### How important can they be?

You've learned how **powerful and flexible** ordinary least squares (**OLS**) regression can be.

::: {.fragment}
However, the results you learned required assumptions.
:::

::: {.fragment}
**Real life often violates these assumptions.**
:::

---

## Assumptions

### How important can they be?

EC421 asks "**What happens when we violate these assumptions?**"

- Can we find a fix? (Especially: How/when is $\beta$ *causal*?)
- What happens if we don't (or can't) apply a fix?

OLS still does some amazing things, but you need to know when to be **cautious, confident, or dubious**.

---

## Not everything is causal {visibility="hidden"}

```{r}
#| label: spurious
#| echo: false
#| fig-align: center
#| fig-width: 10.5
#| fig-height: 7

tmp <- data.frame(
  year = 1999:2009,
  count = c(
    9, 8, 11, 12, 11, 13, 12, 9, 9, 7, 9,
    6, 5, 5, 10, 8, 14, 10, 4, 8, 5, 6
  ),
  type = rep(c("letters", "deaths"), each = 11)
)
ggplot(data = tmp, aes(x = year, y = count, color = type)) +
  geom_path() +
  geom_point(size = 4) +
  xlab("Year") +
  ylab("Count") +
  scale_color_manual(
    "",
    labels = c("Deaths from spiders", "Letters in the winning spelling bee word"),
    values = c(red_pink, "darkslategray")
  ) +
  theme_pander(base_size = 17) +
  theme(legend.position = "bottom")
```

---

## Not everything is causal

### But what _is_?

Suppose you estimate our sales model for your boss.

$$ \textcolor{#6A5ACD}{\text{Sales}}_i = \hat{\beta}_0 + \hat{\beta}_1 \textcolor{#e64173}{\text{Ad}}_i + \hat{\beta}_2 \textcolor{#e64173}{\text{Price}}_i + \hat{\beta}_3 \textcolor{#e64173}{\text{Comp}}_i + e_i $$

Can you trust that $\hat{\beta}_2$ gives you the actual effect of [price]{.pink} on [sales]{.purple}?

::: {.fragment}
You should be asking several questions...
:::

::: {.fragment}
1. [Where]{.it} does the [variation in price]{.it .pink} come from?
   - Is it [random]{.it .pink} ([exogenous]{.it})?
   - [Why]{.it .pink} are some products (or times) [more]{.it .pink} expensive than others?
2. [Whom]{.it .orange} do the data represent? Are they [relevant]{.it .orange} to your setting?
3. How [confident]{.it .orange} are you in your answer?
:::

---

## Econometrics

Applied econometrics, data science, analytics require:

1. Intuition for the __theory__ behind statistics/econometrics<br>(assumptions, results, strengths, weaknesses).

2. Practical knowledge of how to __apply theoretical methods__ to data.

3. Efficient methods for __working with data__<br>(cleaning, aggregating, joining, visualizing).

::: {.fragment}
__This course__ aims to deepen your knowledge in each of these three areas.
:::

::: {.fragment}
- 1: As before.
- 2-3: __R__
:::

---

My "[big-picture takeaways]{.b}" (the [intuition]{.it .slab} that I hope you form)

::: {.columns .small}
::: {.column width="50%"}
::: {.fragment}
- [most]{.it} interesting questions are [causal]{.slab .pink};
- [selection into treatment]{.slab .pink} dominates correlation (esp. cross-sectional);
- [measurement error]{.slab .pink} can too;
- causality comes from [design]{.slab .pink}, not from models/assumptions;
- ask about the [counterfactual]{.slab .pink};
- [non-stationary]{.slab .pink} time series will lead you to bad conclusions;
:::
:::

::: {.column width="50%"}
::: {.fragment}
- quantifying [uncertainty]{.slab .purple} is just as important as the effect estimate;
- consider the [which population]{.slab .purple .it} your data represent;
- the [mean]{.slab .purple} is only one of many ways to summarize a population;
- don't mistake [mean reversion]{.slab .purple} for treatment effects/heterogeneity;
:::
::: {.fragment}
- many [maps]{.slab .orange} are just [population]{.it};
- [graphs]{.slab .orange} should clearly communicate a [message]{.it}... beautifully.
:::
:::
:::

---

## {.center}

[Next:]{.it .slab} [R]{.mono} basics + (More) Metrics review(s)
