---
title: "Panel Estimation"
subtitle: "EC 421, Set 12"
author: "Edward Rubin"
date: ""
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, middle

```{r, setup, include = FALSE}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(
  broom, dplyr, fixest, ggplot2, kableExtra, knitr,
  magrittr, scales, tidyr
)

# Define colors
red_pink = "#e64173"
turquoise = "#20B2AA"
orange = "#FFA500"
red = "#fb6107"
blue = "#2b59c3"
green = "#8bb174"
grey_light = "grey70"
grey_mid = "grey50"
grey_dark = "grey20"
purple = "#6A5ACD"

# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = FALSE,
  message = FALSE
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(crayon.enabled = FALSE)
options(knitr.table.format = "html")

table_style <- function(x, digits = 2, font_size = 22) {
  align <- paste(rep("c", ncol(x)), collapse = "")
  x %>%
    kable(escape = FALSE, digits = digits, align = align) %>%
    kable_styling(font_size = font_size, full_width = FALSE)
}

fmt3 <- function(x) sprintf("%.3f", x)

theme_set(theme_minimal(base_size = 20))

set.seed(421)

# Tiny example panel for early slides
example_panel <- tribble(
  ~Unit, ~Year, ~`y_it`, ~`x_it`,
  "A", 1, 10, 2,
  "A", 2, 11, 3,
  "A", 3, 13, 4,
  "B", 1, 7, 1,
  "B", 2, 8, 1,
  "B", 3, 9, 2
)

# Simulated state panel for minimum-wage example
states <- state.abb[1:40]
years <- 2014:2021
alpha_state <- rnorm(length(states), sd = 3)
policy_pref <- 0.9 * alpha_state + rnorm(length(states), sd = 1)
names(alpha_state) <- states
names(policy_pref) <- states
lambda_t <- c(-1.0, -0.5, -0.2, 0.0, 0.3, 0.6, 0.9, 1.1)

mw_df <- expand_grid(state = states, year = years) %>%
  mutate(
    alpha = alpha_state[state],
    policy = policy_pref[state],
    post = year >= 2018,
    min_wage = 7.25 + 0.35 * policy + 0.18 * (year - min(years)) +
      if_else(policy > median(policy) & post, 1.1, 0) + rnorm(n(), sd = 0.18),
    teen_emp = 58 - 1.6 * min_wage + 2.7 * alpha + lambda_t[match(year, years)] +
      rnorm(n(), sd = 1.4)
  )

plot_ids <- mw_df %>%
  distinct(state, alpha) %>%
  arrange(alpha) %>%
  slice(c(3, 10, 17, 24, 31, 38)) %>%
  pull(state)

mw_plot_df <- mw_df %>%
  filter(state %in% plot_ids) %>%
  mutate(state = factor(state, levels = plot_ids))

mw_df$mw_tw <- resid(feols(min_wage ~ 1 | state + year, data = mw_df))
mw_df$emp_tw <- resid(feols(teen_emp ~ 1 | state + year, data = mw_df))

pool_mod <- feols(teen_emp ~ min_wage, data = mw_df)
fe_mod <- feols(teen_emp ~ min_wage | state + year, cluster = "state", data = mw_df)

pool_se <- sqrt(diag(vcov(pool_mod)))[["min_wage"]]
fe_se <- sqrt(diag(vcov(fe_mod)))[["min_wage"]]
fe_wr2 <- as.numeric(fitstat(fe_mod, "wr2"))

model_compare <- tibble(
  Model = c("Truth", "Pooled OLS", "State + year FE"),
  `Slope on MinWage_it` = c("-1.600", fmt3(coef(pool_mod)[["min_wage"]]), fmt3(coef(fe_mod)[["min_wage"]])),
  `Std. error` = c("", fmt3(pool_se), fmt3(fe_se)),
  `Within R^2` = c("", "", fmt3(fe_wr2))
)

# DiD example from fixest
data(base_did)

did_cells <- base_did %>%
  group_by(treat, post) %>%
  summarize(mean_y = mean(y), .groups = "drop") %>%
  mutate(
    Group = ifelse(treat == 1, "Treated", "Control"),
    Period = ifelse(post == 1, "Post", "Pre")
  )

did_table <- did_cells %>%
  select(Group, Period, mean_y) %>%
  pivot_wider(names_from = Period, values_from = mean_y) %>%
  mutate(
    Change = Post - Pre,
    across(c(Pre, Post, Change), fmt3)
  ) %>%
  arrange(Group)

control_change <- did_cells %>%
  filter(treat == 0) %>%
  summarize(change = mean_y[post == 1] - mean_y[post == 0]) %>%
  pull(change)

treated_change <- did_cells %>%
  filter(treat == 1) %>%
  summarize(change = mean_y[post == 1] - mean_y[post == 0]) %>%
  pull(change)

did_est <- treated_change - control_change

did_plot_df <- base_did %>%
  group_by(treat, period) %>%
  summarize(mean_y = mean(y), .groups = "drop") %>%
  mutate(Group = ifelse(treat == 1, "Treated", "Control"))

did_mod <- feols(y ~ i(post, treat, ref = 0) | id + period, cluster = "id", data = base_did)
did_se <- sqrt(diag(vcov(did_mod)))[["post::1:treat"]]

did_reg_table <- tibble(
  Term = "post x treat",
  Estimate = fmt3(coef(did_mod)[["post::1:treat"]]),
  `Std. error` = fmt3(did_se),
  `Within R^2` = fmt3(as.numeric(fitstat(did_mod, "wr2")))
)
```

# Prologue

---
name: schedule

# Schedule

## Last time

Instrumental variables

## Today

**Panel estimation**

- what panel data buy us,
- fixed effects,
- the difference-in-differences estimator,
- *brief extension:* the synthetic-controls estimator.

---
layout: false
class: inverse, middle

# Panel estimation

---
layout: true

# Panel estimation

---
name: panel_intro

## Introduction

Until now, we mostly used one of two data structures:

--

1. .hi-orange[Cross section:] many units, one period,
--

2. .hi-purple[Time series:] one unit, many periods.

--

A .hi[panel] gives us both:

- .orange[units] indexed by $\color{#FFA500}{i}$,
- .purple[(time) periods] indexed by $\color{#6A5ACD}{t}$,
- outcomes and regressors like $\color{#e64173}{y_{\color{#FFA500}{i}\color{#6A5ACD}{t}}}$ and $\color{#e64173}{x_{\color{#FFA500}{i}\color{#6A5ACD}{t}}}$.

---
## Example

Consider the following panel of six observations:

.pull-left[
```{r, example-panel-1, echo = FALSE}
example_panel |> data.table::as.data.table()
```
]

---
## Example

Consider the following panel of six observations:

.pull-left[
```{r, example-panel-2, echo = FALSE, highlight.output = 3:5}
example_panel |> data.table::as.data.table()
```
]
.pull-right[
We observe the .orange[same unit] repeatedly.
]

We can compare unit A to itself .hi-purple[over time], not just A to B<br>(potentially helping with *selection bias*).

---
## Example

Consider the following panel of six observations:

.pull-left[
```{r, example-panel-3, echo = FALSE, highlight.output = c(3, 6)}
example_panel |> data.table::as.data.table()
```
]
.pull-right[
Similarly, we observe the .purple[period] across .orange[units].
]

This feature lets us control for shocks shared across .orange[units] at a given .purple[time].

---

## The benefits of panels

Panels help solve (some) econometric problems because they allow us to

--

- control for .b.orange[time-invariant differences] across units<br>.grey-light.it[removes selection bias from cross-sectional differences across units]
  
--

- implicitly study .b.purple[changes]<br>.grey-light.it[not just levels]

--

- .note[related:] evaluate policies using their .b.purple[timing]<br>.grey-light.it[policy-timing is often more plausibly exogeneous than policy existence]

--

Panel methods are workhorses of economics and public policy.

---

## Panels are helpful—not magical

Panel data *can* be helpful for estimating causal effects,
<br>but they can still cause problems without careful use.

--

- We need (sufficient) .it.orange[within variation] in $\color{#FFA500}{x_{it}}$.<br>.it[I.e.,] $x_{it}$ must change meaningfully through time for some units.

- .purple[Time-varying] omitted variables can cause bias.

- A .orange[unit]'s disturbances $u_{\color{#FFA500}{i}\color{#6A5ACD}{t}}$ are often correlated over .purple[time].

- Attrition and missing periods can matter—as can measurement error.

.grey-light.it[(Any tool can be misused.)]

---
name: pooled_problem
## The problem with pooled OLS

Suppose you're interested in a simple linear regression model
$$y_{it} = \beta_0 + \beta_1 x_{it} + u_{it}$$
and have .b.slate[panel data] (with individual $i$ and time $t$).

--

Let the disturbance combine (1) .orange[unit]-specific factor $\color{#FFA500}{\alpha_i}$ and (2) error $\color{#e64173}{v_{it}}$
$$
\begin{align}
  u_{it} = \color{#FFA500}{\alpha_i} + \color{#e64173}{v_{it}}
\end{align}
$$

--

If $\mathop{\text{Cov}}\left( x_{it}, \color{#FFA500}{\alpha_i} \right) \neq 0$, then OLS may be biased. .grey-vlight.it[(Think OVB.)]

OLS's estimates of $\beta_1$ will reflect both

- .it[between-unit differences] $\color{#FFA500}{\alpha_i}$,
- .it[actual changes] in $x_{it}$. .note[(We want to isolate this last part.)]
  
---

## Introducing *fixed effects*

.note[The solution:] Let each .orange[unit] have its own intercept $\color{#FFA500}{\alpha_i}$,<br> and each .purple[time period] have its own intercept $\color{#6A5ACD}{\lambda_t}$:
$$y_{it} = \beta_1 x_{it} + \color{#FFA500}{\alpha_i} + \color{#6A5ACD}{\lambda_t} + u_{it}$$

These .orange[unit]- and .purple[period]-specific intercepts are called .attn[fixed effects].

--

```{r, example-panel-4, echo = FALSE}
# Show fixed effects
example_panel |>
  mutate(
    Unit_A = 1 * (Unit == 'A'),
    Unit_B = 1 * (Unit == 'B'),
    Year_1 = 1 * (Year == 1),
    Year_2 = 1 * (Year == 2),
    Year_3 = 1 * (Year == 3)
  ) |>
  data.table::as.data.table()
```

---
## Fixed effects in action

.qa[Q] So what are these fixed effects (FE) doing?

--

Remember the old interpretation of coefficients on binary variables?

--

.qa[A] These fixed effects *absorb*

- $\color{#FFA500}{\alpha_i}$: all time-invariant .orange[differences across units],<br>.grey-light.it[(including selection bias from cross-sectional differences)]

- $\color{#6A5ACD}{\lambda_t}$: all common shocks at a given .purple[time].

--

The .attn[(two-way) fixed-effects estimator] (OLS that includes unit and time FEs)
<br>estimates effects using .pink.it[within-unit variation]—net of common time shocks.

---

## The *within* transformation

For intuition, focus on unit-level fixed effects.
$$y_{it} = \beta_0 + \beta_1 x_{it} + \alpha_i + u_{it}$$

--

Take the average of this equation for unit $i$ across all periods...
$$\overline{y}_i = \beta_0 + \beta_1 \overline{x}_i + \alpha_i + \overline{u}_i$$

--

Finally, subtract the unit mean from both sides of the original equation.
<br>(This step is what our unit-level fixed effects are doing.)
$$y_{it} - \overline{y}_i = \beta_1 \left( x_{it} - \overline{x}_i \right) + \left( u_{it} - \overline{u}_i \right)$$

--

.hi[So what?] Unit-specific FEs identify $\beta_1$ using .attn[within-unit variation] in $x_{it}$.
<br>.grey-vlight.it[(The same logic applies to time FEs.)]

---
## Fixed-effects questions

Fixed effects .b[do not] ask questions like
<br>.note[Why is unit A different from unit B?]
<br>.grey-light.it[In fact, they try to absorb cross-unit differences.]

--

Fixed effects .b[do] ask questions like
<br>.note[When unit A changed its value of] $x_{it}$, .note[, how did its] $y_{it}$ .note[change?]
<br>.grey-light.it[This is the "within" variation.]

---
## Example: State minimum wage laws

Suppose we want to estimate the effect of min. wage laws on employment
$$\text{Employment}_{it} = \beta_1 \text{(Min. Wage)}_{it} + \alpha_i + \lambda_t + u_{it}$$
with a .orange[state] ($\color{#FFA500}{i}$) by .purple[year] ($\color{#6A5ACD}{t}$) panel of employment and min. wage data.

--

- $\alpha_i$ absorbs time-invariant, cross-state differences;
- $\lambda_t$ absorbs national shocks in year $t$.

--

.qa[Q] Should we worry that some states have higher min. wage than others?
--
<br>.qa[A] No. $\alpha_i$ absorbs those differences. The FE estimator estimates using within-state variation in min. wage.

---
## Example: State minimum wage laws

Suppose we want to estimate the effect of min. wage laws on employment
$$\text{Employment}_{it} = \beta_1 \text{(Min. Wage)}_{it} + \alpha_i + \lambda_t + u_{it}$$
with a .orange[state] ($\color{#FFA500}{i}$) by .purple[year] ($\color{#6A5ACD}{t}$) panel of employment and min. wage data.

- $\alpha_i$ absorbs time-invariant, cross-state differences;
- $\lambda_t$ absorbs national shocks in year $t$.


.qa[Q] What if some *years* have lower emp. than others (recessions)?
--
<br>.qa[A] Still not worried! $\lambda_t$ absorbs *shared* shocks.

---
## Example: State minimum wage laws

Suppose we want to estimate the effect of min. wage laws on employment
$$\text{Employment}_{it} = \beta_1 \text{(Min. Wage)}_{it} + \alpha_i + \lambda_t + u_{it}$$
with a .orange[state] ($\color{#FFA500}{i}$) by .purple[year] ($\color{#6A5ACD}{t}$) panel of employment and min. wage data.

- $\alpha_i$ absorbs time-invariant, cross-state differences;
- $\lambda_t$ absorbs national shocks in year $t$.


.qa[Q] So what should we worry about?
--
<br>.qa[A] We should worry about .b[time-varying differences between states].<br>*E.g.*, if some states have faster-growing economies and also raise their minimum wages.

---

## What FE means in that example

In the minimum-wage panel, fixed effects net out factors like

--

- state geography
- long-run industry mix
- baseline labor-market conditions
- permanent political differences

--

So $\beta_1$ comes from .hi[states that change their minimum wage], relative to their own histories and common national trends.

---
layout: false
class: clear, middle

```{r, fe-raw-plot, echo = FALSE}
ggplot(mw_plot_df, aes(x = min_wage, y = teen_emp, color = state)) +
  geom_point(size = 2.4, alpha = 0.9) +
  geom_smooth(method = 'lm', formula = y ~ x, se = FALSE, linewidth = 0.9) +
  geom_smooth(
    data = mw_df,
    aes(x = min_wage, y = teen_emp),
    inherit.aes = FALSE,
    method = 'lm',
    formula = y ~ x,
    se = FALSE,
    color = grey_dark,
    linewidth = 1.4
  ) +
  scale_color_viridis_d(option = 'magma', begin = .2, end = .85) +
  labs(
    x = 'State minimum wage',
    y = 'Teen employment rate',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    # legend.position = 'bottom'
  )
```

The dark line is .hi-orange[pooled OLS].

It slopes .hi-orange[up] because high-wage states also have stronger labor markets for other reasons.

--

That is .hi[cross-sectional selection bias]: states with high minimum wages are not directly comparable to states with low minimum wages.

---

## Minimum-wage example results

```{r, fe-model-table, echo = FALSE}
table_style(model_compare, digits = 3, font_size = 24)
```

--

.hi[Truth:] Raising the minimum wage lowers teen employment in this example.

--

.hi-purple[Fixed effects] flip the sign because they compare each state to itself over time.

---

## After partialling out unit and year effects

```{r, fe-resid-plot, echo = FALSE}
ggplot(mw_df, aes(x = mw_tw, y = emp_tw)) +
  geom_hline(yintercept = 0, color = grey_light, linewidth = 0.5) +
  geom_vline(xintercept = 0, color = grey_light, linewidth = 0.5) +
  geom_point(color = blue, alpha = 0.35, size = 2) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = red_pink,
    linewidth = 1.4
  ) +
  labs(
    x = "Minimum wage residual",
    y = "Employment residual"
  ) +
  theme_minimal(base_size = 21) +
  theme(panel.grid.minor = element_blank())
```

--

This is the .hi[within relationship] after removing state and year averages.

---

## `fixest` syntax

```{r, fe-code, eval = FALSE}
fe_mod <- feols(
  teen_emp ~ min_wage | state + year,
  cluster = "state",
  data = mw_df
)
```

--

- Left of `|`: regressors of interest

--

- Right of `|`: fixed effects

--

- `cluster = "state"`: allow correlation within a state over time

---

## Within $R^2$

For the FE model above:

```{r, fe-wr2, echo = FALSE}
table_style(
  tibble(
    Statistic = c("Slope on min_wage", "Within R^2"),
    Value = c(fmt3(coef(fe_mod)[["min_wage"]]), fmt3(fe_wr2))
  ),
  digits = 3,
  font_size = 25
)
```

--

.hi[Within $R^2$] asks:

> After removing the fixed effects, how much of the remaining variation in teen employment do changes in the minimum wage explain?

--

It is the right fit measure when FE identification comes from .hi-purple[within variation].

---

## What FE does not fix

Fixed effects solve one important problem:

- .hi[time-invariant omitted variables]

--

But FE does .hi-orange[not] solve

- time-varying confounders
- reverse causality
- bad measurement of treatment timing
- weak within variation

---
layout: false
class: inverse, middle

# Difference-in-Differences

---
layout: true
# Difference-in-Differences
## Core Idea

---
name: did_intuition

Difference-in-differences (DiD) is a .hi[panel-data design].

--

We compare

- a .hi-purple[treated group] and
- a .hi-slate[control group]

--

before and after treatment begins.

--

It is fixed effects plus a very specific source of within variation:

.hi-orange[a policy shock that turns on for some units and not others.]

---

## The basic DiD formula

$$
\begin{align}
  \widehat{\delta}_{DiD} =
  \left( \overline{y}_{T,post} - \overline{y}_{T,pre} \right) -
  \left( \overline{y}_{C,post} - \overline{y}_{C,pre} \right)
\end{align}
$$

--

The first difference removes permanent treated-control differences.

--

The second difference removes common time shocks.

---

## A 2 x 2 DiD table

```{r, did-cell-table, echo = FALSE}
table_style(did_table, digits = 3, font_size = 24)
```

--

- Treated change: `r fmt3(treated_change)`
- Control change: `r fmt3(control_change)`
- DiD estimate: `r fmt3(did_est)`

---

## Classic DiD example: Card and Krueger

One famous application compares .hi[fast-food employment]

--

- in .hi-purple[New Jersey], where the minimum wage rose
- to .hi-slate[Pennsylvania], where it did not

--

before and after the policy change.

--

DiD asks whether employment in New Jersey changed .hi-orange[more or less than the counterfactual trend] suggested by Pennsylvania.

---

## DiD as a regression

With two groups and two periods:

$$
\begin{align}
  y_{it} = \beta_0 + \beta_1 \text{Treat}_i + \beta_2 \text{Post}_t + \delta \left( \text{Treat}_i \times \text{Post}_t \right) + u_{it}
\end{align}
$$

--

With panel data, we often write

$$
\begin{align}
  y_{it} = \alpha_i + \lambda_t + \delta D_{it} + u_{it}
\end{align}
$$

--

where $D_{it}=1$ only for treated units after treatment starts.

---

## DiD in pictures

```{r, did-plot, echo = FALSE}
ggplot(did_plot_df, aes(x = period, y = mean_y, color = Group)) +
  geom_vline(xintercept = 5.5, linetype = "dashed", color = grey_mid) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2.7) +
  scale_color_manual(values = c("Control" = purple, "Treated" = red_pink)) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = "Period",
    y = "Average y",
    color = NULL
  ) +
  theme_minimal(base_size = 21) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

--

Treatment begins between periods 5 and 6.

---

## `fixest` DiD code

```{r, did-code, eval = FALSE}
did_mod <- feols(
  y ~ i(post, treat, ref = 0) | id + period,
  cluster = "id",
  data = base_did
)
```

--

```{r, did-reg-table, echo = FALSE}
table_style(did_reg_table, digits = 3, font_size = 24)
```

--

The interaction term is the .hi[DiD estimate].

---

## The key DiD assumption

.hi[Parallel trends:]

In the absence of treatment, treated and control units would have followed the same average trend.

--

What threatens DiD?

--

- treated units were already on a different trend
- anticipation changes behavior before treatment
- another shock hits treated units at the same time
- standard errors ignore within-unit dependence

---

## Where DiD fits

DiD is especially useful when

--

- treatment timing is clear

--

- you have good pre-treatment data

--

- untreated units provide a plausible counterfactual

--

- the treatment effect is tied to a policy or event

---
layout: true
# Panel estimation
## What comes next

---
name: scm

## Synthetic control and friends

If one unit gets treated and many do not, .hi[synthetic control methods] can help.

--

Idea:

- build a weighted average of control units
- match the treated unit's pre-treatment path
- compare treated vs. synthetic control after treatment

--

There are many related panel methods, but for us the core ideas today are still

- .hi[within variation]
- .hi-purple[fixed effects]
- .hi-orange[treated vs. untreated changes over time]

---

## Takeaways

- Panel data let us compare a unit .hi[to itself over time].

--

- Fixed effects remove .hi-purple[time-invariant omitted variables].

--

- The identifying variation is .hi-orange[within variation], so within $R^2$ matters.

--

- DiD is a panel-data strategy built around a treated-control change over time.

--

- Panels help a lot, but they still depend on strong assumptions.
