---
title: "Panel Estimation"
subtitle: "EC 421, Set 12"
author: "Edward Rubin"
date: ""
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, middle

```{r, setup, include = FALSE}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(
  broom, dplyr, fixest, ggplot2, kableExtra, knitr,
  magrittr, scales, tidyr
)

# Define colors
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#fb6107"
blue <- "#2b59c3"
green <- "#8bb174"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"

# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = FALSE,
  message = FALSE
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(crayon.enabled = FALSE)
options(knitr.table.format = "html")

table_style <- function(x, digits = 2, font_size = 22) {
  align <- paste(rep("c", ncol(x)), collapse = "")
  x %>%
    kable(escape = FALSE, digits = digits, align = align) %>%
    kable_styling(font_size = font_size, full_width = FALSE)
}

fmt3 <- function(x) sprintf("%.3f", x)

theme_set(theme_minimal(base_size = 20))

set.seed(421)

# Tiny example panel for early slides
example_panel <- tribble(
  ~Unit, ~Year, ~`y_it`, ~`x_it`,
  "A", 1, 10, 2,
  "A", 2, 11, 3,
  "A", 3, 13, 4,
  "B", 1, 7, 1,
  "B", 2, 8, 1,
  "B", 3, 9, 2
)

# Simulated state panel for minimum-wage example
states <- state.abb[1:40]
years <- 2014:2021
alpha_state <- rnorm(length(states), sd = 3)
policy_pref <- 0.9 * alpha_state + rnorm(length(states), sd = 1)
names(alpha_state) <- states
names(policy_pref) <- states
lambda_t <- c(-1.0, -0.5, -0.2, 0.0, 0.3, 0.6, 0.9, 1.1)

mw_df <- expand_grid(state = states, year = years) %>%
  mutate(
    alpha = alpha_state[state],
    policy = policy_pref[state],
    post = year >= 2018,
    min_wage = 7.25 + 0.35 * policy + 0.18 * (year - min(years)) +
      if_else(policy > median(policy) & post, 1.1, 0) + rnorm(n(), sd = 0.18),
    teen_emp = 58 - 1.6 * min_wage + 2.7 * alpha + lambda_t[match(year, years)] +
      rnorm(n(), sd = 1.4)
  )

plot_ids <- mw_df %>%
  distinct(state, alpha) %>%
  arrange(alpha) %>%
  slice(c(3, 10, 17, 24, 31, 38)) %>%
  pull(state)

mw_plot_df <- mw_df %>%
  filter(state %in% plot_ids) %>%
  mutate(state = factor(state, levels = plot_ids))

mw_df$mw_tw <- resid(feols(min_wage ~ 1 | state + year, data = mw_df))
mw_df$emp_tw <- resid(feols(teen_emp ~ 1 | state + year, data = mw_df))

pool_mod <- feols(teen_emp ~ min_wage, data = mw_df)
fe_mod <- feols(teen_emp ~ min_wage | state + year, cluster = "state", data = mw_df)

pool_se <- sqrt(diag(vcov(pool_mod)))[["min_wage"]]
fe_se <- sqrt(diag(vcov(fe_mod)))[["min_wage"]]
fe_wr2 <- as.numeric(fitstat(fe_mod, "wr2"))

model_compare <- tibble(
  Model = c("Truth", "Pooled OLS", "State + year FE"),
  `Slope on MinWage_it` = c("-1.600", fmt3(coef(pool_mod)[["min_wage"]]), fmt3(coef(fe_mod)[["min_wage"]])),
  `Std. error` = c("", fmt3(pool_se), fmt3(fe_se)),
  `Within R^2` = c("", "", fmt3(fe_wr2))
)

# DiD example from fixest
data(base_did)

did_cells <- base_did %>%
  group_by(treat, post) %>%
  summarize(mean_y = mean(y), .groups = "drop") %>%
  mutate(
    Group = ifelse(treat == 1, "Treated", "Control"),
    Period = ifelse(post == 1, "Post", "Pre")
  )

did_table <- did_cells %>%
  select(Group, Period, mean_y) %>%
  pivot_wider(names_from = Period, values_from = mean_y) %>%
  mutate(
    Change = Post - Pre,
    across(c(Pre, Post, Change), fmt3)
  ) %>%
  arrange(Group)

control_change <- did_cells %>%
  filter(treat == 0) %>%
  summarize(change = mean_y[post == 1] - mean_y[post == 0]) %>%
  pull(change)

treated_change <- did_cells %>%
  filter(treat == 1) %>%
  summarize(change = mean_y[post == 1] - mean_y[post == 0]) %>%
  pull(change)

did_est <- treated_change - control_change

did_plot_df <- base_did %>%
  group_by(treat, period) %>%
  summarize(mean_y = mean(y), .groups = "drop") %>%
  mutate(Group = ifelse(treat == 1, "Treated", "Control"))

did_mod <- feols(y ~ i(post, treat, ref = 0) | id + period, cluster = "id", data = base_did)
did_se <- sqrt(diag(vcov(did_mod)))[["post::1:treat"]]

did_reg_table <- tibble(
  Term = "post x treat",
  Estimate = fmt3(coef(did_mod)[["post::1:treat"]]),
  `Std. error` = fmt3(did_se),
  `Within R^2` = fmt3(as.numeric(fitstat(did_mod, "wr2")))
)
```

# Prologue

---
name: schedule

# Schedule

## Last Time

Instrumental variables

## Today

- What panel data buy us
- Fixed effects
- Difference-in-differences
- Brief extension: synthetic controls

---
layout: false
class: inverse, middle

# Panel Estimation

---
layout: true
# Panel Estimation
## Introduction

---
name: panel_intro

Until now, we mostly used one of two data structures:

--

- .hi-slate[Cross section:] many units, one period

--

- .hi-purple[Time series:] one unit, many periods

--

A .hi[panel] gives us both:

- units indexed by $i$
- periods indexed by $t$
- outcomes and regressors like $\color{#e64173}{y_{it}}$ and $\color{#e64173}{x_{it}}$

---

## Example: A tiny panel

```{r, example-panel, echo = FALSE}
table_style(example_panel, digits = 0, font_size = 24)
```

--

We observe the .hi[same unit repeatedly].

--

That means we can compare .hi-purple[unit A to itself over time], not just A to B.

---

## Why economists like panels

Panels can help because they let us

--

- control for .hi[time-invariant differences] across units

--

- study .hi-purple[changes], not just levels

--

- evaluate policies that start at a known time

--

- use the same logic for workers, firms, counties, states, or countries

---

## Panels are helpful, not magical

Panels still have several traps:

--

- We need .hi-orange[within variation] in $\color{#FFA500}{x_{it}}$.

--

- Time-varying omitted variables can still bias us.

--

- Errors within a unit are often correlated over time.

--

- Attrition and missing periods can matter.

---
layout: true
# Panel Estimation
## Fixed Effects

---
name: pooled_problem

Start with the familiar panel regression

$$
\begin{align}
  y_{it} = \beta_0 + \beta_1 x_{it} + u_{it}
\end{align}
$$

--

Now suppose the error term contains a unit-specific factor:

$$
\begin{align}
  u_{it} = \alpha_i + v_{it}
\end{align}
$$

--

If $\mathop{\text{Cov}}\left( x_{it}, \alpha_i \right) \neq 0$, then pooled OLS mixes

- .hi-slate[between-unit differences] and
- .hi-purple[within-unit changes].

---

## Fixed effects idea

Let each unit have its own intercept:

$$
\begin{align}
  y_{it} = \beta_1 x_{it} + \alpha_i + \lambda_t + u_{it}
\end{align}
$$

--

- $\color{#314f4f}{\alpha_i}$: unit fixed effects

--

- $\color{#6A5ACD}{\lambda_t}$: time fixed effects

--

- $\beta_1$: effect of $x_{it}$ using .hi[within-unit variation], net of common time shocks

---

## The within transformation

For intuition, focus on unit fixed effects:

$$
\begin{align}
  y_{it} &= \beta_0 + \beta_1 x_{it} + \alpha_i + u_{it}
\end{align}
$$

Subtract the unit mean from both sides:

$$
\begin{align}
  y_{it} - \overline{y}_i = \beta_1 \left( x_{it} - \overline{x}_i \right) + \left( u_{it} - \overline{u}_i \right)
\end{align}
$$

--

.hi[Why this matters:] $\alpha_i$ disappears because it is constant within unit $i$.

---

## What fixed effects use

Fixed effects do .hi[not] ask:

- Why is unit A different from unit B?

--

Fixed effects ask:

- When unit A changes $\color{#e64173}{x_{it}}$, how does $\color{#e64173}{y_{it}}$ change?

--

So FE uses .hi-purple[within variation].

--

Time-invariant regressors are dropped because

$$
\begin{align}
  z_i - \overline{z}_i = 0
\end{align}
$$

for every period.

---

## Applied example: state minimum wages

Suppose we want to estimate the effect of the minimum wage on employment.

$$
\begin{align}
  \text{Emp}_{it} = \beta_1 \text{MinWage}_{it} + \alpha_i + \lambda_t + u_{it}
\end{align}
$$

--

Here,

- $i$ is a .hi-purple[state]
- $t$ is a .hi-slate[year or quarter]

--

- $\alpha_i$ absorbs time-invariant state differences
- $\lambda_t$ absorbs national shocks

---

## What FE means in that example

In the minimum-wage panel, fixed effects net out factors like

--

- state geography
- long-run industry mix
- baseline labor-market conditions
- permanent political differences

--

So $\beta_1$ comes from .hi[states that change their minimum wage], relative to their own histories and common national trends.

---

## Example: pooled OLS vs. fixed effects

```{r, fe-raw-plot, echo = FALSE}
ggplot(mw_plot_df, aes(x = min_wage, y = teen_emp, color = state)) +
  geom_point(size = 2.4, alpha = 0.9) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.9) +
  geom_smooth(
    data = mw_df,
    aes(x = min_wage, y = teen_emp),
    inherit.aes = FALSE,
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = grey_dark,
    linewidth = 1.4
  ) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    x = "State minimum wage",
    y = "Teen employment rate",
    color = NULL
  ) +
  theme_minimal(base_size = 21) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

The dark line is .hi-orange[pooled OLS].

--

It slopes .hi-orange[up] because high-wage states also have stronger labor markets for other reasons.

--

That is .hi[cross-sectional selection bias]: states with high minimum wages are not directly comparable to states with low minimum wages.

---

## Minimum-wage example results

```{r, fe-model-table, echo = FALSE}
table_style(model_compare, digits = 3, font_size = 24)
```

--

.hi[Truth:] Raising the minimum wage lowers teen employment in this example.

--

.hi-purple[Fixed effects] flip the sign because they compare each state to itself over time.

---

## After partialling out unit and year effects

```{r, fe-resid-plot, echo = FALSE}
ggplot(mw_df, aes(x = mw_tw, y = emp_tw)) +
  geom_hline(yintercept = 0, color = grey_light, linewidth = 0.5) +
  geom_vline(xintercept = 0, color = grey_light, linewidth = 0.5) +
  geom_point(color = blue, alpha = 0.35, size = 2) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = red_pink,
    linewidth = 1.4
  ) +
  labs(
    x = "Minimum wage residual",
    y = "Employment residual"
  ) +
  theme_minimal(base_size = 21) +
  theme(panel.grid.minor = element_blank())
```

--

This is the .hi[within relationship] after removing state and year averages.

---

## `fixest` syntax

```{r, fe-code, eval = FALSE}
fe_mod <- feols(
  teen_emp ~ min_wage | state + year,
  cluster = "state",
  data = mw_df
)
```

--

- Left of `|`: regressors of interest

--

- Right of `|`: fixed effects

--

- `cluster = "state"`: allow correlation within a state over time

---

## Within $R^2$

For the FE model above:

```{r, fe-wr2, echo = FALSE}
table_style(
  tibble(
    Statistic = c("Slope on min_wage", "Within R^2"),
    Value = c(fmt3(coef(fe_mod)[["min_wage"]]), fmt3(fe_wr2))
  ),
  digits = 3,
  font_size = 25
)
```

--

.hi[Within $R^2$] asks:

> After removing the fixed effects, how much of the remaining variation in teen employment do changes in the minimum wage explain?

--

It is the right fit measure when FE identification comes from .hi-purple[within variation].

---

## What FE does not fix

Fixed effects solve one important problem:

- .hi[time-invariant omitted variables]

--

But FE does .hi-orange[not] solve

- time-varying confounders
- reverse causality
- bad measurement of treatment timing
- weak within variation

---
layout: false
class: inverse, middle

# Difference-in-Differences

---
layout: true
# Difference-in-Differences
## Core Idea

---
name: did_intuition

Difference-in-differences (DiD) is a .hi[panel-data design].

--

We compare

- a .hi-purple[treated group] and
- a .hi-slate[control group]

--

before and after treatment begins.

--

It is fixed effects plus a very specific source of within variation:

.hi-orange[a policy shock that turns on for some units and not others.]

---

## The basic DiD formula

$$
\begin{align}
  \widehat{\delta}_{DiD} =
  \left( \overline{y}_{T,post} - \overline{y}_{T,pre} \right) -
  \left( \overline{y}_{C,post} - \overline{y}_{C,pre} \right)
\end{align}
$$

--

The first difference removes permanent treated-control differences.

--

The second difference removes common time shocks.

---

## A 2 x 2 DiD table

```{r, did-cell-table, echo = FALSE}
table_style(did_table, digits = 3, font_size = 24)
```

--

- Treated change: `r fmt3(treated_change)`
- Control change: `r fmt3(control_change)`
- DiD estimate: `r fmt3(did_est)`

---

## Classic DiD example: Card and Krueger

One famous application compares .hi[fast-food employment]

--

- in .hi-purple[New Jersey], where the minimum wage rose
- to .hi-slate[Pennsylvania], where it did not

--

before and after the policy change.

--

DiD asks whether employment in New Jersey changed .hi-orange[more or less than the counterfactual trend] suggested by Pennsylvania.

---

## DiD as a regression

With two groups and two periods:

$$
\begin{align}
  y_{it} = \beta_0 + \beta_1 \text{Treat}_i + \beta_2 \text{Post}_t + \delta \left( \text{Treat}_i \times \text{Post}_t \right) + u_{it}
\end{align}
$$

--

With panel data, we often write

$$
\begin{align}
  y_{it} = \alpha_i + \lambda_t + \delta D_{it} + u_{it}
\end{align}
$$

--

where $D_{it}=1$ only for treated units after treatment starts.

---

## DiD in pictures

```{r, did-plot, echo = FALSE}
ggplot(did_plot_df, aes(x = period, y = mean_y, color = Group)) +
  geom_vline(xintercept = 5.5, linetype = "dashed", color = grey_mid) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2.7) +
  scale_color_manual(values = c("Control" = purple, "Treated" = red_pink)) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = "Period",
    y = "Average y",
    color = NULL
  ) +
  theme_minimal(base_size = 21) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

--

Treatment begins between periods 5 and 6.

---

## `fixest` DiD code

```{r, did-code, eval = FALSE}
did_mod <- feols(
  y ~ i(post, treat, ref = 0) | id + period,
  cluster = "id",
  data = base_did
)
```

--

```{r, did-reg-table, echo = FALSE}
table_style(did_reg_table, digits = 3, font_size = 24)
```

--

The interaction term is the .hi[DiD estimate].

---

## The key DiD assumption

.hi[Parallel trends:]

In the absence of treatment, treated and control units would have followed the same average trend.

--

What threatens DiD?

--

- treated units were already on a different trend
- anticipation changes behavior before treatment
- another shock hits treated units at the same time
- standard errors ignore within-unit dependence

---

## Where DiD fits

DiD is especially useful when

--

- treatment timing is clear

--

- you have good pre-treatment data

--

- untreated units provide a plausible counterfactual

--

- the treatment effect is tied to a policy or event

---
layout: true
# Panel Estimation
## What Comes Next

---
name: scm

## Synthetic control and friends

If one unit gets treated and many do not, .hi[synthetic control methods] can help.

--

Idea:

- build a weighted average of control units
- match the treated unit's pre-treatment path
- compare treated vs. synthetic control after treatment

--

There are many related panel methods, but for us the core ideas today are still

- .hi[within variation]
- .hi-purple[fixed effects]
- .hi-orange[treated vs. untreated changes over time]

---

## Takeaways

- Panel data let us compare a unit .hi[to itself over time].

--

- Fixed effects remove .hi-purple[time-invariant omitted variables].

--

- The identifying variation is .hi-orange[within variation], so within $R^2$ matters.

--

- DiD is a panel-data strategy built around a treated-control change over time.

--

- Panels help a lot, but they still depend on strong assumptions.
