---
title: 'Panel Estimation'
subtitle: 'EC 421, Set 12'
author:
  - name: Ed Rubin
    email: edwardr@uoregon.edu
editor:
  render-on-save: true
format:
  clean-revealjs:
    slide-level: 2
    theme: [../my.scss, default]
    highlight-style: gruvbox
    embed-resources: true
    slide-number: c/t
    incremental: false
    controls-layout: edges
    transition: none
    preview-links: auto
    html-math-method: mathjax
    code-line-numbers: false
execute:
  warning: false
  message: false

---

## {.center}

[Prologue]{.bigger .it .slab}

---

## Schedule {#schedule}

```{r}
#| label: "setup"
#| include: false
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(
  data.table, broom, dplyr, fixest, ggplot2, kableExtra, knitr,
  magrittr, scales, tidyr
)

# Define colors
red_pink = '#e64173'
turquoise = '#20B2AA'
orange = '#FFA500'
orange_l = '#f8d194'
red = '#fb6107'
blue = '#2b59c3'
green = '#8bb174'
grey_light = 'grey70'
grey_mid = 'grey50'
grey_dark = 'grey20'
purple = '#6A5ACD'
purple_l = '#c5bee9'
slate = '#314f4f'

# Knitr options
opts_chunk$set(
  comment = '#>',
  fig.align = 'center',
  fig.height = 7,
  fig.width = 10.5,
  warning = FALSE,
  message = FALSE
)
opts_chunk$set(dev = 'svg')
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(crayon.enabled = FALSE)
options(knitr.table.format = 'html')

table_style <- function(x, digits = 2, font_size = 22) {
  align <- paste(rep('c', ncol(x)), collapse = '')
  x %>%
    kable(escape = FALSE, digits = digits, align = align) %>%
    kable_styling(font_size = font_size, full_width = FALSE)
}

fmt3 <- function(x) sprintf('%.3f', x)

theme_set(theme_minimal(base_size = 20))

set.seed(421)

# Tiny example panel for early slides
example_panel <- tribble(
  ~Unit, ~Year, ~`y_it`, ~`x_it`,
  'A', 1, 10, 2,
  'A', 2, 11, 3,
  'A', 3, 13, 4,
  'B', 1, 7, 1,
  'B', 2, 8, 1,
  'B', 3, 9, 2
)

# Simulated state panel for minimum-wage example
states <- state.abb[1:40]
years <- 2014:2021
alpha_state <- rnorm(length(states), sd = 3)
policy_pref <- 0.9 * alpha_state + rnorm(length(states), sd = 1)
names(alpha_state) <- states
names(policy_pref) <- states
lambda_t <- c(-1.0, -0.5, -0.2, 0.0, 0.3, 0.6, 0.9, 1.1)

mw_df <- expand_grid(state = states, year = years) %>%
  mutate(
    alpha = alpha_state[state],
    policy = policy_pref[state],
    post = year >= 2018,
    min_wage = 7.25 + 0.35 * policy + 0.18 * (year - min(years)) +
      if_else(policy > median(policy) & post, 1.1, 0) + rnorm(n(), sd = 0.18),
    teen_emp = 58 - 1.6 * min_wage + 2.7 * alpha + lambda_t[match(year, years)] +
      rnorm(n(), sd = 1.4)
  )

plot_ids <- mw_df %>%
  distinct(state, alpha) %>%
  arrange(alpha) %>%
  slice(c(3, 10, 17, 24, 31, 38)) %>%
  pull(state)

mw_plot_df <- mw_df %>%
  filter(state %in% plot_ids) %>%
  mutate(state = factor(state, levels = plot_ids))

mw_df$mw_tw <- resid(feols(min_wage ~ 1 | state + year, data = mw_df))
mw_df$emp_tw <- resid(feols(teen_emp ~ 1 | state + year, data = mw_df))

pool_mod <- feols(teen_emp ~ min_wage, data = mw_df)
fe_mod <- feols(teen_emp ~ min_wage | state + year, cluster = 'state', data = mw_df)

pool_se <- sqrt(diag(vcov(pool_mod)))[['min_wage']]
fe_se <- sqrt(diag(vcov(fe_mod)))[['min_wage']]
fe_wr2 <- as.numeric(fitstat(fe_mod, 'wr2'))

model_compare <- tibble(
  Model = c('Truth', 'Pooled OLS', 'State + year FE'),
  `Slope on MinWage_it` = c('-1.600', fmt3(coef(pool_mod)[['min_wage']]), fmt3(coef(fe_mod)[['min_wage']])),
  `Std. error` = c('', fmt3(pool_se), fmt3(fe_se)),
  `Within R^2` = c('', '', fmt3(fe_wr2))
)

# DiD example from fixest
data(base_did)
# Ammend base_did to add separtion between trt and control
base_did = base_did %>%
  mutate(
    y = y + (post == 0) * (treat == 1) * 1.2,
    y = y + (period == 1) * (treat == 1) * 1.5
  )

did_cells <- base_did %>%
  group_by(treat, post) %>%
  summarize(mean_y = mean(y), .groups = 'drop') %>%
  mutate(
    Group = ifelse(treat == 1, 'Treated', 'Control'),
    Period = ifelse(post == 1, 'Post', 'Pre')
  )

did_table <- did_cells %>%
  select(Group, Period, mean_y) %>%
  pivot_wider(names_from = Period, values_from = mean_y) %>%
  mutate(
    Change = Post - Pre,
    across(c(Pre, Post, Change), fmt3)
  ) %>%
  arrange(Group)

control_change <- did_cells %>%
  filter(treat == 0) %>%
  summarize(change = mean_y[post == 1] - mean_y[post == 0]) %>%
  pull(change)

treated_change <- did_cells %>%
  filter(treat == 1) %>%
  summarize(change = mean_y[post == 1] - mean_y[post == 0]) %>%
  pull(change)

did_est <- treated_change - control_change

did_plot_df <- base_did %>%
  group_by(treat, period) %>%
  summarize(mean_y = mean(y), .groups = 'drop') %>%
  mutate(Group = ifelse(treat == 1, 'Treated', 'Control'))

did_mod <- feols(y ~ i(post, treat, ref = 0) | id + period, cluster = 'id', data = base_did)
did_se <- sqrt(diag(vcov(did_mod)))[['post::1:treat']]

did_reg_table <- tibble(
  Term = 'post × treat',
  Estimate = fmt3(coef(did_mod)[['post::1:treat']]),
  `Std. error` = fmt3(did_se),
  `Within R^2` = fmt3(as.numeric(fitstat(did_mod, 'wr2')))
)
```

### Last time

Instrumental variables

### Today

**Panel estimation**

- what panel data buy us,
- fixed effects,
- the difference-in-differences estimator,
- *brief extension:* the synthetic-controls estimator.

---

## {.center}

[Panel estimation]{.bigger .it .slab}

---

## Panel estimation {#panel_intro}

### Introduction

Until now, we mostly used one of two data structures:

::: {.fragment}

1. [Cross section:]{.b .orange} many units, one period,

:::

::: {.fragment}

2. [Time series:]{.b .purple} one unit, many periods.

:::

::: {.fragment}

A [panel]{.b .pink} gives us both:

:::

::: {.fragment}

- [units]{.orange} indexed by $\textcolor{#FFA500}{i}$,
- [(time) periods]{.purple} indexed by $\textcolor{#6A5ACD}{t}$,
- outcomes and regressors like $\textcolor{#e64173}{y_{\textcolor{#FFA500}{i}\textcolor{#6A5ACD}{t}}}$ and $\textcolor{#e64173}{x_{\textcolor{#FFA500}{i}\textcolor{#6A5ACD}{t}}}$.

:::

---

## Panel estimation

### Example

Consider the following panel of six observations:

```{r, example-panel-1, echo = FALSE}
example_panel |> data.table::as.data.table()
```

---

## Panel estimation

### Example

Consider the following panel of six observations:

::: columns
::: {.column width="50%"}
```{r}
#| label: "example-panel-2"
#| echo: false
#| highlight-output: [3, 5]
example_panel |> data.table::as.data.table()
```
:::
::: {.column width="50%"}
We observe the [same unit]{.orange} repeatedly.
:::
:::

We can compare unit A to itself [over time]{.b .purple}, not just A to B<br>(potentially helping with *selection bias*).

---

## Panel estimation

### Example

Consider the following panel of six observations:

::: columns
::: {.column width="50%"}
```{r}
#| label: "example-panel-3-6"
#| echo: false
#| highlight-output: "c(3"
example_panel |> data.table::as.data.table()
```
:::
::: {.column width="50%"}
Similarly, we observe the [period]{.purple} across [units]{.orange}.
:::
:::

This feature lets us control for shocks shared across [units]{.orange} at a given [time]{.purple}.

---

## Panel estimation

### The benefits of panels

Panels help solve (some) econometric problems because they allow us to

::: {.fragment}

- control for [time-invariant differences]{.b .orange} across units<br>[removes selection bias from cross-sectional differences across units]{.grey-light .it}

:::

::: {.fragment}

- implicitly study [changes]{.b .purple}<br>[not just levels]{.grey-light .it}

:::

::: {.fragment}

- [related:]{.note} evaluate policies using their [timing]{.b .purple}<br>[policy-timing is often more plausibly exogeneous than policy existence]{.grey-light .it}

:::

::: {.fragment}

Panel methods are workhorses of economics and public policy.

:::

---

## Panel estimation

### Panels are helpful—not magical

Panel data *can* be helpful for estimating causal effects,
<br>but they can still cause problems without careful use.

::: {.fragment}

- We need (sufficient) [within variation]{.it .orange} in $\textcolor{#FFA500}{x_{it}}$.<br>[I.e.,]{.it} $x_{it}$ must change meaningfully through time for some units.

:::

::: {.fragment}

- [Time-varying]{.purple} omitted variables can cause bias.

:::

::: {.fragment}

- A [unit]{.orange}'s disturbances $u_{\textcolor{#FFA500}{i}\textcolor{#6A5ACD}{t}}$ are often correlated over [time]{.purple}.

:::

::: {.fragment}

- Attrition and missing periods can matter—as can measurement error.

:::

::: {.fragment}

[(Any tool can be misused.)]{.grey-light .it}

:::

---

## Panel estimation {#pooled_problem}

### The problem with pooled OLS

Suppose you're interested in a simple linear regression model
$$y_{it} = \beta_0 + \textcolor{#e64173}{\beta_1} x_{it} + u_{it}$$
and have [panel data]{.b .slate} (with individual $i$ and time $t$).

::: {.fragment}

Let the disturbance combine (1) [unit]{.orange}-specific factor $\textcolor{#FFA500}{\alpha_i}$ and (2) error $\textcolor{#2b59c3}{v_{it}}$
$$
\begin{align}
  u_{it} = \textcolor{#FFA500}{\alpha_i} + \textcolor{#2b59c3}{v_{it}}
\end{align}
$$

:::

::: {.fragment}

If $\mathop{\text{Cov}}\left( x_{it}, \textcolor{#FFA500}{\alpha_i} \right) \neq 0$, then OLS may be biased. [(Think OVB.)]{.grey-vlight .it}

:::

::: {.fragment}

[Standard OLS estimates]{.attn}

:::

::: {.fragment}

- [between-unit differences]{.it} $\textcolor{#FFA500}{\alpha_i}$,
- [actual changes]{.it} in $x_{it}$. [(We want to isolate this last part.)]{.note}

:::

---

## Panel estimation

### Introducing *fixed effects*

[The solution:]{.note}

$$y_{it} = \beta_1 x_{it} + \textcolor{#FFA500}{\alpha_i} + \textcolor{#6A5ACD}{\lambda_t} + u_{it}$$

These [unit]{.orange}- and [period]{.purple}-specific intercepts are called [fixed effects]{.attn}.

::: {.fragment}

```{r}
#| label: "example-panel-4"
#| echo: false
# Show fixed effects
example_panel |>
  mutate(
    Unit_A = 1 * (Unit == 'A'),
    Unit_B = 1 * (Unit == 'B'),
    Year_1 = 1 * (Year == 1),
    Year_2 = 1 * (Year == 2),
    Year_3 = 1 * (Year == 3)
  ) |>
  data.table::as.data.table()
```

:::

---

## Panel estimation

### Fixed effects in action

[Q]{.qa} So what are these fixed effects (FE) doing?

::: {.fragment}

Remember the old interpretation of coefficients on binary variables?

:::

::: {.fragment}

[A]{.qa} These fixed effects *absorb*

:::

::: {.fragment}

- $\textcolor{#FFA500}{\alpha_i}$: all time-invariant [differences across units]{.orange},<br>[(including selection bias from cross-sectional differences)]{.grey-light .it}

:::

::: {.fragment}

- $\textcolor{#6A5ACD}{\lambda_t}$: all common shocks at a given [time]{.purple}.

:::

::: {.fragment}

The [(two-way) fixed-effects estimator]{.attn} (OLS that includes unit and time FEs)
<br>estimates effects using [within-unit variation]{.pink .it}—net of common time shocks.

:::

---

## Panel estimation

### The *within* transformation

For intuition, focus on unit-level fixed effects.
$$y_{it} = \beta_0 + \beta_1 x_{it} + \alpha_i + u_{it}$$

::: {.fragment}

Take the average of this equation for unit $i$ across all periods...
$$\overline{y}_i = \beta_0 + \beta_1 \overline{x}_i + \alpha_i + \overline{u}_i$$

:::

::: {.fragment}

Finally, subtract the unit mean from both sides of the original equation.
<br>(This step is what our unit-level fixed effects are doing.)
$$y_{it} - \overline{y}_i = \beta_1 \left( x_{it} - \overline{x}_i \right) + \left( u_{it} - \overline{u}_i \right)$$

:::

::: {.fragment}

[So what?]{.b .pink}

:::

::: {.fragment}

<br>[(The same logic applies to time FEs.)]{.grey-vlight .it}

:::

---

## Panel estimation

### Fixed-effects questions

Fixed effects [do not]{.b} ask questions like
<br>[FE]{.white} [Why is unit A different from unit B?]{.note}
<br>[In fact, they try to absorb cross-unit differences.]{.grey-light .it .slab}

::: {.fragment}

Fixed effects also [do not]{.b} ask questions like
<br>[FE]{.white} [When period 1 had a shock, how did all units' outcomes change?]{.note}
<br>[They try to absorb common (shared) time shocks.]{.grey-light .it .slab}

:::

::: {.fragment}

Fixed effects [do]{.b} ask questions like
<br>[FE]{.white} [When unit A changed its value of]{.note} $x_{it}$, [how did its]{.note} $y_{it}$ [change?]{.note}
<br>[This is the "within" variation.]{.grey-light .it .slab}

:::

---

## Panel estimation

### Example: State minimum wage laws

Suppose we want to estimate the effect of min. wage laws on employment
$$\text{Employment}_{it} = \beta_1 \text{(Min. Wage)}_{it} + \alpha_i + \lambda_t + u_{it}$$
with a [state]{.orange} ($\textcolor{#FFA500}{i}$) by [year]{.purple} ($\textcolor{#6A5ACD}{t}$) panel of employment and min. wage data.

::: {.fragment}

- $\alpha_i$ absorbs time-invariant, cross-state differences;
- $\lambda_t$ absorbs national shocks in year $t$.

:::

::: {.fragment}

[Q]{.qa} Should we worry that some states have higher min. wage than others?

:::

::: {.fragment}

<br>[A]{.qa} No. $\alpha_i$ absorbs those differences. The FE estimator estimates using within-state variation in min. wage.

:::

---

## Panel estimation

### Example: State minimum wage laws

Suppose we want to estimate the effect of min. wage laws on employment
$$\text{Employment}_{it} = \beta_1 \text{(Min. Wage)}_{it} + \alpha_i + \lambda_t + u_{it}$$
with a [state]{.orange} ($\textcolor{#FFA500}{i}$) by [year]{.purple} ($\textcolor{#6A5ACD}{t}$) panel of employment and min. wage data.

- $\alpha_i$ absorbs time-invariant, cross-state differences;
- $\lambda_t$ absorbs national shocks in year $t$.

[Q]{.qa} What if some *years* have lower emp. than others (recessions)?<br>[[A]{.qa} Still not worried! $\lambda_t$ absorbs *shared* shocks.]{.fragment}

---

## Panel estimation

### Example: State minimum wage laws

Suppose we want to estimate the effect of min. wage laws on employment
$$\text{Employment}_{it} = \beta_1 \text{(Min. Wage)}_{it} + \alpha_i + \lambda_t + u_{it}$$
with a [state]{.orange} ($\textcolor{#FFA500}{i}$) by [year]{.purple} ($\textcolor{#6A5ACD}{t}$) panel of employment and min. wage data.

- $\alpha_i$ absorbs time-invariant, cross-state differences;
- $\lambda_t$ absorbs national shocks in year $t$.

[Q]{.qa} So what should we worry about?<br>[[A]{.qa} We should worry about [time-varying differences between states]{.attn}.]{.fragment}

::: {.fragment}

*E.g.*, if some states have faster-growing economies and also raise their minimum wages.

:::

---

## Panel estimation

### What FE means in that example

In the minimum-wage panel, fixed effects (FEs) net out factors like

- state geography,
- long-run industry mix,
- baseline labor-market conditions,
- permanent political differences.

::: {.fragment}

$$\text{Employment}_{it} = \textcolor{#e64173}{\beta_1} \text{(Min. Wage)}_{it} + \alpha_i + \lambda_t + u_{it}$$

:::

::: {.fragment}

With FEs, $\textcolor{#e64173}{\hat{\beta}_1}$ comes from [states that change their minimum wage]{.b .pink}, relative to their own histories and common national trends.

:::

---

## {.clear}

Let's see a graphical example.<br>[(Simulated data but real FE issues.)]{.grey-vlight}

---

## {.clear}

```{r}
#| label: "fe-raw-plot1"
#| echo: false
ggplot(mw_plot_df, aes(x = min_wage, y = teen_emp, color = state)) +
  geom_line(
    stat = 'smooth', method = 'lm', formula = y ~ x, se = FALSE,
    linewidth = 1.2, color = slate
  ) +
  geom_point(size = 2.4, alpha = 0.9) +
  scale_color_viridis_d(option = 'magma', begin = .2, end = .85) +
  labs(
    x = 'State minimum wage',
    y = 'Teen employment rate',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    # legend.position = 'bottom'
  )
```

The dark line is [pooled OLS]{.b .slate}—a regression without fixed effects.
<br>
Here, states with higher min. wage laws have higher employment rates.

---

## {.clear}

```{r}
#| label: "fe-raw-plot2"
#| echo: false
ggplot(mw_plot_df, aes(x = min_wage, y = teen_emp, color = state)) +
  geom_line(
    stat = 'smooth', method = 'lm', formula = y ~ x, se = FALSE,
    linewidth = 1.2, color = slate, alpha = 0.3
  ) +
  geom_point(size = 2.4, alpha = 0.9) +
  geom_smooth(
    data = mw_plot_df,
    aes(x = min_wage, y = teen_emp, color = state),
    inherit.aes = FALSE,
    method = 'lm',
    formula = y ~ x,
    se = FALSE,
    linewidth = 1.4
  ) +
  scale_color_viridis_d(option = 'magma', begin = .2, end = .85) +
  labs(
    x = 'State minimum wage',
    y = 'Teen employment rate',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    # legend.position = 'bottom'
  )
```

However, the [within-state relationships]{.b .purple} tell another story.
<br>
The FE estimator focuses on this var.—[absorbing]{.it} cross-state differences.

---

## {.clear}

```{r}
#| label: "fe-raw-plot3"
#| echo: false
# Absorb state FEs
mw_fe_dt = mw_plot_df
data.table::setDT(mw_fe_dt)
mw_fe_dt[, `:=`(
  min_wage_fe = min_wage - mean(min_wage),
  teen_emp_fe = teen_emp - mean(teen_emp)
), by = state]
# Plot
ggplot(mw_fe_dt, aes(x = min_wage_fe, y = teen_emp_fe, color = state)) +
  geom_line(
    stat = 'smooth', method = 'lm', formula = y ~ x, se = FALSE,
    linewidth = 1.2, color = slate
  ) +
  geom_point(size = 2.4, alpha = 0.9) +
  geom_line(
    data = mw_fe_dt,
    aes(x = min_wage_fe, y = teen_emp_fe, color = state),
    stat = 'smooth',
    method = 'lm',
    formula = y ~ x,
    se = FALSE,
    linewidth = 1.4,
    alpha = .1
  ) +
  scale_color_viridis_d(option = 'magma', begin = .2, end = .85) +
  labs(
    x = 'State minimum wage',
    y = 'Teen employment rate',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    # legend.position = 'bottom'
  )
```

Integrating [fixed effects]{.b .purple}, the relationship flips.<br>States that raise their minimum wage see lower teen employment.

---

## {.clear}

```{r}
#| label: "fe-raw-plot4"
#| echo: false
ggplot(mw_plot_df, aes(x = min_wage, y = teen_emp, color = state)) +
  geom_line(
    stat = 'smooth', method = 'lm', formula = y ~ x, se = FALSE,
    linewidth = 1.2, color = slate, alpha = 0.3
  ) +
  geom_point(size = 2.4, alpha = 0.9) +
  geom_smooth(
    data = mw_plot_df,
    aes(x = min_wage, y = teen_emp, color = state),
    inherit.aes = FALSE,
    method = 'lm',
    formula = y ~ x,
    se = FALSE,
    linewidth = 1.4
  ) +
  scale_color_viridis_d(option = 'magma', begin = .2, end = .85) +
  labs(
    x = 'State minimum wage',
    y = 'Teen employment rate',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    # legend.position = 'bottom'
  )
```

This *flipped* relationship is an example of [Simpson's paradox]{.attn}—when a relationship reverses after controlling for *"group"*.

---

## Panel estimation
### `fixest` syntax

Now you know the intution, logic, and mechanics of fixed effects.

You also now understand the `fix` part of the `fixest` package.

To specify FEs: use `lm` syntax along with `|` followed by the FEs.

```{r}
#| label: "fe-code"
#| echo: true
#| eval: false
feols(
{{ teen_emp ~ min_wage | state + year, }}
  cluster = ~ state,
  data = mw_df
)
```

Here we have fixed effects for [state]{.orange} and [year]{.purple}.

---

## {.clear}

[Output]{.b .mono .pink}

```{r}
#| label: "fe-reg-table1"
#| echo: false
#| highlight-output: [1]
# Compare pooled OLS and FE estimates
est_ols =
  feols(
    teen_emp ~ min_wage,
    cluster = ~ state,
    data = mw_df
  )
est_fe =
  feols(
    teen_emp ~ min_wage | state + year,
    cluster = ~ state,
    data = mw_df
  )
etable(est_ols, est_fe, digits = 3)
```

---

## {.clear}

[Output]{.b .mono .pink}

```{r}
#| label: "fe-reg-table2"
#| echo: false
#| highlight-output: [5]
etable(est_ols, est_fe, digits = 3)
```

Both $\textcolor{#e64173}{\hat{\beta}_1}$ are stat. significant, but they tell [very different stories]{.pink}.

---

## {.clear}

[Output]{.b .mono .pink}

```{r}
#| label: "fe-reg-table3"
#| echo: false
#| highlight-output: [4]
etable(est_ols, est_fe, digits = 3)
```

[Q]{.qa} Why is there [no intercept]{.pink} $(\beta_0)$ in the FE model?<br>[[A]{.qa} The [fixed effects are collinear]{.pink} with the intercept. They each provide a separate intercept for each unit and time period.]{.fragment}

---

## {.clear}

[Output]{.b .mono .pink}

```{r}
#| label: "fe-reg-table4"
#| echo: false
#| highlight-output: [13]
etable(est_ols, est_fe, digits = 3)
```

[Q]{.qa} What is [*within* $R^2$]{.b .pink}?<br>[[A]{.qa} It is the $R^2$ of the regression [after accounting for the FEs]{.pink}: how much of the outcome's variation regressors explain, *after controlling for FEs*.]{.fragment}

---

## {.clear}

[Output]{.b .mono .pink}

```{r}
#| label: "fe-reg-table5"
#| echo: false
#| highlight-output: [12]
etable(est_ols, est_fe, digits = 3)
```

The FE model has as high $\textcolor{#e64173}{R^2}$ because fixed effects explain a lot of variation (in essence, means of states and years).

---

## Panel estimation
### Summary

Fixed effects help one key problem: [time-invariant omitted variables]{.b .pink}.

::: {.fragment}

But FEs [do not]{.b .slate} solve *all* causal problems, *e.g.*,

:::

::: {.fragment}

- time-varying confounders,
- reverse causality
- measurement error,
- poor measurement of treatment timing,
- weak within variation.

:::

::: {.fragment}

As with *all* econometric tools, careful use is required.

:::

---

## Difference-in-differences {#did_intuition}

### Core idea

[Difference-in-differences]{.attn}

^[[†]{.pink} Also called *double-differences* or *diff-in-diff*. Increasingly combined with *two-way fixed effects (TWFE)*.]

DiD compares observations across two dimensions:

1. [Group:]{.b .purple} comparing
  - [treated]{.purple .it} individuals (potentially affected by the policy)
  - [control]{.purple .it} individuals (unaffected by the policy)
2. [Time:]{.b .orange} comparing
  - [before]{.orange .it} the policy takes effect [(pre-treatment)]{.grey-vlight .it}
  - [after]{.orange .it} the policy takes effect [(post-treatment)]{.grey-vlight .it}

---

## Difference-in-differences

### Intuition

DiD functions by taking the difference between two differences:

::: {.fragment}

1. Δ[Y̅[Pre]{.sub .orange .tran50}]{.b} $=$ [Y̅]{.b}([Trt]{.purple}; [Pre]{.orange .tran50}) $-$ [Y̅]{.b}([Ctrl]{.purple .tran50}; [Pre]{.orange .tran50}) <br> the average difference between [trt.]{.purple} and [control]{.purple .tran50} [before]{.b .orange .tran50} the policy, <br> accounting for pre-existing differences between the groups;

:::

::: {.fragment}

2. Δ[Y̅[Post]{.sub .orange}]{.b} $=$ [Y̅]{.b}([Trt]{.purple}; [Post]{.orange}) $-$ [Y̅]{.b}([Ctrl]{.purple .tran50}; [Post]{.orange}) <br> the average difference between [trt.]{.purple} and [control]{.purple .tran50} [after]{.b .orange} the policy, <br> which includes pre-existing differences *and* the [treatment effect]{.pink}.

:::

::: {.fragment}

The [DiD estimate]{.attn} is the difference between these differences
<br>

:::

::: {.fragment}

[τ̂]{.pink}

:::

::: {.fragment}

<br>

:::

::: {.fragment}

[I.e.,]{.note}

:::

---

## {.clear}

[Ex.]{.mono .b .slate}

```{r}
#| label: "did-cell-table1"
#| echo: false
# Copy the did_cells to data.table
did_dt = copy(did_cells)
setDT(did_dt)
# Arrange by period and group
setorder(did_dt, -Period, -Group)
# Set column order
setcolorder(
  did_dt,
  c('Group', 'Period', 'mean_y')
)
# Set names
setnames(did_dt, c('Group', 'Period', 'Mean', 'trt', 'post'))
# Simplify means
did_dt[, Mean := Mean |> round(3)]
# Show the table
did_dt[, 1:3]
```

---

## {.clear}

[Ex.]{.mono .b .slate}

```{r}
#| label: "did-cell-table2"
#| echo: false
#| highlight-output: [3, 4]
did_dt[, 1:3]
```

1. Δ[Y̅[Pre]{.sub .orange .tran50}]{.b} $=$ [Y̅]{.b}([Trt]{.purple}; [Pre]{.orange .tran50}) $-$ [Y̅]{.b}([Ctrl]{.purple .tran50}; [Pre]{.orange .tran50}) $=$ `r did_dt$Mean[2:1] |> diff()`.

---

## {.clear}

[Ex.]{.mono .b .slate}

```{r}
#| label: "did-cell-table3"
#| echo: false
#| highlight-output: [5, 6]
did_dt[, 1:3]
```

1. Δ[Y̅[Pre]{.sub .orange .tran50}]{.b} $=$ [Y̅]{.b}([Trt]{.purple}; [Pre]{.orange .tran50}) $-$ [Y̅]{.b}([Ctrl]{.purple .tran50}; [Pre]{.orange .tran50}) $=$ `r did_dt$Mean[2:1] |> diff()`.

2. Δ[Y̅[Post]{.sub .orange}]{.b} $=$ [Y̅]{.b}([Trt]{.purple}; [Post]{.orange}) $-$ [Y̅]{.b}([Ctrl]{.purple .tran50}; [Post]{.orange}) $=$ `r did_dt$Mean[4:3] |> diff()`

---

## {.clear}

[Ex.]{.mono .b .slate}

```{r}
#| label: "did-cell-table4"
#| echo: false
did_dt[, 1:3]
```

1. Δ[Y̅[Pre]{.sub .orange .tran50}]{.b} $=$ [Y̅]{.b}([Trt]{.purple}; [Pre]{.orange .tran50}) $-$ [Y̅]{.b}([Ctrl]{.purple .tran50}; [Pre]{.orange .tran50}) $=$ `r did_dt$Mean[2:1] |> diff()`.

2. Δ[Y̅[Post]{.sub .orange}]{.b} $=$ [Y̅]{.b}([Trt]{.purple}; [Post]{.orange}) $-$ [Y̅]{.b}([Ctrl]{.purple .tran50}; [Post]{.orange}) $=$ `r did_dt$Mean[4:3] |> diff()`

The [DiD estimate]{.attn}—the difference between these differences
<br>

[τ̂]{.pink}

---

## {.clear}

[Note]{.note}

[τ̂]{.pink}[[τ̂]{.white}]{.fragment}

::: {.fragment}

$-$
$\big[$ [Y̅]{.b}([Trt]{.purple}; [Pre]{.orange .tran50}) $-$ [Y̅]{.b}([Ctrl]{.purple .tran50}; [Pre]{.orange .tran50}) $\big]$

:::

::: {.fragment}

[τ̂]{.white}

:::

::: {.fragment}

[τ̂]{.white}

:::

::: {.fragment}

[τ̂]{.white}

:::

::: {.fragment}

[I.e.,]{.note}

:::

::: {.fragment}

We're using the [control group's [change]{.b}]{.purple .tran50} as a [counterfactual]{.b} for <br> the [treated group's [change]{.b}]{.purple} had it not received treatment.

:::

---

## Difference-in-differences

### Classic example: Card and Krueger (1994)

One famous DiD [application](https://davidcard.berkeley.edu/papers/njmin-aer.pdf) estimates the [impact of increasing the minimum wage on fast-food employment]{.b .slate}.

- [Who:]{.note} [New Jersey (trt)]{.purple} *vs.* [Pennsylvania (ctrl)]{.purple .tran50}
- [When:]{.note} [before 01 April 1992 (pre)]{.orange .tran50} *vs.* [after 01 April 1992 (post)]{.orange}

DiD estimates [(both questions = same answer)]{.grey-vlight .it}

1. How did NJ-PA empl. diff. change after the policy, relative to before?
2. How did NJ's empl. change after the policy relative to PA's change?

::: {.fragment}

[Key assumption:]{.note}

:::

---

## Difference-in-differences

### DiD as a regression

With two groups and two periods
$$y_{it} = \beta_0 + \beta_1 \textcolor{#6A5ACD}{\text{Trt}_i} + \beta_2 \textcolor{#FFA500}{\text{Post}_t} + \delta \left( \textcolor{#6A5ACD}{\text{Trt}_i} \times \textcolor{#FFA500}{\text{Post}_t} \right) + u_{it}$$
where $\textcolor{#6A5ACD}{\text{Trt}_i}$ and $\textcolor{#FFA500}{\text{Post}_t}$ are binary indicators for the [treated group]{.purple} and [post-treatment period]{.orange}.

The interaction asks if if the [post]{.orange} change in the [treated group]{.purple} differs from the [post]{.orange} change in the [control group]{.purple .tran50}: the [treatment effect]{.attn}.

::: {.fragment}

With panel data, we often write

:::

::: {.fragment}

$$y_{it} = \alpha_i + \lambda_t + \delta D_{it} + u_{it}$$

:::

::: {.fragment}

where $D_{it}=1$ only for treated units after treatment starts.

:::

---

## {.clear}

```{r}
#| label: "did-plot1"
#| echo: false
# Add post indicator
did_plot_dt = copy(did_plot_df)
setDT(did_plot_dt)
did_plot_dt[, post := 1 * (period > 5)]
# Add factor var for interaction
did_plot_dt[, int := interaction(treat, post)]
# The first plot
ggplot(did_plot_dt, aes(x = period, y = mean_y, color = Group)) +
  geom_vline(xintercept = 5.5, linetype = 'dashed', color = grey_mid) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2.7) +
  scale_color_manual(values = c('Control' = purple_l, 'Treated' = purple)) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = 'Time period',
    y = 'Average Y',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = 'bottom'
  )
```

[DiD in pictures]{.b .slate}

<br>

[(Treatment starts between periods 5 and 6.)]{.grey-vlight .it}

---

## {.clear}

```{r}
#| label: "did-plot2"
#| echo: false
ggplot(did_plot_dt, aes(x = period, y = mean_y, color = Group)) +
  geom_vline(xintercept = 5.5, linetype = 'dashed', color = grey_mid) +
  geom_line(linewidth = 1.3, alpha = .15) +
  geom_point(size = 2.7, alpha = .15) +
  geom_line(
    aes(group = int),
    stat = 'smooth', method = 'lm', formula = y ~ 1, se = FALSE,
    linewidth = 1.3
  ) +
  scale_color_manual(values = c('Control' = purple_l, 'Treated' = purple)) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = 'Time period',
    y = 'Average Y',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = 'bottom'
  )
```

Here we have [group]{.b .purple}-by-[time]{.b .orange} means.

---

## {.clear}

```{r}
#| label: "did-plot3"
#| echo: false
ggplot(did_plot_dt, aes(x = period, y = mean_y, color = Group)) +
  geom_vline(xintercept = 5.5, linetype = 'dashed', color = grey_mid) +
  geom_line(linewidth = 1.3, alpha = .15) +
  geom_point(size = 2.7, alpha = .15) +
  geom_line(
    aes(group = int),
    stat = 'smooth', method = 'lm', formula = y ~ 1, se = FALSE,
    linewidth = 1.3
  ) +
  annotate(
    'segment',
    x = 3, xend = 3,
    y = did_plot_dt[post == 0 & treat == 0, mean(mean_y)],
    yend = did_plot_dt[post == 0 & treat == 1, mean(mean_y)],
    color = orange_l, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  annotate(
    'segment',
    x = 8, xend = 8,
    y = did_plot_dt[post == 1 & treat == 0, mean(mean_y)],
    yend = did_plot_dt[post == 1 & treat == 1, mean(mean_y)],
    color = orange, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  scale_color_manual(values = c('Control' = purple_l, 'Treated' = purple)) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = 'Time period',
    y = 'Average Y',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = 'bottom'
  )
```

The [DiD estimate]{.attn} compares the [pre-trt]{.b .orange .tran50} difference to the [post-trt]{.b .orange} diff.

---

## {.clear}

```{r}
#| label: "did-plot4"
#| echo: false
# Calculate pre- and post-trt diffs
pre_diff =
  did_plot_dt[post == 0 & treat == 1, mean(mean_y)] -
  did_plot_dt[post == 0 & treat == 0, mean(mean_y)]
post_diff =
  did_plot_dt[post == 1 & treat == 1, mean(mean_y)] -
  did_plot_dt[post == 1 & treat == 0, mean(mean_y)]
# The plot
ggplot(did_plot_dt, aes(x = period, y = mean_y, color = Group)) +
  geom_vline(xintercept = 5.5, linetype = 'dashed', color = grey_mid) +
  geom_line(linewidth = 1.3, alpha = .15) +
  geom_point(size = 2.7, alpha = .15) +
  geom_line(
    aes(group = int),
    stat = 'smooth', method = 'lm', formula = y ~ 1, se = FALSE,
    linewidth = 1.3
  ) +
  annotate(
    'segment',
    x = 3, xend = 3,
    y = did_plot_dt[post == 0 & treat == 0, mean(mean_y)],
    yend = did_plot_dt[post == 0 & treat == 1, mean(mean_y)],
    color = orange_l, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  annotate(
    'segment',
    x = 7.5, xend = 7.5,
    y = did_plot_dt[post == 1 & treat == 0, mean(mean_y)],
    yend = did_plot_dt[post == 1 & treat == 0, mean(mean_y)] + pre_diff,
    color = orange_l, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  annotate(
    'segment',
    x = 8, xend = 8,
    y = did_plot_dt[post == 1 & treat == 0, mean(mean_y)],
    yend = did_plot_dt[post == 1 & treat == 1, mean(mean_y)],
    color = orange, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  annotate(
    'segment',
    x = 7.5, xend = 7.5,
    y = did_plot_dt[post == 1 & treat == 0, mean(mean_y)] + pre_diff,
    yend = did_plot_dt[post == 1 & treat == 1, mean(mean_y)],
    color = red_pink, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  scale_color_manual(values = c('Control' = purple_l, 'Treated' = purple)) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = 'Time period',
    y = 'Average Y',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = 'bottom'
  )
```

The difference in the [pre-trt]{.b .orange .tran50} and [post-trt]{.b .orange} diffs is the [DiD estimate]{.attn}.

---

## {.clear}

[DiD in `fixest`]{.b}

```{r}
#| label: "did-code"
#| echo: true
did_mod =
  feols(
    y ~ i(post, treat, ref = 0) | id + period,
    cluster = 'id',
    data = base_did
  )
```

::: {.fragment}

```{r}
#| label: "did-reg-table"
#| echo: false
#| highlight-output: [6]
did_mod
```

:::

::: {.fragment}

The interaction btn [post]{.orange} and [treat]{.purple} is the [DiD estimate]{.attn} of the trt effect.

:::

---

## Difference-in-differences

### The key DiD assumption

DiD relies upon a critical assumption often called [parallel trends]{.attn}:

> In the absence of treatment, treated and control units would have followed the [same average trend]{.pink}.

Again, we're assuming that the trend of the control units provide <br> a [good counterfactual]{.b .slate} for the treated group.

---

## Difference-in-differences

### What threatens DiD?

So what violates this [parallel trends]{.note} assumption?

::: {.fragment}

- treated units were already on a different trend,
- anticipation changes behavior before treatment,
- other shocks hit treated units at the same time as treatment.

:::

::: {.fragment}

What else can *mess up* DiD estimates?

:::

::: {.fragment}

- standard errors ignore within-unit dependence,
- measurement error,
- spillovers from trt to control,
- dynamic treatment effects,
- het trt effects *and* het trt timing...

:::

---

## Difference-in-differences

### DiD's place in the toolbox

DiD is especially useful when

- treatment [timing]{.orange} is clear,
- [treatment definition]{.purple} is clear,
- you have good [pre-treatment]{.orange} data
- [ctrl group]{.purple .tran50} provide a plausible counterfactual for [trt group]{.purple},
- the [treatment effect]{.pink} is tied to a discrete policy or event.

::: {.fragment}

[Q]{.qa} What are the other options?

:::

::: {.fragment}

<br>[A]{.qa} Synthetic control and friends build a weighted average of control units to better match the treated unit's pre-treatment path.

:::

---

## Difference-in-differences

### Takeaways

Key concepts from panel estimation, FEs, and DiD:

- [Panel data]{.b .slate} let us compare a unit to itself [over time]{.b .purple}.

- [Fixed effects]{.b .slate} remove [time-invariant omitted variables]{.b .orange}.

- $\implies$ [Identifying variation]{.b .slate} is [within variation]{.b .purple}.

- [DiD]{.b .slate} focuses on how trt-ctrl [differences]{.b .slate} change over time.

Panels can certainly help, but they still depend on strong assumptions.
