---
title: "Panel Estimation"
subtitle: "EC 421, Set 12"
author: "Edward Rubin"
date: ""
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, middle

```{r, setup, include = FALSE}
options(htmltools.dir.version = FALSE)
library(pacman)
p_load(
  data.table, broom, dplyr, fixest, ggplot2, kableExtra, knitr,
  magrittr, scales, tidyr
)

# Define colors
red_pink = '#e64173'
turquoise = '#20B2AA'
orange = '#FFA500'
orange_l = '#f8d194'
red = '#fb6107'
blue = '#2b59c3'
green = '#8bb174'
grey_light = 'grey70'
grey_mid = 'grey50'
grey_dark = 'grey20'
purple = '#6A5ACD'
purple_l = '#c5bee9'
slate = '#314f4f'

# Knitr options
opts_chunk$set(
  comment = '#>',
  fig.align = 'center',
  fig.height = 7,
  fig.width = 10.5,
  warning = FALSE,
  message = FALSE
)
opts_chunk$set(dev = 'svg')
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(crayon.enabled = FALSE)
options(knitr.table.format = 'html')

table_style <- function(x, digits = 2, font_size = 22) {
  align <- paste(rep('c', ncol(x)), collapse = '')
  x %>%
    kable(escape = FALSE, digits = digits, align = align) %>%
    kable_styling(font_size = font_size, full_width = FALSE)
}

fmt3 <- function(x) sprintf('%.3f', x)

theme_set(theme_minimal(base_size = 20))

set.seed(421)

# Tiny example panel for early slides
example_panel <- tribble(
  ~Unit, ~Year, ~`y_it`, ~`x_it`,
  'A', 1, 10, 2,
  'A', 2, 11, 3,
  'A', 3, 13, 4,
  'B', 1, 7, 1,
  'B', 2, 8, 1,
  'B', 3, 9, 2
)

# Simulated state panel for minimum-wage example
states <- state.abb[1:40]
years <- 2014:2021
alpha_state <- rnorm(length(states), sd = 3)
policy_pref <- 0.9 * alpha_state + rnorm(length(states), sd = 1)
names(alpha_state) <- states
names(policy_pref) <- states
lambda_t <- c(-1.0, -0.5, -0.2, 0.0, 0.3, 0.6, 0.9, 1.1)

mw_df <- expand_grid(state = states, year = years) %>%
  mutate(
    alpha = alpha_state[state],
    policy = policy_pref[state],
    post = year >= 2018,
    min_wage = 7.25 + 0.35 * policy + 0.18 * (year - min(years)) +
      if_else(policy > median(policy) & post, 1.1, 0) + rnorm(n(), sd = 0.18),
    teen_emp = 58 - 1.6 * min_wage + 2.7 * alpha + lambda_t[match(year, years)] +
      rnorm(n(), sd = 1.4)
  )

plot_ids <- mw_df %>%
  distinct(state, alpha) %>%
  arrange(alpha) %>%
  slice(c(3, 10, 17, 24, 31, 38)) %>%
  pull(state)

mw_plot_df <- mw_df %>%
  filter(state %in% plot_ids) %>%
  mutate(state = factor(state, levels = plot_ids))

mw_df$mw_tw <- resid(feols(min_wage ~ 1 | state + year, data = mw_df))
mw_df$emp_tw <- resid(feols(teen_emp ~ 1 | state + year, data = mw_df))

pool_mod <- feols(teen_emp ~ min_wage, data = mw_df)
fe_mod <- feols(teen_emp ~ min_wage | state + year, cluster = 'state', data = mw_df)

pool_se <- sqrt(diag(vcov(pool_mod)))[['min_wage']]
fe_se <- sqrt(diag(vcov(fe_mod)))[['min_wage']]
fe_wr2 <- as.numeric(fitstat(fe_mod, 'wr2'))

model_compare <- tibble(
  Model = c('Truth', 'Pooled OLS', 'State + year FE'),
  `Slope on MinWage_it` = c('-1.600', fmt3(coef(pool_mod)[['min_wage']]), fmt3(coef(fe_mod)[['min_wage']])),
  `Std. error` = c('', fmt3(pool_se), fmt3(fe_se)),
  `Within R^2` = c('', '', fmt3(fe_wr2))
)

# DiD example from fixest
data(base_did)
# Ammend base_did to add separtion between trt and control
base_did = base_did %>%
  mutate(
    y = y + (post == 0) * (treat == 1) * 1.2,
    y = y + (period == 1) * (treat == 1) * 1.5
  )

did_cells <- base_did %>%
  group_by(treat, post) %>%
  summarize(mean_y = mean(y), .groups = 'drop') %>%
  mutate(
    Group = ifelse(treat == 1, 'Treated', 'Control'),
    Period = ifelse(post == 1, 'Post', 'Pre')
  )

did_table <- did_cells %>%
  select(Group, Period, mean_y) %>%
  pivot_wider(names_from = Period, values_from = mean_y) %>%
  mutate(
    Change = Post - Pre,
    across(c(Pre, Post, Change), fmt3)
  ) %>%
  arrange(Group)

control_change <- did_cells %>%
  filter(treat == 0) %>%
  summarize(change = mean_y[post == 1] - mean_y[post == 0]) %>%
  pull(change)

treated_change <- did_cells %>%
  filter(treat == 1) %>%
  summarize(change = mean_y[post == 1] - mean_y[post == 0]) %>%
  pull(change)

did_est <- treated_change - control_change

did_plot_df <- base_did %>%
  group_by(treat, period) %>%
  summarize(mean_y = mean(y), .groups = 'drop') %>%
  mutate(Group = ifelse(treat == 1, 'Treated', 'Control'))

did_mod <- feols(y ~ i(post, treat, ref = 0) | id + period, cluster = 'id', data = base_did)
did_se <- sqrt(diag(vcov(did_mod)))[['post::1:treat']]

did_reg_table <- tibble(
  Term = 'post × treat',
  Estimate = fmt3(coef(did_mod)[['post::1:treat']]),
  `Std. error` = fmt3(did_se),
  `Within R^2` = fmt3(as.numeric(fitstat(did_mod, 'wr2')))
)
```

# Prologue

---
name: schedule

# Schedule

## Last time

Instrumental variables

## Today

**Panel estimation**

- what panel data buy us,
- fixed effects,
- the difference-in-differences estimator,
- *brief extension:* the synthetic-controls estimator.

---
layout: false
class: inverse, middle

# Panel estimation

---
layout: true

# Panel estimation

---
name: panel_intro

## Introduction

Until now, we mostly used one of two data structures:

--

1. .hi-orange[Cross section:] many units, one period,
--

2. .hi-purple[Time series:] one unit, many periods.

--

A .hi[panel] gives us both:

- .orange[units] indexed by $\color{#FFA500}{i}$,
- .purple[(time) periods] indexed by $\color{#6A5ACD}{t}$,
- outcomes and regressors like $\color{#e64173}{y_{\color{#FFA500}{i}\color{#6A5ACD}{t}}}$ and $\color{#e64173}{x_{\color{#FFA500}{i}\color{#6A5ACD}{t}}}$.

---
## Example

Consider the following panel of six observations:

.pull-left[
```{r, example-panel-1, echo = FALSE}
example_panel |> data.table::as.data.table()
```
]

---
## Example

Consider the following panel of six observations:

.pull-left[
```{r, example-panel-2, echo = FALSE, highlight.output = 3:5}
example_panel |> data.table::as.data.table()
```
]
.pull-right[
We observe the .orange[same unit] repeatedly.
]

We can compare unit A to itself .hi-purple[over time], not just A to B<br>(potentially helping with *selection bias*).

---
## Example

Consider the following panel of six observations:

.pull-left[
```{r, example-panel-3, echo = FALSE, highlight.output = c(3, 6)}
example_panel |> data.table::as.data.table()
```
]
.pull-right[
Similarly, we observe the .purple[period] across .orange[units].
]

This feature lets us control for shocks shared across .orange[units] at a given .purple[time].

---

## The benefits of panels

Panels help solve (some) econometric problems because they allow us to

--

- control for .b.orange[time-invariant differences] across units<br>.grey-light.it[removes selection bias from cross-sectional differences across units]
  
--

- implicitly study .b.purple[changes]<br>.grey-light.it[not just levels]

--

- .note[related:] evaluate policies using their .b.purple[timing]<br>.grey-light.it[policy-timing is often more plausibly exogeneous than policy existence]

--

Panel methods are workhorses of economics and public policy.

---

## Panels are helpful—not magical

Panel data *can* be helpful for estimating causal effects,
<br>but they can still cause problems without careful use.

--

- We need (sufficient) .it.orange[within variation] in $\color{#FFA500}{x_{it}}$.<br>.it[I.e.,] $x_{it}$ must change meaningfully through time for some units.

- .purple[Time-varying] omitted variables can cause bias.

- A .orange[unit]'s disturbances $u_{\color{#FFA500}{i}\color{#6A5ACD}{t}}$ are often correlated over .purple[time].

- Attrition and missing periods can matter—as can measurement error.

.grey-light.it[(Any tool can be misused.)]

---
name: pooled_problem
## The problem with pooled OLS

Suppose you're interested in a simple linear regression model
$$y_{it} = \beta_0 + \color{#e64173}{\beta_1} x_{it} + u_{it}$$
and have .b.slate[panel data] (with individual $i$ and time $t$).

--

Let the disturbance combine (1) .orange[unit]-specific factor $\color{#FFA500}{\alpha_i}$ and (2) error $\color{#2b59c3}{v_{it}}$
$$
\begin{align}
  u_{it} = \color{#FFA500}{\alpha_i} + \color{#2b59c3}{v_{it}}
\end{align}
$$

--

If $\mathop{\text{Cov}}\left( x_{it}, \color{#FFA500}{\alpha_i} \right) \neq 0$, then OLS may be biased. .grey-vlight.it[(Think OVB.)]

--

.attn[Standard OLS estimates] of $\color{#e64173}{\beta_1}$ (*i.e.*, without FEs) will reflect both

- .it[between-unit differences] $\color{#FFA500}{\alpha_i}$,
- .it[actual changes] in $x_{it}$. .note[(We want to isolate this last part.)]
  
---

## Introducing *fixed effects*

.note[The solution:] Let each .orange[unit] have its own intercept $\color{#FFA500}{\alpha_i}$,<br> and each .purple[time period] have its own intercept $\color{#6A5ACD}{\lambda_t}$:
$$y_{it} = \beta_1 x_{it} + \color{#FFA500}{\alpha_i} + \color{#6A5ACD}{\lambda_t} + u_{it}$$

These .orange[unit]- and .purple[period]-specific intercepts are called .attn[fixed effects].

--

```{r, example-panel-4, echo = FALSE}
# Show fixed effects
example_panel |>
  mutate(
    Unit_A = 1 * (Unit == 'A'),
    Unit_B = 1 * (Unit == 'B'),
    Year_1 = 1 * (Year == 1),
    Year_2 = 1 * (Year == 2),
    Year_3 = 1 * (Year == 3)
  ) |>
  data.table::as.data.table()
```

---
## Fixed effects in action

.qa[Q] So what are these fixed effects (FE) doing?

--

Remember the old interpretation of coefficients on binary variables?

--

.qa[A] These fixed effects *absorb*

- $\color{#FFA500}{\alpha_i}$: all time-invariant .orange[differences across units],<br>.grey-light.it[(including selection bias from cross-sectional differences)]

- $\color{#6A5ACD}{\lambda_t}$: all common shocks at a given .purple[time].

--

The .attn[(two-way) fixed-effects estimator] (OLS that includes unit and time FEs)
<br>estimates effects using .pink.it[within-unit variation]—net of common time shocks.

---

## The *within* transformation

For intuition, focus on unit-level fixed effects.
$$y_{it} = \beta_0 + \beta_1 x_{it} + \alpha_i + u_{it}$$

--

Take the average of this equation for unit $i$ across all periods...
$$\overline{y}_i = \beta_0 + \beta_1 \overline{x}_i + \alpha_i + \overline{u}_i$$

--

Finally, subtract the unit mean from both sides of the original equation.
<br>(This step is what our unit-level fixed effects are doing.)
$$y_{it} - \overline{y}_i = \beta_1 \left( x_{it} - \overline{x}_i \right) + \left( u_{it} - \overline{u}_i \right)$$

--

.hi[So what?] Unit-specific FEs identify $\beta_1$ using .attn[within-unit variation] in $x_{it}$.
<br>.grey-vlight.it[(The same logic applies to time FEs.)]

---
## Fixed-effects questions

Fixed effects .b[do not] ask questions like
<br>.white[FE] .note[Why is unit A different from unit B?]
<br>.grey-light.it.slab[In fact, they try to absorb cross-unit differences.]

--

Fixed effects also .b[do not] ask questions like
<br>.white[FE] .note[When period 1 had a shock, how did all units' outcomes change?]
<br>.grey-light.it.slab[They try to absorb common (shared) time shocks.]

--

Fixed effects .b[do] ask questions like
<br>.white[FE] .note[When unit A changed its value of] $x_{it}$, .note[how did its] $y_{it}$ .note[change?]
<br>.grey-light.it.slab[This is the "within" variation.]

---
## Example: State minimum wage laws

Suppose we want to estimate the effect of min. wage laws on employment
$$\text{Employment}_{it} = \beta_1 \text{(Min. Wage)}_{it} + \alpha_i + \lambda_t + u_{it}$$
with a .orange[state] ($\color{#FFA500}{i}$) by .purple[year] ($\color{#6A5ACD}{t}$) panel of employment and min. wage data.

--

- $\alpha_i$ absorbs time-invariant, cross-state differences;
- $\lambda_t$ absorbs national shocks in year $t$.

--

.qa[Q] Should we worry that some states have higher min. wage than others?
--
<br>.qa[A] No. $\alpha_i$ absorbs those differences. The FE estimator estimates using within-state variation in min. wage.

---
## Example: State minimum wage laws

Suppose we want to estimate the effect of min. wage laws on employment
$$\text{Employment}_{it} = \beta_1 \text{(Min. Wage)}_{it} + \alpha_i + \lambda_t + u_{it}$$
with a .orange[state] ($\color{#FFA500}{i}$) by .purple[year] ($\color{#6A5ACD}{t}$) panel of employment and min. wage data.

- $\alpha_i$ absorbs time-invariant, cross-state differences;
- $\lambda_t$ absorbs national shocks in year $t$.


.qa[Q] What if some *years* have lower emp. than others (recessions)?
--
<br>.qa[A] Still not worried! $\lambda_t$ absorbs *shared* shocks.

---
## Example: State minimum wage laws

Suppose we want to estimate the effect of min. wage laws on employment
$$\text{Employment}_{it} = \beta_1 \text{(Min. Wage)}_{it} + \alpha_i + \lambda_t + u_{it}$$
with a .orange[state] ($\color{#FFA500}{i}$) by .purple[year] ($\color{#6A5ACD}{t}$) panel of employment and min. wage data.

- $\alpha_i$ absorbs time-invariant, cross-state differences;
- $\lambda_t$ absorbs national shocks in year $t$.


.qa[Q] So what should we worry about?
--
<br>.qa[A] We should worry about .attn[time-varying differences between states].

*E.g.*, if some states have faster-growing economies and also raise their minimum wages.

---

## What FE means in that example

In the minimum-wage panel, fixed effects (FEs) net out factors like

- state geography,
- long-run industry mix,
- baseline labor-market conditions,
- permanent political differences.

--

$$\text{Employment}_{it} = \color{#e64173}{\beta_1} \text{(Min. Wage)}_{it} + \alpha_i + \lambda_t + u_{it}$$

With FEs, $\color{#e64173}{\hat{\beta}_1}$ comes from .b.pink[states that change their minimum wage], relative to their own histories and common national trends.

---
layout: false
class: clear, middle

Let's see a graphical example.<br>.grey-vlight[(Simulated data but real FE issues.)]

---
layout: false
class: clear, middle

```{r, fe-raw-plot1, echo = FALSE}
ggplot(mw_plot_df, aes(x = min_wage, y = teen_emp, color = state)) +
  geom_line(
    stat = 'smooth', method = 'lm', formula = y ~ x, se = FALSE,
    linewidth = 1.2, color = slate
  ) +
  geom_point(size = 2.4, alpha = 0.9) +
  scale_color_viridis_d(option = 'magma', begin = .2, end = .85) +
  labs(
    x = 'State minimum wage',
    y = 'Teen employment rate',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    # legend.position = 'bottom'
  )
```

The dark line is .b.slate[pooled OLS]—a regression without fixed effects.
<br>
Here, states with higher min. wage laws have higher employment rates.

---
layout: false
class: clear, middle

```{r, fe-raw-plot2, echo = FALSE}
ggplot(mw_plot_df, aes(x = min_wage, y = teen_emp, color = state)) +
  geom_line(
    stat = 'smooth', method = 'lm', formula = y ~ x, se = FALSE,
    linewidth = 1.2, color = slate, alpha = 0.3
  ) +
  geom_point(size = 2.4, alpha = 0.9) +
  geom_smooth(
    data = mw_plot_df,
    aes(x = min_wage, y = teen_emp, color = state),
    inherit.aes = FALSE,
    method = 'lm',
    formula = y ~ x,
    se = FALSE,
    linewidth = 1.4
  ) +
  scale_color_viridis_d(option = 'magma', begin = .2, end = .85) +
  labs(
    x = 'State minimum wage',
    y = 'Teen employment rate',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    # legend.position = 'bottom'
  )
```

However, the .b.purple[within-state relationships] tell another story.
<br>
The FE estimator focuses on this var.—.it[absorbing] cross-state differences.

---
layout: false
class: clear, middle

```{r, fe-raw-plot3, echo = FALSE}
# Absorb state FEs
mw_fe_dt = mw_plot_df
data.table::setDT(mw_fe_dt)
mw_fe_dt[, `:=`(
  min_wage_fe = min_wage - mean(min_wage),
  teen_emp_fe = teen_emp - mean(teen_emp)
), by = state]
# Plot
ggplot(mw_fe_dt, aes(x = min_wage_fe, y = teen_emp_fe, color = state)) +
  geom_line(
    stat = 'smooth', method = 'lm', formula = y ~ x, se = FALSE,
    linewidth = 1.2, color = slate
  ) +
  geom_point(size = 2.4, alpha = 0.9) +
  geom_line(
    data = mw_fe_dt,
    aes(x = min_wage_fe, y = teen_emp_fe, color = state),
    stat = 'smooth',
    method = 'lm',
    formula = y ~ x,
    se = FALSE,
    linewidth = 1.4,
    alpha = .1
  ) +
  scale_color_viridis_d(option = 'magma', begin = .2, end = .85) +
  labs(
    x = 'State minimum wage',
    y = 'Teen employment rate',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    # legend.position = 'bottom'
  )
```

Integrating .b.purple[fixed effects], the relationship flips.<br>States that raise their minimum wage see lower teen employment.

---
layout: false
class: clear, middle

```{r, fe-raw-plot4, echo = FALSE}
ggplot(mw_plot_df, aes(x = min_wage, y = teen_emp, color = state)) +
  geom_line(
    stat = 'smooth', method = 'lm', formula = y ~ x, se = FALSE,
    linewidth = 1.2, color = slate, alpha = 0.3
  ) +
  geom_point(size = 2.4, alpha = 0.9) +
  geom_smooth(
    data = mw_plot_df,
    aes(x = min_wage, y = teen_emp, color = state),
    inherit.aes = FALSE,
    method = 'lm',
    formula = y ~ x,
    se = FALSE,
    linewidth = 1.4
  ) +
  scale_color_viridis_d(option = 'magma', begin = .2, end = .85) +
  labs(
    x = 'State minimum wage',
    y = 'Teen employment rate',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    # legend.position = 'bottom'
  )
```

This *flipped* relationship is an example of .attn[Simpson's paradox]—when a relationship reverses after controlling for *"group"*.

---
# Panel estimation
## `fixest` syntax

Now you know the intution, logic, and mechanics of fixed effects.

You also now understand the `fix` part of the `fixest` package.

To specify FEs: use `lm` syntax along with `|` followed by the FEs.

```{r, fe-code, eval = FALSE}
feols(
{{ teen_emp ~ min_wage | state + year, }}
  cluster = ~ state,
  data = mw_df
)
```
Here we have fixed effects for .orange[state] and .purple[year].

---
layout: false
class: clear

.b.mono.pink[Output] `feols` output for the *plain* OLS model .it[vs.] the FE model.

```{r, fe-reg-table1, echo = FALSE, highlight.output = 1}
# Compare pooled OLS and FE estimates
est_ols =
  feols(
    teen_emp ~ min_wage,
    cluster = ~ state,
    data = mw_df
  )
est_fe =
  feols(
    teen_emp ~ min_wage | state + year,
    cluster = ~ state,
    data = mw_df
  )
etable(est_ols, est_fe, digits = 3)
```

---
layout: false
class: clear

.b.mono.pink[Output] `feols` output for the *plain* OLS model .it[vs.] the FE model.

```{r, fe-reg-table2, echo = FALSE, highlight.output = 5}
etable(est_ols, est_fe, digits = 3)
```

Both $\color{#e64173}{\hat{\beta}_1}$ are stat. significant, but they tell .pink[very different stories].

---
layout: false
class: clear

.b.mono.pink[Output] `feols` output for the *plain* OLS model .it[vs.] the FE model.

```{r, fe-reg-table3, echo = FALSE, highlight.output = 4}
etable(est_ols, est_fe, digits = 3)
```

.qa[Q] Why is there .pink[no intercept] $(\beta_0)$ in the FE model?
--
<br>
.qa[A] The .pink[fixed effects are collinear] with the intercept—they each provide a separate intercept for each unit and time period.

---
layout: false
class: clear

.b.mono.pink[Output] `feols` output for the *plain* OLS model .it[vs.] the FE model.

```{r, fe-reg-table4, echo = FALSE, highlight.output = 13}
etable(est_ols, est_fe, digits = 3)
```

.qa[Q] What is .pink.b[*within* R.super[2]]?
--
<br>
.qa[A] It is the $R^2$ of the regression .pink[after accouting for the FEs]: how much of the outcome's variation regressors explain, *after controlling for FEs*.

---
layout: false
class: clear

.b.mono.pink[Output] `feols` output for the *plain* OLS model .it[vs.] the FE model.

```{r, fe-reg-table5, echo = FALSE, highlight.output = 12}
etable(est_ols, est_fe, digits = 3)
```

The FE model has as high $\color{#e64173}{R^2}$ because fixed effects explain a lot of variation (in essence, means of states and years).

---
layout: false

# Panel estimation
## Summary

Fixed effects help one key problem: .hi[time-invariant omitted variables].

--

But FEs .b.slate[do not] solve *all* causal problems, *e.g.*,

- time-varying confounders,
- reverse causality
- measurement error,
- poor measurement of treatment timing,
- weak within variation.
  
As with *all* econometric tools, careful use is required.

---
layout: true
# Difference-in-differences

---
class: inverse, middle

---
name: did_intuition

## Core idea

.attn[Difference-in-differences] (DiD).super.pink[†] is a popular panel-data approach for estimating causal effects of policies/events.

.footnote[.pink[†] Also called .it[double-differences] or .it[diff-in-diff]. Increasingly combined with .it[two-way fixed effects (TWFE)].]

DiD compares observations across two dimensions:

1. .b.purple[Group:] comparing
  - .purple.it[treated] individuals (potentially affected by the policy)
  - .purple.it[control] individuals (unaffected by the policy)
2. .b.orange[Time:] comparing
  - .orange.it[before] the policy takes effect .grey-vlight.it[(pre-treatment)]
  - .orange.it[after] the policy takes effect .grey-vlight.it[(post-treatment)]

---
## Intuition

DiD functions by taking the difference between two differences:

--

1. Δ.b[Y̅.sub.orange.tran50[Pre]] $=$ .b[Y̅](.purple[Trt]; .orange.tran50[Pre]) $-$ .b[Y̅](.purple.tran50[Ctrl]; .orange.tran50[Pre]) <br> the average difference between .purple[trt.] and .purple.tran50[control] .b.orange.tran50[before] the policy, <br> accounting for pre-existing differences between the groups;

--

2. Δ.b[Y̅.sub.orange[Post]] $=$ .b[Y̅](.purple[Trt]; .orange[Post]) $-$ .b[Y̅](.purple.tran50[Ctrl]; .orange[Post]) <br> the average difference between .purple[trt.] and .purple.tran50[control] .b.orange[after] the policy, <br> which includes pre-existing differences *and* the .pink[treatment effect].

--

The .attn[DiD estimate] is the difference between these differences
<br>
   .pink[τ̂] $=$ Δ.b[Y̅.sub.orange[Post]] $-$ Δ.b[Y̅.sub.orange.tran50[Pre]]
<br>
.note[I.e.,] how much did the .purple[trt]-.purple.tran50[control] difference change from .orange.tran50[pre] to .orange[post]?

---
layout: false
class: clear

.mono.b.slate[Ex.] Two .purple[groups (trt/ctrl)] and two .orange[time periods (pre/post policy)].

```{r, did-cell-table1, echo = FALSE}
# Copy the did_cells to data.table
did_dt = copy(did_cells)
setDT(did_dt)
# Arrange by period and group
setorder(did_dt, -Period, -Group)
# Set column order
setcolorder(
  did_dt,
  c('Group', 'Period', 'mean_y')
)
# Set names
setnames(did_dt, c('Group', 'Period', 'Mean', 'trt', 'post'))
# Simplify means
did_dt[, Mean := Mean |> round(3)]
# Show the table
did_dt[, 1:3]
```

---
layout: false
class: clear

.mono.b.slate[Ex.] Two .purple[groups (trt/ctrl)] and two .orange[time periods (pre/post policy)].

```{r, did-cell-table2, echo = FALSE, highlight.output = 3:4}
did_dt[, 1:3]
```

1. Δ.b[Y̅.sub.orange.tran50[Pre]] $=$ .b[Y̅](.purple[Trt]; .orange.tran50[Pre]) $-$ .b[Y̅](.purple.tran50[Ctrl]; .orange.tran50[Pre]) $=$ `r did_dt$Mean[2:1] |> diff()`.

---
layout: false
class: clear

.mono.b.slate[Ex.] Two .purple[groups (trt/ctrl)] and two .orange[time periods (pre/post policy)].

```{r, did-cell-table3, echo = FALSE, highlight.output = 5:6}
did_dt[, 1:3]
```

1. Δ.b[Y̅.sub.orange.tran50[Pre]] $=$ .b[Y̅](.purple[Trt]; .orange.tran50[Pre]) $-$ .b[Y̅](.purple.tran50[Ctrl]; .orange.tran50[Pre]) $=$ `r did_dt$Mean[2:1] |> diff()`.

2. Δ.b[Y̅.sub.orange[Post]] $=$ .b[Y̅](.purple[Trt]; .orange[Post]) $-$ .b[Y̅](.purple.tran50[Ctrl]; .orange[Post]) $=$ `r did_dt$Mean[4:3] |> diff()`

---
layout: false
class: clear

.mono.b.slate[Ex.] Two .purple[groups (trt/ctrl)] and two .orange[time periods (pre/post policy)].

```{r, did-cell-table4, echo = FALSE}
did_dt[, 1:3]
```

1. Δ.b[Y̅.sub.orange.tran50[Pre]] $=$ .b[Y̅](.purple[Trt]; .orange.tran50[Pre]) $-$ .b[Y̅](.purple.tran50[Ctrl]; .orange.tran50[Pre]) $=$ `r did_dt$Mean[2:1] |> diff()`.

2. Δ.b[Y̅.sub.orange[Post]] $=$ .b[Y̅](.purple[Trt]; .orange[Post]) $-$ .b[Y̅](.purple.tran50[Ctrl]; .orange[Post]) $=$ `r did_dt$Mean[4:3] |> diff()`

The .attn[DiD estimate]—the difference between these differences
<br>
   .pink[τ̂] $=$ Δ.b[Y̅.sub.orange[Post]] $-$ Δ.b[Y̅.sub.orange.tran50[Pre]] $=$ `r (did_dt$Mean[4:3] |> diff()) - (did_dt$Mean[2:1] |> diff())`

---
layout: false
class: clear

.note[Note] You can also write the DiD estimator as two other differences,


   .pink[τ̂] $=$ Δ.b[Y̅.sub.orange[Post]] $-$ Δ.b[Y̅.sub.orange.tran50[Pre]] 

--

   .white[τ̂] $=$ $\big[$ .b[Y̅](.purple[Trt]; .orange[Post]) $-$ .b[Y̅](.purple.tran50[Ctrl]; .orange[Post]) $\big]$
$-$
$\big[$ .b[Y̅](.purple[Trt]; .orange.tran50[Pre]) $-$ .b[Y̅](.purple.tran50[Ctrl]; .orange.tran50[Pre]) $\big]$

--

   .white[τ̂] $=$.b[Y̅](.purple[Trt]; .orange[Post]) $-$ .b[Y̅](.purple.tran50[Ctrl]; .orange[Post]) $-$ .b[Y̅](.purple[Trt]; .orange.tran50[Pre]) $+$ .b[Y̅](.purple.tran50[Ctrl]; .orange.tran50[Pre])

--

   .white[τ̂] $=$ $\big[$ .b[Y̅](.purple[Trt]; .orange[Post]) $-$ .b[Y̅](.purple[Trt]; .orange.tran50[Pre]) $\big]$ $-$ $\big[$ .b[Y̅](.purple.tran50[Ctrl]; .orange[Post]) $-$ .b[Y̅](.purple.tran50[Ctrl]; .orange.tran50[Pre]) $\big]$

--

   .white[τ̂] $=$ Δ.b[Y̅].sub.purple[Trt] $-$ Δ.b[Y̅].sub.purple.tran50[Ctrl]

.note[I.e.,] the change in the .purple[treated group] (.orange[post] $-$ .orange.tran50[pre]) relative to <br> the change in the .purple.tran50[control group].

We're using the .purple.tran50[control group's .b[change]] as a .b[counterfactual] for <br> the .purple[treated group's .b[change]] had it not received treatment.

---
layout: true
# Difference-in-differences

---
## Classic example: Card and Krueger (1994)

One famous DiD [application](https://davidcard.berkeley.edu/papers/njmin-aer.pdf) estimates the .b.slate[impact of increasing the minimum wage on fast-food employment].

- .note[Who:] .purple[New Jersey (trt)] *vs.* .purple.tran50[Pennsylvania (ctrl)]
- .note[When:] .orange.tran50[before 01 April 1992 (pre)] *vs.* .orange[after 01 April 1992 (post)]

DiD estimates .grey-vlight.it[(both questions = same answer)]

1. How did NJ-PA empl. diff. change after the policy, relative to before?
2. How did NJ's empl. change after the policy relative to PA's change?

--

.note[Key assumption:] PA needs to be a good counterfactual for NJ. In the absence of the policy, NJ and PA would have followed the same empl. trends.

---

## DiD as a regression

With two groups and two periods
$$y_{it} = \beta_0 + \beta_1 \color{#6A5ACD}{\text{Trt}_i} + \beta_2 \color{#FFA500}{\text{Post}_t} + \delta \left( \color{#6A5ACD}{\text{Trt}_i} \times \color{#FFA500}{\text{Post}_t} \right) + u_{it}$$
where $\color{#6A5ACD}{\text{Trt}_i}$ and $\color{#FFA500}{\text{Post}_t}$ are binary indicators for the .purple[treated group] and .orange[post-treatment period]. 

The interaction asks if if the .orange[post] change in the .purple[treated group] differs from the .orange[post] change in the .purple.tran50[control group]: the .attn[treatment effect].

--

With panel data, we often write

$$y_{it} = \alpha_i + \lambda_t + \delta D_{it} + u_{it}$$

where $D_{it}=1$ only for treated units after treatment starts.

---
layout: false
class: clear

```{r, did-plot1, echo = FALSE}
# Add post indicator
did_plot_dt = copy(did_plot_df)
setDT(did_plot_dt)
did_plot_dt[, post := 1 * (period > 5)]
# Add factor var for interaction
did_plot_dt[, int := interaction(treat, post)]
# The first plot
ggplot(did_plot_dt, aes(x = period, y = mean_y, color = Group)) +
  geom_vline(xintercept = 5.5, linetype = 'dashed', color = grey_mid) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 2.7) +
  scale_color_manual(values = c('Control' = purple_l, 'Treated' = purple)) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = 'Time period',
    y = 'Average Y',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = 'bottom'
  )
```

.b.slate[DiD in pictures] Starting with time series for each unit.
<br>
.grey-vlight.it[(Treatment starts between periods 5 and 6.)]

---
class: clear

```{r, did-plot2, echo = FALSE}
ggplot(did_plot_dt, aes(x = period, y = mean_y, color = Group)) +
  geom_vline(xintercept = 5.5, linetype = 'dashed', color = grey_mid) +
  geom_line(linewidth = 1.3, alpha = .15) +
  geom_point(size = 2.7, alpha = .15) +
  geom_line(
    aes(group = int),
    stat = 'smooth', method = 'lm', formula = y ~ 1, se = FALSE,
    linewidth = 1.3
  ) +
  scale_color_manual(values = c('Control' = purple_l, 'Treated' = purple)) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = 'Time period',
    y = 'Average Y',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = 'bottom'
  )
```

Here we have .b.purple[group]-by-.b.orange[time] means.

---
class: clear

```{r, did-plot3, echo = FALSE}
ggplot(did_plot_dt, aes(x = period, y = mean_y, color = Group)) +
  geom_vline(xintercept = 5.5, linetype = 'dashed', color = grey_mid) +
  geom_line(linewidth = 1.3, alpha = .15) +
  geom_point(size = 2.7, alpha = .15) +
  geom_line(
    aes(group = int),
    stat = 'smooth', method = 'lm', formula = y ~ 1, se = FALSE,
    linewidth = 1.3
  ) +
  annotate(
    'segment',
    x = 3, xend = 3,
    y = did_plot_dt[post == 0 & treat == 0, mean(mean_y)],
    yend = did_plot_dt[post == 0 & treat == 1, mean(mean_y)],
    color = orange_l, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  annotate(
    'segment',
    x = 8, xend = 8,
    y = did_plot_dt[post == 1 & treat == 0, mean(mean_y)],
    yend = did_plot_dt[post == 1 & treat == 1, mean(mean_y)],
    color = orange, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  scale_color_manual(values = c('Control' = purple_l, 'Treated' = purple)) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = 'Time period',
    y = 'Average Y',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = 'bottom'
  )
```

The .attn[DiD estimate] compares the .b.orange.tran50[pre-trt] difference to the .b.orange[post-trt] diff.

---
class: clear

```{r, did-plot4, echo = FALSE}
# Calculate pre- and post-trt diffs
pre_diff =
  did_plot_dt[post == 0 & treat == 1, mean(mean_y)] -
  did_plot_dt[post == 0 & treat == 0, mean(mean_y)]
post_diff =
  did_plot_dt[post == 1 & treat == 1, mean(mean_y)] -
  did_plot_dt[post == 1 & treat == 0, mean(mean_y)]
# The plot
ggplot(did_plot_dt, aes(x = period, y = mean_y, color = Group)) +
  geom_vline(xintercept = 5.5, linetype = 'dashed', color = grey_mid) +
  geom_line(linewidth = 1.3, alpha = .15) +
  geom_point(size = 2.7, alpha = .15) +
  geom_line(
    aes(group = int),
    stat = 'smooth', method = 'lm', formula = y ~ 1, se = FALSE,
    linewidth = 1.3
  ) +
  annotate(
    'segment',
    x = 3, xend = 3,
    y = did_plot_dt[post == 0 & treat == 0, mean(mean_y)],
    yend = did_plot_dt[post == 0 & treat == 1, mean(mean_y)],
    color = orange_l, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  annotate(
    'segment',
    x = 7.5, xend = 7.5,
    y = did_plot_dt[post == 1 & treat == 0, mean(mean_y)],
    yend = did_plot_dt[post == 1 & treat == 0, mean(mean_y)] + pre_diff,
    color = orange_l, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  annotate(
    'segment',
    x = 8, xend = 8,
    y = did_plot_dt[post == 1 & treat == 0, mean(mean_y)],
    yend = did_plot_dt[post == 1 & treat == 1, mean(mean_y)],
    color = orange, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  annotate(
    'segment',
    x = 7.5, xend = 7.5,
    y = did_plot_dt[post == 1 & treat == 0, mean(mean_y)] + pre_diff,
    yend = did_plot_dt[post == 1 & treat == 1, mean(mean_y)],
    color = red_pink, linewidth = 1.9,
    arrow = arrow(length = unit(.2, 'inches'), ends = 'both'),
    lineend = 'butt', linejoin = 'mitre'
  ) +
  scale_color_manual(values = c('Control' = purple_l, 'Treated' = purple)) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = 'Time period',
    y = 'Average Y',
    color = NULL
  ) +
  theme_minimal(base_size = 21, base_family = 'Fira Sans Condensed') +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = 'bottom'
  )
```

The difference in the .b.orange.tran50[pre-trt] and .b.orange[post-trt] diffs is the .attn[DiD estimate].

---
class: clear

.b[DiD in `fixest`]

```{r, did-code}
did_mod =
  feols(
    y ~ i(post, treat, ref = 0) | id + period,
    cluster = 'id',
    data = base_did
  )
```

--

```{r, did-reg-table, echo = FALSE, highlight.output = 6}
did_mod
```

The interaction btn .orange[post] and .purple[treat] is the .attn[DiD estimate] of the trt effect.

---
layout: true
# Difference-in-differences

---
## The key DiD assumption

DiD relies upon a critical assumption often called .attn[parallel trends]:

> In the absence of treatment, treated and control units would have followed the .pink[same average trend].

Again, we're assuming that the trend of the control units provide <br> a .b.slate[good counterfactual] for the treated group.

---
## What threatens DiD?

So what violates this .note[parallel trends] assumption?

--

- treated units were already on a different trend,
- anticipation changes behavior before treatment,
- other shocks hit treated units at the same time as treatment.

--

What else can *mess up* DiD estimates?

- standard errors ignore within-unit dependence,
- measurement error,
- spillovers from trt to control,
- dynamic treatment effects,
- het trt effects *and* het trt timing...

---

## DiD's place in the toolbox

DiD is especially useful when

- treatment .orange[timing] is clear,
- .purple[treatment definition] is clear,
- you have good .orange[pre-treatment] data
- .purple.tran50[ctrl group] provide a plausible counterfactual for .purple[trt group],
- the .pink[treatment effect] is tied to a discrete policy or event.
  
--

.qa[Q] What are the other options?
--
<br>
.qa[A] Synthetic control and friends—build a weighted average of control units to better match the treated unit's pre-treatment path.

---
## Takeaways

Key concepts from panel estimation, FEs, and DiD:

- .b.slate[Panel data] let us compare a unit to itself .b.purple[over time].

- .b.slate[Fixed effects] remove .b.orange[time-invariant omitted variables].

- $\implies$ .b.slate[Identifying variation] is .b.purple[within variation].

- .b.slate[DiD] focuses on how trt-ctrl .b.slate[differences] change over time.

Panels can certainly help, but they still depend on strong assumptions.
