---
title: 'Instrumental Variables'
subtitle: 'EC 421, Set 11'
author:
  - name: Ed Rubin
    email: edwardr@uoregon.edu
editor:
  render-on-save: true
format:
  clean-revealjs:
    slide-level: 2
    theme: [../my.scss, default]
    highlight-style: gruvbox
    embed-resources: true
    slide-number: c/t
    incremental: false
    controls-layout: edges
    transition: none
    preview-links: auto
    html-math-method: mathjax
    code-line-numbers: false
execute:
  warning: false
  message: false

---

## {.center}

[Prologue]{.bigger .it .slab}

---

## Schedule {#schedule}

```{r}
#| label: "setup"
#| include: false
# devtools::install_github("dill/emoGG")
library(pacman)
p_load(
  broom, here, tidyverse,
  emoGG, latex2exp, ggplot2, ggthemes, ggforce, viridis, extrafont, gridExtra,
  wooldridge,
  kableExtra, snakecase,
  data.table, dplyr, estimatr,
  lubridate,
  magrittr, knitr, parallel
)
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#fb6107"
blue <- "#2b59c3"
green <- "#8bb174"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(crayon.enabled = F)
options(knitr.table.format = "html")
# Column names for regression results
reg_columns <- c("Term", "Est.", "S.E.", "t stat.", "p-Value")
# Function for formatting p values
format_pvi <- function(pv) {
  return(ifelse(
    pv < 0.0001,
    "<0.0001",
    round(pv, 4) %>% format(scientific = F)
  ))
}
format_pv <- function(pvs) lapply(X = pvs, FUN = format_pvi) %>% unlist()
# Tidy regression results table
tidy_table <- function(x, terms, highlight_row = 1, highlight_color = "black", highlight_bold = T, digits = c(NA, 3, 3, 2, 5)) {
  x %>%
    tidy() %>%
    select(1:5) %>%
    mutate(
      term = terms,
      p.value = p.value %>% format_pv()
    ) %>%
    kable(
      col.names = reg_columns,
      escape = F,
      digits = digits
    ) %>%
    kable_styling(font_size = 20) %>%
    row_spec(1:nrow(tidy(x)), background = "white") %>%
    row_spec(highlight_row, bold = highlight_bold, color = highlight_color)
}
```

### Last Time

Causality

### Today

- Review: Causality
- New: Instrumental variables

---

## {.center}

[Causality]{.bigger .it .slab}

[Review]{.it .slab .grey-vlight}

---

## Causality {#causality_review}
### Review

In our last lecture, we returned to the concept of [causality]{.b .pink}.

::: {.fragment}

We worked through the *Rubin causal model*, in which we defined

:::

::: {.fragment}

- $\textcolor{#e64173}{y_{1i}:}$ the outcome for individual $i$ if she had received treatment

:::

::: {.fragment}

- $\textcolor{#6A5ACD}{y_{0i}:}$ the outcome for individual $i$ if she had not received treatment

:::

::: {.fragment}

and we referred to individuals who did not receive treatment as *control*.

:::

::: {.fragment}

***If*** we were able to know both $\textcolor{#e64173}{y_{1i}}$ ***and*** $\textcolor{#6A5ACD}{y_{0i}}$, we could calculate the causal effect of treatment for individual $i$, _i.e._,

:::

::: {.fragment}

$$
\begin{align}
  \tau_i = \textcolor{#e64173}{y_{1i}} - \textcolor{#6A5ACD}{y_{0i}}
\end{align}
$$

:::

---

## Causality
### Review

[Fundamental problem of causal inference:]{.b .slate}

<br>We cannot simultaneously know $\textcolor{#e64173}{y_{1i}}$ and $\textcolor{#6A5ACD}{y_{0i}}$.

::: {.fragment}

Either we observe individual $i$ in the treatment group, _i.e._,
$$
\begin{align}
  \tau_i = \textcolor{#e64173}{y_{1i}} - \textcolor{#6A5ACD}{?}
\end{align}
$$

:::

::: {.fragment}

or we observe $i$ in the control group, _i.e._,
$$
\begin{align}
  \tau_i = \textcolor{#e64173}{?} - \textcolor{#6A5ACD}{y_{0i}}
\end{align}
$$

:::

::: {.fragment}

but never both at the same time.

:::

---

## Causality
### Review

If we want to know $\tau_i$ (or at least $\overline{\tau}$), what can we do?

::: {.fragment}

**Idea:** Estimate the [average treatment effect]{.b .slate} as the difference between the average outcomes in the treatment group and the control group, _i.e._,

:::

::: {.fragment}

$$
\begin{align}
  \textcolor{#e64173}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)}
\end{align}
$$
where $D_i=1$ if $i$ received treatment, and $D_i=0$ if $i$ is in the control group.

:::

---

## Causality
### Review

**Result:** We showed that even when the treatment effect is constant (meaning $\tau_i=\tau$ for all $i$),
$$
\begin{align}
  &\textcolor{#e64173}{\mathop{Avg}\left( y_i\mid D_i = 1 \right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_i\mid D_i =0 \right)} \\
  &\quad\quad = \tau + \underbrace{\textcolor{#e64173}{\mathop{Avg}\left(\textcolor{#6A5ACD}{y_{0,i}} \mid D_i = 1 \right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}}_{\textcolor{#FFA500}{\text{Selection bias}}}
\end{align}
$$

which says that the difference in the groups' means will give us a **biased estimate** for the causal effect of treatment [if we have selection bias.]{.b .orange}

---

## Causality
### Review

**Q:** What is this [selection bias]{.b .orange}?

::: {.fragment}

**A:** **(Informal)** We have selection bias when our control group doesn't offer a good comparison for our treatment group.

:::

::: {.fragment}

Specifically, the control group doesn't give us a good [counterfactual]{.b .orange} for [what our treatment group would have looked like if the members had not received treatment.]{.orange}

:::

::: {.fragment}

Basically, the groups are different.

:::

::: {.fragment}

**A:** **(Formal)** The [average *untreated* outcome for a member of our **treatment group**]{.pink} (which we cannot observe) differs from the [average *untreated* outcome for a member of our **control group**]{.purple}, _i.e._,
$$
\begin{align}
  \textcolor{#e64173}{\mathop{Avg}\left(\textcolor{#6A5ACD}{y_{0,i}} \mid D_i = 1 \right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}
\end{align}
$$

:::

---

## Causality
### Review

[Practical problem:]{.b .slate}

$$
\begin{align}
  \underbrace{\textcolor{#e64173}{\mathop{Avg}\left(\textcolor{#6A5ACD}{y_{0,i}} \mid D_i = 1 \right)}}_{\textcolor{#e64173}{\text{Unobservable}}} - \textcolor{#6A5ACD}{\mathop{Avg}\left( y_{0,i}\mid D_i =0 \right)}
\end{align}
$$

(back to the *fundamental problem of causal inference*)

::: {.fragment}

[Bigger problem:]{.b .slate}

:::

::: {.fragment}

Sounds a bit like omitted-variable bias, right?

:::

::: {.fragment}

Our [treatment]{.pink} variable is correlated with something that makese the two groups different.

:::

---

## Causality
### Review

**Example:** Imagine we have two people—Al and Bri—and a single binary treatment, college. We interested in the effect of college on earnings.

::: columns
::: {.column width="50%"}
::: {.center}
[Earn[1,Al]{.sub}]{.pink}

<br>[Earn[0,Al]{.sub}]{.purple} = [$30K]{.purple}
:::
:::
::: {.column width="50%"}
[Earn[1,Bri]{.sub}]{.pink}

<br>[Earn[0,Bri]{.sub}]{.purple} = [$110K]{.purple}
:::
:::

::: {.fragment}

They both have the same treatment effect (return to college)

:::

::: {.fragment}

<br> $\quad\quad\tau$[Al]{.sub} = [Earn[1,Al]{.sub}]{.pink} - [Earn[0,Al]{.sub}]{.purple} = [$60K]{.pink} - [$30K]{.purple} = $30K

:::

::: {.fragment}

<br> $\quad\quad\tau$[Bri]{.sub} = [Earn[1,Bri]{.sub}]{.pink} - [Earn[0,Bri]{.sub}]{.purple} = [$140K]{.pink} - [$100K]{.purple} = $30K

:::

::: {.fragment}

but any real-world estimate would have serious selection issues since [Earn[0,Al]{.sub}]{.purple} ≠ [Earn[0,Bri]{.sub}.]{.purple}

:::

---

## Causality
### Review

**Example:** Imagine we have two people—Al and Bri—and a single binary treatment, college. We interested in the effect of college on earnings.

::: columns
::: {.column width="50%"}
::: {.center}
[Earn[1,Al]{.sub}]{.pink}

<br>[Earn[0,Al]{.sub}]{.purple} = [$30K]{.purple}
:::
:::
::: {.column width="50%"}
[Earn[1,Bri]{.sub}]{.pink}

<br>[Earn[0,Bri]{.sub}]{.purple} = [$110K]{.purple}
:::
:::

The selection bias...

::: {.fragment}

If Bri attended college (D[Bri]{.sub}=1) and Al did not (D[Al]{.sub}=0):
<br>
$\quad\quad\hat{\tau}$ = [Earn[1,Bri]{.sub}]{.pink} - [Earn[0,Al]{.sub}]{.purple} = [$140K]{.pink} - [$30K]{.purple} = $110K

:::

::: {.fragment}

If Al attended college (D[Al]{.sub}=1) and Bri did not (D[Bri]{.sub}=0):
<br>
$\quad\quad\hat{\tau}$ = [Earn[1,Al]{.sub}]{.pink} - [Earn[0,Bri]{.sub}]{.purple} = [$60K]{.pink} - [$110K]{.purple} = -$50K

:::

---

## Causality
### Review

We have (at least) two problems...

1. Selection bias is difficult to observe

2. If selection bias is present, our estimate for $\tau$ is biased, preventing us from understanding the causal effect of treatment.

::: {.fragment}

[Solution:]{.b .slate}

:::

::: {.fragment}

- [Option 1:]{.b .slate} [Distribute treatment]{.b .pink} in a way such that the treatment and control groups are essentially identical

:::

::: {.fragment}

(experiments).

:::

::: {.fragment}

- [Option 2:]{.b .slate} [Build a control]{.b .purple} group that *matches* the treatment group

:::

::: {.fragment}

<br>(life with observational data).

:::

---

## Instrumental variables {#intro}

### Intro

[Instrumental variables (IV)]{.b .pink}

::: {.fragment}

*Recall:* [Selection bias]{.b .orange} means our [treatment]{.pink} and [control]{.purple} groups differ on some unobserved/omitted dimension. ([Endogeneity]{.b .slate})

:::

::: {.fragment}

[Instrumental variables]{.b .pink}

:::

::: {.fragment}

- the [exogenous]{.b .slate} part of $x$, which gives us unbiased estimates
- the [endogenous]{.b .slate} part of $x$, which biases our results

:::

::: {.fragment}

If we use only the exogenous (*good*) variation in $x$, then we can avoid selection bias/omitted-variable bias.

:::

---

## Instrumental variables

### Introductory example

*Example:* If we want to estimate the effect of veteran status on earnings,
$$
\begin{align}
  \text{Earnings}_i = \beta_0 + \beta_1 \text{Veteran}_i + u_i \tag{1}
\end{align}
$$

::: {.fragment}

We would love to calculate $\textcolor{#e64173}{\text{Earnings}_{1i}} - \textcolor{#6A5ACD}{\text{Earnings}_{0i}}$, but we can't.

:::

::: {.fragment}

And OLS will likely be biased for $(1)$ due to selection/omitted-variable bias.

:::

---

## Instrumental variables

### Introductory example

Imagine that we can split veteran status into an exogenous part and an endogenous part...

::: {.fragment}

$$
\begin{align}
  \text{Earnings}_i
  &= \beta_0 + \beta_1 \text{Veteran}_i + u_i \tag{1} \\
  &= \beta_0 + \beta_1 \left(\text{Veteran}_i^{\text{Exog.}} + \text{Veteran}_i^{\text{Endog.}}\right) + u_i \\
  &= \beta_0 + \beta_1 \text{Veteran}_i^{\text{Exog.}} + \underbrace{\beta_1 \text{Veteran}_i^{\text{Endog.}} + u_i}_{w_i} \\
  &= \beta_0 + \beta_1 \text{Veteran}_i^{\text{Exog.}} + w_i
\end{align}
$$

:::

::: {.fragment}

We could use this exogenous variation in veteran status to consistently estimate $\beta_1$.

:::

::: {.fragment}

**Q:** What would exogenous variation in veteran status mean?

:::

---

## Instrumental variables

### Introductory example

**Q:** What would exogenous variation in veteran status mean?

::: {.fragment}

**A[1]{.sub}:** Choices to enlist in the military that are essentially random—or at least uncorrelated with omitted variables and the disturbance.

:::

::: {.fragment}

**A[2]{.sub}:** [No selection bias:]{.b .orange}
$$
\begin{align}
  \textcolor{#e64173}{\mathop{Avg}\left(\text{Earnings}_{0i}\mid\text{Veteran}_i = 1\right)} - \textcolor{#6A5ACD}{\mathop{Avg}\left( \text{Earnings}_{0i} \mid \text{Veteran}_i = 0 \right)} = 0
\end{align}
$$

:::

---

## Instrumental variables {#instruments}

### Instruments

**Q:** How do we isolate this *exogenous variation* in our explanatory variable?

::: {.fragment}

<br>**A:** Find an instrument (an instrumental variable).

:::

::: {.fragment}

**Q:** What's an instrument?

:::

::: {.fragment}

<br>**A:** An [instrument]{.b .pink} is a variable that is

:::

::: {.fragment}

1. **correlated** with the **explanatory variable** of interest ([relevant]{.b .pink}),
2. **uncorrelated** with the **disturbance** ([exogenous]{.b .pink}).

:::

---

## Instrumental variables

### Instruments

>**Q:** What's an instrument?
><br>**A:** An [instrument]{.b .pink} is a variable that is
>
>1. **correlated** with the **explanatory variable** of interest ([relevant]{.b .pink}),
>2. **uncorrelated** with the **disturbance** ([exogenous]{.b .pink}).

::: {.fragment}

So if we want an instrument $z_i$ for endogenous veteran status in
$$
\begin{align}
  \text{Earnings}_i = \beta_0 + \beta_1 \text{Veteran}_i + u_i
\end{align}
$$

:::

::: {.fragment}

1. [Relevant:]{.b .pink} $\mathop{\text{Cov}} \left( \text{Veteran}_i,\, z_i \right) \neq 0$
2. [Exogenous:]{.b .pink} $\mathop{\text{Cov}} \left( z_i,\, u_i \right) = 0$

:::

---

## Instrumental variables {#relevant}

### Instruments: Relevance

[Relevance:]{.b .pink}

We can actually test this requirement using regression and a *t* test.

::: {.fragment}

***Example:*** For the [veteran status]{.pink}, consider three potential instruments:

:::

::: columns
::: {.column width="50%"}
::: {.fragment}

1\. Social security number<br>[blank]{.white}

:::
:::
::: {.column width="50%"}
::: {.fragment}

[Probably not relevant]{.b .slate}

:::
:::
:::

::: columns
::: {.column width="50%"}
::: {.fragment}

2\. Physical fitness<br>[blank]{.white}

:::
:::
::: {.column width="50%"}
::: {.fragment}

[Potentially relevant]{.b .pink}

:::
:::
:::

::: columns
::: {.column width="50%"}
::: {.fragment}

3\. Vietnam War draft

:::
:::
::: {.column width="50%"}
::: {.fragment}

[Relevant]{.b .pink}

:::
:::
:::

---

## Instrumental variables {#exogenous}

### Instruments: Exogeneity

[Exogeneity:]{.b .pink}

$z_i$ must be uncorrelated with our disturbance $u_i$. [Not testable.]{.b .pink}

::: {.fragment}

***Example:*** For the [veteran status]{.pink}, consider three potential instruments:

:::

::: columns
::: {.column width="50%"}
::: {.fragment}

1\. Social security number<br>[blank]{.white}

:::
:::
::: {.column width="50%"}
::: {.fragment}

[Exogenous]{.b .pink}

:::
:::
:::

::: columns
::: {.column width="50%"}
::: {.fragment}

2\. Physical fitness<br>[blank]{.white}

:::
:::
::: {.column width="50%"}
::: {.fragment}

[Not exogenous]{.b .slate}

:::
:::
:::

::: columns
::: {.column width="50%"}
::: {.fragment}

3\. Vietnam War draft

:::
:::
::: {.column width="50%"}
::: {.fragment}

[Exogenous]{.b .pink}

:::
:::
:::

---

## Instrumental variables

### Instrumental review

Let's recap...

- Our instrument must be [correlated with our endogenous variable]{.b .pink}.

- Our instrument must be [uncorrelated with any other variable that affects the outcome]{.b .pink}.

::: {.fragment}

[In other words:]{.b .slate}

:::

::: {.fragment}

<br>The instrument only affects our outcome through the endogenous variable.

:::

---

## Instrumental variables

### Back to our example

For [veteran status]{.pink} we considered three potential instruments:

::: columns
::: {.column width="50%"}
::: {.fragment}

1\. Social security number<br>[blank]{.white}<br>[blank]{.white}

:::
:::
::: {.column width="50%"}
::: {.fragment}

[Not relevant]{.b .slate}

:::
:::
:::

::: columns
::: {.column width="50%"}
::: {.fragment}

2\. Physical fitness<br>[blank]{.white}<br>[blank]{.white}

:::
:::
::: {.column width="50%"}
::: {.fragment}

[Probably relevant]{.b .pink}

:::
:::
:::

::: columns
::: {.column width="50%"}
::: {.fragment}

3\. Vietnam War draft<br>[blank]{.white}<br>[blank]{.white}

:::
:::
::: {.column width="50%"}
::: {.fragment}

[Relevant]{.b .pink}

:::
:::
:::

::: {.fragment}

Thus, only the Vietnam War's draft lottery appears to be a [*valid* instrument]{.b .pink}.

:::

---

## {.clear}

If we have a *valid* instrument (_e.g._, the draft lottery), how do we use it?

---

## Instrumental variables {#iv_estimation}
### Estimation

*Recall:* We want to estimate the effect of veteran status on earnings.
$$
\begin{align}
  \textcolor{#FFA500}{\text{Earnings}_i} = \beta_0 + \beta_1 \textcolor{#6A5ACD}{\text{Veteran}_i} + u_i
\end{align}
$$

::: {.fragment}

Let's consider two related effects:

:::

::: {.fragment}

1. The effect of the [instrument]{.b .pink} on the [endogenous variable]{.b .purple}, _e.g._,
$$
\begin{align}
  \textcolor{#6A5ACD}{\text{Veteran}_i} = \gamma_0 + \gamma_1 \textcolor{#e64173}{\text{Draft}_i} + v_i
\end{align}
$$

:::

::: {.fragment}

1. The effect of the [instrument]{.b .pink} on the [outcome variable]{.b .orange}, _e.g._,
$$
\begin{align}
  \textcolor{#FFA500}{\text{Earnings}_i} = \pi_0 + \pi_1 \textcolor{#e64173}{\text{Draft}_i} + w_i
\end{align}
$$

:::

---

## Instrumental variables
### Estimation

*Recall:* We want to estimate the effect of veteran status on earnings.
$$
\begin{align}
  \textcolor{#FFA500}{\text{Earnings}_i} = \beta_0 + \beta_1 \textcolor{#6A5ACD}{\text{Veteran}_i} + u_i
\end{align}
$$
and we know that the draft affected veteran status.

::: {.fragment}

[[Draft]{.b .pink} ⟶ [Veteran status]{.b .purple} ⟶ [Earnings]{.b .orange}]{.center}

:::

::: {.fragment}

Using our assumptions on independence and exogeneity:

:::

::: {.fragment}

(Effect of [the draft]{.b .pink} on [earnings]{.b .orange}) =
<br>  (Effect of [the draft]{.b .pink} on [veteran status]{.b .purple})×
<br>  (Effect of [veteran status]{.b .purple} on [earnings]{.b .orange})

:::

---

## Instrumental variables
### Estimation

We just wrote out an expression for the effect of [the draft]{.b .pink} on [earnings]{.b .orange}, _i.e._,

(Effect of [the draft]{.b .pink} on [earnings]{.b .orange}) =
<br>  (Effect of [the draft]{.b .pink} on [veteran status]{.b .purple})×
<br>  (Effect of [veteran status]{.b .purple} on [earnings]{.b .orange})[but we want to know the effect of [veteran status]{.b .purple} on [earnings]{.b .orange}.]{.fragment}

::: {.fragment}

Rearrange!

:::

::: {.fragment}

(Effect of [veteran status]{.b .purple} on [earnings]{.b .orange}) =
<br>  [(Effect of [the draft]{.b .pink} on [earnings]{.b .orange})]{.top}
<br>  [(Effect of [the draft]{.b .pink} on [veteran status]{.b .purple})]{.bottom}

:::

::: {.fragment}

Our [instrument]{.b .pink} consistently estimates both parts of this fraction!

:::

---

## Instrumental variables
### Estimation: Bring it all together

By estimating two regressions involving our [instrument]{.b .pink},

1. The effect of the [instrument]{.b .pink} on the [endogenous variable]{.b .purple}, _e.g._,
$$
\begin{align}
  \textcolor{#6A5ACD}{\text{Veteran}_i} = \gamma_0 + \gamma_1 \textcolor{#e64173}{\text{Draft}_i} + v_i
\end{align}
$$

1. The effect of the [instrument]{.b .pink} on the [outcome variable]{.b .orange}, _e.g._,
$$
\begin{align}
  \textcolor{#FFA500}{\text{Earnings}_i} = \pi_0 + \pi_1 \textcolor{#e64173}{\text{Draft}_i} + w_i
\end{align}
$$

::: {.fragment}

we can estimate our desired effect:

:::

::: {.fragment}

(Effect of [veteran status]{.b .purple} on [earnings]{.b .orange}) =
<br>  [(Effect of [the draft]{.b .pink} on [earnings]{.b .orange})]{.top}
<br>  [(Effect of [the draft]{.b .pink} on [veteran status]{.b .purple})]{.bottom}

:::

---

## Instrumental variables
### Estimation: Bring it all together

By estimating two regressions involving our [instrument]{.b .pink},

1. The effect of the [instrument]{.b .pink} on the [endogenous variable]{.b .purple}, _e.g._,
$$
\begin{align}
  \textcolor{#6A5ACD}{\text{Veteran}_i} = \gamma_0 + \gamma_1 \textcolor{#e64173}{\text{Draft}_i} + v_i
\end{align}
$$

1. The effect of the [instrument]{.b .pink} on the [outcome variable]{.b .orange}, _e.g._,
$$
\begin{align}
  \textcolor{#FFA500}{\text{Earnings}_i} = \pi_0 + \pi_1 \textcolor{#e64173}{\text{Draft}_i} + w_i
\end{align}
$$

we can estimate our desired effect:

(Effect of [veteran status]{.b .purple} on [earnings]{.b .orange}) = $\dfrac{\pi_1}{\gamma_1}$

---

## Instrumental variables
### Estimation: Bring it all together

So with instrumental variables, we estimate $\beta_1$ using
$$
\begin{align}
  \hat{\beta}_1^\text{IV} = \dfrac{\hat{\pi}_1}{\hat{\gamma}_1}
\end{align}
$$
where $\hat{\pi}_1$ and $\hat{\gamma}_1$ come from the two equations we just discussed.

::: {.fragment}

**Q:** Can we trust $\hat{\beta}_1^\text{IV}$?

:::

::: {.fragment}

<br>**A:** Yes... **if we have a valid instrument.**

:::

::: {.fragment}

$$
\begin{align}
  \mathop{\text{plim}}\left( \hat{\beta}_1^\text{IV} \right) = \beta_1 + \dfrac{\mathop{\text{Cov}} \left( \textcolor{#e64173}{\text{Instrument}},\, u \right)}{\mathop{\text{Cov}} \left( \textcolor{#e64173}{\text{Instrument}},\, \textcolor{#6A5ACD}{\text{Endog. variable}} \right)}
\end{align}
$$

:::

::: {.fragment}

which equals $\beta_1$ as long as our instrument is [exogenous]{.b .pink} (numerator) and [relevant]{.b .purple} (denominator).

:::

---

## {.clear}

```{r}
#| label: "venn_iv"
#| echo: false
#| fig-height: 7.5
# Colors (order: x1, x2, x3, y, z)
venn_colors <- c(purple, red, "grey60", orange, red_pink)
# Line types (order: x1, x2, x3, y, z)
venn_lines <- c("solid", "dotted", "dotted", "solid", "solid")
# Locations of circles
venn_df <- tibble(
  x  = c( 0.0,   -0.5,    1.5,   -1.0, -1.4),
  y  = c( 0.0,   -2.5,   -1.8,    2.0, -2.6),
  r  = c( 1.9,    1.5,    1.5,    1.3,  1.3),
  l  = c( "Y", "X[1]", "X[2]", "X[3]",  "Z"),
  xl = c( 0.0,    0.7,    1.6,   -1.0, -2.9),
  yl = c( 0.0,   -3.8,   -1.9,    2.2, -2.6)
)
# Venn
ggplot(data = venn_df, aes(x0 = x, y0 = y, r = r, fill = l, color = l)) +
geom_circle(aes(linetype = l), alpha = 0.3, size = 0.75) +
theme_void() +
theme(legend.position = "none") +
scale_fill_manual(values = venn_colors) +
scale_color_manual(values = venn_colors) +
scale_linetype_manual(values = venn_lines) +
geom_text(aes(x = xl, y = yl, label = l), size = 9, family = "Fira Sans Book", parse = T) +
annotate(
  x = -5.5, y = 3.3,
  geom = "text", label = "Figure 1", size = 10, family = "Fira Sans Book", hjust = 0
) +
xlim(-5.5, 4.5) +
ylim(-4.2, 3.4) +
coord_equal()
```

---

## {.clear}

```{r}
#| label: "venn_iv_endog"
#| echo: false
#| fig-height: 7.5
# Change locations of circles
venn_df %>%
mutate(
  x = x +   c(0, 0, 0, 0, 0),
  xl = xl + c(0, 0, 0, 0, 0),
  y = y +   c(0, 0, 0, 0, 1),
  yl = yl + c(0, 0, 0, 0, 1)
) %>%
# Venn
ggplot(data = ., aes(x0 = x, y0 = y, r = r, fill = l, color = l)) +
geom_circle(aes(linetype = l), alpha = 0.3, size = 0.75) +
theme_void() +
theme(legend.position = "none") +
scale_fill_manual(values = venn_colors) +
scale_color_manual(values = venn_colors) +
scale_linetype_manual(values = venn_lines) +
geom_text(aes(x = xl, y = yl, label = l), size = 9, family = "Fira Sans Book", parse = T) +
annotate(
  x = -5.5, y = 3.3,
  geom = "text", label = "Figure 2", size = 10, family = "Fira Sans Book", hjust = 0
) +
xlim(-5.5, 4.5) +
ylim(-4.2, 3.4) +
coord_equal()
```

---

## {.clear}

```{r}
#| label: "venn_iv_irrelevant"
#| echo: false
#| fig-height: 7.5
# Change locations of circles
venn_df %>%
mutate(
  x = x +   c(0, 0, 0, 0,-1),
  xl = xl + c(0, 0, 0, 0,-1),
  y = y +   c(0, 0, 0, 0, 2.3),
  yl = yl + c(0, 0, 0, 0, 2.3)
) %>%
# Venn
ggplot(data = ., aes(x0 = x, y0 = y, r = r, fill = l, color = l)) +
geom_circle(aes(linetype = l), alpha = 0.3, size = 0.75) +
theme_void() +
theme(legend.position = "none") +
scale_fill_manual(values = venn_colors) +
scale_color_manual(values = venn_colors) +
scale_linetype_manual(values = venn_lines) +
geom_text(aes(x = xl, y = yl, label = l), size = 9, family = "Fira Sans Book", parse = T) +
annotate(
  x = -5.5, y = 3.3,
  geom = "text", label = "Figure 3", size = 10, family = "Fira Sans Book", hjust = 0
) +
xlim(-5.5, 4.5) +
ylim(-4.2, 3.4) +
coord_equal()
```

---

## {.clear}

```{r}
#| label: "venn_iv_endog2"
#| echo: false
#| fig-height: 7.5
# Change locations of circles
venn_df %>%
mutate(
  x = x +   c(0,    0,   0, 0,    2),
  xl = xl + c(0, -2.4, 0.8, 0,  4.6),
  y = y +   c(0,    0,   0, 0,    0),
  yl = yl + c(0,    0,   0, 0, -1.1)
) %>%
# Venn
ggplot(data = ., aes(x0 = x, y0 = y, r = r, fill = l, color = l)) +
geom_circle(aes(linetype = l), alpha = 0.3, size = 0.75) +
theme_void() +
theme(legend.position = "none") +
scale_fill_manual(values = venn_colors) +
scale_color_manual(values = venn_colors) +
scale_linetype_manual(values = venn_lines) +
geom_text(aes(x = xl, y = yl, label = l), size = 9, family = "Fira Sans Book", parse = T) +
annotate(
  x = -5.5, y = 3.3,
  geom = "text", label = "Figure 4", size = 10, family = "Fira Sans Book", hjust = 0
) +
xlim(-5.5, 4.5) +
ylim(-4.2, 3.4) +
coord_equal()
```

---

## Venn diagram explanation

In these figures (Venn diagrams)

- Each circle illustrates a variable.
- Overlap gives the share of correlatation between two variables.
- Dotted borders denote *omitted* variables.

Take-aways

- Figure 1: [Valid instrument]{.b .pink} (relevant; exogenous)
- Figure 2: [Invalid instrument]{.b .slate} (relevant; not exogenous)
- Figure 3: [Invalid instrument]{.b .slate} (not relevant; not exogenous)
- Figure 4: [Invalid instrument]{.b .slate} (relevant; not exogenous)

---

## {.clear}

Let's work an example in [R]{.mono}.

---

## Instrumental variables {#r_example}
### Example in [R]{.mono}

Back to our age-old battle to estimate the returns to education.

```{r}
#| label: "wooldridge_data"
#| echo: false
# Grab desired variables
wage_df <- wage2 %>% transmute(wage, education = educ, education_dad = feduc, education_mom = meduc) %>% na.omit() %>% as_tibble()
# Print to screen
wage_df
```

---

## Instrumental variables
### Example in [R]{.mono}

OLS for the returns to education with will likely (definitely) be biased...
$$
\begin{align}
  \textcolor{#FFA500}{\text{Wage}_i} = \beta_0 + \beta_1 \textcolor{#6A5ACD}{\text{Education}}_i + u_i
\end{align}
$$

::: {.fragment}

[(Likely biased) OLS results:]{.b .slate}

:::

::: {.fragment}

```{r}
#| label: "ols"
#| echo: false
reg_ols <- lm(wage ~ education, wage_df)
reg_ols %>%
  tidy_table(
    terms = c("Intercept", "Education"),
    highlight_row = 2,
    highlight_color = purple
  )
```

:::

::: {.fragment}

but what if mother's education provides a valid instrument?

:::

---

## Instrumental variables
### Example in [R]{.mono}

We can check/test the *relevance* of [mother's education]{.b .pink} for [education]{.b .purple}.

::: {.fragment}

This regression is known as the [*first stage*:]{.b .slate}
<br> The effect of the [instrument]{.pink} on our [endogenous explanatory variable]{.purple}.

:::

::: {.fragment}

$$
\begin{align}
  \textcolor{#6A5ACD}{\text{Education}_i} = \gamma_0 + \gamma_1 \textcolor{#e64173}{\left( \text{Mother's Education} \right)_i} + v_i
\end{align}
$$

:::

::: {.fragment}

[First-stage results:]{.b .slate}

:::

::: {.fragment}

```{r}
#| label: "first_stage"
#| echo: false
reg_iv1 <- lm(education ~ education_mom, wage_df)
reg_iv1 %>%
  tidy_table(
    terms = c("Intercept", "Mother's Education"),
    highlight_row = 2,
    highlight_color = red_pink
  )
```

:::

::: {.fragment}

The *p*-value suggests a very strong relationship (very *relevant*).

:::

---

## Instrumental variables
### Visualizing the first stage

```{r}
#| label: "first_stage_plot"
#| echo: false
#| fig-height: 5.75
ggplot(data = wage_df, aes(x = education_mom, y = education)) +
geom_point(alpha = 0.15, size = 3.5) +
labs(x = "Mother's education (years)", y = "Education (years)") +
theme_pander(base_size = 20, base_family = "Fira Sans Book")
```

---

## Instrumental variables
### Visualizing the first stage

```{r}
#| label: "first_stage_plot2"
#| echo: false
#| fig-height: 5.75
ggplot(data = wage_df, aes(x = education_mom, y = education)) +
geom_point(alpha = 0.15, size = 3.5) +
geom_smooth(method = lm, se = F, color = red_pink) +
labs(x = "Mother's education (years)", y = "Education (years)") +
theme_pander(base_size = 20, base_family = "Fira Sans Book")
```

---

## Instrumental variables
### Exogeneity

**Q:** What does [exogeneity]{.b .pink} mean in this case?

::: {.fragment}

<br>**A:** We need two things

:::

::: {.fragment}

1. [Mother's education (our instrument)]{.pink} must only affect earnings through [education (our endogenous explanatory variable)]{.purple}.
2. [Mother's education]{.pink} must be uncorrelated with other factors that affect [wages (our outcome variable)]{.orange}.

:::

::: {.fragment}

We want to be able to compare two people (*A* and *B*) whose mothers have different levels of education and say that the only differences between the two people (*A* and *B*) are due to their mothers' educational levels.

:::

::: {.fragment}

**Q:** Does *mother's education* seem likely to satisfy exogeneity?

:::

---

## Instrumental variables
### Example in [R]{.mono}

Now, let's estimate the [*reduced form*]{.b .turquoise}:
<br> The effect of the [instrument]{.pink} on our [outcome variable]{.orange}.

$$
\begin{align}
  \textcolor{#FFA500}{\text{Wage}_i} = \pi_0 + \pi_1 \textcolor{#e64173}{\left( \text{Mother's Education} \right)_i} + w_i
\end{align}
$$

::: {.fragment}

[Reduced-form results:]{.b .turquoise}

:::

::: {.fragment}

```{r}
#| label: "reduced_form"
#| echo: false
reg_rf <- lm(wage ~ education_mom, wage_df)
reg_rf %>%
  tidy_table(
    terms = c("Intercept", "Mother's Education"),
    highlight_row = 2,
    highlight_color = red_pink,
    digits = c(NA, 2, 2, 2, 5)
  )
```

:::

::: {.fragment}

**Q[1]{.sub}:** How do we interpret this estimated coefficient $\left( \hat{\pi}_1 \right)$?

:::

::: {.fragment}

<br>**Q[2]{.sub}:** If our instrument is *valid*, can we interpret these estimates as [causal]{.b .pink}?

:::

---

## Instrumental variables
### Example in [R]{.mono}

So what is our IV-based estimate for the returns to education?
$$
\begin{align}
  \textcolor{#FFA500}{\text{Wage}_i} = \beta_0 + \beta_1 \textcolor{#6A5ACD}{\text{Education}}_i + u_i
\end{align}
$$

::: {.fragment}

We know that the IV estimate for $\beta_1$ is

:::

::: {.fragment}

$$
\begin{align}
  \hat{\beta}_1^\text{IV} = \dfrac{\textcolor{#20B2AA}{\hat{\pi}_1}}{\textcolor{#314f4f}{\hat{\gamma}_1}}
\end{align}
$$

:::

::: {.fragment}

1. In the [reduced-form equation]{.b .turquoise}, we estimated $\textcolor{#20B2AA}{\hat{\pi}_1 \approx `r reg_rf$coef[2] %>% round(2)`}$.
2. In the [first-stage equation]{.b .slate}, we estimated $\textcolor{#314f4f}{\hat{\gamma}_1 \approx `r reg_iv1$coef[2] %>% round(3)`}$.

:::

::: {.fragment}

$$
\begin{align}
  \implies\hat{\beta}_1^\text{IV} = \dfrac{\textcolor{#20B2AA}{\hat{\pi}_1}}{\textcolor{#314f4f}{\hat{\gamma}_1}} = \dfrac{\textcolor{#20B2AA}{`r reg_rf$coef[2] %>% round(2)`}}{\textcolor{#314f4f}{`r reg_iv1$coef[2] %>% round(3)`}} \approx `r ((reg_rf$coef[2] %>% round(2))/(reg_iv1$coef[2] %>% round(3))) %>% round(1)`
\end{align}
$$

:::

---

## Instrumental variables
### Example in [R]{.mono}

**Alternative:** Use the function `iv_robust()` from the `estimatr` package.

This new function `iv_robust` works very similar to our good friend `lm`:

`iv_robust(y ~ x | z, data = dataset)`

- `formula` Specify the regression followed by `|` and your instrument (`z`).
- `data` You still need a dataset.

***Note:*** As you might guess by its name, `iv_robust` calculates heteroskedasticity-robust standard errors by default.

---

## Instrumental variables
### Example in [R]{.mono}

In practice...

```{r}
#| label: "iv_reg"
#| echo: true
# Estimate our IV regression
iv_est <- iv_robust(wage ~ education | education_mom, data = wage_df)
```

```{r}
#| label: "iv_reg_results"
#| echo: false
iv_est %>%
  tidy_table(
    terms = c("Intercept", "Education"),
    highlight_row = 2,
    highlight_color = purple
  )
```

---

## Instrumental variables
### More

So now we know how to "do" instrumental variables

::: {.fragment}

*when we have one endogenous variable and one exogenous variable.*

:::

::: {.fragment}

1. Estimate the reduced form (regress [outcome var.]{.orange} on [instrument]{.pink}).
2. Estimate the first stage (regress [expl. var.]{.purple} on [instrument]{.pink}).
3. Calculate the IV estimate using the estimates from (1) and (2).

:::

::: {.fragment}

Our magical [instrument]{.pink} isolates the exogenous variation in our [endogenous variable]{.purple}.

:::

::: {.fragment}

**Q:** What if we want more?

:::

::: {.fragment}

(_E.g._, more instruments or endog. variables)

:::

::: {.fragment}

<br>**A:** Too bad.

:::

---

## Instrumental variables
### More

So now we know how to "do" instrumental variables *when we have one endogenous variable and one exogenous variable.*

1. Estimate the reduced form (regress [outcome var.]{.orange} on [instrument]{.pink}).
2. Estimate the first stage (regress [expl. var.]{.purple} on [instrument]{.pink}).
3. Calculate the IV estimate using the estimates from (1) and (2).

Our magical [instrument]{.pink} isolates the exogenous variation in our [endogenous variable]{.purple}.

**Q:** What if we want more? (_E.g._, more instruments or endog. variables)
<br>**A:** [Too bad.]{.st} Extend IV to [two-stage least squares (2SLS)]{.b .pink}.

---

## {.center}

[Two-stage least squares]{.bigger .it .slab}

---

## Two-stage least squares {#2sls_intro}
### Intro

The intuition and insights of IV carry over into two-stage least squares.

::: {.fragment}

**Plus:** The *first stage* that we've been discussing is actually the *first* of the *two stages* in two-stage least squares.

:::

::: {.fragment}

$$
\begin{align}
  {\textcolor{#c5c5c5}{\text{Endogenous model}}}& &\textcolor{#FFA500}{\text{Outcome}_i} &= \beta_0 + \beta_1 \textcolor{#6A5ACD}{\left( \text{Endog. var.} \right)_i} + u_i\\[0.5em]
  {\text{First stage}}& &\textcolor{#6A5ACD}{\left( \text{Endog. var.} \right)_i} &= \pi_0 + \pi_1 \textcolor{#e64173}{\text{Instrument}_i} + v_i\\[0.25em]
  {\text{Second stage}}& &\textcolor{#FFA500}{\text{Outcome}_i} &= \delta_0 + \delta_1 \textcolor{#6A5ACD}{\widehat{\left( \text{Endog. var.} \right)}_i} + \varepsilon_i\\[0.5em]
  {\textcolor{#c5c5c5}{\text{Reduced form}}}& &\textcolor{#FFA500}{\text{Outcome}_i} &= \pi_0 + \pi_1 \textcolor{#e64173}{\text{Instrument}_i} + w_i\\[0.25em]
\end{align}
$$

:::

::: {.fragment}

where $\textcolor{#6A5ACD}{\widehat{\left( \text{Endog. var.} \right)}_i}$ denotes the predicted values (*fitted values*) from the first-stage regression.

:::

---

## Two-stage least squares
### Intro

Two-stage least squares is very flexible—we include other controls, additional endogenous variables, *and* have multiple instruments.

**But** don't get too distracted by this fancy flexiblity.

We still need [valid]{.b .pink} instruments.

---

## Two-stage least squares {#2sls_r}
### In [R]{.mono}

Back to our *returns to education* example.

$$
\begin{align}
  \textcolor{#FFA500}{\text{Wage}_i} = \beta_0 + \beta_1 \textcolor{#6A5ACD}{\text{Education}}_i + u_i
\end{align}
$$

Imagine that mother's *and* father's education are both valid instruments.

Then our [first-stage regression]{.b .slate} is
$$
\begin{align}
  \textcolor{#6A5ACD}{\text{Education}}_i = \gamma_0 + \gamma_1 \textcolor{#e64173}{\left( \text{Mother's education} \right)}_i + \gamma_2 \textcolor{#e64173}{\left( \text{Father's education} \right)}_i + v_i
\end{align}
$$
which we can estimate via OLS.

::: {.fragment}

**Q:** Why?

:::

---

## Two-stage least squares
### In [R]{.mono}

$$
\begin{align}
  \textcolor{#6A5ACD}{\text{Education}}_i = \gamma_0 + \gamma_1 \textcolor{#e64173}{\left( \text{Mother's education} \right)}_i + \gamma_2 \textcolor{#e64173}{\left( \text{Father's education} \right)}_i + v_i
\end{align}
$$

```{r}
#| label: "stage1_code"
#| echo: true
stage1 <- lm(education ~ education_mom + education_dad, wage_df)
```

[First-stage results:]{.b .slate}

```{r}
#| label: "stage1_table"
#| echo: false
stage1 %>%
  tidy_table(
    terms = c("Intercept", "Mother's Education", "Father's Education"),
    highlight_row = 2:3,
    highlight_color = red_pink
  )
```

::: {.fragment}

Our instruments each appear to be *relevant*.

:::

::: {.fragment}

<br>Formally, we should jointly test them (_e.g._, $F$ test).

:::

---

## Two-stage least squares
### In [R]{.mono}

Using our [estimated first stage]{.slate}, we grab the *fitted* [endogenous variable]{.purple}
$$
\begin{align}
  \textcolor{#6A5ACD}{\widehat{\text{Education}}}_i = \widehat{\gamma}_0 + \widehat{\gamma}_1 \textcolor{#e64173}{\left( \text{Mother's education} \right)}_i + \widehat{\gamma}_2 \textcolor{#e64173}{\left( \text{Father's education} \right)}_i
\end{align}
$$

::: {.fragment}

```{r}
#| label: "add_predications"
#| echo: true
# Add fitted values from first stage
wage_df$education_hat <- stage1$fitted.values
```

:::

::: {.fragment}

Now we use OLS (again) to estimate the [second-stage regression]{.b .green}
$$
\begin{align}
  \textcolor{#FFA500}{\text{Wage}_i} = \delta_0 + \delta_1 \textcolor{#6A5ACD}{\widehat{\text{Education}}}_i + \varepsilon_i
\end{align}
$$

:::

---

## Two-stage least squares
### In [R]{.mono}

$$
\begin{align}
  \textcolor{#FFA500}{\text{Wage}_i} = \delta_0 + \delta_1 \textcolor{#6A5ACD}{\widehat{\text{Education}}}_i + \varepsilon_i
\end{align}
$$

```{r}
#| label: "stage2_code"
#| echo: true
stage2 <- lm(wage ~ education_hat, wage_df)
```

[Second-stage results:]{.b .green}

```{r}
#| label: "stage2_table"
#| echo: false
stage2 %>%
  tidy_table(
    terms = c("Intercept", "Fitted Education"),
    highlight_row = 2,
    highlight_color = purple
  )
```

---

## {.clear}

[Ordinary least squares]{.purple}

```{r}
#| label: "results_ols"
#| echo: false
reg_ols %>%
  tidy_table(
    terms = c("Intercept", "Education"),
    highlight_row = 2,
    highlight_color = purple
  )
```

<br>[Instrumental variables]{.slate}

```{r}
#| label: "results_iv"
#| echo: false
iv_est %>%
  tidy_table(
    terms = c("Intercept", "Education"),
    highlight_row = 2,
    highlight_color = "darkslategrey"
  )
```

<br>[Two-stage least squares w/ two instruments]{.green}

```{r}
#| label: "results_2sls"
#| echo: false
stage2 %>%
  tidy_table(
    terms = c("Intercept", "Education"),
    highlight_row = 2,
    highlight_color = "#8bb174"
  )
```

---

## Two-stage least squares
### In [R]{.mono}

As you probably guessed, [R]{.mono} will do both of the stages for you.

::: {.fragment}

`iv_robust(y ~ x1 + x2 + ⋯ | z1 + z2 + ⋯, data)`

:::

::: {.fragment}

In our case, we have

<ul>
<li>one explanatory variable (`x`) ([education]{.purple})</li>
<li>two instruments (`z`) ([parents' educations]{.pink})</li>
</ul>

:::

::: {.fragment}

```{r}
#| label: "iv_robust_2sls"
#| echo: true
#| eval: false
iv_robust(wage ~ education | education_mom + education_dad, data = wage_df)
```

:::

::: {.fragment}

```{r}
#| label: "iv_robus_2sls_table"
#| echo: false
iv_robust(wage ~ education | education_mom + education_dad, data = wage_df) %>%
  tidy_table(
    terms = c("Intercept", "Education, fitted"),
    highlight_row = 2,
    highlight_color = purple
  )
```

:::

---

## Two-stage least squares {#more-apps}
### There's more!

Because 2SLS [isolates exogenous variation in an endogenous variable]{.b .pink}, we apply it in other settings that are biased from *endogenous* relationships.

::: {.fragment}

[Common applications]{.b .pink}

:::

::: {.fragment}

- **General causal inference** for observational data (as we've seen).
- **Experiments:** Randomize a treatment that affects an endog. variable.
- **Measurement error:** Regress noisy $x_1$ on noisy $x_2$ to capture signal.
- **Simultaneous relationships** (_e.g._, $p$ and $q$ from supply and demand).

:::

::: {.fragment}

However, in any 2SLS/IV setting, you need to mind the requirements for [valid instruments]{.b .pink}—[exogeneity]{.pink} and [relevance]{.pink}.

:::

---

## Table of contents

::: columns
::: {.column width="50%"}
#### Admin

::: {.smallest}
1. [Schedule](#schedule)
1. [Causality review](#causality_review)
:::
:::
::: {.column width="50%"}
#### Instrumental variables

::: {.smallest}
1. [Introduction](#intro)
1. [What is an instrument?](#instruments)
  - [Relevant](#relevant)
  - [Exogenous](#exogenous)
1. [IV Estimation](#iv_estimation)
1. [Venn diagrams](#venn)
1. [Example in [R]{.mono}](#r_example)
1. [Two-stage least squares](#2sls_intro)
  - [Introduction](#2sls_intro)
  - [Back to [R]{.mono}](#2sls_r)
1. [More applications](#more)
:::
:::
:::
